### 2015-02-04 draft.tex
Initial commit of Abstract Data Type Inference.

\documentclass[10pt,preprint,nocopyrightspace]{sigplanconf}

\usepackage[utf8]{inputenc}
\usepackage{microtype}
\usepackage{natbib}
\usepackage{xspace}
\usepackage{balance}
\usepackage{hyperref}
\usepackage{url}
\usepackage{breakurl}
\usepackage{amsmath}

\begin{document}

\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country} 
\copyrightyear{20yy} 
\copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
\doi{nnnnnnn.nnnnnnn}

% Uncomment one of the following two, if you are not going for the 
% traditional copyright transfer agreement.

%\exclusivelicense                % ACM gets exclusive license to publish, 
                                  % you retain copyright

%\permissiontopublish             % ACM gets nonexclusive license to publish
                                  % (paid open-access papers, 
                                  % short abstracts)

% \titlebanner{banner above paper title}        % These are ignored unless
% \preprintfooter{short description of paper}   % 'preprint' option specified.

\title{Abstract Data Type Inference}
% \subtitle{}

\authorinfo{Michael Emmi}
           {IMDEA Software Institute}
           {michael.emmi@imdea.org}
\authorinfo{Constantin Enea\and Jad Hamza}
           {Université Paris Diderot}
           {\{cenea,jhamza\}@univ-paris-diderot.fr}

\maketitle

\begin{abstract}

  TODO WRITE THE ABSTRACT

\end{abstract}

\category{F.3.1}{Specifying and Verifying and Reasoning about Programs}{Mechanical verification}

\terms
Reliability, Verification

\keywords
Concurrency; Refinement; Linearizability

\input intro
\input inference
\input embedding
\input computing
\input empirical
\input related

% \appendix

% \acks …

\bibliographystyle{abbrvnat}
\bibliography{violin}

% % The bibliography should be embedded for final submission.
%
% \begin{thebibliography}{}
% \softraggedright
%
% \bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
% P. Q. Smith, and X. Y. Jones. ...reference text...
%
% \end{thebibliography}

\end{document}

### 2015-02-04 inference.tex
Initial commit of Abstract Data Type Inference.

\section{ADT Inference}
\label{sec:inference}

TODO DEFINE REFINEMENT AND SET THE ADT INFERENCE PROBLEM

### 2015-02-04 intro.tex
Initial commit of Abstract Data Type Inference.

\section{Introduction}
\label{sec:intro}

TODO WRITE THE STORY

### 2015-02-04 related.tex
Initial commit of Abstract Data Type Inference.

\section{Related Work}
\label{sec:related}

TODO COMPARE WITH OTHER REFINEMENT CHECKING APPROACHES AND FIGURE OUT HOW THIS
MIGHT BE RELATED TO OTHER INFERENCE ALGORITHMS

### 2015-02-27 intro.tex
Initial draft of introduction.

@@
 
-TODO WRITE THE STORY
+TODO DISCUSS THE IMPORTANCE OF OBSERVATIONAL REFINEMENT, AND THAT IT IS HARD
+
+Recent work has demonstrated that checking observational refinement can be
+performed efficiently in practice, given symbolic ADT representations,
+e.g.,~using first-order logic. However, writing accurate symbolic
+specifications is an expected challenge for the designers of concurrent ADTs.
+Common practice is to write reference implementations. Mapping these imperative
+implementations to the logical formulæ describing valid executions is a great
+challenge for human developers.
+
+In this work we demonstrate that effective symbolic ADT representations can be
+extracted from the executions of reference implementations. Our approach
+exploits two key features of concurrent object ADTs: that violations of each
+ADT can be decomposed into a small set of representative patterns, and that
+these patterns manifest in executions with few operations. The first feature
+allows us to represent symbolic ADTs finitely, as exclusions of violation
+patterns. The second allows us to extract violation patterns from finite
+enumerations of executions.
+
+The fundamental challenge is in identifying the algebraic properties of ADTs
+which allow us to characterize an infinite set of violating executions with a
+finite set of patterns. This characterization is non-trivial since an execution
+with more operations than a given violating execution is not necessarily a
+violation itself. For instance, an execution which contains only a single {\tt
+pop} operation returning the value {\tt 1} is a violation to the atomic stack
+ADT, whereas an execution containing an additional {\tt push(1)} operation,
+overlapping in time with the {\tt pop}, is not. Further complication arises
+from the infinite set of possible data values, i.e.,~method argument and return
+values. Our patterns must be sensitive to the \emph{relation} among data values
+without being sensitive to the data values themselves. For instance, a
+sequential execution in which {\tt 1} and {\tt 2} are pushed and subsequently
+popped in the same order violates the atomic stack ADT. Yet, while replacing
+both values {\tt 1} and {\tt 2} with the value {\tt 1} results in a
+\emph{valid} stack execution, replacing them with {\tt 3} and {\tt 4} results
+in a violation.
+
+Our algebraic insight is based on grouping the operations of an execution into
+\emph{matchings}. Intuitively, operations which refer to the same instances of
+values belong to the same matching. For example, a {\tt pop} operation
+returning the value {\tt 1} matches a preceding {\tt push(1)} operation. By
+comparing executions by the characteristics of their matchings, rather than the
+actual data values they use, we capture the relation among data values
+independently of the data values themselves. Furthermore, the notion of
+matchings provides a key algebraic property of ADTs: the executions of
+naturally-occurring concurrent object ADTs are closed under the removal of
+matchings. For example, any execution of the atomic stack ADT using values {\tt
+1}, {\tt 2}, and {\tt 3} would remain a valid execution were all operations
+using the value {\tt 2} deleted. Conversely, any execution which extends a
+violating execution with additional matchings is itself a violation. This
+property, along with analogous algebraic properties concerning operation order
+and completion, allow us to characterize the violations of ADTs with finite
+basis sets, ultimately leading to finite symbolic ADT representations.
+
+Computing the basis sets of ADT violation patterns is a challenging problem,
+requiring the computation of global properties of an infinite number of
+executions — analogously to the inference of inductive invariants. Exploiting a
+hypothesis that violation patterns manifest in executions with few operations,
+we propose an under-approximating algorithm which extracts the patterns
+observed in all violating executions up to a given number of operations. In
+theory, for an arbitrary ADT, this is clearly incomplete: any violation which
+only surfaces with a greater number of operations would not be captured, thus
+resulting in a symbolic ADT representation which can fail to identify
+violations — though still guarantees never to classify a valid execution as a
+violation. Empirically, however, we demonstrate that our hypothesis holds: the
+patterns emerging from executions with few operations are complete in
+characterizing all violations of naturally-occurring concurrent object ADTs,
+thus allowing us to compute complete symbolic ADT representations in practice.
+
+The contributions and outline of this work are as follows:
+\begin{itemize}
+
+  \item The statement of the ADT inference problem (§\ref{sec:inference}).
+
+  \item Identification of algebraic properties allowing finite
+  characterizations of infinite ADT violations sets (§\ref{sec:embedding}).
+
+  \item An under-approximating algorithm for the ADT inference problem
+  (§\ref{sec:computing}).
+
+  \item An algorithm to infer the matching functions required for our algebraic
+  characterization of ADTs (§\ref{sec:matching}).
+
+  \item Empirical validation that our ADT-inference approximation is complete
+  in practice, and effective for the detection of violations to observational
+  refinement~(§\ref{sec:empirical}).
+
+\end{itemize}
+We conclude with a discussion of related work (§\ref{sec:related}).

### 2015-02-27 algebra.tex
Section titles.

\section{Algebraic Properties of ADTs}
\label{sec:algebra}

TODO DEFINE EMBEDDING AND PROVE OUR ADTS ARE CLOSED

\begin{theorem}
  “Our ADTs are closed under embedding.”
\end{theorem}

### 2015-02-27 algorithm.tex
Section titles.

\section{Approximating ADT Inference}
\label{sec:algorithm}

TODO SHOW HOW TO COMPUTE MINIMAL, FINITE, AND HOPEFULLY COMPLETE ADT
REPRESENTATIONS WHICH CAN BE USED FOR REFINEMENT CHECKING

\begin{theorem}
  “Our algorithm is sound; maybe it is also complete under certain conditions.”
\end{theorem}

### 2015-02-27 intro.tex
Section titles.

@@
   \item Identification of algebraic properties allowing finite
-  characterizations of infinite ADT violations sets (§\ref{sec:embedding}).
+  characterizations of infinite ADT violations sets (§\ref{sec:algebra}).
 
   \item An under-approximating algorithm for the ADT inference problem
-  (§\ref{sec:computing}).
+  (§\ref{sec:algorithm}).
 

### 2015-02-27 matching.tex
Section titles.

\section{Inferring Matching Functions}
\label{sec:matching}

TODO SHOW HOW TO COMPUTE MATCHING FUNCTIONS FROM THE SPECIFICATION THAT OBJECTS
SHOULD BE MATCH-REMOVAL/EMBEDDING CLOSED WRT A GIVEN MATCHING FUNCTION

\begin{theorem}
  “Our algorithm is sound; maybe it is also complete under certain conditions.”
\end{theorem}

### 2015-02-27 draft.tex
Added the major theorems.
Optimized packages.
New section for computing matching functions.
Section titles.
Abstract plus a couple citations.

@@
-\documentclass[10pt,preprint,nocopyrightspace]{sigplanconf}
+\documentclass[10pt,preprint,nocopyrightspace,natbib,authoryear]{sigplanconf}
 
@@
 \usepackage{microtype}
-\usepackage{natbib}
-\usepackage{xspace}
-\usepackage{balance}
 \usepackage{hyperref}
-\usepackage{url}
-\usepackage{breakurl}
 \usepackage{amsmath}
+\usepackage{amsthm}
+
+\newtheorem{theorem}{Theorem}
+\newtheorem{lemma}{Lemma}
 
@@
 
-  TODO WRITE THE ABSTRACT
+  Checking that efficient implementations of concurrent objects adhere to their
+  abstract data types (ADTs) can be made tractable when ADTs are given by
+  symbolic, logical representations. Such representations are however difficult
+  to obtain. Common practice is to design reference implementations.
+
+  In this work we demonstrate that effective symbolic ADT representations can
+  be automatically extracted from the executions of reference implementations.
+  Our approach exploits key features of naturally-occurring ADTs: violations
+  can be decomposed into a small set of representative patterns, and these
+  patterns manifest in executions with few operations. By identifying certain
+  algebraic properties of naturally-occurring ADTs, and exhaustively sampling
+  executions up to a small number of operations, we extract concise symbolic
+  ADT representations which are complete in practice.
 
@@
 \input inference
-\input embedding
-\input computing
+\input algebra
+\input algorithm
+\input matching
 \input empirical
@@
 % \acks …
+
+% \balancecolumns
 

### 2015-02-27 intro.tex
Abstract plus a couple citations.

@@
 performed efficiently in practice, given symbolic ADT representations,
-e.g.,~using first-order logic. However, writing accurate symbolic
-specifications is an expected challenge for the designers of concurrent ADTs.
-Common practice is to write reference implementations. Mapping these imperative
-implementations to the logical formulæ describing valid executions is a great
-challenge for human developers.
+e.g.,~using first-order logic~\cite{conf/popl/BouajjaniEEH15,
+conf/pldi/EmmiEH15}. However, writing accurate symbolic specifications is an
+expected challenge for the designers of concurrent ADTs. Common practice is to
+write reference implementations. Mapping these imperative implementations to
+the logical formulæ describing valid executions is a great challenge for human
+developers.
 

### 2015-02-28 draft.tex
Updated intro, and added related work.

@@
   abstract data types (ADTs) can be made tractable when ADTs are given by
-  symbolic, logical representations. Such representations are however difficult
-  to obtain. Common practice is to design reference implementations.
+  symbolic, logical representations. Obtaining these symbolic representations,
+  however, is a complex task requiring rare expertise. In practice, concurrent
+  objects are specified indirectly by possibly-abstract imperative reference
+  implementations.
 
   In this work we demonstrate that effective symbolic ADT representations can
-  be automatically extracted from the executions of reference implementations.
+  be automatically generated from the executions of reference implementations.
   Our approach exploits key features of naturally-occurring ADTs: violations
@@
   algebraic properties of naturally-occurring ADTs, and exhaustively sampling
-  executions up to a small number of operations, we extract concise symbolic
+  executions up to a small number of operations, we generate concise symbolic
   ADT representations which are complete in practice.

### 2015-02-28 intro.tex
Updated intro, and added related work.

--- 2015-02-27 intro.tex
@@
 
-TODO DISCUSS THE IMPORTANCE OF OBSERVATIONAL REFINEMENT, AND THAT IT IS HARD
+Many modern software systems rely on efficient implementations of abstract data
+types (ADTs) like atomic collections, semaphores, and locks. Efficient
+implementation is difficult though, because efficiency entails maximizing
+concurrency, and concurrency jeopardizes atomicity. Consequently,
+implementations are vulnerable to subtle and elusive bugs compromising
+adherence to their ADTs~\cite{journals/cacm/staff11}, thus necessitating
+automated verification. Adherence of an implementation to an ADT is captured
+formally by a property called \emph{observational refinement}: any observable
+behavior over the method call and return actions of the implementation is also
+admitted by the the ADT.
 
-Recent work has demonstrated that checking observational refinement can be
-performed efficiently in practice, given symbolic ADT representations,
-e.g.,~using first-order logic. However, writing accurate symbolic
-specifications is an expected challenge for the designers of concurrent ADTs.
-Common practice is to write reference implementations. Mapping these imperative
-implementations to the logical formulæ describing valid executions is a great
-challenge for human developers.
+Although checking observational refinement is a fundamentally hard
+problem\footnote{See Section~\ref{sec:related} for further discussion.} for
+arbitrary ADTs~\cite{journals/siamcomp/GibbonsK97, conf/esop/BouajjaniEEH13},
+recent work demonstrates that naturally-occurring ADTs like atomic collections,
+semaphores, and locks can be checked efficiently given effective symbolic ADT
+representations~\cite{conf/popl/BouajjaniEEH15, journals/arxiv/BouajjaniEEH15,
+conf/pldi/EmmiEH15}, e.g.,~specified in first-order logic. However, writing
+accurate symbolic specifications is an expected challenge for ADT designers. In
+practice, ADTs are specified indirectly as the observable behaviors of
+possibly-abstract imperative \emph{reference implementations}. Mapping these
+reference implementations to logical formulæ describing the same sets of
+behaviors is a complex task requiring rare expertise.
 
 In this work we demonstrate that effective symbolic ADT representations can be
-extracted from the executions of reference implementations. Our approach
-exploits two key features of concurrent object ADTs: that violations of each
-ADT can be decomposed into a small set of representative patterns, and that
-these patterns manifest in executions with few operations. The first feature
-allows us to represent symbolic ADTs finitely, as exclusions of violation
-patterns. The second allows us to extract violation patterns from finite
-enumerations of executions.
+generated automatically from the executions of reference implementations. Our
+approach exploits two key features of concurrent-object ADTs: that violations
+of each ADT can be decomposed into a small set of representative patterns, and
+that these patterns manifest in executions with few operations. The first
+feature allows us to represent symbolic ADTs finitely, as exclusions of
+violation patterns. The second allows us to extract violation patterns from
+finite enumerations of executions.
 
@@
 property, along with analogous algebraic properties concerning operation order
-and completion, allow us to characterize the violations of ADTs with finite
-basis sets, ultimately leading to finite symbolic ADT representations.
+and completion, allow us to compare executions via a violation-preserving
+embedding relation. This relation is a well-founded partial order on
+executions, and thus allows us to characterize the infinite set of violations
+to an ADT with a finite basis set, ultimately leading to a finite symbolic
+representation.
 
@@
   \item Empirical validation that our ADT-inference approximation is complete
-  in practice, and effective for the detection of violations to observational
-  refinement~(§\ref{sec:empirical}).
+  in practice, and effective for the detection of observational-refinement
+  violations~(§\ref{sec:empirical}).
 

### 2015-02-28 related.tex
Updated intro, and added related work.

@@
 MIGHT BE RELATED TO OTHER INFERENCE ALGORITHMS
+
+The idea of decomposing ADT violations into a finite set of patterns has been
+proposed for atomic collections~\cite{conf/concur/HenzingerSV13,
+conf/popl/DoddsHK15, conf/popl/BouajjaniEEH15}, and generalized to ADTs which
+can be expressed with a particular form of recursive
+definitions~\cite{journals/arxiv/BouajjaniEEH15}. These works suggest patterns
+which directly recognize violations across the entire set of \emph{concurrent}
+executions, effectively reducing observational refinement to the verification
+of classical temporal-logic properties. Some of these patterns are expressible
+only as infinite, though \emph{regular}, languages. On the contrary, in this
+work we suggest patterns expressed as \emph{finite} languages, which recognize
+violations in ADT \emph{kernels} — i.e.,~the sequential executions of atomic
+objects, or the “concurrency-aware” executions of “concurrency-aware”
+specifications~\cite{conf/podc/HemedR14}. Though by themselves these patterns
+are incomplete in recognizing violations across the entire set of concurrent
+executions, they are complete when considered as negative conditions on
+concurrent executions’ linearization orders, effectively reducing verification
+of a single concurrent execution to satisfiability
+checking~\cite{conf/pldi/EmmiEH15}.

### 2015-02-28 draft.tex
Added to related work.

@@
   however, is a complex task requiring rare expertise. In practice, concurrent
-  objects are specified indirectly by possibly-abstract imperative reference
-  implementations.
+  objects are specified indirectly by imperative reference implementations.
 
@@
   executions up to a small number of operations, we generate concise symbolic
-  ADT representations which are complete in practice.
+  ADT representations which are complete in practice, enabling the application
+  of efficient symbolic verification algorithms without the burden of manual
+  specification.
 

### 2015-02-28 intro.tex
Added to related work.

@@
 arbitrary ADTs~\cite{journals/siamcomp/GibbonsK97, conf/esop/BouajjaniEEH13},
-recent work demonstrates that naturally-occurring ADTs like atomic collections,
-semaphores, and locks can be checked efficiently given effective symbolic ADT
-representations~\cite{conf/popl/BouajjaniEEH15, journals/arxiv/BouajjaniEEH15,
-conf/pldi/EmmiEH15}, e.g.,~specified in first-order logic. However, writing
-accurate symbolic specifications is an expected challenge for ADT designers. In
-practice, ADTs are specified indirectly as the observable behaviors of
-possibly-abstract imperative \emph{reference implementations}. Mapping these
-reference implementations to logical formulæ describing the same sets of
-behaviors is a complex task requiring rare expertise.
+recent work demonstrates that implementations of naturally-occurring ADTs like
+atomic collections, semaphores, and locks can be checked efficiently given
+effective symbolic ADT representations~\cite{conf/popl/BouajjaniEEH15,
+journals/arxiv/BouajjaniEEH15, conf/pldi/EmmiEH15}, e.g.,~specified in
+first-order logic. However, writing accurate symbolic specifications is an
+expected challenge for ADT designers. In practice, ADTs are specified
+indirectly as the observable behaviors of possibly-abstract imperative
+\emph{reference implementations}. Mapping these reference implementations to
+logical formulæ describing the same sets of behaviors is a complex task
+requiring rare expertise.
 
 In this work we demonstrate that effective symbolic ADT representations can be
-generated automatically from the executions of reference implementations. Our
-approach exploits two key features of concurrent-object ADTs: that violations
-of each ADT can be decomposed into a small set of representative patterns, and
-that these patterns manifest in executions with few operations. The first
-feature allows us to represent symbolic ADTs finitely, as exclusions of
-violation patterns. The second allows us to extract violation patterns from
-finite enumerations of executions.
+generated automatically from the executions of reference implementations,
+enabling the application of efficient symbolic refinement-checking algorithms
+without the burden of writing formal specifications manually. Our approach
+exploits two key features of concurrent-object ADTs: that violations of each
+ADT can be decomposed into a small set of representative patterns, and that
+these patterns manifest in executions with few operations. The first feature
+allows us to represent symbolic ADTs finitely, as exclusions of violation
+patterns. The second allows us to extract violation patterns from finite
+enumerations of executions.
 
@@
 We conclude with a discussion of related work (§\ref{sec:related}).
+
+As far as we are aware, this work is the first to suggest the automatic
+generation of symbolic ADT representations for concurrent objects. By removing
+the burden of writing formal ADT specifications manually, this work broadens
+the scope of efficient symbolic refinement-checking algorithms to
+newly-designed ADTs, apart from those few traditionally studied in the
+literature.

### 2015-02-28 related.tex
Added to related work.

--- 2015-02-04 related.tex
@@
 
-TODO COMPARE WITH OTHER REFINEMENT CHECKING APPROACHES AND FIGURE OUT HOW THIS
-MIGHT BE RELATED TO OTHER INFERENCE ALGORITHMS
+Inference of symbolic abstract data types is a recently-pertinent problem
+arising from the the emergence of efficient symbolic refinement-checking
+algorithms~\cite{conf/popl/BouajjaniEEH15, conf/pldi/EmmiEH15} which require
+symbolic, logical representations of ADT specifications. Though approaches
+based on the explicit enumeration of execution
+linearizations~\cite{journals/jpdc/WingG93, conf/pldi/BurckhardtDMT10,
+conf/asplos/BurnimNS11, conf/kbse/ZhangCW13} do not require symbolic ADT
+representations, and work directly with given reference implementations, these
+approaches are intractable, since the number of linearizations is exponential
+in execution length.
+
+Other approaches to tractable refinement-checking are based on annotating
+method bodies with \emph{linearization
+points}~\cite{journals/toplas/HerlihyW90, conf/cav/AmitRRSY07,
+conf/fm/LiuCLS09, conf/cav/Vafeiadis10, conf/podc/OHearnRVYY10,
+conf/icse/Zhang11a, conf/oopsla/ShachamBASVY11, conf/cav/DragoiGH13,
+conf/pldi/LiangF13} to reduce the otherwise exponential number of possible
+linearizations to one single linearization. This approach, however, does not
+apply to all implementations~\cite{journals/toplas/HerlihyW90}, and
+applications of this approach often rely on manual annotation of the
+implementation source code. Furthermore, this approach does not admit
+conclusive evidence of a refinement violation in case of failure.
+
+The idea of decomposing ADT violations into a finite set of patterns has been
+proposed for atomic collections~\cite{conf/tacas/AbdullaHHJR13,
+conf/concur/HenzingerSV13, conf/popl/DoddsHK15, conf/popl/BouajjaniEEH15}, and
+generalized to ADTs which can be expressed with a particular form of recursive
+definition~\cite{journals/arxiv/BouajjaniEEH15}. These works suggest patterns
+which directly recognize violations in \emph{concurrent} executions,
+effectively reducing observational refinement to the verification of classical
+temporal-logic properties. Some of these patterns are expressible only as
+\emph{infinite}, though regular, languages. On the contrary, in this work we
+suggest patterns expressed as \emph{finite} languages, which recognize
+violations in ADT \emph{kernels} — i.e.,~the sequential executions of atomic
+objects, or the “concurrency-aware” executions of “concurrency-aware”
+specifications~\cite{conf/podc/HemedR14}. Though by themselves these patterns
+are incomplete in recognizing violations in concurrent executions, they are
+complete when considered as negative conditions on the linearizations of
+concurrent executions, effectively reducing verification of a single execution
+to satisfiability checking~\cite{conf/pldi/EmmiEH15}.

### 2015-03-07 algebra.tex
Some algebra.

@@
 
-TODO DEFINE EMBEDDING AND PROVE OUR ADTS ARE CLOSED
+Two histories $h_1$ and $h_2$ are related by $\to_x$, for $x = \mathrm{o},
+\mathrm{c}, \mathrm{p}$, when $h_2$ is obtained from $h_1$ by:
+\begin{itemize}
 
-\begin{theorem}
-  “Our ADTs are closed under embedding.”
-\end{theorem}
+  \item removing an order constraint (o),
+
+  \item making a completed operation pending (c), or
+
+  \item adding a pending operation (p).
+
+\end{itemize}
+A set of histories $H$ is \emph{closed} under a relation $\to$ when $h_2 \in H$
+whenever $h_1 \to h_2$ and $h_1 \in H$. A fundamental property of concurrent
+object implementations is that their histories are closed under weakening via
+less ordering, fewer operations completed, and additional pending
+operations~\cite{conf/popl/BouajjaniEEH15}.
+
+\begin{lemma}
+
+  $H(\mathcal{I})$ is closed under $\to_\mathrm{o}$, $\to_\mathrm{p}$, 
+  $\to_\mathrm{c}$.
+
+\end{lemma}
+
+The ADTs we consider in this work are also closed under the additional
+relations $\to_\mathrm{m}$ and $\to_\mathrm{d}$ which relate two histories
+$h_1$ and $h_2$ when $h_2$ is obtained from $h_1$ by
+\begin{itemize}
+
+  \item removing a match (m), or
+
+  \item removing a duplicate operation (d).
+
+\end{itemize}
+We say an ADT closed under $\to_\mathrm{m}$ and $\to_\mathrm{d}$ is
+\emph{normal}. Defining the relation $\succeq$ as the reflexive and transitive
+closure of the above relations
+\begin{align*}
+  \succeq = (\to_\mathrm{o} \cup \to_\mathrm{p} \cup \to_\mathrm{c} \cup 
+  \to_\mathrm{m} \cup \to_\mathrm{d})^\ast,
+\end{align*}
+closure under $\succeq$ follows immediately.
+
+\begin{lemma}
+
+  Normal ADTs are closed under $\succeq$.
+
+\end{lemma}
+
+TODO DEFINE WQO, FINITELY-REPRESENTABLE, COMPUTABLE
+
+\begin{example}
+
+  $\preceq$ is not a wqo.
+
+\end{example}
+
+\begin{lemma}
+
+  $\preceq$ is a wqo on the set of histories with at most $k$ pending
+  operations, for any $k \in \mathbb{N}$.
+
+\end{lemma}
+
+For the remainder of this section, we fix an ADT $A$, and let $B$ be the
+complement of $\ker A$.
+
+\begin{lemma}
+
+  $B$ is finitely representable if $A$ is normal.
+
+\end{lemma}
+
+For all $i \in \mathbb{N}$, let
+\begin{align*}
+  B_i &= \set{ h \in B : \_ \le i } \\
+  B_i' &= \set{ h' \in B :
+    \exists h \in B_i.\ h \preceq h' \text{ and } \_ \le i+1
+  }
+\end{align*}
+be, respectively, the histories of $B$ with at most $i$ matches, and those
+derived from $B_i$ with at most $i\!+\!1$ matches. We say that $A$ is
+\emph{predictable} if $B_i = B$ whenever $B_i' = B_{i+1}$.
+
+\begin{lemma}
+
+  $B_i$ is computable, for all $i \in \mathbb{N}$, if $A$ is normal.
+
+\end{lemma}
+
+\begin{lemma}
+
+  $B$ is computable if $A$ is normal and predictable.
+
+\end{lemma}

### 2015-03-07 inference.tex
Section titles.
Some algebra.

@@
-\section{ADT Inference}
+\section{Symbolic ADT Inference}
 \label{sec:inference}
@@
 TODO DEFINE REFINEMENT AND SET THE ADT INFERENCE PROBLEM
+
+A \emph{history} $h = \tup{O,<,\ell,f,m}$ is a partial order $<$ on a set $O$
+of operations with method labeling $\ell: O \to M$, completion labeling $f: O
+\to \mathbb{B}$ — an operation $o \in O$ is \emph{completed} if $f(o)$
+otherwise \emph{pending} — and partial matching function $m: O \rightharpoonup
+O$ such that pending operations are maximal. We say $h$ is \emph{sequential}
+when $<$ is a total order on $O$, and \emph{(in)complete} when (not) all
+operations are completed. An operation $o \in \rng(m)$ is a \emph{match
+target}, and the set $\set{ o' \in O : m(o') = o}$ of operations targeting $o$
+is a \emph{match}.

### 2015-03-07 algebra.tex
Added not-a-wqo example.

@@
 
-  $\preceq$ is not a wqo.
+  Consider the infinite history sequence $h_1 h_2 \ldots$ where each $h_i$
+  contains $2i$ operations $o_1, o_1', \ldots, o_i, o_i'$ where each $o_j$ is a
+  completed {\tt push} operation matching itself, and each $o_j'$ is a pending
+  {\tt pop} operation with undefined matching. Because each successive $h_i$
+  has both more matches and more pending operations, no two histories of the
+  sequence are related by $\preceq$. Thus $\preceq$ is not a wqo.
 
 \end{example}
+
+The previous example demonstrates that $\preceq$ is not a wqo by allowing each
+history $h_i$ of the infinite sequence to contain more and more pending
+operations in order to ensure that $h_j \not\preceq h_i$ for every $j < i$.
+Curbing this ability by limiting the maximum amount of pending operations per
+history makes $\preceq$ a wqo. Although limiting to $k$ pending operations
+essentially limits us to histories of executions with at most $k$ operations
+parallel at any moment — e.g.,~of programs with at most $k$ threads — this
+restriction comes at no loss of completeness when considering atomic ATD
+kernels, whose histories are limited to $k\!=\!1$ pending operation.
+
+TODO COMMENT ON CONCURRENCY-AWARE KERNELS
 

### 2015-03-07 inference.tex
Added preliminary definitions.

--- 2015-02-04 inference.tex
@@
-\section{ADT Inference}
+\section{Symbolic ADT Inference}
 \label{sec:inference}
 
-TODO DEFINE REFINEMENT AND SET THE ADT INFERENCE PROBLEM
+TODO MOTIVATE THIS SECTION
+
+We fix an arbitrary infinite set $\mathbb{O}$ of operation identifiers, and
+given sets $\mathbb{M}$ and $\mathbb{V}$ of method names and values. A
+\emph{call action} binds an operation identifier $o \in \mathbb{O}$ with a
+method name $p \in \mathbb{M}$ and argument value $v \in \mathbb{V}$, while a
+\emph{return action} binds an operation identifier $o$ with a return value $v$.
+An \emph{execution} $e$ is a sequence of call and return actions where
+\begin{itemize}
+
+  \item each operation identifier is used in at most one call action, and in at
+  most one return action, and
+
+  \item each return action is preceded by a call action with the same operation
+  identifier.
+
+\end{itemize}
+A \emph{matching scheme} $M$ associates to each execution $e$ with operations
+$O$ a partial matching function $M(e): O \rightharpoonup O$.
+
+\begin{example}
+
+  TODO MATCHING SCHEME EXAMPLE
+
+\end{example}
+
+In this work, histories are abstractions of executions which retain method
+names yet ignore argument and return values, and retain the relative order of
+operations yet ignore the exact sequence of call and return actions. Formally,
+a \emph{history} $h = \tup{O,<,c,f,m}$ is a tuple where
+\begin{itemize}
+
+  \item $O \subseteq \mathbb{O}$ is a set of \emph{operations},
+
+  \item $<$ is a partial \emph{happens-before} order on $O$,
+
+  \item $c: O \to \mathbb{B}$ labels operations as \emph{completed}, or not,
+
+  \item $f: O \to \mathbb{M}$ labels operations with method names, and
+
+  \item $m: O \rightharpoonup O$ is a partial \emph{matching} function.
+
+\end{itemize}
+Non-completed operations are \emph{pending}, and are maximal in the
+happens-before order. The operations of a \emph{sequential} history are totally
+ordered. An operation $o \in \rng(m)$ is a \emph{match target}, and the set
+$\set{ o' \in O : m(o') = o}$ of operations targeting $o$ is a \emph{match}. We
+assume targets target themselves: $o \in \set{ o' \in O : m(o') = o}$.
+
+The history $H(M,e)$ of an execution $e$ under matching scheme $M$ is the tuple
+$\tup{O,<,c,f,M(e)}$ where
+\begin{itemize}
+
+  \item $O$ are the operations of $e$,
+
+  \item $o_1 < o_2$ if{f} operation $o_1$ returns before $o_2$ is called in $e$,
+
+  \item $c(o)$ if{f} operation $o$ returns in $e$, and
+
+  \item $f(o)$ is the name of the method executed by $o$ in $e$.
+
+\end{itemize}
+We denote the set $\set{H(M,e) : e \in E}$ of histories of an execution set $E$
+by $H(M,E)$.
+
+A matching scheme $M$ is \emph{faithful} to a set $E$ of executions when $e \in
+E$ if{f} $e' \in E$ for any two executions $e$ and $e'$ such that $H(M,e) =
+H(M,e')$. A set of executions $E$ is \emph{data independent}\footnote{TODO NOTE
+THAT THIS IS NOT THE STANDARD DEFINITION…} when there exists a faithful
+matching scheme. By definition, abstracting executions to histories comes at no
+loss of precision up to set inclusion for data-independent execution sets.
+
+\begin{lemma}
+
+  $H(M,e) \in H(M,E)$ if and only if $e \in E$, for any faithful matching
+  scheme $M$.
+
+\end{lemma}
+
+For the remainder of this work, we assume each set of executions $E$ comes
+equipped with a matching scheme $M$, and abbreviate $H(M,E)$ by $H(E)$.
+
+An \emph{implementation} $\mathcal{I}$ is a labeled transition system whose
+traces are executions. The histories $H(\mathcal{I})$ of an implementation are
+the histories of its traces. An \emph{abstract data type (ADT)} $A$ is the set
+$H(\mathcal{I})$ of histories of \emph{some} implementation $\mathcal{I}$. We
+define \emph{observational refinement} between two implementations, or between
+an implementation and an ADT, as history-set inclusion. We say $\mathcal{I}_1$
+\emph{refines} $\mathcal{I}_2$ if{f} $H(\mathcal{I}_1) \subseteq
+H(\mathcal{I}_2)$, and $\mathcal{I}$ \emph{refines} $A$ if{f} $H(\mathcal{I})
+\subseteq A$.
+
+TODO DEFINE THE ADT INFERENCE PROBLEM

### 2015-03-07 inference.tex
Defined symbolic ADT inference problem.

@@
 For the remainder of this work, we assume each set of executions $E$ comes
-equipped with a matching scheme $M$, and abbreviate $H(M,E)$ by $H(E)$.
+equipped with a faithful matching scheme $M$, and abbreviate $H(M,E)$ by $H(E)$.
 
@@
 
-TODO DEFINE THE ADT INFERENCE PROBLEM
+TODO DEFINE HISTORY LOGIC
+
+The \emph{symbolic ADT inference problem} is to compute a first-order history
+formula $F$ for a given ADT $A$ such that
+\begin{itemize}
+
+  \item $h \models F$ for all $h \in \ker A$, and
+
+  \item $h \not\models F$ for all $h \in B$,
+
+\end{itemize}
+where $B$ is the complement of $\ker A$.

### 2015-03-07 inference.tex
Interpretation of history formulas.

@@
 
-An \emph{implementation} $\mathcal{I}$ is a labeled transition system whose
-traces are executions. The histories $H(\mathcal{I})$ of an implementation are
-the histories of its traces. An \emph{abstract data type (ADT)} $A$ is the set
-$H(\mathcal{I})$ of histories of \emph{some} implementation $\mathcal{I}$. We
-define \emph{observational refinement} between two implementations, or between
-an implementation and an ADT, as history-set inclusion. We say $\mathcal{I}_1$
-\emph{refines} $\mathcal{I}_2$ if{f} $H(\mathcal{I}_1) \subseteq
-H(\mathcal{I}_2)$, and $\mathcal{I}$ \emph{refines} $A$ if{f} $H(\mathcal{I})
-\subseteq A$.
+An \emph{implementation} $\mathcal{I}$ is a prefixed-closed set of executions.
+An \emph{abstract data type (ADT)} $A$ is the set $H(\mathcal{I})$ of histories
+of an implementation $\mathcal{I}$. We define \emph{observational refinement}
+between two implementations, or between an implementation and an ADT, as
+history-set inclusion. We say $\mathcal{I}_1$ \emph{refines} $\mathcal{I}_2$
+if{f} $H(\mathcal{I}_1) \subseteq H(\mathcal{I}_2)$, and $\mathcal{I}$
+\emph{refines} $A$ if{f} $H(\mathcal{I}) \subseteq A$.
 
-TODO DEFINE HISTORY LOGIC
+In the following we fix a symbolic representation of ADTs on which to base our
+symbolic ADT inference problem. A \emph{history formula} $F$ is a first-order
+logic formula with
+\begin{itemize}
 
-The \emph{symbolic ADT inference problem} is to compute a first-order history
-formula $F$ for a given ADT $A$ such that
+  \item variables ranging over operation identifiers,
+
+  \item constants from $\mathbb{M}$ for method names,
+
+  \item a function symbol $\mathsf{m}$ for matching, and
+
+  \item predicate symbols $\mathsf{c}$ and $\mathsf{<}$ for completion and
+  order.
+
+\end{itemize}
+A history formula $F$ is interpreted over a history $h$ in the natural way, by
+binding variables to the operations of $h$, and binding function and predicate
+symbols to their interpretations in $h$. We write $h \models F$ when $h$ is a
+model of $F$, and $h \not\models F$ otherwise.
+
+\begin{example}
+
+  TODO HISTORY FORMULA EXAMPLE
+
+\end{example}
+
+TODO DEFINE KERNEL AND COMPLEMENT
+
+A history formula $F$ \emph{represents} an ADT $A$ when:
 \begin{itemize}
@@
 \end{itemize}
-where $B$ is the complement of $\ker A$.
+where $B$ is the complement of $\ker A$. The \emph{symbolic ADT inference
+problem} is to compute a history formula $F$ representing the ADT of a given
+implementation $\mathcal{I}$.

### 2015-03-07 inference.tex
Added references to nature.

@@
 
-For the remainder of this work, we assume each set of executions $E$ comes
-equipped with a faithful matching scheme $M$, and abbreviate $H(M,E)$ by $H(E)$.
+In Section~\ref{sec:nature} we demonstrate faithful matching schemes for the
+executions of naturally-occurring concurrent objects, and in
+Section~\ref{sec:matching} we demonstrate how to infer faithful matching
+schemes. Otherwise, for the remainder of this work, we assume each set of
+executions comes equipped with a faithful matching scheme $M$, and abbreviate
+$H(M,E)$ by $H(E)$.
 

### 2015-03-07 nature.tex
Added references to nature.

\section{Naturally-Occurring ADTs}
\label{sec:nature}

TODO DEMONSTRATE FAITHFUL MATCHING SCHEMES, NORMALITY, AND PREDICTABILITY FOR
EACH NATURALLY-OCCURRING ADT

\subsection{The Atomic Queue}

\subsection{The Atomic Stack}

\subsection{The Atomic Set}

\subsection{The Atomic Register}

\subsection{The Atomic Lock}

\subsection{The Atomic Semaphore}

\subsection{The Rendezvous Synchronizer}

\subsection{The Exchanger}

\subsection{The Barrier}

ANY OTHERS??

### 2015-03-10 algebra.tex
Added references to nature.
Consistent Section 2.

@@
 
-Two histories $h_1$ and $h_2$ are related by $\to_x$, for $x = \mathrm{o},
-\mathrm{c}, \mathrm{p}$, when $h_2$ is obtained from $h_1$ by:
-\begin{itemize}
-
-  \item removing an order constraint (o),
-
-  \item making a completed operation pending (c), or
-
-  \item adding a pending operation (p).
-
-\end{itemize}
-A set of histories $H$ is \emph{closed} under a relation $\to$ when $h_2 \in H$
-whenever $h_1 \to h_2$ and $h_1 \in H$. A fundamental property of concurrent
-object implementations is that their histories are closed under weakening via
-less ordering, fewer operations completed, and additional pending
-operations~\cite{conf/popl/BouajjaniEEH15}.
-
-\begin{lemma}
-
-  $H(\mathcal{I})$ is closed under $\to_\mathrm{o}$, $\to_\mathrm{p}$, 
-  $\to_\mathrm{c}$.
-
-\end{lemma}
+TODO MOTIVATE THIS SECTION
 
@@
 We say an ADT closed under $\to_\mathrm{m}$ and $\to_\mathrm{d}$ is
-\emph{normal}. Defining the relation $\succeq$ as the reflexive and transitive
-closure of the above relations
+\emph{normal}. In Section~\ref{sec:nature} we demonstrate that
+naturally-occurring ADTs are normal. Defining the relation $\succeq$ as the
+reflexive and transitive closure of the above relations
 \begin{align*}
@@
 derived from $B_i$ with at most $i\!+\!1$ matches. We say that $A$ is
-\emph{predictable} if $B_i = B$ whenever $B_i' = B_{i+1}$.
+\emph{predictable} if $B_i = B$ whenever $B_i' = B_{i+1}$. In 
+Section~\ref{sec:nature} we demonstrate that naturally-occurring ADTs are
+predictable.
 

### 2015-03-10 inference.tex
Consistent Section 2.

@@
 \end{itemize}
-A \emph{matching scheme} $M$ associates to each execution $e$ with operations
-$O$ a partial matching function $M(e): O \rightharpoonup O$.
+An \emph{implementation} $\mathcal{I}$ is a prefix-closed set of executions
+which is additionally closed under
+\begin{itemize}
 
-\begin{example}
+  \item appending call actions (of fresh operations),
 
-  TODO MATCHING SCHEME EXAMPLE
+  \item permuting call actions backward in time, and
 
-\end{example}
+  \item permuting return actions forward in time.
+
+\end{itemize}
+These conditions capture the environment in which implementations execute:
+calls are always enabled in their client programs, and thread schedulers may
+induce arbitrary delay between implementation code and the associated call and
+return actions~\cite{conf/popl/BouajjaniEEH15}.
 
@@
 Non-completed operations are \emph{pending}, and are maximal in the
-happens-before order. The operations of a \emph{sequential} history are totally
-ordered. An operation $o \in \rng(m)$ is a \emph{match target}, and the set
-$\set{ o' \in O : m(o') = o}$ of operations targeting $o$ is a \emph{match}. We
-assume targets target themselves: $o \in \set{ o' \in O : m(o') = o}$.
+happens-before order. The \emph{width} of $h$ is the maximal size of a
+subset of $O$ for which each pair of operations is unordered. The \emph{width}
+of a history set is the maximal width of its elements. Width-$1$ histories are
+\emph{sequential}. An operation $o \in \rng(m)$ is a \emph{match target}, and
+the set $\set{ o' \in O : m(o') = o}$ of operations targeting $o$ is a
+\emph{match}. We assume targets target themselves: $o \in \set{ o' \in O :
+m(o') = o}$.
+
+Our abstraction from executions to histories relies correlating the operations
+associated with the same values via matching. A \emph{matching scheme} $M$
+associates to each execution $e$ with operations $O$ a partial matching
+function $M(e): O \rightharpoonup O$.
+
+\begin{example}
+
+  TODO MATCHING SCHEME EXAMPLE
+
+\end{example}
 
@@
 
-An \emph{implementation} $\mathcal{I}$ is a prefixed-closed set of executions.
+Two histories $h_1$ and $h_2$ are related by $\to_x$, for $x = \mathrm{o},
+\mathrm{c}, \mathrm{p}$, when $h_2$ is obtained from $h_1$ by:
+\begin{itemize}
+
+  \item removing an order constraint (o),
+
+  \item making a completed operation pending (c), or
+
+  \item adding a pending operation (p).
+
+\end{itemize}
+A set of histories $H$ is \emph{closed} under a relation $\to$ when $h_2 \in H$
+whenever $h_1 \to h_2$ and $h_1 \in H$. A fundamental property of
+implementations is that their histories are closed under weakening via less
+ordering, fewer operations completed, and additional pending
+operations~\cite{conf/popl/BouajjaniEEH15}.
+
+\begin{lemma}
+
+  $H(\mathcal{I})$ is closed under $\to_\mathrm{o}$, $\to_\mathrm{p}$, 
+  $\to_\mathrm{c}$.
+
+\end{lemma}
+
 An \emph{abstract data type (ADT)} $A$ is the set $H(\mathcal{I})$ of histories
-of an implementation $\mathcal{I}$. We define \emph{observational refinement}
-between two implementations, or between an implementation and an ADT, as
-history-set inclusion. We say $\mathcal{I}_1$ \emph{refines} $\mathcal{I}_2$
-if{f} $H(\mathcal{I}_1) \subseteq H(\mathcal{I}_2)$, and $\mathcal{I}$
-\emph{refines} $A$ if{f} $H(\mathcal{I}) \subseteq A$.
+of an implementation $\mathcal{I}$. A \emph{kernel} of an ADT $A$ is a minimal
+set $H$ of histories such that
+\begin{align*}
+  A = \set{ h' : \exists h \in H.\ h \to^\ast h' }
+\end{align*}
+where $\to = (\to_\mathrm{o} \cup \to_\mathrm{p} \cup \to_\mathrm{c})$. In
+Section~\ref{sec:nature} we demonstrate that naturally-occurring ADTs have
+unique kernels. For the remainder of this work, we assume that ADTs have unique
+kernels, and denote \emph{the kernel} of $A$ by $\ker A$. We say that $A$ is
+\emph{atomic} when the width of $\ker A$ is $1$.
+
+We define \emph{observational refinement} between two implementations, or
+between an implementation and an ADT, as history-set inclusion. We say
+$\mathcal{I}_1$ \emph{refines} $\mathcal{I}_2$ if{f} $H(\mathcal{I}_1)
+\subseteq H(\mathcal{I}_2)$, and $\mathcal{I}$ \emph{refines} $A$ if{f}
+$H(\mathcal{I}) \subseteq A$.
 
@@
 
-TODO DEFINE KERNEL AND COMPLEMENT
-
-A history formula $F$ \emph{represents} an ADT $A$ when:
+The \emph{complement} of an ADT $A$ with width $k \in \mathbb{N}$ is the set of
+histories of width at most $k$ which are excluded from $A$. A history formula
+$F$ \emph{represents} an ADT $A$ when:
 \begin{itemize}

### 2015-03-10 algebra.tex
Note on width-bounded kernels.

@@
 history makes $\preceq$ a wqo. Although limiting to $k$ pending operations
-essentially limits us to histories of executions with at most $k$ operations
-parallel at any moment — e.g.,~of programs with at most $k$ threads — this
-restriction comes at no loss of completeness when considering atomic ATD
-kernels, whose histories are limited to $k\!=\!1$ pending operation.
+essentially limits us to width-$k$ histories, of executions with at most $k$
+operations parallel at any moment — e.g.,~of programs with at most $k$ threads
+— this restriction comes at no loss of completeness when considering atomic ADT
+kernels, which have width $1$.
 
@@
 
+TODO DEFINE A TERM FOR ADTS WHOSE KERNELS HAVE BOUNDED WIDTH
+
 \begin{lemma}
 
-  $\preceq$ is a wqo on the set of histories with at most $k$ pending
-  operations, for any $k \in \mathbb{N}$.
+  $\preceq$ is a wqo on width-$k$ histories, for any $k \in \mathbb{N}$.
 

### 2015-03-10 inference.tex
Added notion of “weight”.

@@
 Non-completed operations are \emph{pending}, and are maximal in the
-happens-before order. The \emph{width} of $h$ is the maximal size of a
-subset of $O$ for which each pair of operations is unordered. The \emph{width}
-of a history set is the maximal width of its elements. Width-$1$ histories are
-\emph{sequential}. An operation $o \in \rng(m)$ is a \emph{match target}, and
-the set $\set{ o' \in O : m(o') = o}$ of operations targeting $o$ is a
-\emph{match}. We assume targets target themselves: $o \in \set{ o' \in O :
-m(o') = o}$.
+happens-before order. The \emph{width} of a history is the maximum size of an
+unordered subset of its operations. The \emph{width} of a history set is the
+maximal width of its elements. Width-$1$ histories are \emph{sequential}. An
+operation $o \in \rng(m)$ is a \emph{match target}, and the set $\set{ o' \in O
+: m(o') = o}$ of operations targeting $o$ is a \emph{match}. We assume targets
+target themselves: $m(o) = o$ for $o \in \rng(m)$. The \emph{weight} of $h$ is
+the amount $|\rng(m)|$ of matches.
 

### 2015-03-11 inference.tex
Better Section 2.

--- 2015-03-10 inference.tex
@@
 given sets $\mathbb{M}$ and $\mathbb{V}$ of method names and values. A
-\emph{call action} binds an operation identifier $o \in \mathbb{O}$ with a
-method name $p \in \mathbb{M}$ and argument value $v \in \mathbb{V}$, while a
-\emph{return action} binds an operation identifier $o$ with a return value $v$.
-An \emph{execution} $e$ is a sequence of call and return actions where
+\emph{call action} binds an operation $o \in \mathbb{O}$ with a method name $p
+\in \mathbb{M}$ and argument value $v \in \mathbb{V}$, while a \emph{return
+action} binds an operation $o \in \mathbb{O}$ with a return value $v \in
+\mathbb{V}$. An \emph{execution} $e$ is a sequence of call and return actions
+where
 \begin{itemize}
@@
 
-  \item permuting call actions backward in time, and
+  \item permuting call actions backward, and
 
-  \item permuting return actions forward in time.
+  \item permuting return actions forward.
 
@@
 In this work, histories are abstractions of executions which retain method
-names yet ignore argument and return values, and retain the relative order of
-operations yet ignore the exact sequence of call and return actions. Formally,
-a \emph{history} $h = \tup{O,<,c,f,m}$ is a tuple where
+names yet lose the exact argument and return values, and retain the relative
+order of operations yet lose the exact sequence of call and return actions.
+Formally, a \emph{history} $h = \tup{O,<,c,f,m}$ is a tuple where
 \begin{itemize}
@@
 Non-completed operations are \emph{pending}, and are maximal in the
-happens-before order. The \emph{width} of $h$ is the maximal size of a
-subset of $O$ for which each pair of operations is unordered. The \emph{width}
-of a history set is the maximal width of its elements. Width-$1$ histories are
+happens-before order. The \emph{width} of a history is the maximum number of
+its operations which are mutually unordered, and the \emph{width} of a history
+set is the maximum width of its elements. Histories with width $1$ are
 \emph{sequential}. An operation $o \in \rng(m)$ is a \emph{match target}, and
 the set $\set{ o' \in O : m(o') = o}$ of operations targeting $o$ is a
-\emph{match}. We assume targets target themselves: $o \in \set{ o' \in O :
-m(o') = o}$.
+\emph{match}. We assume targets target themselves: $m(o) = o$ for $o \in
+\rng(m)$. The \emph{weight} of $h$ is the number $|\rng(m)|$ of matches.
+Since the operation identifiers of a history have no intrinsic meaning, we
+consider equality between histories up to renaming of operation identifiers.
 
-Our abstraction from executions to histories relies correlating the operations
-associated with the same values via matching. A \emph{matching scheme} $M$
-associates to each execution $e$ with operations $O$ a partial matching
-function $M(e): O \rightharpoonup O$.
+\begin{example}
+
+  TODO HISTORY EXAMPLE
+
+\end{example}
+
+Our abstraction of executions relies on correlating the operations associated
+with the same values in an execution via the partial matching function of a
+history. We construct partial matching functions systematically. A
+\emph{matching scheme} $M$ associates to each execution $e$ with operations $O$
+a partial matching function $M(e): O \rightharpoonup O$.
 
@@
 
-The history $H(M,e)$ of an execution $e$ under matching scheme $M$ is the tuple
+The history $H(e,M)$ of an execution $e$ under matching scheme $M$ is the tuple
 $\tup{O,<,c,f,M(e)}$ where
@@
 \end{itemize}
-We denote the set $\set{H(M,e) : e \in E}$ of histories of an execution set $E$
-by $H(M,E)$.
+We denote the set $\set{H(e,M) : e \in E}$ of histories of an execution set $E$
+by $H(E,M)$.
+
+\begin{example}
+
+  TODO HISTORY OF AN EXECUTION
+
+\end{example}
 
 A matching scheme $M$ is \emph{faithful} to a set $E$ of executions when $e \in
-E$ if{f} $e' \in E$ for any two executions $e$ and $e'$ such that $H(M,e) =
-H(M,e')$. A set of executions $E$ is \emph{data independent}\footnote{TODO NOTE
+E$ if{f} $e' \in E$ for any two executions $e$ and $e'$ such that $H(e,M) =
+H(e',M)$. A set of executions $E$ is \emph{data independent}\footnote{TODO NOTE
 THAT THIS IS NOT THE STANDARD DEFINITION…} when there exists a faithful
-matching scheme. By definition, abstracting executions to histories comes at no
-loss of precision up to set inclusion for data-independent execution sets.
+matching scheme. By definition, our abstraction of executions incurs no loss of
+precision for data-independent execution sets.
 
@@
 
-  $H(M,e) \in H(M,E)$ if and only if $e \in E$, for any faithful matching
+  $H(e,M) \in H(E,M)$ if and only if $e \in E$, for any faithful matching
   scheme $M$.
@@
 In Section~\ref{sec:nature} we demonstrate faithful matching schemes for the
-executions of naturally-occurring concurrent objects, and in
-Section~\ref{sec:matching} we demonstrate how to infer faithful matching
-schemes. Otherwise, for the remainder of this work, we assume each set of
-executions comes equipped with a faithful matching scheme $M$, and abbreviate
-$H(M,E)$ by $H(E)$.
+executions of naturally-occurring ADTs, and in Section~\ref{sec:matching} we
+demonstrate how to infer faithful matching schemes. Otherwise, for the
+remainder of this work, we assume each set of executions comes equipped with a
+faithful matching scheme $M$, and abbreviate $H(E,M)$ by $H(E)$.
 
@@
 of an implementation $\mathcal{I}$. A \emph{kernel} of an ADT $A$ is a minimal
-set $H$ of histories such that
+set $H$ of histories generating $A$, i.e.,~such that
 \begin{align*}
@@
 
-We define \emph{observational refinement} between two implementations, or
-between an implementation and an ADT, as history-set inclusion. We say
-$\mathcal{I}_1$ \emph{refines} $\mathcal{I}_2$ if{f} $H(\mathcal{I}_1)
-\subseteq H(\mathcal{I}_2)$, and $\mathcal{I}$ \emph{refines} $A$ if{f}
-$H(\mathcal{I}) \subseteq A$.
+We define \emph{observational refinement} between implementations and ADTs as
+history-set inclusion. Although this refinement is typically defined with
+respect to the admissibility of program executions, recent work demonstrates
+that these definitions are equivalent~\cite{conf/popl/BouajjaniEEH15}.
 
-In the following we fix a symbolic representation of ADTs on which to base our
-symbolic ADT inference problem. A \emph{history formula} $F$ is a first-order
-logic formula with
+\begin{definition}
+
+  We say an implementation $\mathcal{I}$ \emph{refines} an ADT $A$ when
+  $H(\mathcal{I}) \subseteq A$. We say an implementation $\mathcal{I}_1$
+  \emph{refines} another implementation $\mathcal{I}_2$ when $\mathcal{I}_1$
+  refines $H(\mathcal{I}_2)$.
+
+\end{definition}
+
+In order to frame the symbolic ADT inference problem, we fix a natural symbolic
+representation of ADTs. A \emph{history formula} $F$ is a first-order logic
+formula with
 \begin{itemize}
@@
 
-The \emph{complement} of an ADT $A$ with width $k \in \mathbb{N}$ is the set of
-histories of width at most $k$ which are excluded from $A$. A history formula
-$F$ \emph{represents} an ADT $A$ when:
+The \emph{complement} of a history set $H$ of width $k \in \mathbb{N} \cup
+\set{\omega}$ is the set of histories of width at most $k$ which are excluded
+from $H$. Let $A$ be an ADT and $B$ the complement of $\ker A$. We say that
+a history formula $F$ \emph{represents} $A$ when
 \begin{itemize}
@@
 
-  \item $h \not\models F$ for all $h \in B$,
+  \item $h \not\models F$ for all $h \in B$.
 
 \end{itemize}
-where $B$ is the complement of $\ker A$. The \emph{symbolic ADT inference
-problem} is to compute a history formula $F$ representing the ADT of a given
-implementation $\mathcal{I}$.
+The inference problem is thus to compute a formula representing an ADT.
+
+\begin{definition}
+
+  The \emph{symbolic ADT inference problem} is to compute a history formula $F$
+  representing the ADT $H(\mathcal{I})$ of a given implementation $\mathcal{I}$.
+
+\end{definition}

### 2015-03-11 histories.tex
Better organization.

--- 2015-03-11 inference.tex
@@
-\section{Symbolic ADT Inference}
-\label{sec:inference}
+\section{Histories of ADT Implementations}
+\label{sec:histories}
 
@@
 
+\begin{example}
+
+  TODO EXAMPLE EXECUTION
+
+\end{example}
+
 In this work, histories are abstractions of executions which retain method
@@
 its operations which are mutually unordered, and the \emph{width} of a history
-set is the maximum width of its elements. Histories with width $1$ are
+set is the maximum width of its elements. Width-$1$ histories are
 \emph{sequential}. An operation $o \in \rng(m)$ is a \emph{match target}, and
@@
 \end{lemma}
-
-An \emph{abstract data type (ADT)} $A$ is the set $H(\mathcal{I})$ of histories
-of an implementation $\mathcal{I}$. A \emph{kernel} of an ADT $A$ is a minimal
-set $H$ of histories generating $A$, i.e.,~such that
-\begin{align*}
-  A = \set{ h' : \exists h \in H.\ h \to^\ast h' }
-\end{align*}
-where $\to = (\to_\mathrm{o} \cup \to_\mathrm{p} \cup \to_\mathrm{c})$. In
-Section~\ref{sec:nature} we demonstrate that naturally-occurring ADTs have
-unique kernels. For the remainder of this work, we assume that ADTs have unique
-kernels, and denote \emph{the kernel} of $A$ by $\ker A$. We say that $A$ is
-\emph{atomic} when the width of $\ker A$ is $1$.
-
-We define \emph{observational refinement} between implementations and ADTs as
-history-set inclusion. Although this refinement is typically defined with
-respect to the admissibility of program executions, recent work demonstrates
-that these definitions are equivalent~\cite{conf/popl/BouajjaniEEH15}.
-
-\begin{definition}
-
-  We say an implementation $\mathcal{I}$ \emph{refines} an ADT $A$ when
-  $H(\mathcal{I}) \subseteq A$. We say an implementation $\mathcal{I}_1$
-  \emph{refines} another implementation $\mathcal{I}_2$ when $\mathcal{I}_1$
-  refines $H(\mathcal{I}_2)$.
-
-\end{definition}
-
-In order to frame the symbolic ADT inference problem, we fix a natural symbolic
-representation of ADTs. A \emph{history formula} $F$ is a first-order logic
-formula with
-\begin{itemize}
-
-  \item variables ranging over operation identifiers,
-
-  \item constants from $\mathbb{M}$ for method names,
-
-  \item a function symbol $\mathsf{m}$ for matching, and
-
-  \item predicate symbols $\mathsf{c}$ and $\mathsf{<}$ for completion and
-  order.
-
-\end{itemize}
-A history formula $F$ is interpreted over a history $h$ in the natural way, by
-binding variables to the operations of $h$, and binding function and predicate
-symbols to their interpretations in $h$. We write $h \models F$ when $h$ is a
-model of $F$, and $h \not\models F$ otherwise.
-
-\begin{example}
-
-  TODO HISTORY FORMULA EXAMPLE
-
-\end{example}
-
-The \emph{complement} of a history set $H$ of width $k \in \mathbb{N} \cup
-\set{\omega}$ is the set of histories of width at most $k$ which are excluded
-from $H$. Let $A$ be an ADT and $B$ the complement of $\ker A$. We say that
-a history formula $F$ \emph{represents} $A$ when
-\begin{itemize}
-
-  \item $h \models F$ for all $h \in \ker A$, and
-
-  \item $h \not\models F$ for all $h \in B$.
-
-\end{itemize}
-The inference problem is thus to compute a formula representing an ADT.
-
-\begin{definition}
-
-  The \emph{symbolic ADT inference problem} is to compute a history formula $F$
-  representing the ADT $H(\mathcal{I})$ of a given implementation $\mathcal{I}$.
-
-\end{definition}

### 2015-03-11 inference.tex
Better organization.

@@
-\section{Symbolic ADT Inference}
+\section{Refinement \& ADT Inference}
 \label{sec:inference}
@@
 TODO MOTIVATE THIS SECTION
-
-We fix an arbitrary infinite set $\mathbb{O}$ of operation identifiers, and
-given sets $\mathbb{M}$ and $\mathbb{V}$ of method names and values. A
-\emph{call action} binds an operation $o \in \mathbb{O}$ with a method name $p
-\in \mathbb{M}$ and argument value $v \in \mathbb{V}$, while a \emph{return
-action} binds an operation $o \in \mathbb{O}$ with a return value $v \in
-\mathbb{V}$. An \emph{execution} $e$ is a sequence of call and return actions
-where
-\begin{itemize}
-
-  \item each operation identifier is used in at most one call action, and in at
-  most one return action, and
-
-  \item each return action is preceded by a call action with the same operation
-  identifier.
-
-\end{itemize}
-An \emph{implementation} $\mathcal{I}$ is a prefix-closed set of executions
-which is additionally closed under
-\begin{itemize}
-
-  \item appending call actions (of fresh operations),
-
-  \item permuting call actions backward, and
-
-  \item permuting return actions forward.
-
-\end{itemize}
-These conditions capture the environment in which implementations execute:
-calls are always enabled in their client programs, and thread schedulers may
-induce arbitrary delay between implementation code and the associated call and
-return actions~\cite{conf/popl/BouajjaniEEH15}.
-
-In this work, histories are abstractions of executions which retain method
-names yet lose the exact argument and return values, and retain the relative
-order of operations yet lose the exact sequence of call and return actions.
-Formally, a \emph{history} $h = \tup{O,<,c,f,m}$ is a tuple where
-\begin{itemize}
-
-  \item $O \subseteq \mathbb{O}$ is a set of \emph{operations},
-
-  \item $<$ is a partial \emph{happens-before} order on $O$,
-
-  \item $c: O \to \mathbb{B}$ labels operations as \emph{completed}, or not,
-
-  \item $f: O \to \mathbb{M}$ labels operations with method names, and
-
-  \item $m: O \rightharpoonup O$ is a partial \emph{matching} function.
-
-\end{itemize}
-Non-completed operations are \emph{pending}, and are maximal in the
-happens-before order. The \emph{width} of a history is the maximum number of
-its operations which are mutually unordered, and the \emph{width} of a history
-set is the maximum width of its elements. Histories with width $1$ are
-\emph{sequential}. An operation $o \in \rng(m)$ is a \emph{match target}, and
-the set $\set{ o' \in O : m(o') = o}$ of operations targeting $o$ is a
-\emph{match}. We assume targets target themselves: $m(o) = o$ for $o \in
-\rng(m)$. The \emph{weight} of $h$ is the number $|\rng(m)|$ of matches.
-Since the operation identifiers of a history have no intrinsic meaning, we
-consider equality between histories up to renaming of operation identifiers.
-
-\begin{example}
-
-  TODO HISTORY EXAMPLE
-
-\end{example}
-
-Our abstraction of executions relies on correlating the operations associated
-with the same values in an execution via the partial matching function of a
-history. We construct partial matching functions systematically. A
-\emph{matching scheme} $M$ associates to each execution $e$ with operations $O$
-a partial matching function $M(e): O \rightharpoonup O$.
-
-\begin{example}
-
-  TODO MATCHING SCHEME EXAMPLE
-
-\end{example}
-
-The history $H(e,M)$ of an execution $e$ under matching scheme $M$ is the tuple
-$\tup{O,<,c,f,M(e)}$ where
-\begin{itemize}
-
-  \item $O$ are the operations of $e$,
-
-  \item $o_1 < o_2$ if{f} operation $o_1$ returns before $o_2$ is called in $e$,
-
-  \item $c(o)$ if{f} operation $o$ returns in $e$, and
-
-  \item $f(o)$ is the name of the method executed by $o$ in $e$.
-
-\end{itemize}
-We denote the set $\set{H(e,M) : e \in E}$ of histories of an execution set $E$
-by $H(E,M)$.
-
-\begin{example}
-
-  TODO HISTORY OF AN EXECUTION
-
-\end{example}
-
-A matching scheme $M$ is \emph{faithful} to a set $E$ of executions when $e \in
-E$ if{f} $e' \in E$ for any two executions $e$ and $e'$ such that $H(e,M) =
-H(e',M)$. A set of executions $E$ is \emph{data independent}\footnote{TODO NOTE
-THAT THIS IS NOT THE STANDARD DEFINITION…} when there exists a faithful
-matching scheme. By definition, our abstraction of executions incurs no loss of
-precision for data-independent execution sets.
-
-\begin{lemma}
-
-  $H(e,M) \in H(E,M)$ if and only if $e \in E$, for any faithful matching
-  scheme $M$.
-
-\end{lemma}
-
-In Section~\ref{sec:nature} we demonstrate faithful matching schemes for the
-executions of naturally-occurring ADTs, and in Section~\ref{sec:matching} we
-demonstrate how to infer faithful matching schemes. Otherwise, for the
-remainder of this work, we assume each set of executions comes equipped with a
-faithful matching scheme $M$, and abbreviate $H(E,M)$ by $H(E)$.
-
-Two histories $h_1$ and $h_2$ are related by $\to_x$, for $x = \mathrm{o},
-\mathrm{c}, \mathrm{p}$, when $h_2$ is obtained from $h_1$ by:
-\begin{itemize}
-
-  \item removing an order constraint (o),
-
-  \item making a completed operation pending (c), or
-
-  \item adding a pending operation (p).
-
-\end{itemize}
-A set of histories $H$ is \emph{closed} under a relation $\to$ when $h_2 \in H$
-whenever $h_1 \to h_2$ and $h_1 \in H$. A fundamental property of
-implementations is that their histories are closed under weakening via less
-ordering, fewer operations completed, and additional pending
-operations~\cite{conf/popl/BouajjaniEEH15}.
-
-\begin{lemma}
-
-  $H(\mathcal{I})$ is closed under $\to_\mathrm{o}$, $\to_\mathrm{p}$, 
-  $\to_\mathrm{c}$.
-
-\end{lemma}
 

### 2015-03-11 inference.tex
Segue between refinement and inference.

@@
 
-In order to frame the symbolic ADT inference problem, we fix a natural symbolic
-representation of ADTs. A \emph{history formula} $F$ is a first-order logic
-formula with
+Recent works demonstrate efficient algorithms\footnote{In time polynomial in
+the number of operations, per execution.} for checking observational
+refinement~\cite{conf/popl/BouajjaniEEH15, conf/pldi/EmmiEH15}, yet rely on
+hand-written symbolic ADT representations. In order to frame the problem of
+computing these automatically, we fix a language for symbolic representation. A
+\emph{history formula} $F$ is a first-order logic formula with
 \begin{itemize}

### 2015-03-11 algebra.tex
Added notion of “weight”.
Better story in Section 4.

@@
 
-The ADTs we consider in this work are also closed under the additional
-relations $\to_\mathrm{m}$ and $\to_\mathrm{d}$ which relate two histories
-$h_1$ and $h_2$ when $h_2$ is obtained from $h_1$ by
+In addition to the relations $\to_\mathrm{o}$, $\to_\mathrm{p}$, and
+$\to_\mathrm{c}$ of Section~\ref{sec:histories} under which all ADTs are
+closed, the ADTs we consider in this work are also closed under additional
+relations which remove matches, and duplicate operations. Given a history
+$\tup{O,<,c,f,m}$, we say an operation $o \in O$ is a \emph{duplicate} if there
+exists another operation $o' \in O$ with the same labels and matching: $c(o) =
+c(o')$, $f(o) = f(o')$, and $m(o) = m(o')$. The relations $\to_\mathrm{m}$ and
+$\to_\mathrm{d}$ relate two histories $h_1$ and $h_2$ when $h_2$ is obtained
+from $h_1$ by
 \begin{itemize}
@@
 naturally-occurring ADTs are normal. Defining the relation $\succeq$ as the
-reflexive and transitive closure of the above relations
+reflexive and transitive closure of the above relations,
 \begin{align*}
   \succeq = (\to_\mathrm{o} \cup \to_\mathrm{p} \cup \to_\mathrm{c} \cup 
-  \to_\mathrm{m} \cup \to_\mathrm{d})^\ast,
+  \to_\mathrm{m} \cup \to_\mathrm{d})^\ast
 \end{align*}
@@
 
-TODO DEFINE WQO, FINITELY-REPRESENTABLE, COMPUTABLE
+TODO MOTIVATE WQO
+
+A \emph{well-quasi-ordering (wqo)} $R$ on a set $X$ is a reflexive, transitive
+binary relation on $X$ for which in every infinite sequence $x_0 x_1 \ldots$ of
+elements from $X$, there exists a pair $x_i, x_j$, for $i \neq j$, related by
+$R$.
 
@@
 essentially limits us to width-$k$ histories, of executions with at most $k$
-operations parallel at any moment — e.g.,~of programs with at most $k$ threads
-— this restriction comes at no loss of completeness when considering atomic ADT
-kernels, which have width $1$.
-
-TODO COMMENT ON CONCURRENCY-AWARE KERNELS
-
-TODO DEFINE A TERM FOR ADTS WHOSE KERNELS HAVE BOUNDED WIDTH
+operations parallel at any moment, e.g.,~of programs with at most $k$ threads,
+this restriction comes at no loss of completeness when considering only the
+histories of bounded-width ADT kernels, which is the subject of the remainder
+of this section.
 
@@
 For the remainder of this section, we fix an ADT $A$, and let $B$ be the
-complement of $\ker A$.
+complement of $\ker A$. When $\ker A$ has bounded width, so does $B$, and thus
+$\preceq$ is a wqo on $B$. Furthermore, when $A$ normal, it is closed under
+$\succeq$, and thus $B$ is closed under $\preceq$. Closure under a wqo implies
+representation by a finite set. Formally, we say a set $X$ is \emph{finitely
+representable} if there exists a finite set $Y$ and a relation $R \subseteq Y
+\times X$ such that $X = \set{ x : \exists y\in Y.\ R(y,x) }$. In our case, we
+obtain a finite set from which exactly the elements of $B$ are related by
+$\preceq$.
 
@@
 
-For all $i \in \mathbb{N}$, let
+Computing $B$ thus amounts to computing its finite representation. In general,
+this is achieved by enumerating the elements of $B$ while maintaining the
+$\preceq$-minimal elements, and recognizing a condition under which all
+elements of $B$ are related to the current set of
+minimals~\cite{conf/lics/AbdullaCJT96, journals/tcs/FinkelS01}. In our case, we
+recognize a condition on the weight-increasing enumeration of $B$ which
+guarantees coverage by the current set of minimals. Formally, for all $i \in
+\mathbb{N}$, we define
 \begin{align*}
-  B_i &= \set{ h \in B : \_ \le i } \\
+  B_i &= \set{ h \in B : \weight h \le i } \\
   B_i' &= \set{ h' \in B :
-    \exists h \in B_i.\ h \preceq h' \text{ and } \_ \le i+1
+    \exists h \in B_i.\ h \preceq h' \text{ and } \weight h \le i+1
   }
 \end{align*}
-be, respectively, the histories of $B$ with at most $i$ matches, and those
+as, respectively, the histories of $B$ with at most $i$ matches, and those
 derived from $B_i$ with at most $i\!+\!1$ matches. We say that $A$ is
@@
 Section~\ref{sec:nature} we demonstrate that naturally-occurring ADTs are
-predictable.
+predictable. Since the number of weight-$i$ histories is finite\footnote{As
+noted in Section~\ref{sec:histories}, we consider equality between histories up
+to renaming of operation identifiers.} for each $i \in \mathbb{N}$, each $B_i$
+is computable.
 
@@
 
+Predictability ensures that once $B_{i+1}$ does not alter the set of computed
+$\preceq$-minimals, with respect to those of $B_i$, then the minimals computed
+for $B_i$ are exactly those of $B$.
+
 \begin{lemma}
@@
 \end{lemma}
+
+Note that for non-predictable ADTs, computing $B_i$, for any $i \in
+\mathbb{N}$, under-approximates $B$. In practice, computing $B_i$ can be used
+to identify all violations which surface in histories with $i$ matches,
+overlooking violations which only surface with more than $i$ matches. The
+resulting symbolic ADT representation would still be sound in only identifying
+\emph{actual} violations, yet incomplete in identifying \emph{all} violations.

### 2015-03-11 inference.tex
Better story in Section 4.

@@
 Section~\ref{sec:nature} we demonstrate that naturally-occurring ADTs have
-unique kernels. For the remainder of this work, we assume that ADTs have unique
-kernels, and denote \emph{the kernel} of $A$ by $\ker A$. We say that $A$ is
-\emph{atomic} when the width of $\ker A$ is $1$.
+unique, bounded-width kernels. For the remainder of this work, we assume ADTs
+have unique, bounded-width kernels, denoting \emph{the kernel} of $A$ by
+$\ker A$. When $\ker A$ has width $1$, we say $A$ is \emph{atomic}.
 

### 2015-03-11 algorithm.tex
Better titles.

@@
-\section{Approximating ADT Inference}
+\section{Computing Symbolic Representations}
 \label{sec:algorithm}
 
-TODO SHOW HOW TO COMPUTE MINIMAL, FINITE, AND HOPEFULLY COMPLETE ADT
-REPRESENTATIONS WHICH CAN BE USED FOR REFINEMENT CHECKING
+TODO MOTIVATE THIS SECTION
 
-\begin{theorem}
-  “Our algorithm is sound; maybe it is also complete under certain conditions.”
-\end{theorem}
+In addition to the relations $\to_\mathrm{o}$, $\to_\mathrm{p}$, and
+$\to_\mathrm{c}$ of Section~\ref{sec:histories} under which all ADTs are
+closed, the ADTs we consider in this work are also closed under additional
+relations which remove matches, and duplicate operations. Given a history
+$\tup{O,<,c,f,m}$, we say an operation $o \in O$ is a \emph{duplicate} if there
+exists another operation $o' \in O$ with the same labels and matching: $c(o) =
+c(o')$, $f(o) = f(o')$, and $m(o) = m(o')$. The relations $\to_\mathrm{m}$ and
+$\to_\mathrm{d}$ relate two histories $h_1$ and $h_2$ when $h_2$ is obtained
+from $h_1$ by
+\begin{itemize}
+
+  \item removing a match (m), or
+
+  \item removing a duplicate operation (d).
+
+\end{itemize}
+We say an ADT closed under $\to_\mathrm{m}$ and $\to_\mathrm{d}$ is
+\emph{normal}. In Section~\ref{sec:nature} we demonstrate that
+naturally-occurring ADTs are normal. Defining the relation $\succeq$ as the
+reflexive and transitive closure of the above relations,
+\begin{align*}
+  \succeq = (\to_\mathrm{o} \cup \to_\mathrm{p} \cup \to_\mathrm{c} \cup 
+  \to_\mathrm{m} \cup \to_\mathrm{d})^\ast
+\end{align*}
+closure under $\succeq$ follows immediately.
+
+\begin{lemma}
+
+  Normal ADTs are closed under $\succeq$.
+
+\end{lemma}
+
+TODO MOTIVATE WQO
+
+A \emph{well-quasi-ordering (wqo)} $R$ on a set $X$ is a reflexive, transitive
+binary relation on $X$ for which in every infinite sequence $x_0 x_1 \ldots$ of
+elements from $X$, there exists a pair $x_i, x_j$, for $i \neq j$, related by
+$R$.
+
+\begin{example}
+
+  Consider the infinite history sequence $h_1 h_2 \ldots$ where each $h_i$
+  contains $2i$ operations $o_1, o_1', \ldots, o_i, o_i'$ where each $o_j$ is a
+  completed {\tt push} operation matching itself, and each $o_j'$ is a pending
+  {\tt pop} operation with undefined matching. Because each successive $h_i$
+  has both more matches and more pending operations, no two histories of the
+  sequence are related by $\preceq$. Thus $\preceq$ is not a wqo.
+
+\end{example}
+
+The previous example demonstrates that $\preceq$ is not a wqo by allowing each
+history $h_i$ of the infinite sequence to contain more and more pending
+operations in order to ensure that $h_j \not\preceq h_i$ for every $j < i$.
+Curbing this ability by limiting the maximum amount of pending operations per
+history makes $\preceq$ a wqo. Although limiting to $k$ pending operations
+essentially limits us to width-$k$ histories, of executions with at most $k$
+operations parallel at any moment, e.g.,~of programs with at most $k$ threads,
+this restriction comes at no loss of completeness when considering only the
+histories of bounded-width ADT kernels, which is the subject of the remainder
+of this section.
+
+\begin{lemma}
+
+  $\preceq$ is a wqo on width-$k$ histories, for any $k \in \mathbb{N}$.
+
+\end{lemma}
+
+For the remainder of this section, we fix an ADT $A$, and let $B$ be the
+complement of $\ker A$. When $\ker A$ has bounded width, so does $B$, and thus
+$\preceq$ is a wqo on $B$. Furthermore, when $A$ normal, it is closed under
+$\succeq$, and thus $B$ is closed under $\preceq$. Closure under a wqo implies
+representation by a finite set. Formally, we say a set $X$ is \emph{finitely
+representable} if there exists a finite set $Y$ and a relation $R \subseteq Y
+\times X$ such that $X = \set{ x : \exists y\in Y.\ R(y,x) }$. In our case, we
+obtain a finite set from which exactly the elements of $B$ are related by
+$\preceq$.
+
+\begin{lemma}
+
+  $B$ is finitely representable if $A$ is normal.
+
+\end{lemma}
+
+Computing $B$ thus amounts to computing its finite representation. In general,
+this is achieved by enumerating the elements of $B$ while maintaining the
+$\preceq$-minimal elements, and recognizing a condition under which all
+elements of $B$ are related to the current set of
+minimals~\cite{conf/lics/AbdullaCJT96, journals/tcs/FinkelS01}. In our case, we
+recognize a condition on the weight-increasing enumeration of $B$ which
+guarantees coverage by the current set of minimals. Formally, for all $i \in
+\mathbb{N}$, we define
+\begin{align*}
+  B_i &= \set{ h \in B : \weight h \le i } \\
+  B_i' &= \set{ h' \in B :
+    \exists h \in B_i.\ h \preceq h' \text{ and } \weight h \le i+1
+  }
+\end{align*}
+as, respectively, the histories of $B$ with at most $i$ matches, and those
+derived from $B_i$ with at most $i\!+\!1$ matches. We say that $A$ is
+\emph{predictable} if $B_i = B$ whenever $B_i' = B_{i+1}$. In 
+Section~\ref{sec:nature} we demonstrate that naturally-occurring ADTs are
+predictable. Since the number of weight-$i$ histories is finite\footnote{As
+noted in Section~\ref{sec:histories}, we consider equality between histories up
+to renaming of operation identifiers.} for each $i \in \mathbb{N}$, each $B_i$
+is computable.
+
+\begin{lemma}
+
+  $B_i$ is computable, for all $i \in \mathbb{N}$, if $A$ is normal.
+
+\end{lemma}
+
+Predictability ensures that once $B_{i+1}$ does not alter the set of computed
+$\preceq$-minimals, with respect to those of $B_i$, then the minimals computed
+for $B_i$ are exactly those of $B$.
+
+\begin{lemma}
+
+  $B$ is computable if $A$ is normal and predictable.
+
+\end{lemma}
+
+Note that for non-predictable ADTs, computing $B_i$, for any $i \in
+\mathbb{N}$, under-approximates $B$. In practice, computing $B_i$ can be used
+to identify all violations which surface in histories with $i$ matches,
+overlooking violations which only surface with more than $i$ matches. The
+resulting symbolic ADT representation would still be sound in only identifying
+\emph{actual} violations, yet incomplete in identifying \emph{all} violations.

### 2015-03-11 intro.tex
Better titles.

@@
 
-  \item Identification of algebraic properties allowing finite
-  characterizations of infinite ADT violations sets (§\ref{sec:algebra}).
-
-  \item An under-approximating algorithm for the ADT inference problem
-  (§\ref{sec:algorithm}).
+  \item Identification of algebraic properties allowing finite characterization
+  and computability of infinite ADT violations sets (§\ref{sec:algorithm}).
 

### 2015-03-12 algorithm.tex
Added Theorem 1.
Better definition of weight, includes duplicates.

@@
 closed, the ADTs we consider in this work are also closed under additional
-relations which remove matches, and duplicate operations. Given a history
-$\tup{O,<,c,f,m}$, we say an operation $o \in O$ is a \emph{duplicate} if there
-exists another operation $o' \in O$ with the same labels and matching: $c(o) =
-c(o')$, $f(o) = f(o')$, and $m(o) = m(o')$. The relations $\to_\mathrm{m}$ and
-$\to_\mathrm{d}$ relate two histories $h_1$ and $h_2$ when $h_2$ is obtained
-from $h_1$ by
+relations which remove matches, and duplicate operations. The relations
+$\to_\mathrm{m}$ and $\to_\mathrm{d}$ relate two histories $h_1$ and $h_2$ when
+$h_2$ is obtained from $h_1$ by
 \begin{itemize}
@@
 \begin{align*}
-  B_i &= \set{ h \in B : \weight h \le i } \\
+  B_i &= \set{ h \in B : \weight(h) \le i } \\
   B_i' &= \set{ h' \in B :
-    \exists h \in B_i.\ h \preceq h' \text{ and } \weight h \le i+1
+    \exists h \in B_i.\ h \preceq h' \text{ and } \weight(h) \le i+1
   }
 \end{align*}
-as, respectively, the histories of $B$ with at most $i$ matches, and those
-derived from $B_i$ with at most $i\!+\!1$ matches. We say that $A$ is
-\emph{predictable} if $B_i = B$ whenever $B_i' = B_{i+1}$. In 
+as, respectively, the histories of $B$ with at most $i$ matches and duplicates,
+and those derived from $B_i$ with at most $i\!+\!1$ matches and duplicates. We
+say that $A$ is \emph{predictable} if $B_i = B$ whenever $B_i' = B_{i+1}$. In
 Section~\ref{sec:nature} we demonstrate that naturally-occurring ADTs are
@@
 \emph{actual} violations, yet incomplete in identifying \emph{all} violations.
+
+TODO DESCRIBE HOW TO GET THE FORMULA $F$ FROM $B$, AND ARGUE THAT WE HAVE
+SOLVED THE PROBLEM.
+
+\begin{theorem}
+
+  “The symbolic ADT inference problem is computable for normal ADTs.”
+
+\end{theorem}

### 2015-03-12 histories.tex
Better definition of weight, includes duplicates.

@@
 set is the maximum width of its elements. Width-$1$ histories are
-\emph{sequential}. An operation $o \in \rng(m)$ is a \emph{match target}, and
-the set $\set{ o' \in O : m(o') = o}$ of operations targeting $o$ is a
-\emph{match}. We assume targets target themselves: $m(o) = o$ for $o \in
-\rng(m)$. The \emph{weight} of $h$ is the number $|\rng(m)|$ of matches.
-Since the operation identifiers of a history have no intrinsic meaning, we
-consider equality between histories up to renaming of operation identifiers.
+\emph{sequential}. Two operations with the same labels and matching are
+\emph{identical}, and the \emph{frequency} of an operation $o$, denoted
+$\freq(o)$, is the number of operations identical to $o$, including $o$ itself.
+An operation is a \emph{duplicate} when $\freq(o) > 1$. An operation $o \in
+\rng(m)$ is a \emph{match target}, and the set $\set{ o' \in O : m(o') = o}$ of
+operations targeting $o$ is a \emph{match}. We assume targets target
+themselves: $m(o) = o$ for $o \in \rng(m)$. The \emph{weight} of $h$ is the
+maximum $\max(\set{ \freq(o) : o \in O } \cup \set{|\rng(m)|})$ between
+operation frequencies and the number of matches. Since the operation
+identifiers of a history have no intrinsic meaning, we consider equality
+between histories up to renaming of operation identifiers.
 

### 2015-03-13 algorithm.tex
Fixed typo.
Fixed a couple of definitions.

@@
 binary relation on $X$ for which in every infinite sequence $x_0 x_1 \ldots$ of
-elements from $X$, there exists a pair $x_i, x_j$, for $i \neq j$, related by
-$R$.
+elements from $X$, there exists indices $i < j$ such that $R(x_i,x_j)$.
 
@@
 
-  $\preceq$ is a wqo on width-$k$ histories, for any $k \in \mathbb{N}$.
+  $\preceq$ is a wqo on width-bounded histories.
 
@@
   B_i' &= \set{ h' \in B :
-    \exists h \in B_i.\ h \preceq h' \text{ and } \weight(h) \le i+1
+    \exists h \in B_i.\ h \preceq h' \text{ and } \weight(h') \le i+1
   }

### 2015-03-13 histories.tex
Fixed a couple of definitions.

--- 2015-03-11 histories.tex
@@
 set is the maximum width of its elements. Width-$1$ histories are
-\emph{sequential}. An operation $o \in \rng(m)$ is a \emph{match target}, and
-the set $\set{ o' \in O : m(o') = o}$ of operations targeting $o$ is a
-\emph{match}. We assume targets target themselves: $m(o) = o$ for $o \in
-\rng(m)$. The \emph{weight} of $h$ is the number $|\rng(m)|$ of matches.
-Since the operation identifiers of a history have no intrinsic meaning, we
-consider equality between histories up to renaming of operation identifiers.
+\emph{sequential}.
+Two operations $o, o'$ are \emph{identical}, written $o \sim o'$, when they
+have the same labels: $c(o) = c(o')$, $f(o) = f(o')$, and matching: $m(o) =
+m(o')$. The \emph{frequency} of an operation $o$ is the number $\freq(o) =
+|\set{ o' \in O : o \sim o' }|$ of operations identical to $o$, and $o$ has
+\emph{duplicates} when $\freq(o) > 1$. An operation $o \in \rng(m)$ is a
+\emph{match target}, and the set $\set{ o' \in O : m(o') = o}$ of operations
+targeting $o$ is a \emph{match}. We assume targets target themselves: $m(o) =
+o$ for $o \in \rng(m)$. The \emph{weight} of $h$ is the maximum $\max(\set{
+\freq(o) : o \in O } \cup \set{|\rng(m)|})$ between operation frequencies and
+the number of matches. Since the operation identifiers of a history have no
+intrinsic meaning, we consider equality between histories up to renaming of
+operation identifiers.
 

### 2015-03-16 algorithm.tex
formulas

@@
+%!TEX root = draft.tex
 \section{Computing Symbolic Representations}
@@
 
+Given a history $h = \tup{O,<,c,f,m}$, we define a formula $\textsc{Stronger}(h)$
+describing all histories stronger than $h$ w.r.t. $\preceq$. The set of variables $X_O$
+contains one variable $x_o$ for each $o\in O$.
+
+\begin{align*}
+\exists X_O. & \bigwedge_{o\in O} {\sf meth}(x_o)=f(o) \land \bigwedge_{o_1<o_2} x_{o_1} < x_{o_2}\land \bigwedge_{c(o)} {\sf c}(x_o) \\
+& \land \bigwedge_{m(o_1)=o_2} {\sf match}(x_{o_1}) = x_{o_2} \\
+& \land  \bigwedge_{o\in O} \forall x.\ {\sf match}(x) = x_o\Rightarrow \textsc{Dupl}(x,x_o)\vee {\sf rd}(x) 
+\end{align*}
+
+\begin{align*}
+\textsc{Dupl}(x,x_o) \triangleq \bigvee_{o_1\in O} {\sf meth}(x_{o_1}) = {\sf meth}(x) \land {\sf match}(x_{o_1})=x_o
+\end{align*}
+
+\begin{lemma}
+
+For every two histories $h$ and $h'$,
+\begin{align*}
+h'\models \textsc{Stronger}(h) \mbox{ implies } h\preceq h'
+\end{align*}
+
+\end{lemma}
+
+Let $Min(B)$ be the set of minimal elements of $B$. Then, the formula $F$ representing the ADT A is defined as:
+\begin{align*}
+F \triangleq \neg \bigvee_{h\in Min(B)} \textsc{Stronger}(h)
+\end{align*}
+
+\begin{theorem}
+
+  The formula $F$ represents $A$.
+
+\end{theorem}
+
 \begin{theorem}

### 2015-03-16 algorithm.tex
Soft description of the formulas.

--- 2015-03-13 algorithm.tex
@@
 
-TODO DESCRIBE HOW TO GET THE FORMULA $F$ FROM $B$, AND ARGUE THAT WE HAVE
-SOLVED THE PROBLEM.
+WRITE A CLEAN SEGUE TO THE NEXT PART
+
+Let $h = \tup{O,<,c,f,m}$ be a history with operations $O = \set{ o_1, \ldots,
+o_n }$. Without loss of generality, we assume the match targets $\set{o_1,
+\ldots, o_k}$ are indexed consecutively from $1$ to $k$, for some $k < n$. We
+define the macro $\textsc{Embed}_o(x,Y,z)$:
+\begin{align*}
+  & (c(o) \implies {\sf c}(x)) \land {\sf f}(x) = f(o) \\
+  & \land (o \in \dom(m) \implies \lnot{\sf um}(x) \land {\sf m}(x) = z) \\
+  & \land (o \not\in \dom(m) \implies {\sf um}(x)) \land \bigwedge_{y \in Y} x < y
+\end{align*}
+capturing the correspondence between the logical variable $x$ and operation
+$o$. The variables $Y$ and $z$ represent the operations ordered after $x$, and
+the operation which $x$ matches. The macro $\textsc{Identical}(x,y)$, defined as
+\begin{align*}
+  {\sf f}(x) = {\sf f}(y)
+  \land {\sf c}(x) \Leftrightarrow {\sf c}(y)
+  \land {\sf m}(x) = {\sf m}(y)
+\end{align*}
+captures whether the operations bound to $x$ and $y$ are identical.
+To express the constraints among the matches of an embedded history, we define
+the macro $\textsc{Match}(Y,z)$:
+\begin{align*}
+  \forall x.\ {\sf m}(x) = z \implies
+  \textsc{ReadOnly}(x) \lor \bigvee_{y \in Y} \textsc{Identical}(x,y)
+\end{align*}
+which requires any operation which matches $z$ to be either read-only, or
+identical to some operation in $Y$, which represents the operations of $h$
+which match $z$. Finally, we express the embedding of $h$ with the macro
+$\textsc{Embed}_h$:
+\begin{align*}
+  \exists x_1, \ldots, x_n.\ \bigwedge_{i=1}^{n} \textsc{Embed}_{o_i}(x_i,Y_i,z_i)
+  \land \bigwedge_{i=1}^{k} \textsc{Match}(M_i, x_i)
+\end{align*}
+where $Y_i = \set{ x_j : o_i < o_j }$ are the variables corresponding to
+operations ordered after $o_i$, and $z_i$ is the variable corresponding to
+$m(o_i)$, if defined, and $M_i = \set{ x_j : m(o_j) = o_i }$ are the variables
+corresponding to operations matching $o_i$.
+
+\begin{lemma}
+
+  $h_1 \models \textsc{Embed}_{h_2}$ if{f} $h_1 \succeq h_2$.
+
+\end{lemma}
+
+We obtain a formula representing an ADT $A$ by taking the conjunction of the
+negative embedding formulæ $\set{ \lnot\textsc{Embed}_h : h \in H }$ from any
+set $H$ that generates the complement of $A$.
+
+\begin{lemma}
+
+  The formula $\bigwedge_{h \in H} \neg \textsc{Embed}_h$ represents $A$,
+  for any set $H$ which generates the compliment of $A$.
+
+\end{lemma}
+
+For normal and predictable ADTs, such sets are computable, and so our inference
+problem is also computable.
 
@@
 
-  “The symbolic ADT inference problem is computable for normal ADTs.”
+  The symbolic ADT inference problem is computable for normal and predictable ADTs.
 

### 2015-03-16 formula.tex
Split algorithm section in two.

--- 2015-03-16 algorithm.tex
@@
-%!TEX root = draft.tex
 \section{Computing Symbolic Representations}
-\label{sec:algorithm}
+\label{sec:formula}
 
 TODO MOTIVATE THIS SECTION
-
-In addition to the relations $\to_\mathrm{o}$, $\to_\mathrm{p}$, and
-$\to_\mathrm{c}$ of Section~\ref{sec:histories} under which all ADTs are
-closed, the ADTs we consider in this work are also closed under additional
-relations which remove matches, and duplicate operations. The relations
-$\to_\mathrm{m}$ and $\to_\mathrm{d}$ relate two histories $h_1$ and $h_2$ when
-$h_2$ is obtained from $h_1$ by
-\begin{itemize}
-
-  \item removing a match (m), or
-
-  \item removing a duplicate operation (d).
-
-\end{itemize}
-We say an ADT closed under $\to_\mathrm{m}$ and $\to_\mathrm{d}$ is
-\emph{normal}. In Section~\ref{sec:nature} we demonstrate that
-naturally-occurring ADTs are normal. Defining the relation $\succeq$ as the
-reflexive and transitive closure of the above relations,
-\begin{align*}
-  \succeq = (\to_\mathrm{o} \cup \to_\mathrm{p} \cup \to_\mathrm{c} \cup 
-  \to_\mathrm{m} \cup \to_\mathrm{d})^\ast
-\end{align*}
-closure under $\succeq$ follows immediately.
-
-\begin{lemma}
-
-  Normal ADTs are closed under $\succeq$.
-
-\end{lemma}
-
-TODO MOTIVATE WQO
-
-A \emph{well-quasi-ordering (wqo)} $R$ on a set $X$ is a reflexive, transitive
-binary relation on $X$ for which in every infinite sequence $x_0 x_1 \ldots$ of
-elements from $X$, there exists indices $i < j$ such that $R(x_i,x_j)$.
-
-\begin{example}
-
-  Consider the infinite history sequence $h_1 h_2 \ldots$ where each $h_i$
-  contains $2i$ operations $o_1, o_1', \ldots, o_i, o_i'$ where each $o_j$ is a
-  completed {\tt push} operation matching itself, and each $o_j'$ is a pending
-  {\tt pop} operation with undefined matching. Because each successive $h_i$
-  has both more matches and more pending operations, no two histories of the
-  sequence are related by $\preceq$. Thus $\preceq$ is not a wqo.
-
-\end{example}
-
-The previous example demonstrates that $\preceq$ is not a wqo by allowing each
-history $h_i$ of the infinite sequence to contain more and more pending
-operations in order to ensure that $h_j \not\preceq h_i$ for every $j < i$.
-Curbing this ability by limiting the maximum amount of pending operations per
-history makes $\preceq$ a wqo. Although limiting to $k$ pending operations
-essentially limits us to width-$k$ histories, of executions with at most $k$
-operations parallel at any moment, e.g.,~of programs with at most $k$ threads,
-this restriction comes at no loss of completeness when considering only the
-histories of bounded-width ADT kernels, which is the subject of the remainder
-of this section.
-
-\begin{lemma}
-
-  $\preceq$ is a wqo on width-bounded histories.
-
-\end{lemma}
-
-For the remainder of this section, we fix an ADT $A$, and let $B$ be the
-complement of $\ker A$. When $\ker A$ has bounded width, so does $B$, and thus
-$\preceq$ is a wqo on $B$. Furthermore, when $A$ normal, it is closed under
-$\succeq$, and thus $B$ is closed under $\preceq$. Closure under a wqo implies
-representation by a finite set. Formally, we say a set $X$ is \emph{finitely
-representable} if there exists a finite set $Y$ and a relation $R \subseteq Y
-\times X$ such that $X = \set{ x : \exists y\in Y.\ R(y,x) }$. In our case, we
-obtain a finite set from which exactly the elements of $B$ are related by
-$\preceq$.
-
-\begin{lemma}
-
-  $B$ is finitely representable if $A$ is normal.
-
-\end{lemma}
-
-Computing $B$ thus amounts to computing its finite representation. In general,
-this is achieved by enumerating the elements of $B$ while maintaining the
-$\preceq$-minimal elements, and recognizing a condition under which all
-elements of $B$ are related to the current set of
-minimals~\cite{conf/lics/AbdullaCJT96, journals/tcs/FinkelS01}. In our case, we
-recognize a condition on the weight-increasing enumeration of $B$ which
-guarantees coverage by the current set of minimals. Formally, for all $i \in
-\mathbb{N}$, we define
-\begin{align*}
-  B_i &= \set{ h \in B : \weight(h) \le i } \\
-  B_i' &= \set{ h' \in B :
-    \exists h \in B_i.\ h \preceq h' \text{ and } \weight(h') \le i+1
-  }
-\end{align*}
-as, respectively, the histories of $B$ with at most $i$ matches and duplicates,
-and those derived from $B_i$ with at most $i\!+\!1$ matches and duplicates. We
-say that $A$ is \emph{predictable} if $B_i = B$ whenever $B_i' = B_{i+1}$. In
-Section~\ref{sec:nature} we demonstrate that naturally-occurring ADTs are
-predictable. Since the number of weight-$i$ histories is finite\footnote{As
-noted in Section~\ref{sec:histories}, we consider equality between histories up
-to renaming of operation identifiers.} for each $i \in \mathbb{N}$, each $B_i$
-is computable.
-
-\begin{lemma}
-
-  $B_i$ is computable, for all $i \in \mathbb{N}$, if $A$ is normal.
-
-\end{lemma}
-
-Predictability ensures that once $B_{i+1}$ does not alter the set of computed
-$\preceq$-minimals, with respect to those of $B_i$, then the minimals computed
-for $B_i$ are exactly those of $B$.
-
-\begin{lemma}
-
-  $B$ is computable if $A$ is normal and predictable.
-
-\end{lemma}
-
-Note that for non-predictable ADTs, computing $B_i$, for any $i \in
-\mathbb{N}$, under-approximates $B$. In practice, computing $B_i$ can be used
-to identify all violations which surface in histories with $i$ matches,
-overlooking violations which only surface with more than $i$ matches. The
-resulting symbolic ADT representation would still be sound in only identifying
-\emph{actual} violations, yet incomplete in identifying \emph{all} violations.
-
-WRITE A CLEAN SEGUE TO THE NEXT PART
 
@@
 \begin{align*}
-  {\sf f}(x) = {\sf f}(y)
-  \land {\sf c}(x) \Leftrightarrow {\sf c}(y)
+  ({\sf c}(x) \Leftrightarrow {\sf c}(y))
+  \land {\sf f}(x) = {\sf f}(y)
   \land {\sf m}(x) = {\sf m}(y)
@@
 
+\begin{example}
+
+  TODO EXAMPLE FORMULÆ FROM PREVIOUS PATTERNS
+
+\end{example}
+
 \begin{lemma}
@@
 
+\begin{example}
+
+  TODO EXAMPLE OF COMPLETE FORMULA FROM PREVIOUS COMPLETE PATTERNS/FORMULÆ
+
+\end{example}
+
 For normal and predictable ADTs, such sets are computable, and so our inference

### 2015-03-16 intro.tex
Split algorithm section in two.

@@
   \item Identification of algebraic properties allowing finite characterization
-  and computability of infinite ADT violations sets (§\ref{sec:algorithm}).
+  and computability of infinite ADT violations sets
+  (§\ref{sec:patterns}–\ref{sec:formula}).
 

### 2015-03-16 formula.tex
Formula readability.

@@
 o_n }$. Without loss of generality, we assume the match targets $\set{o_1,
-\ldots, o_k}$ are indexed consecutively from $1$ to $k$, for some $k < n$. We
-define the macro $\textsc{Embed}_o(x,Y,z)$:
+\ldots, o_k}$ are indexed consecutively from $1$ to $k$, for some $k \le n$. 
+For each $o \in O$, we define the macro $\textsc{Embed}_o(x,Y,z)$:
 \begin{align*}
-  & (c(o) \implies {\sf c}(x)) \land {\sf f}(x) = f(o) \\
-  & \land (o \in \dom(m) \implies \lnot{\sf um}(x) \land {\sf m}(x) = z) \\
-  & \land (o \not\in \dom(m) \implies {\sf um}(x)) \land \bigwedge_{y \in Y} x < y
+  & \big(c(o) \implies {\sf c}(x)\big) \land {\sf f}(x) = f(o) \\
+  & \land \big(o \in \dom(m)
+    \implies \lnot{\sf um}(x) \land {\sf m}(x) = z\big) \\
+  & \land \big(o \not\in \dom(m) \implies {\sf um}(x)\big)
+    \land \bigwedge_{y \in Y} x < y
 \end{align*}
-capturing the correspondence between the logical variable $x$ and operation
-$o$. The variables $Y$ and $z$ represent the operations ordered after $x$, and
-the operation which $x$ matches. The macro $\textsc{Identical}(x,y)$, defined as
+capturing the correspondence between the operation $o$ and the logical variable
+$x$. The variables $Y$ and $z$ represent the operations ordered after $x$, and
+the operation which $x$ matches. The macro $\textsc{Identical}(x,y)$:
 \begin{align*}
-  ({\sf c}(x) \Leftrightarrow {\sf c}(y))
-  \land {\sf f}(x) = {\sf f}(y)
+  & \big({\sf c}(x) \Leftrightarrow {\sf c}(y)\big)
+  \land {\sf f}(x) = {\sf f}(y) \\
+  & \land \big({\sf um}(x) \Leftrightarrow {\sf um}(y)\big)
   \land {\sf m}(x) = {\sf m}(y)
@@
 captures whether the operations bound to $x$ and $y$ are identical.
-To express the constraints among the matches of an embedded history, we define
+To express the constraints among the matches of embedded operations, we define
 the macro $\textsc{Match}(Y,z)$:

### 2015-03-16 histories.tex
some small stuff

@@
+%!TEX root = draft.tex
 \section{Histories of ADT Implementations}
@@
 have the same labels: $c(o) = c(o')$, $f(o) = f(o')$, and matching: $m(o) =
-m(o')$. The \emph{frequency} of an operation $o$ is the number $\freq(o) =
+m(o')$ (TODO: if matching is defined).
+
+The \emph{frequency} of an operation $o$ is the number $\freq(o) =
 |\set{ o' \in O : o \sim o' }|$ of operations identical to $o$, and $o$ has

### 2015-03-16 formula.tex
Tiny typo fixes.

@@
 o_n }$. Without loss of generality, we assume the match targets $\set{o_1,
-\ldots, o_k}$ are indexed consecutively from $1$ to $k$, for some $k \le n$. 
-For each $o \in O$, we define the macro $\textsc{Embed}_o(x,Y,z)$:
+\ldots, o_k}$ of $h$ are indexed consecutively from $1$ to $k$, for some $k \le
+n$. For each $o \in O$, we define the macro $\textsc{Embed}_o(x,Y,z)$:
 \begin{align*}
@@
 \begin{align*}
-  \forall x.\ {\sf m}(x) = z \implies
+  \forall x.
+  {\sf m}(x) = z \implies
   \textsc{ReadOnly}(x) \lor \bigvee_{y \in Y} \textsc{Identical}(x,y)
@@
 \begin{align*}
-  \exists x_1, \ldots, x_n.\ \bigwedge_{i=1}^{n} \textsc{Embed}_{o_i}(x_i,Y_i,z_i)
-  \land \bigwedge_{i=1}^{k} \textsc{Match}(M_i, x_i)
+  \exists x_1, \ldots, x_n.
+  \bigwedge_{i=1}^{n} \textsc{Embed}_{o_i}(x_i,Y_i,z_i)
+  \land \bigwedge_{i=1}^{k} \textsc{Match}(W_i, x_i)
 \end{align*}
@@
 operations ordered after $o_i$, and $z_i$ is the variable corresponding to
-$m(o_i)$, if defined, and $M_i = \set{ x_j : m(o_j) = o_i }$ are the variables
+$m(o_i)$, if defined, and $W_i = \set{ x_j : m(o_j) = o_i }$ are the variables
 corresponding to operations matching $o_i$.

### 2015-03-17 matching.tex
Better titles.
Definitions for matching functions.

@@
-\section{Inferring Matching Functions}
+\section{Computing Matching Schemes}
 \label{sec:matching}
 
-TODO SHOW HOW TO COMPUTE MATCHING FUNCTIONS FROM THE SPECIFICATION THAT OBJECTS
-SHOULD BE MATCH-REMOVAL/EMBEDDING CLOSED WRT A GIVEN MATCHING FUNCTION
+TODO MOTIVATE THIS SECTION
+
+When $P$ is a $k$-length sequence of $n$-ary predicates, we write $P(x_1,
+\ldots, x_n)$ to denote the sequence
+\begin{align*}
+  P_1(x_1, \ldots, x_n) \ldots P_k(x_1, \ldots, x_n)
+\end{align*}
+Given finite predicate sequences $P$ and $Q$ and a set $E$ of executions, a
+matching scheme $M$ is \emph{simple} when there exists an $n$-ary Boolean
+function $G$, for $n = 2\cdot|P|+|Q|$, such that for each execution $e \in E$
+\begin{itemize}
+
+  \item $G(P(e,o_1),P(e,o_2),Q(e,o_1,o_2))$ is satisfied for at most one
+  operation $o_1$, for any operation $o_2$,
+
+  \item $M(e)(o_2)$ is undefined unless there exists an operation $o_1$
+  for which $G(P(e,o_1),P(e,o_2),Q(e,o_1,o_2))$, and
+
+  \item $M(e)(o_2) = o_1$ if{f} $G(P(e,o_1),P(e,o_2),P(e,o_1,o_2))$,
+
+\end{itemize}
+where $o_1$ and $o_2$ range over the operations of $e$.
+
+TODO QUALIFY THE EXECUTION SETS, E.G., TO HAVE UNIQUE VALUES, AND SO ON
+
+\begin{example}
+
+  TODO SHOW A FINITARY SIMPLE MATCHING SCHEME
+
+\end{example}
+
+\begin{definition}
+
+  The \emph{matching scheme inference problem} is to compute a faithful, simple
+  matching scheme $M$ for a given set $E$ of executions and finite predicate
+  sequences $P$ and $Q$.
+
+\end{definition}
 
 \begin{theorem}
-  “Our algorithm is sound; maybe it is also complete under certain conditions.”
+
+  The matching scheme inference problem is computable for \ldots
+
 \end{theorem}

### 2015-03-17 matching.tex
Example for simple matching scheme.

@@
 When $P$ is a $k$-length sequence of $n$-ary predicates, we write $P(x_1,
-\ldots, x_n)$ to denote the sequence
+\ldots, x_n)$ to denote the valuation sequence
 \begin{align*}
-  P_1(x_1, \ldots, x_n) \ldots P_k(x_1, \ldots, x_n)
+  P_1(x_1, \ldots, x_n) \ldots P_k(x_1, \ldots, x_n).
 \end{align*}
@@
 
-  TODO SHOW A FINITARY SIMPLE MATCHING SCHEME
+  We say an execution of the atomic register ADT \emph{writes unique values}
+  if the argument value to each write operation is unique. Consider the set $E$
+  of executions which write unique values, and the following predicates:
+  \begin{align*}
+    & \mathsf{write}(e,o) & \text{$o$ is a {\sf write} operation in $e$}, \\
+    & \mathsf{read}(e,o) & \text{$o$ is a {\sf read} operation in $e$,} \\
+    & \mathsf{eq}(e,o_1,o_2) & \text{$o_1$ and $o_2$ the same operation, and} \\
+    & \mathsf{veq}(e,o_1,o_2) & \text{$o_1$ and $o_2$ read/write the same values.}
+  \end{align*}
+  Given a valuation $\tup{x_\mathsf{w}, x_\mathsf{r}, y_\mathsf{w},
+  y_\mathsf{r}, z_\mathsf{=}, z_\mathsf{v}}$ of the predicates above, we define
+  the function $G(x_\mathsf{w}, x_\mathsf{r}, y_\mathsf{w}, y_\mathsf{r},
+  z_\mathsf{=}, z_\mathsf{v})$ to be satisfied if and only if:
+  \begin{align*}
+    \big( x_\mathsf{w} \land z_\mathsf{=} \big)
+    \lor \big( x_\mathsf{w} \land y_\mathsf{r} \land z_\mathsf{v} \big)
+  \end{align*}
+  Intuitively, this defines a simple matching scheme for which write operations
+  match themselves, and read operations match the write which wrote the value
+  read. In the case such a write exists, it is unique in any execution which
+  writes unique values. The match is otherwise undefined.
+
+  TODO ARGUE THAT THIS IS A FAITHFUL MATCHING SCHEME
 

### 2015-03-17 matching.tex
Matching schemes normalize implementations.

--- 2015-03-17 matching.tex
@@
 
-When $P$ is a $k$-length sequence of $n$-ary predicates, we write $P(x_1,
-\ldots, x_n)$ to denote the sequence
+A \emph{language} $L = \tup{E,P,Q}$ is a set $E$ of executions, along with
+finite sequences $P = P_1 P_2 \ldots$ and $Q = Q_1 Q_2 \ldots$ of binary and
+ternary predicates $P_i(e,o_1)$ and $Q_i(e,o_1,o_2)$ ranging over the
+executions of $E$ and their operations. When $P$ is a $k$-length sequence of
+$n$-ary predicates, we write $P(x_1, \ldots, x_n)$ to denote the valuation
+sequence
 \begin{align*}
-  P_1(x_1, \ldots, x_n) \ldots P_k(x_1, \ldots, x_n)
+  P_1(x_1, \ldots, x_n) \ldots P_k(x_1, \ldots, x_n).
 \end{align*}
-Given finite predicate sequences $P$ and $Q$ and a set $E$ of executions, a
-matching scheme $M$ is \emph{simple} when there exists an $n$-ary Boolean
-function $G$, for $n = 2\cdot|P|+|Q|$, such that for each execution $e \in E$
+Given a language $L = \tup{E,P,Q}$, we say a matching scheme $M$ is
+\emph{simple} when there exists an $n$-ary Boolean function $G$, for $n =
+2\cdot|P|+|Q|$, such that for each execution $e \in E$
 \begin{itemize}
@@
 
-TODO QUALIFY THE EXECUTION SETS, E.G., TO HAVE UNIQUE VALUES, AND SO ON
-
 \begin{example}
 
-  TODO SHOW A FINITARY SIMPLE MATCHING SCHEME
+  We say an execution with methods $\set{\mathsf{write}, \mathsf{read}}$
+  \emph{writes unique values} if the argument value to each write operation is
+  unique. Consider the language $L$ whose executions write unique values, with
+  the following predicates:
+  \begin{align*}
+    & \mathsf{write}(e,o) & \text{$o$ is a {\sf write} operation in $e$}, \\
+    & \mathsf{read}(e,o) & \text{$o$ is a {\sf read} operation in $e$,} \\
+    & \mathsf{eq}(e,o_1,o_2) & \text{$o_1$ and $o_2$ are the same operation, and} \\
+    & \mathsf{veq}(e,o_1,o_2) & \text{$o_1$ and $o_2$ read/write the same values.}
+  \end{align*}
+  Given a valuation $\tup{x_\mathsf{w}, x_\mathsf{r}, y_\mathsf{w},
+  y_\mathsf{r}, z_\mathsf{=}, z_\mathsf{v}}$ of the predicates above, we define
+  the function $G(x_\mathsf{w}, x_\mathsf{r}, y_\mathsf{w}, y_\mathsf{r},
+  z_\mathsf{=}, z_\mathsf{v})$ to be satisfied if and only if:
+  \begin{align*}
+    ( x_\mathsf{w} \land z_\mathsf{=} )
+    \lor ( x_\mathsf{w} \land y_\mathsf{r} \land z_\mathsf{v} )
+  \end{align*}
+  Intuitively, this defines a simple matching scheme for $L$ for which write
+  operations match themselves, and read operations match the write which wrote
+  the value read. In the case such a write exists, it is unique in any
+  execution which writes unique values. The match is otherwise undefined.
+
+  TODO ARGUE THAT THIS IS A FAITHFUL MATCHING SCHEME
 
@@
 
+An implementation $\mathcal{I}$ \emph{adheres} to a language $L = \tup{E,P,Q}$
+if $\mathcal{I} \subseteq E$. A match scheme $M$ \emph{normalizes} an
+implementation $\mathcal{I}$ when $M$ is faithful to $\mathcal{I}$ and
+$H(\mathcal{I},M)$ is normal.
+
 \begin{definition}
 
-  The \emph{matching scheme inference problem} is to compute a faithful, simple
-  matching scheme $M$ for a given set $E$ of executions and finite predicate
-  sequences $P$ and $Q$.
+  The \emph{matching scheme inference problem} is to compute a simple matching
+  scheme $M$ which normalizes a given implementation $\mathcal{I}$ adhering to
+  a given language $L$.
 
 \end{definition}
+
+TODO ARGUE THAT SIMPLE NORMALIZING MATCHING SCHEMES CAN BE
+COMPUTED/APPROXIMATED BY CHECKING WHETHER FINITE SETS OF EXECUTIONS AND THEIR
+PROJECTIONS ARE MATCH-REMOVAL CLOSED
 

### 2015-03-17 matching.tex
Added placeholder for normalization example.
Quick fixes.

@@
 
-  We say an execution with methods $\set{\mathsf{write}, \mathsf{read}}$
-  \emph{writes unique values} if the argument value to each write operation is
-  unique. Consider the language $L$ whose executions write unique values, with
-  the following predicates:
+  We say an execution of read and write methods \emph{writes unique values} if
+  the argument value to each write operation is unique. Consider the language
+  whose executions write unique values, with the following predicates:
   \begin{align*}
@@
   \end{align*}
-  Intuitively, this defines a simple matching scheme for $L$ for which write
+  Intuitively, this defines a simple matching scheme for which write
   operations match themselves, and read operations match the write which wrote
@@
   execution which writes unique values. The match is otherwise undefined.
-
-  TODO ARGUE THAT THIS IS A FAITHFUL MATCHING SCHEME
 
@@
 $H(\mathcal{I},M)$ is normal.
+
+\begin{example}
+
+  TODO SHOW THAT PREVIOUS SIMPLE MATCHING SCHEME NORMALIZES SOMETHING
+
+\end{example}
 

### 2015-03-17 patterns.tex
Split algorithm section in two.
Added a figure.

--- 2015-03-13 algorithm.tex
@@
-\section{Computing Symbolic Representations}
-\label{sec:algorithm}
+\section{Computing ADT Complements}
+\label{sec:patterns}
 
-TODO MOTIVATE THIS SECTION
+TODO MOTIVATE THIS SECTION; ALSO, RETHINK THE TITLE OF THIS SECTION
 
@@
 
-TODO DESCRIBE HOW TO GET THE FORMULA $F$ FROM $B$, AND ARGUE THAT WE HAVE
-SOLVED THE PROBLEM.
+\begin{example}
 
-\begin{theorem}
+  TODO EXAMPLE OF COMPLETE PATTERNS
 
-  “The symbolic ADT inference problem is computable for normal ADTs.”
+  \begin{figure}[t]
+    \centering
+    \includegraphics[scale=0.2]{figures/register-pattern-1},
+    \includegraphics[scale=0.2]{figures/register-pattern-2},
+    \includegraphics[scale=0.2]{figures/register-pattern-3}
+    \caption{Three histories which generate the complement of the atomic
+      register ADT kernel.}
+    \label{fig:register-patterns}
+  \end{figure}
 
-\end{theorem}
+\end{example}

### 2015-03-17 matching.tex
Outline of matching scheme inference.

@@
 
+$\textsc{IsMatch}_{e,O}$:
+\begin{align*}
+  \bigvee_{o_1 \in O} \bigwedge_{o_2 \in O}
+  \mathsf{g}(P(e,o_1), P(e,o_2), Q(e,o_1,o_2))
+\end{align*}
+represents whether $O$ is a match of $e$ according to the simple matching
+scheme generated by $\mathsf{g}$, and
+\begin{align*}
+  \bigwedge_{\tup{e,O}\in \_} \lnot \textsc{IsMatch}_{e,O}
+\end{align*}
+insists that for any execution $e \in \mathcal{I}$ whose projection to
+operations not in $O$ is not in $\mathcal{I}$, then $O$ must not be a match.
+
+ENUMERATE LOTS OF EXECUTION PAIRS $\tup{e,e'}$ FOR WHICH $e'$ IS A PROJECTION
+OF $e'$ WITH $e \in \mathcal{I}$ AND $e' \not\in \mathcal{I}$, AND LET $O$ BE
+THE OPERATIONS IN $e$ BUT NOT $e'$. HOPEFULLY IF THIS FORMULA IS SATISFIABLE, A
+MODEL FOR {\sf g} CAN BE A GOOD SIMPLE MATCHING SCHEME
+
 \begin{theorem}

### 2015-03-18 matching.tex
A reasonable story for matching scheme computation.

--- 2015-03-17 matching.tex
@@
 
-TODO ARGUE THAT SIMPLE NORMALIZING MATCHING SCHEMES CAN BE
-COMPUTED/APPROXIMATED BY CHECKING WHETHER FINITE SETS OF EXECUTIONS AND THEIR
-PROJECTIONS ARE MATCH-REMOVAL CLOSED
+We automate the computation of matching schemes by considering the function
+${\sf g}$ underlying a simple matching scheme as an uninterpreted function in a
+logical satisfiability problem, and constrain ${\sf g}$ to normalize the
+executions of an implementation. To this end, we fix an implementation
+$\mathcal{I}$ and a language $L = \tup{E,P,Q}$ to which $\mathcal{I}$ adheres,
+then consider any enumeration $F$ of execution pairs $\tup{e,e'} \in E^2$ such
+that
+\begin{itemize}
 
-\begin{theorem}
+  \item $e \in \mathcal{I}$ and $e' \not\in \mathcal{I}$, and
 
-  The matching scheme inference problem is computable for \ldots
+  \item $e'$ is obtained by deleting operations of $e$.
 
-\end{theorem}
+\end{itemize}
+Any such pair of executions can be used to rule out several possibilities,
+for instance, that the operations removed from $e$ to obtain $e'$:
+\begin{itemize}
+
+  \item do not constitute a match in $e$,
+
+  \item are not all duplicate operations,
+
+  \item are not all read-only operations,
+
+  \item do not constitute multiple matches in $e$,
+
+\end{itemize}
+and so on. Ruling out these possibilities is sound since, for example, a
+normalized scheme $M$ could not consider those operations a match: otherwise
+the history abstraction $H(\mathcal{I},M)$ which includes $H(e,M)$ must also
+include $H(e',M)$, being normal, and in particular closed under match removal.
+Such an $M$ would thus not be faithful. For simplicity, in what follows we
+consider ruling out only the first possibility: that the operations removed
+from $e$ are not a match. In principle, the approach extends to rule out all
+possibilities.
+
+In what follows, we denote the operations of an execution $e$ by $O_e$.
+To consider whether a given pair $o_1, o_2$ of operations of an execution $e$
+is a match according to the simple matching scheme based on ${\sf g}$, for
+the given language $L$, we define the macro $\textsc{IsMatch}_{e,o_1,o_2}$:
+\begin{align*}
+  \mathsf{g}\big( P(e,o_1), P(e,o_2), Q(e,o_1,o_2) \big).
+\end{align*}
+Then a given set $O \subseteq O_e$ constitutes a match according to
+${\sf g}$ when all operations $o_2 \in O$ target some operation $o_1 \in O$,
+and no other operation $o_2 \in O \setminus O_e$ does. We express this with the
+macro $\textsc{EntireMatch}_{e,O}$:
+\begin{align*}
+  \bigvee_{o_1 \in O}
+  \Big(
+  \bigwedge_{o_2 \in O} \textsc{IsMatch}_{e,o_1,o_2}
+  \land \bigwedge_{o_2 \in O_e \setminus O} \lnot \textsc{IsMatch}_{e,o_1,o_2}
+  \Big)
+\end{align*}
+Finally, given a pair $\tup{e,e'} \in F$, we prohibit the operations $O_e
+\setminus O_{e'}$ from constituting a match according to ${\sf g}$, since if
+${\sf g}$ normalized $\mathcal{I}$, and $e \in \mathcal{I}$, then $e'$ should
+also be in $\mathcal{I}$. We express this exclusion for all pairs of $F$
+with the macro $\textsc{Normalizes}_{F}$:
+\begin{align*}
+  \bigwedge_{\tup{e,e'}\in F} \lnot \textsc{EntireMatch}_{e,(O_e \setminus O_{e'})}
+\end{align*}
+We thus check satisfiability for the conjunction of non-matches $O_e \setminus
+O_{e'}$ over all pairs $\tup{e,e'} \in F$.
+
+On the one hand, checking satisfiability of $\textsc{Normalizes}_F$ can be used
+to conclude the impossibility of a good matching scheme — at least for the
+given language.
+
+\begin{lemma}
+
+  If $\textsc{Normalizes}_F$ is unsatisfiable, then there exists no simple
+  matching scheme that normalizes $\mathcal{I}$ for the language $L$.
+
+\end{lemma}
+
+In this case, the reason for unsatisfiability can be used as a counterexample
+to refine the given language, for example, by adding additional predicates.
+
+\begin{example}
+
+  TODO SHOW THAT THE REGISTER LANGUAGE EXAMPLE WITHOUT ONE OF THE PREDICATES
+  WILL LEAD TO UNSAT.
+
+\end{example}
+
+On the other hand, the satisfiability of $\textsc{Normalizes}_F$ does not
+necessarily lead to a unique matching scheme, since $\textsc{Normalize}_F$ can
+have multiple satisfying assignments. Furthermore, $\textsc{Normalizes}_F$ does
+not necessarily normalize $\mathcal{I}$. For one reason, $F$ may not be a
+complete set of examples of executions and invalid projections. Second, our
+simple characterization of $\textsc{Normalize}_F$ does not rule out the other
+reasons for a given example $\tup{e,e'} \in F$ to be an invalid projection,
+e.g.,~that the removed operations do not constitute \emph{multiple} matches.
+However, we believe that checking satisfiability of $\textsc{Normalize}_F$ is
+useful nonetheless: at the very least, satisfying assignments can be used as
+\emph{assistance} in constructing normalizing matching schemes.

### 2015-03-18 patterns.tex
started wqo proof

@@
+%!TEX root = draft.tex
 \section{Computing ADT Complements}
@@
 \end{lemma}
+
+\begin{proof}
+
+%We prove that $\preceq$ is well-founded, i.e., there is no infinite strictly decreasing sequence $h_1 \succ h_2\succ \ldots$\footnote{We define $\succ$ as usual by $h\succ h'$ iff $h\succeq h'$ and $h\neq h'$.},
+%and that $\preceq$ has no infinite antichain, i.e., there is no infinite set $\{h_1,h_2,\ldots\}$ of mutually incomparable elements 
+%$h_i\not\preceq h_j$ and $h_j\not\preceq h_i$ when $i\neq j$.
+%
+%Concerning well-foundedness, note that a given history $h_1$ can be weaken only finitely many times by removing matches, duplicates in a match, or happens-before ordering constraints, or by turning completed operations into pending operations.
+Since we restrict ourselves to width-bounded histories, $h_1$ can also be weaken only finitely many times by adding pending operations (adding a pending operation increases the width of a history by 1).
+%
+%TODO PROBLEM WITH PENDING OPS
+
+%To prove that $\preceq$ has no infinite antichain, 
+%we use the following definitions.
+Let $h_1 h_2\ldots $ be an infinite sequence of histories. We prove that there exists $i<j$ such that $h_i\preceq h_j$.
+
+The \emph{signature} of an operation $o$ in a history $h$ is the pair $\sigma(o)=(c(o),f(o))$.
+Then, the \emph{signature} of a match $\mu=\set{ o' \in O : m(o') = o}$ is the pair 
+\[
+\sigma(\mu)=(\sigma(o),\set{\sigma(o') : m(o')=o, o'\neq o})
+\]
+containing the signature of the match target and the set of the signatures of the other operations.
+The \emph{signature} of a history $h$ is the set 
+\[
+\sigma(h)=\set{\sigma(\mu): \mu\mbox{ a match of $h$}}\cup\set{\sigma(o): o\mbox{ unmatched}}.
+\]
+of match signatures and signatures of unmatched operations.
+
+%Assume by contradiction that $\preceq$ has an infinite antichain.
+Since the set of history signatures is bounded (because the set of methods is bounded)
+and the set of pending operations is also bounded (since we consider only bounded width histories),
+the sequence $h_1 h_2\ldots $ contains an infinite sequence of histories $h_1' h_2' \ldots$ that have 
+the same signature and the same number of pending operations with the same signature.
+
+%The \emph{vector} of a match $\mu$ is the pair
+%\[
+%\nu(\mu)=(\sigma(o),\mset{\sigma(o') : m(o')=o, o'\neq o})
+%\]
+%containing the signature of the match target and the multiset of the signatures of the other operations.
+%Also, 
+The \emph{vector} of a history $h$ is the multiset 
+\[
+\nu(h)=\mset{\sigma(\mu): \mu\mbox{ a match of $h$}}
+\]
+of match signatures in $h$.
+
+Given a finite alphabet $\Sigma$, a domain $\mathbb{D}$ enhanced with a wqo $\leq$, and two functions 
+$\alpha,\beta:\Sigma\rightarrow \mathbb{D}$, let $\sqsubseteq$ be an order relation defined by
+\[
+\alpha\sqsubseteq\beta\mbox{ iff for every $a\in \Sigma$, $\alpha(a)\leq\beta(a)$}.
+\]
+Note that $\sqsubseteq$ is a wqo.
+
+The relation $\sqsubseteq$ on history vectors (mappings from the alphabet $\Sigma$ of match signatures to $\mathbb{N}$), 
+where $\leq$ is the order relation on natural numbers, is a wqo.
+Hence, the set $H$ contains an infinite sequence $h_1',h_2',\ldots$ such that $\nu(h_1')\sqsubseteq \nu(h_2')\sqsubseteq \ldots$.
+
+A \emph{nested vector} of a history $h$ is the nested multiset
+\[
+\nu^2(h)=\mset{\mset{\nu(\mu): \sigma(\mu)=\sigma} : \sigma\mbox{ the signature of a match in $h$}}.
+\]
+
+Again using the fact that $\leq$ is a wqo, the sequence $h_1',h_2',\ldots$ contains a history $h'_k$ and an infinite set of histories 
+$\set{h_1'',h_2'',\ldots}$ such that for every $i$ and every match $\mu$ of $h_i''$,
+$\nu(\mu_k)\leq \nu(\mu)$, for some match $\mu_k$ of $h_k$ such that $\sigma(\mu_k)=\sigma(\mu)$.
+
+\end{proof}
 

### 2015-03-18 histories.tex
Closure under unmatched-removal.

--- 2015-03-12 histories.tex
@@
 order of operations yet lose the exact sequence of call and return actions.
-Formally, a \emph{history} $h = \tup{O,<,c,f,m}$ is a tuple where
+Formally, a \emph{history} is a tuple $h = \tup{O,<,c,f,m}$ where
 \begin{itemize}
@@
 Non-completed operations are \emph{pending}, and are maximal in the
-happens-before order. The \emph{width} of a history is the maximum number of
-its operations which are mutually unordered, and the \emph{width} of a history
-set is the maximum width of its elements. Width-$1$ histories are
-\emph{sequential}. Two operations with the same labels and matching are
-\emph{identical}, and the \emph{frequency} of an operation $o$, denoted
-$\freq(o)$, is the number of operations identical to $o$, including $o$ itself.
-An operation is a \emph{duplicate} when $\freq(o) > 1$. An operation $o \in
-\rng(m)$ is a \emph{match target}, and the set $\set{ o' \in O : m(o') = o}$ of
-operations targeting $o$ is a \emph{match}. We assume targets target
-themselves: $m(o) = o$ for $o \in \rng(m)$. The \emph{weight} of $h$ is the
-maximum $\max(\set{ \freq(o) : o \in O } \cup \set{|\rng(m)|})$ between
-operation frequencies and the number of matches. Since the operation
-identifiers of a history have no intrinsic meaning, we consider equality
-between histories up to renaming of operation identifiers.
+happens-before order.
+
+Two operations $o_1, o_2 \in O$ are \emph{identical} when they have
+\begin{itemize}
+
+  \item the same labels: $c(o_1) = c(o_2)$, $f(o_1)= f(o_2)$, and
+
+  \item the same matching: either $m(o_1) = m(o_2)$, or both $m(o_1)$
+  and $m(o_2)$ are undefined.
+
+\end{itemize} 
+The \emph{frequency} of an operation $o$ is the number of operations identical
+to $o$, denoted $\freq(o)$. We say that $o$ has \emph{duplicates} when
+$\freq(o) > 1$. An operation $o \in \img(m)$ in the image of $m$ is a
+\emph{match target}, and the inverse set $m^{-1}(o)$ of a target is a
+\emph{match}. We assume $o \in m^{-1}(o)$. A operation $o$ is \emph{unmatched}
+when $o$ is completed and $m(o)$ is undefined.
+
+The \emph{width} of a history is the maximum number of its operations which are
+mutually unordered, and the \emph{width} of a history set is the maximum width
+of its elements. Width-$1$ histories are \emph{sequential}. The \emph{weight}
+of a history is the maximum among operation frequencies and the number of
+matches. Since the operation identifiers of a history have no intrinsic
+meaning, we consider equality between histories up to renaming of operation
+identifiers.
 

### 2015-03-18 patterns.tex
Closure under unmatched-removal.

@@
 closed, the ADTs we consider in this work are also closed under additional
-relations which remove matches, and duplicate operations. The relations
-$\to_\mathrm{m}$ and $\to_\mathrm{d}$ relate two histories $h_1$ and $h_2$ when
-$h_2$ is obtained from $h_1$ by
+relations which remove read-only operations, unmatched operations, entire
+matches, and duplicate operations. The relations $\to_\mathrm{r}$,
+$\to_\mathrm{u}$, $\to_\mathrm{m}$ and $\to_\mathrm{d}$ relate two histories
+$h_1$ and $h_2$ when $h_2$ is obtained from $h_1$ by
 \begin{itemize}
+
+  \item removing a read-only operation (r),
+
+  \item removing an unmatched operation (u),
 
@@
 \end{itemize}
-We say an ADT closed under $\to_\mathrm{m}$ and $\to_\mathrm{d}$ is
-\emph{normal}. In Section~\ref{sec:nature} we demonstrate that
-naturally-occurring ADTs are normal. Defining the relation $\succeq$ as the
-reflexive and transitive closure of the above relations,
+We say an ADT closed under $\to_\mathrm{r}$, $\to_\mathrm{u}$, $\to_\mathrm{m}$
+and $\to_\mathrm{d}$ is \emph{normal}. In Section~\ref{sec:nature} we
+demonstrate that naturally-occurring ADTs are normal. Defining the relation
+$\succeq$ as the reflexive and transitive closure of the above relations,
 \begin{align*}
-  \succeq = (\to_\mathrm{o} \cup \to_\mathrm{p} \cup \to_\mathrm{c} \cup 
-  \to_\mathrm{m} \cup \to_\mathrm{d})^\ast
+  \mathord{\succeq} = (
+    \to_\mathrm{o} \cup \to_\mathrm{p} \cup \to_\mathrm{c} \cup 
+    \to_\mathrm{r} \cup \to_\mathrm{u} \cup \to_\mathrm{m} \cup \to_\mathrm{d}
+  )^\ast
 \end{align*}

### 2015-03-18 patterns.tex
some wqo proof - missing final part

@@
 containing the signature of the match target and the set of the signatures of the other operations.
-The \emph{signature} of a history $h$ is the set 
-\[
-\sigma(h)=\set{\sigma(\mu): \mu\mbox{ a match of $h$}}\cup\set{\sigma(o): o\mbox{ unmatched}}.
-\]
-of match signatures and signatures of unmatched operations.
+The \emph{signature} of a history $h$ is the tuple 
+\begin{align*}
+\sigma(h)&=&(\set{\sigma(\mu): \mu\mbox{ a match}}, \set{\sigma(o): o\mbox{ read-only}}, \\
+&& \set{\sigma(o): o\mbox{ completed and unmatched}}, \\
+&& \mset{\sigma(o): o\mbox{ pending and unmatched}}).
+\end{align*}
+
+The history signature contains sets of signatures 
+(for matches, read-only and respectively, completed and unmatched operations) 
+and the multiset of signatures of the pending and unmatched operations.
 
@@
 the sequence $h_1 h_2\ldots $ contains an infinite sequence of histories $h_1' h_2' \ldots$ that have 
-the same signature and the same number of pending operations with the same signature.
+the same signature.
 
-%The \emph{vector} of a match $\mu$ is the pair
-%\[
-%\nu(\mu)=(\sigma(o),\mset{\sigma(o') : m(o')=o, o'\neq o})
-%\]
-%containing the signature of the match target and the multiset of the signatures of the other operations.
+%
 %Also,
-The \emph{vector} of a history $h$ is the multiset 
+
+The \emph{vector} of a history $h$ is the tuple
+\begin{align*}
+\nu(h)&=&(\mset{\sigma(\mu): \mu\mbox{ a match of $h$}}, \mset{\sigma(o): o\mbox{ read-only}}, \\
+&& \mset{\sigma(o): o\mbox{ completed and unmatched}}).
+\end{align*}
+
+Let $\leq$ be an order relation on multisets $\alpha:\Sigma\rightarrow\mathbb{N}$ of (match or operation) signatures from a finite set $\Sigma$ 
+defined by $\alpha\leq\alpha'$ iff $\alpha(\sigma)\leq\alpha'(\sigma)$, for every $\sigma\in\Sigma$.
+Clearly, $\leq$ is a wqo and so is the component-wise extension of $\leq$ to history vectors.
+Therefore, by known results, the sequence of histories $h_1' h_2' \ldots$ contains 
+an infinite sequence $h_1'',h_2'',\ldots$ such that $\nu(h_1'')\leq \nu(h_2'')\leq \ldots$.
+
+The \emph{vector} of a match $\mu$ is the pair
 \[
-\nu(h)=\mset{\sigma(\mu): \mu\mbox{ a match of $h$}}
+\nu(\mu)=(\sigma(o),\mset{\sigma(o') : m(o')=o, o'\neq o})
 \]
-of match signatures in $h$.
+containing the signature of the match target and the multiset of the signatures of the other operations.
 
-Given a finite alphabet $\Sigma$, a domain $\mathbb{D}$ enhanced with a wqo $\leq$, and two functions 
-$\alpha,\beta:\Sigma\rightarrow \mathbb{D}$, let $\sqsubseteq$ be an order relation defined by
+For every history $h$ and match signature $\sigma$, the \emph{$\sigma$-vector} of a history $h$ is the multiset
 \[
-\alpha\sqsubseteq\beta\mbox{ iff for every $a\in \Sigma$, $\alpha(a)\leq\beta(a)$}.
+\nu_\sigma(h)=\mset{\nu(\mu): \sigma(\mu)=\sigma}
 \]
-Note that $\sqsubseteq$ is a wqo.
+of match vectors of signature $\sigma$.
 
-The relation $\sqsubseteq$ on history vectors (mappings from the alphabet $\Sigma$ of match signatures to $\mathbb{N}$), 
-where $\leq$ is the order relation on natural numbers, is a wqo.
-Hence, the set $H$ contains an infinite sequence $h_1',h_2',\ldots$ such that $\nu(h_1')\sqsubseteq \nu(h_2')\sqsubseteq \ldots$.
+Let $\leq_v$ be an order relation on $\sigma$-vectors defined by $\nu_\sigma(h)\leq_v \nu_\sigma(h')$ iff
+\begin{itemize}
+	\item for every match vector $\nu$ in $\nu_\sigma(h)$ there exists a match vector $\nu'$ in $\nu_\sigma(h')$
+	such that $\nu\leq \nu'$ (here, $\leq$ is the order on multisets defined above).
+\end{itemize}
+By known results, $\leq_v$ is a wqo.
 
-A \emph{nested vector} of a history $h$ is the nested multiset
-\[
-\nu^2(h)=\mset{\mset{\nu(\mu): \sigma(\mu)=\sigma} : \sigma\mbox{ the signature of a match in $h$}}.
-\]
+Let $\sigma_1$, $\sigma_2$,$\ldots$, $\sigma_n$ be the match signatures defined over a set of methods $\mathbb{M}$.
+Since $\leq_v$ is a wqo, the sequence of histories $h_1'',h_2'',\ldots$ contains an 
+infinite sequence $h_{1}^1,h_{2}^1,\ldots$ such that 
+$\nu_{\sigma_1}(h_{1}^1)\leq_v \nu_{\sigma_1}(h_{2}^1)\leq_v \ldots$.
+Then, the sequence $h_{1}^1,h_{2}^1,\ldots$ contains 
+an infinite sequence $h_{1}^2,h_{2}^2,\ldots$ such that 
+$\nu_{\sigma_2}(h_{1}^2)\leq_v \nu_{\sigma_2}(h_{2}^2)\leq_v \ldots$.
+Applying a similar reasoning for the rest of the match signatures, 
+we obtain that $h_1'',h_2'',\ldots$ contains an 
+infinite sequence $h_{1}^n,h_{2}^n,\ldots$
+such that 
+$\nu_{\sigma}(h_{1}^n)\leq_v \nu_{\sigma}(h_{2}^n)\leq_v \ldots$, for every signature $\sigma$.
+Recall that we also have that
+$\nu(h_1^n)\leq\nu(h_2^n)\leq\ldots$ since $h_{1}^n,h_{2}^n,\ldots$ is a sub-sequence of $h_1'',h_2'',\ldots$.
 
-Again using the fact that $\leq$ is a wqo, the sequence $h_1',h_2',\ldots$ contains a history $h'_k$ and an infinite set of histories 
-$\set{h_1'',h_2'',\ldots}$ such that for every $i$ and every match $\mu$ of $h_i''$,
-$\nu(\mu_k)\leq \nu(\mu)$, for some match $\mu_k$ of $h_k$ such that $\sigma(\mu_k)=\sigma(\mu)$.
+Since the sequence of histories $h_{1}^n,h_{2}^n,\ldots$ is infinite there must exist
 

### 2015-03-18 histories.tex

@@
 order of operations yet lose the exact sequence of call and return actions.
-Formally, a \emph{history} is a tuple $h = \tup{O,<,c,f,m}$ where
+Formally, a \emph{history} is a tuple $h = \tup{O,<,c,f,m,r}$ where
 \begin{itemize}
@@
 
-  \item $f: O \to \mathbb{M}$ labels operations with method names, and
+  \item $f: O \to \mathbb{M}$ labels operations with method names,
 
-  \item $m: O \rightharpoonup O$ is a partial \emph{matching} function.
+  \item $m: O \rightharpoonup O$ is a partial \emph{matching} function, and
+
+  \item $r: O \to \mathbb{B}$ labels operations as \emph{read-only}, or not.
 
@@
 
-Two operations $o_1, o_2 \in O$ are \emph{identical} when they have
+Two operations $o_1, o_2 \in O$ are \emph{identical} when
 \begin{itemize}
 
-  \item the same labels: $c(o_1) = c(o_2)$, $f(o_1)= f(o_2)$, and
+  \item they have the same labels: $c(o_1) = c(o_2)$, $f(o_1)= f(o_2)$, and
+  $r(o_1) = r(o_2)$, and
 
-  \item the same matching: either $m(o_1) = m(o_2)$, or both $m(o_1)$
+  \item they have the same matching: either $m(o_1) = m(o_2)$, or both $m(o_1)$
   and $m(o_2)$ are undefined.
@@
 history. We construct partial matching functions systematically. A
-\emph{matching scheme} $M$ associates to each execution $e$ with operations $O$
-a partial matching function $M(e): O \rightharpoonup O$.
+\emph{matching scheme} $\tup{M,R}$ associates to each execution $e$ with
+operations $O$ a partial matching function $M(e): O \rightharpoonup O$ and a
+read-only operation labeling $R(e): O \to \mathbb{B}$.
 
@@
 
-The history $H(e,M)$ of an execution $e$ under matching scheme $M$ is the tuple
-$\tup{O,<,c,f,M(e)}$ where
+The history $H(e,M,R)$ of an execution $e$ under matching scheme $\tup{M,R}$ is
+the tuple $\tup{O,<,c,f,M(e),R(e)}$ where
 \begin{itemize}
@@
 \end{itemize}
-We denote the set $\set{H(e,M) : e \in E}$ of histories of an execution set $E$
-by $H(E,M)$.
+We denote the set $\set{H(e,M,R) : e \in E}$ of histories of an execution set
+$E$ by $H(E,M,R)$.
 
@@
 
-A matching scheme $M$ is \emph{faithful} to a set $E$ of executions when $e \in
-E$ if{f} $e' \in E$ for any two executions $e$ and $e'$ such that $H(e,M) =
-H(e',M)$. A set of executions $E$ is \emph{data independent}\footnote{TODO NOTE
-THAT THIS IS NOT THE STANDARD DEFINITION…} when there exists a faithful
-matching scheme. By definition, our abstraction of executions incurs no loss of
-precision for data-independent execution sets.
+A matching scheme $\tup{M,R}$ is \emph{faithful} to a set $E$ of executions
+when $e \in E$ if{f} $e' \in E$ for any two executions $e$ and $e'$ such that
+$H(e,M,R) = H(e',M,R)$. A set of executions $E$ is \emph{data
+independent}\footnote{TODO NOTE THAT THIS IS NOT THE STANDARD DEFINITION…} when
+there exists a faithful matching scheme. By definition, our abstraction of
+executions into histories incurs no loss of precision for data-independent
+execution sets.
 
@@
 
-  $H(e,M) \in H(E,M)$ if and only if $e \in E$, for any faithful matching
-  scheme $M$.
+  $H(e,M,R) \in H(E,M,R)$ if and only if $e \in E$, for any faithful matching
+  scheme $\tup{M,R}$.
 
@@
 remainder of this work, we assume each set of executions comes equipped with a
-faithful matching scheme $M$, and abbreviate $H(E,M)$ by $H(E)$.
+faithful matching scheme $\tup{M,R}$, and abbreviate $H(E,M,R)$ by $H(E)$.
 

### 2015-03-20 matching.tex
Matching schemes include read-only labelings.
Thorough examples for matching scheme inference.

@@
 
-TODO MOTIVATE THIS SECTION
+TODO MOTIVATE THIS SECTION, AND ALSO ACCOUNT FOR THE READ-ONLY OPERATION
+LABELING TOO.
 
@@
 \begin{example}
+  \label{ex:register-language}
 
@@
   the argument value to each write operation is unique. Consider the language
+  $L_\mathrm{reg}$
   whose executions write unique values, with the following predicates:
   \begin{align*}
-    & \mathsf{write}(e,o) & \text{$o$ is a {\sf write} operation in $e$}, \\
-    & \mathsf{read}(e,o) & \text{$o$ is a {\sf read} operation in $e$,} \\
-    & \mathsf{eq}(e,o_1,o_2) & \text{$o_1$ and $o_2$ are the same operation, and} \\
-    & \mathsf{veq}(e,o_1,o_2) & \text{$o_1$ and $o_2$ read/write the same values.}
+    & \mathsf{w}(e,o) & \text{$o$ is a {\sf write} operation in $e$, and}, \\
+    & \mathsf{v}(e,o_1,o_2) & \text{$o_1$ and $o_2$ read/write the same values.}
   \end{align*}
-  Given a valuation $\tup{x_\mathsf{w}, x_\mathsf{r}, y_\mathsf{w},
-  y_\mathsf{r}, z_\mathsf{=}, z_\mathsf{v}}$ of the predicates above, we define
-  the function $G(x_\mathsf{w}, x_\mathsf{r}, y_\mathsf{w}, y_\mathsf{r},
-  z_\mathsf{=}, z_\mathsf{v})$ to be satisfied if and only if:
+  We define the function $G(x_\mathsf{w}, y_\mathsf{w}, z_\mathsf{v})$ over
+  valuations to the predicates above to be satisfied if and only if:
   \begin{align*}
-    ( x_\mathsf{w} \land z_\mathsf{=} )
-    \lor ( x_\mathsf{w} \land y_\mathsf{r} \land z_\mathsf{v} )
+    x_\mathsf{w} \land z_\mathsf{v}
   \end{align*}
@@
 
+\begin{example}
+
+  Consider again the language $L_\mathrm{reg}$ of read and write methods of
+  Example~\ref{ex:register-language} with predicates ${\sf w}(e,o)$ and ${\sf
+  v}(e,o_1,o_2)$ whose executions write unique values. The following table
+  lists all possible predicate valuations $\tup{x_{\sf w}, y_{\sf w}, z_{\sf
+  v}}$, and for each valuation a valid execution which excludes the positive
+  valuation of ${\sf g}(x_{\sf w}, y_{\sf w}, z_{\sf v})$ in the satisfaction
+  of $\textsc{Normalize}_F$, in the case such an execution exists.
+  \begin{center}
+    \begin{tabular}{ccc|ll}
+      $x_\mathsf{w}$ & $y_\mathsf{w}$ & $z_\mathsf{v}$ & counterexample & reason \\
+      \hline
+      $0$ & $0$ & $0$ & ${\rm w}(1) \ {\rm r}(1) \ {\rm r}(1) \ {\rm w}(2) \ {\rm r}(2)$ & not unique \\
+      $0$ & $0$ & $1$ & ${\rm w}(1) \ {\rm r}(1) \ {\rm r}(1)$ & not unique \\
+      $0$ & $1$ & $0$ & ${\rm w}(1) \ \underline{{\rm r}(1) \ {\rm w}(2)} \ {\rm r}(2)$ & $\to_\mathrm{m}$ \\
+      $0$ & $1$ & $1$ & $\underline{{\rm w}(1) \ {\rm r}(1)} \ {\rm r}(1)$ & $\to_\mathrm{m}$ \\
+      $1$ & $0$ & $0$ & ${\rm w}(1) \ {\rm w}(2) \ {\rm w}(3) \ {\rm r}(3)$ & not unique \\
+      $1$ & $0$ & $1$ & & \\
+      $1$ & $1$ & $0$ & ${\rm w}(1) \ {\rm w}(2) \ {\rm w}(3)$ & not unique \\
+      $1$ & $1$ & $1$ & & \\
+    \end{tabular}
+  \end{center}
+  For example, the first valuation $000$ must be excluded since it allows the
+  read ${\rm r}(2)$ to match both reads ${\rm r}(1)$, violating
+  $\textsc{UniqueMatch}$. The second valuation $001$ must also be excluded,
+  since it allows either read ${\rm r}(1)$ to match both itself and the other
+  read ${\rm r}(1)$. The third valuation $010$ must be excluded since it allows
+  the write ${\rm w}(2)$ to match the read ${\rm r}(1)$, in which case removing
+  the match $\set{ {\rm r}(1), {\rm w}(2) }$ results in the invalid execution
+  ${\rm w}(1)\ {\rm r}(2)$. Reasoning follows similarly for the other rows.
+  In this way, any enumeration $F$ which includes the above executions
+  will exclude all valuations except for $101$ and $111$, ultimately resulting
+  in a satisfiable $\textsc{Normalize}_F$ in which ${\sf g}(x_{\sf w}, y_{\sf
+  w}, z_{\sf v})$ is the same Boolean function $x_{\sf w} \land z_{\sf v}$
+  given in Example~\ref{ex:register-language}.
+
+\end{example}
+
 On the one hand, checking satisfiability of $\textsc{Normalizes}_F$ can be used
@@
 
-  TODO SHOW THAT THE REGISTER LANGUAGE EXAMPLE WITHOUT ONE OF THE PREDICATES
-  WILL LEAD TO UNSAT.
+  Consider again the language $L_\mathrm{reg}$ of read and write methods of
+  Example~\ref{ex:register-language}, yet this time without the predicate ${\sf
+  v}(e,o_1,o_2)$. The following table lists all possible predicate valuations
+  $\tup{x_{\sf w}, y_{\sf w}}$, and for each valuation a valid execution which
+  excludes the positive valuation of ${\sf g}(x_{\sf w}, y_{\sf w})$ in the
+  satisfaction of $\textsc{Normalize}_F$.
+  \begin{center}
+    \begin{tabular}{cc|ll}
+      $x_{\sf w}$ & $y_{\sf w}$ & counterexample & reason \\
+      \hline
+      0 & 0 & ${\rm w}(1)\ {\rm r}(1)\ {\rm r}(1)$ & not unique \\
+      0 & 1 & $\underline{{\rm w}(1)\ {\rm r}(1)}\ {\rm r}(1)$ & $\to_\mathrm{m}$ \\
+      1 & 0 & ${\rm w}(1)\ {\rm w}(2)\ {\rm r}(2)$ & not unique \\
+      1 & 1 & ${\rm w}(1)\ {\rm w}(2)$ & not unique \\
+    \end{tabular}
+  \end{center}
+  Thus any enumeration $F$ including the executions above excludes every
+  possible valuations of $\tup{x_{\sf w}, y_{\sf w}}$, resulting in an
+  unsatisfiable $\textsc{Normalize}_F$.
 

### 2015-03-20 matching.tex
Added UniqueMatch formula.

@@
 ${\sf g}$ underlying a simple matching scheme as an uninterpreted function in a
-logical satisfiability problem, and constrain ${\sf g}$ to normalize the
-executions of an implementation. To this end, we fix an implementation
-$\mathcal{I}$ and a language $L = \tup{E,P,Q}$ to which $\mathcal{I}$ adheres,
-then consider any enumeration $F$ of execution pairs $\tup{e,e'} \in E^2$ such
-that
+logical satisfiability problem which uniquely determines the match of each
+operation, and constrain ${\sf g}$ to normalize the executions of an
+implementation. To this end, we fix an implementation $\mathcal{I}$ and a
+language $L = \tup{E,P,Q}$ to which $\mathcal{I}$ adheres, then consider any
+enumeration $F$ of execution pairs $\tup{e,e'} \in E^2$ such that
 \begin{itemize}
@@
 \end{align*}
+To enforce that each operation of an execution $e$ will have a unique match,
+we define the macro $\textsc{UniqueMatch}_e$:
+\begin{align*}
+  \bigwedge_{o_1,o_2 \in O_e} \Big(
+    \textsc{IsMatch}_{e,o_2,o_2} \implies
+    \bigwedge_{o_1' \neq o_1} \lnot \textsc{IsMatch}_{e,o_1',o_2}
+  \Big).
+\end{align*}
 Then a given set $O \subseteq O_e$ constitutes a match according to
@@
 \begin{align*}
-  \bigwedge_{\tup{e,e'}\in F} \lnot \textsc{EntireMatch}_{e,(O_e \setminus O_{e'})}
+  \bigwedge_{\tup{e,e'}\in F} \Big(
+    \lnot \textsc{EntireMatch}_{e,(O_e \setminus O_{e'})}
+    \land \textsc{UniqueMatch}_{e}
+  \Big).
 \end{align*}

### 2015-03-20 matching.tex
Tiny typo fixes.

@@
   \bigwedge_{\tup{e,e'}\in F} \Big(
-    \lnot \textsc{EntireMatch}_{e,(O_e \setminus O_{e'})}
-    \land \textsc{UniqueMatch}_{e}
+    \textsc{UniqueMatch}_{e}
+    \land \lnot \textsc{EntireMatch}_{e,(O_e \setminus O_{e'})}
   \Big).
@@
   Thus any enumeration $F$ including the executions above excludes every
-  possible valuations of $\tup{x_{\sf w}, y_{\sf w}}$, resulting in an
+  possible valuation of $\tup{x_{\sf w}, y_{\sf w}}$, resulting in an
   unsatisfiable $\textsc{Normalize}_F$.

### 2015-03-21 nature.tex
Note on width-bounded kernels.
Better Section 2.
Better titles.
Can we treat the priority queue?
The nature of registers.

@@
-\section{Naturally-Occurring ADTs}
+\section{Properties of Naturally-Occurring ADTs}
 \label{sec:nature}
 
-TODO DEMONSTRATE FAITHFUL MATCHING SCHEMES, NORMALITY, AND PREDICTABILITY FOR
-EACH NATURALLY-OCCURRING ADT
+TODO DEMONSTRATE FAITHFUL MATCHING SCHEMES, UNIQUE AND WIDTH-BOUNDED KERNELS,
+NORMALITY, AND PREDICTABILITY FOR EACH NATURALLY-OCCURRING ADT
+
+TODO EXPLAIN “ATOMIC”
+
+\subsection{The Atomic Register}
+
+Atomic register objects implement an atomic single-value store, and provide two
+methods:
+\begin{itemize}
+
+  \item write$(v)$ stores the value $v$, and
+
+  \item read $\Rightarrow v$ returns the last-stored value $v$, or $\perp$
+  if no value has been stored.
+
+\end{itemize}
+We say an execution of read and write methods \emph{writes unique values} if
+the argument value to each write operation is unique. A faithful matching
+scheme is given over the set of executions which write unique values as follows:
+\begin{itemize}
+
+  \item write$(v)$ operations match themselves, and
+
+  \item read $\Rightarrow v$ operations are read-only, and match themselves if
+  $v = \perp$, or the unique write$(v)$ operation, if one exists, and otherwise 
+  have no match.
+
+\end{itemize}
+The kernel of atomic registers is the set of sequential histories given by
+sequences
+\begin{align*}
+  \bigcup_{0 \le k \le n < \omega}
+  \Big( \underline{\mathrm{read}:i} \Big)_{i=1}^{k}\ 
+  \Big( \underline{\mathrm{write}:i}\ (\mathrm{read}:i)^\ast \Big)_{i=k}^{n}
+\end{align*}
+in which the initial read operations match themselves, write operations match
+themselves, and each subsequent read operation matches the nearest preceding
+write operation.
+
+TODO ARGUE THAT THIS IS FAITHFUL
+
+It is easy to see that the histories of atomic registers normal. They are
+closed under removal of read-only and duplicate operations, since the each
+match can contain an arbitrary number of read operations. Since the histories
+of atomic registers do not contain unmatched operations, they are also closed
+under their removal. Finally, since the kernel contains an arbitrary number of
+sequenced matches, atomic register histories are also closed under match
+removal.
+
+THIS IS THE SET OF PATTERNS
+\begin{align*}
+  & \mathrm{read}:\_ \\
+  & \underline{\mathrm{write}:1}\ \underline{\mathrm{read}:2} \\
+  & \mathrm{read}:1\ \underline{\mathrm{write}:1} \\
+  & \underline{\mathrm{write}:1}\ \underline{\mathrm{write}:2}\ \mathrm{read}:1
+\end{align*}
+The atomic register ADT is predictable \ldots
 
@@
 
-\subsection{The Atomic Register}
+\subsection{The Priority Queue (?)}
 

### 2015-03-21 histories.tex
Added executions example.

@@
 
-  TODO EXAMPLE EXECUTION
+  Consider the following three executions of a single-value register object
+  with read and write methods
+  \begin{center}
+    \setlength{\tabcolsep}{1em}
+    \begin{tabular}{l|l|l}
+      $e_1$                           & $e_2$                           & $e_3$ \\
+      $1:\mathrm{write}(a)$           & $1:\mathrm{write}(a)$           & $1:\mathrm{write}(a)$ \\
+      $1:\mathbf{ret}$                & $2:\mathrm{write}(b)$           & $1:\mathbf{ret}$ \\
+      $2:\mathrm{write}(b)$           & $1:\mathbf{ret}$                & $2:\mathrm{write}(b)$ \\
+      $2:\mathbf{ret}$                & $2:\mathbf{ret}$                & $3:\mathrm{read}$ \\
+      $3:\mathrm{read}$               & $3:\mathrm{read}$               & $2:\mathbf{ret}$ \\
+      $3:\mathbf{ret} \Rightarrow a$  & $3:\mathbf{ret} \Rightarrow a$  & $3:\mathbf{ret} \Rightarrow a$ \\
+    \end{tabular}
+  \end{center}
+  in which actions are preceded by their operation identifiers. Executions
+  $e_2$ and $e_3$ are obtained from $e_1$ by permuting the return actions of
+  operations $1$ and $2$, respectively, forward, or symmetrically, the call
+  actions of operations $2$ and $3$, respectively, backward. Thus while the
+  operations of Execution $e_1$ are sequential, each following the previous in
+  time, those of $e_2$ and $e_3$ overlap. While $e_1$ should not be admitted
+  by an \emph{atomic} register, since the read of Operation~$3$ does not
+  return the most-recently-written value, both $e_2$ and $e_3$ could be
+  admitted, since there exist linearizations of the partially-ordered operations
+  such that Operation~$3$ does return the most-recently-written value.
 

### 2015-03-21 histories.tex
Added possible history notation.

@@
 \begin{example}
+  \label{ex:executions}
 
@@
 
-  TODO HISTORY OF AN EXECUTION
+  The histories of the Executions $e_1$, $e_2$, and $e_3$ of
+  Example~\ref{ex:executions} are
+  \begin{align*}
+    & H(e_3) = \big( \underline{\mathrm{write}:1} \big) \big( \underline{\mathrm{write}:2} \big) \big( \mathrm{read}:1 \big) \\
+    & H(e_2) = \big( \underline{\mathrm{write}:1} \parallel \underline{\mathrm{write}:2} \big) \big( \mathrm{read}:1 \big) \\
+    & H(e_3) = \big( \underline{\mathrm{write}:1} \big) \big( \underline{\mathrm{write}:2} \parallel \mathrm{read}:1 \big)
+  \end{align*}
+  TODO IS THIS NOTATION GOOD ENOUGH?
 

### 2015-03-21 histories.tex
Added history examples.

--- 2015-03-21 histories.tex
@@
 
-  TODO HISTORY EXAMPLE
+  We draw histories by writing one operation per line, starting with its
+  operation identifier and match target, followed by its method label,
+  possibly a read-only marker, and its happens-before interval. For instance,
+  in the history
+\begin{verbatim}
+    [1:1] write(a)        #
+    [2:1] read => a (RO)    #
+\end{verbatim}
+  Operation~$1$ is a write operation which matches itself, and precedes a
+  read-only read operation which also targets Operation~$1$. Note that we
+  only label method argument and return values for illustrative purposes;
+  histories do not keep them. In the following history, Operations~$1$ and~$2$
+  execute concurrently
+\begin{verbatim}
+    [1:1] write(a)        #
+    [2:1] read => a (RO)  #
+    [3:X] read => b (RO)    #
+\end{verbatim}
+  and Operation~$3$ has no match.
 
@@
 
-  TODO HISTORY OF AN EXECUTION
-
+  The histories of the Executions of Example~\ref{ex:executions} are
+\begin{verbatim}
+    [1:1] write(a)        #
+    [2:2] write(b)          #
+    [3:1] read => a (RO)      #
+\end{verbatim}
+  in which all three operations are sequential,
+\begin{verbatim}
+    [1:1] write(a)        #
+    [2:2] write(b)        #
+    [3:1] read => a (RO)    #
+\end{verbatim}
+  in which the first two operations overlap, and
+\begin{verbatim}
+    [1:1] write(a)        #
+    [2:2] write(b)          #
+    [3:1] read => a (RO)    #
+\end{verbatim}
+  in which the last two operations overlap.
+  TODO SAY SOMETHING INTERESTING IN THIS EXAMPLE
 \end{example}

### 2015-03-21 histories.tex
Better wording.

@@
   read-only read operation which also targets Operation~$1$. Note that we
-  only label method argument and return values for illustrative purposes;
+  only label method argument and return values for illustrative purposes only;
   histories do not keep them. In the following history, Operations~$1$ and~$2$
@@
     [3:X] read => b (RO)    #
+    [4:_] read*     (RO)      #
 \end{verbatim}
-  and Operation~$3$ has no match.
+  Operation~$3$ is unmatched, and Operation~$4$ is pending.
 

### 2015-03-22 histories.tex
Small typo fix.
More examples + citation to interval orders.

@@
 
-  \item $<$ is a partial \emph{happens-before} order on $O$,
+  \item $<$ is a \emph{happens-before} interval order\footnote{An interval
+  order~\cite{books/Fishburn85} is a partial order whose elements can be mapped
+  to integral intervals preserving the order relation, or equivalently, a
+  partial order for which $w < x$ and $y < z$ implies $w < z$ or $y < x$.} on
+  $O$,
 
@@
 \begin{example}
+  \label{ex:histories}
 
@@
   read-only read operation which also targets Operation~$1$. Note that we
-  only label method argument and return values for illustrative purposes only;
+  label method argument and return values for illustrative purposes only;
   histories do not keep them. In the following history, Operations~$1$ and~$2$
@@
 \end{verbatim}
-  Operation~$3$ is unmatched, and Operation~$4$ is pending.
+  Operation~$3$ is unmatched, indicated by the {\tt X}, and Operation~$4$ is
+  pending, indicated by the {\tt *} on its label.
 
@@
 \begin{example}
+  \label{ex:matching}
 
-  TODO MATCHING SCHEME EXAMPLE
+  Consider the following matching scheme for the read and write operations of
+  a single-value register object:
+  \begin{itemize}
+
+    \item write$(v)$ operations match themselves, and
+
+    \item read $\Rightarrow v$ operations are read-only, and match themselves
+    when they return $v = \mathord\perp$, or a write$(v)$ operation, if one
+    exists, and otherwise have no match,
+
+  \end{itemize}
+  which is well defined when each value $v \in \mathbb{V}$ appears as the
+  argument of at most one write operation in any execution. This matching
+  scheme corresponds to the matching functions in the histories of
+  Example~\ref{ex:histories}.
 
@@
 We denote the set $\set{H(e,M,R) : e \in E}$ of histories of an execution set
-$E$ by $H(E,M,R)$.
+$E$ by $H(E,M,R)$. When the matching scheme $\tup{M,E}$ is clear from the
+context, we abbreviate $H(e,M,R)$ and $H(E,M,R)$ by $H(e)$ and $H(E)$.
 
@@
 
-  The histories of the Executions of Example~\ref{ex:executions} are
+  The histories of the executions of Example~\ref{ex:executions} according to
+  the matching scheme of Example~\ref{ex:matching} are $H(e_1)$:
 \begin{verbatim}
@@
 \end{verbatim}
-  in which all three operations are sequential,
+  in which all three operations are sequential, $H(e_2)$:
 \begin{verbatim}
@@
 \end{verbatim}
-  in which the first two operations overlap, and
+  in which the first two operations overlap, and $H(e_3)$:
 \begin{verbatim}
@@
   in which the last two operations overlap.
-  TODO SAY SOMETHING INTERESTING IN THIS EXAMPLE
+
 \end{example}
@@
 remainder of this work, we assume each set of executions comes equipped with a
-faithful matching scheme $\tup{M,R}$, and abbreviate $H(E,M,R)$ by $H(E)$.
+faithful matching scheme $\tup{M,R}$.
 

### 2015-03-22 inference.tex
Better titles.
Soft description of the formulas.
Formula example.

@@
-\section{Refinement \& ADT Inference}
+\section{The Symbolic ADT Inference Problem}
 \label{sec:inference}
@@
 
-  \item a function symbol $\mathsf{m}$ for matching, and
+  \item function symbols ${\sf f}$ and ${\sf m}$ for labels and matching, and
 
-  \item predicate symbols $\mathsf{c}$ and $\mathsf{<}$ for completion and
-  order.
+  \item predicate symbols ${\sf c}$, ${\sf um}$, and ${\sf <}$ for completion,
+  non-matching, and order.
 
@@
 
-  TODO HISTORY FORMULA EXAMPLE
+  The following history formula is satisfied by histories in which no write
+  operation happens between a pair of matching write and read operations:
+  \begin{align*}
+    \forall x, y, z.\ 
+    & {\sf c}(x) \land {\sf f}(x) = \mathrm{write} \land {\sf m}(x) = x \\
+    & \land {\sf c}(y) \land {\sf f}(y) = \mathrm{write} \land {\sf m}(y) = y \\
+    & \land {\sf c}(z) \land {\sf f}(z) = \mathrm{read} \land {\sf m}(z) = x \\
+    & \land x < y 
+      \implies z < y
+  \end{align*}
+  This is one of several requirements of atomic single-value register ADTs,
+  which is satisfied by certain linearizations of histories $H(e_2)$ and
+  $H(e_3)$ of Example~\ref{ex:histories-of-executions}, yet not $H(e_1)$.
 

### 2015-03-22 patterns.tex
Added patterns example.

@@
 
-  TODO EXAMPLE OF COMPLETE PATTERNS
-
-  \begin{figure}[t]
-    \centering
-    \includegraphics[scale=0.2]{figures/register-pattern-1},
-    \includegraphics[scale=0.2]{figures/register-pattern-2},
-    \includegraphics[scale=0.2]{figures/register-pattern-3}
-    \caption{Three histories which generate the complement of the atomic
-      register ADT kernel.}
-    \label{fig:register-patterns}
-  \end{figure}
+  The following four histories generate the complement of the atomic
+  single-value register ADT, witnessing either an unmatched read operation:
+\begin{verbatim}
+  [1:X] read => 1 (RO)  #
+\end{verbatim}
+  a read of an uninitialized register occurring after some write:
+\begin{verbatim}
+  [1:1] write(1)            #
+  [2:2] read => empty (RO)    #
+\end{verbatim}
+  a read which happens before its matching write operation:
+\begin{verbatim}
+  [1:2] read => 1 (RO)  #
+  [2:2] write(1)          #
+\end{verbatim}
+  and a read matching a write which is not the most recent:
+\begin{verbatim}
+  [1:1] write(1)        #
+  [2:2] write(2)          #
+  [3:1] read => 1 (RO)      #
+\end{verbatim}
+  Every sequential history not admitted by the atomic register ADT embeds at
+  least one of these four histories.
 

### 2015-03-22 formula.tex
A few informative words.
Formula examples.

@@
 
-Let $h = \tup{O,<,c,f,m}$ be a history with operations $O = \set{ o_1, \ldots,
+Let $h = \tup{O,<,c,f,m,r}$ be a history with operations $O = \set{ o_1, \ldots,
 o_n }$. Without loss of generality, we assume the match targets $\set{o_1,
@@
 \begin{align*}
-  & \big(c(o) \implies {\sf c}(x)\big) \land {\sf f}(x) = f(o) \\
+  & \big(c(o) \implies {\sf c}(x)\big) \land {\sf f}(x) = f(o) \land {\sf r}(x) = r(o) \\
   & \land \big(o \in \dom(m)
@@
 capturing the correspondence between the operation $o$ and the logical variable
-$x$. The variables $Y$ and $z$ represent the operations ordered after $x$, and
-the operation which $x$ matches. The macro $\textsc{Identical}(x,y)$:
+$x$ representing $o$. The variables $Y$ and $z$ represent the operations
+ordered after $x$, and the operation which $x$ matches. The macro
+$\textsc{Identical}(x,y)$:
 \begin{align*}
   & \big({\sf c}(x) \Leftrightarrow {\sf c}(y)\big)
-  \land {\sf f}(x) = {\sf f}(y) \\
+    \land {\sf f}(x) = {\sf f}(y) 
+    \land \big({\sf r}(x) \Leftrightarrow {\sf r}(y)\big) \\
   & \land \big({\sf um}(x) \Leftrightarrow {\sf um}(y)\big)
@@
   {\sf m}(x) = z \implies
-  \textsc{ReadOnly}(x) \lor \bigvee_{y \in Y} \textsc{Identical}(x,y)
+  {\sf r}(x) \lor \bigvee_{y \in Y} \textsc{Identical}(x,y)
 \end{align*}
@@
 \begin{example}
+  \label{ex:formulas}
 
-  TODO EXAMPLE FORMULÆ FROM PREVIOUS PATTERNS
+  Consider again the histories of Example~\ref{ex:patterns} which generate the
+  complement of the atomic register ADT. The $\textsc{Embed}$ formula for the
+  first history,
+\begin{verbatim}
+  [1:X] read => 1 (RO)  #
+\end{verbatim}
+  after simplifications, like replacing $\mathrm{true} \implies p$ with $p$, is
+  \begin{align*}
+    \exists x_1.\ {\sf c}(x_1) \land {\sf f}(x_1) = \mathrm{read} \land {\sf um}(x_1).
+  \end{align*}
+  The $\textsc{Embed}$ formula for second history,
+\begin{verbatim}
+  [1:1] write(1)            #
+  [2:2] read => empty (RO)    #    
+\end{verbatim}
+  is similarly given by
+  \begin{align*}
+    & \exists x_1, x_2.\ x_1 < x_2 \\
+    & \quad \land {\sf c}(x_1) \land {\sf f}(x_1) = \mathrm{write}
+      \land \lnot{\sf um}(x_1) \land {\sf m}(x_1) = x_1 \\
+    & \quad \land {\sf c}(x_2) \land {\sf f}(x_2) = \mathrm{read}
+      \land \lnot{\sf um}(x_2) \land {\sf m}(x_2) = x_2 \\
+    & \quad \land \big(\forall x.\ {\sf m}(x) = x_1 \implies {\sf r}(x) \lor \textsc{Identical}(x,x_1)\big) \\
+    & \quad \land \big(\forall x.\ {\sf m}(x) = x_2 \implies {\sf r}(x) \lor \textsc{Identical}(x,x_2)\big).
+  \end{align*}
+  The formulas for the other histories are similarly obtained.
 
@@
 We obtain a formula representing an ADT $A$ by taking the conjunction of the
-negative embedding formulæ $\set{ \lnot\textsc{Embed}_h : h \in H }$ from any
+negative embedding formulas $\set{ \lnot\textsc{Embed}_h : h \in H }$ from any
 set $H$ that generates the complement of $A$.
@@
 
-  TODO EXAMPLE OF COMPLETE FORMULA FROM PREVIOUS COMPLETE PATTERNS/FORMULÆ
+  The conjunction of negations of the $\textsc{Embed}$ formula for the
+  histories of Example~\ref{ex:patterns}, which are partially written in
+  Example~\ref{ex:formulas}, represents the atomic register ADT.
 

### 2015-03-22 nature.tex
Added lots of patterns.

@@
 THIS IS THE SET OF PATTERNS
-\begin{align*}
-  & \mathrm{read}:\_ \\
-  & \underline{\mathrm{write}:1}\ \underline{\mathrm{read}:2} \\
-  & \mathrm{read}:1\ \underline{\mathrm{write}:1} \\
-  & \underline{\mathrm{write}:1}\ \underline{\mathrm{write}:2}\ \mathrm{read}:1
-\end{align*}
+
+\begin{verbatim}
+  [1:X] read => 1 (RO)  #
+  --
+  [1:1] write(1)            #
+  [2:2] read => empty (RO)    #
+  --
+  [1:2] read => 1 (RO)  #
+  [2:2] write(1)          #
+  --
+  [1:1] write(1)        #
+  [2:2] write(2)          #
+  [3:1] read => 1 (RO)      #
+\end{verbatim}
+
 The atomic register ADT is predictable \ldots
@@
 
+\begin{verbatim}
+  [1:_] remove => 1  #
+  --
+  [1:1] add(1)                #
+  [2:2] remove => empty (RO)    #
+  --
+  [1:2] remove => 1  #
+  [2:2] add(1)         #
+  --
+  [1:1] add(1)       #
+  [2:2] add(2)         #
+  [3:2] remove => 2      #
+  --
+  [1:1] add(1)       #
+  [2:1] remove => 1    #
+  [3:1] remove => 1      #
+  --
+  [1:1] add(1)                #
+  [2:2] remove => empty (RO)    #
+  [3:1] remove => 1               #
+  --
+  [1:1] add(1)       #
+  [2:2] add(2)         #
+  [3:2] remove => 2      #
+  [4:1] remove => 1        #
+\end{verbatim}
+
 \subsection{The Atomic Stack}
+
+\begin{verbatim}
+  [1:_] remove => 1  #
+  --
+  [1:1] add(1)                #
+  [2:2] remove => empty (RO)    #
+  --
+  [1:2] remove => 1  #
+  [2:2] add(1)         #
+  --
+  [1:1] add(1)       #
+  [2:2] add(2)         #
+  [3:1] remove => 1      #
+  --
+  [1:1] add(1)       #
+  [2:1] remove => 1    #
+  [3:1] remove => 1      #
+  --
+  [1:1] add(1)                #
+  [2:2] remove => empty (RO)    #
+  [3:1] remove => 1               #
+  --
+  [1:1] add(1)       #
+  [2:2] add(2)         #
+  [3:1] remove => 1      #
+  [4:2] remove => 2        #
+\end{verbatim}
 
@@
 
-\subsection{The Priority Queue (?)}
+\begin{verbatim}
+  [1:_] insert(1) => 1 (RO)  #
+  --
+  [1:_] remove(1) => 1  #
+  --
+  [1:_] contains(1) => 1 (RO)  #
+  --
+  [1:1] insert(1) => empty       #
+  [2:2] remove(1) => empty (RO)    #
+  --
+  [1:1] insert(1) => empty         #
+  [2:2] contains(1) => empty (RO)    #
+  --
+  [1:1] insert(1) => empty  #
+  [2:1] remove(1) => 1        #
+  [3:1] remove(1) => 1          #
+  --
+  [1:1] insert(1) => empty     #
+  [2:1] remove(1) => 1           #
+  [3:1] contains(1) => 1 (RO)      #
+  --
+  [1:1] insert(1) => empty       #
+  [2:2] remove(1) => empty (RO)    #
+  [3:1] remove(1) => 1               #
+  --
+  [1:1] insert(1) => empty         #
+  [2:2] contains(1) => empty (RO)    #
+  [3:1] remove(1) => 1                 #
+\end{verbatim}
 
@@
 
-\subsection{The Atomic Semaphore}
+\begin{verbatim}
+  [1:_] lock(1) => 1 (RO)  #
+  --
+  [1:_] unlock => 1  #
+  --
+  [1:1] lock(1) => empty  #
+  [2:2] lock(2) => empty    #
+  --
+  [1:1] lock(1) => empty      #
+  [2:2] unlock => empty (RO)    #
+  --
+  [1:1] lock(1) => empty  #
+  [2:2] lock(2) => empty    #
+  [3:1] unlock => 1           #
+  --
+  [1:1] lock(1) => empty  #
+  [2:2] lock(2) => empty    #
+  [3:2] unlock => 2           #
+  --
+  [1:1] lock(1) => empty   #
+  [2:1] unlock => 1          #
+  [3:1] lock(2) => 1 (RO)      #
+  --
+  [1:1] lock(1) => empty  #
+  [2:1] unlock => 1         #
+  [3:1] unlock => 1           #
+  --
+  [1:1] lock(1) => empty      #
+  [2:2] unlock => empty (RO)    #
+  [3:1] unlock => 1               #
+  --
+  [1:1] lock(1) => empty  #
+  [2:2] lock(2) => empty    #
+  [3:1] unlock => 1           #
+  [4:2] unlock => 2             #
+  --
+  [1:1] lock(1) => empty  #
+  [2:2] lock(2) => empty    #
+  [3:2] unlock => 2           #
+  [4:1] unlock => 1             #
+\end{verbatim}
 
-\subsection{The Rendezvous Synchronizer}
+\subsection{Work-Stealing Queue}
 
-\subsection{The Exchanger}
+TODO THERE ARE 24 PATTERNS FOR EACH OF THE 4 VARIATIONS
 
-\subsection{The Barrier}
-
-ANY OTHERS??
+NOTE LIMITATIONS: PRIORITY QUEUE, SEMAPHORE, RENDEZVOUS SYNCHRONIZER,
+EXCHANGER, BARRIER, …

### 2015-03-22 patterns.tex
Formula examples.
full wqo proof

--- 2015-03-18 patterns.tex
@@
 of this section.
+
+We first prove that the embedding order, defined hereafter, is a wqo on width-bounded
+labeled interval orders.
+
+Let $\Sigma$ be a finite alphabet. A labeled interval order is a triple $(A,\leq,\ell)$, where $(A,\leq)$ is an interval 
+order, and $\ell:A\rightarrow \Sigma$ is a labeling function. The embedding order $\subseteq$ between labeled interval orders is defined as
+usual by:
+\begin{itemize}
+	\item $(A_1,\leq_1,\ell_1)\subseteq (A_2,\leq_2,\ell_2)$ iff there exists an injective function $h:A_1\rightarrow A_2$ that 
+	preserves the labeling, i.e., $\ell_1(x)=\ell_2(h(x))$, for every $x\in A_1$, and the order constraints, i.e.,
+	for every $x,y\in A_1$, if $x \leq_1 y$, then $h(x)\leq_2 h(y)$.
+\end{itemize}
+
+The width of an interval order $(A,\leq,\ell)$ is the maximum number of elements which are mutually unordered.
+
+\begin{lemma}
+$\subseteq$ is a wqo on width-bounded interval orders.
+\end{lemma}
+\begin{proof}
+By definition, for every interval order $(A,\leq,\ell)$, there exists a mapping $I$ from elements of $A$ to intervals over $\mathbb{N}$ such that for every $x,y\in A$, if $x\leq y$, then the interval $I(x)$ ends before the interval $I(y)$ (i.e., the upper bound of $I(x)$ is strictly smaller than the lower bound of $I(y)$).
+%In general, there are more than one mappings $I_A$ that satisfy these constraints.
+In the following, we assume that the mapping $I_A$ is arbitrary but fixed.
+Therefore, in the following, we assume that an interval order is a multiset $\Gamma$ of triples of the form $[i,j,a]$ where $[i,j]$ is an interval on the integer line and $a$ is a symbol in $\Sigma$.
+
+We prove that another order on interval orders, denoted by $\subseteqq$, which is stronger than the embedding order $\subseteq$, is a wqo.
+The order $\subseteqq$ is defined by: $\Gamma_1\subseteqq \Gamma_2$ iff there exists an injective function $h:\Gamma_1\rightarrow \Gamma_2$ such that
+\begin{enumerate}
+	\item $h$ preserves the labeling, i.e., for any triple $[i,j,a]$, $h([i,j,a])=[i',j',a]$, for some $i'$, $j'$, 
+	\item $h$ preserves the ordering constraints, i.e., for every two triples $[i,j,a]$ and $[k,l,b]$ such that $j< k$, $h([i,j,a])=[i',j',a]$, $h([k,l,b])=[k',l',b]$, and $j'< k'$.
+	\item for every two incomparable elements $x$ and $y$ of $\Gamma_1$, if the interval of $x$ starts before the interval of $y$, then the interval of $h(x)$ also starts before the interval of $h(y)$.
+Formally, for every two triples $[i,j,a]$ and $[k,l,b]$ such that $i\leq k$ (i.e., the first interval starts before the second interval), $h([i,j,a])=[i',j',a]$, $h([k,l,b])=[k',l',b]$, and $i'\leq k'$.
+\end{enumerate}
+
+Assume $\Gamma_1$, $\Gamma_2$,$\ldots$ is a \emph{bad} sequence, i.e., an infinite sequence of interval orders s.t.
+there exists no $i < j$ with $\Gamma_i\subseteqq \Gamma_j$.
+Also, assume that $\Gamma_1$ is the minimal size interval order that can start a bad sequence, $\Gamma_2$ is the minimal interval order that can continue a bad sequence starting with $\Gamma_1$, and so on.
+
+For each $\Gamma_k$, let $Min(\Gamma_k)$ be a triple $[i,j,a]$ such that 
+\begin{itemize}
+	\item $i$ is the minimal lower bound of an interval in $\Gamma_k$, and 
+	\item $j$ is the minimal upper bound of an interval in $\Gamma_k$ with lower bound $i$.
+\end{itemize}
+%In general, this triple is not unique but it’s not important which one we choose.
+
+Also, let $Init(\Gamma_k) = (Min(\Gamma_k), P_k)$, where $P_k$ is the multiset of symbols labeling elements of $\Gamma_k$ that are incomparable to $Min(\Gamma_k)$.
+Note that $P_k$ is bounded since we assume width-bounded interval orders.
+
+The infinite sequence $\Gamma_1$, $\Gamma_2$,$\ldots$ contains an infinite sequence $\Gamma_{k_1}$, $\Gamma_{k_2}$,$\ldots$ which, ignoring the integer interval, have the same value for $Init$.
+
+For each $\Gamma_k$, let $\Lambda_k$ be the interval order obtained from $\Gamma_k$ by removing the triple $Min(\Gamma_k)$.
+
+By the minimality assumptions, the infinite sequence $\Gamma_1$, $\Gamma_2$,$\ldots$,$\Gamma_{k_1-1}$, $\Lambda_{k_1}$, $\Lambda_{k_2}$,$\ldots$ is not bad. 
+Therefore, there exists $m<n$ such that $\Lambda_{k_m}\subseteqq \Lambda_{k_n}$.
+
+It remains to prove that:
+\begin{itemize}
+	\item if $Init(\Gamma_{k_m})=Init(\Gamma_{k_n})$ and $\Lambda_{k_m}\subseteqq \Lambda_{k_n}$, then $\Gamma_{k_m}\subseteqq \Gamma_{k_n}$.
+\end{itemize}
+
+Let $h$ be the injective function witnessing $\Lambda_{k_m}\subseteqq \Lambda_{k_n}$.
+We prove that the extension $h'$ of $h$ between $\Gamma_{k_m}$ and $\Gamma_{k_n}$, defined by $h'(Min(\Gamma_{k_m}))=Min(\Gamma_{k_n}$ and $h'(t)=h(t)$, for all $t\in \Lambda_{k_m}$ witnesses $\Gamma_{k_m}\subseteqq \Gamma_{k_n}$.
+Clearly, $h'$ preserves the labeling.
+
+To prove that $h'$ preserves the ordering constraints, let $Min(\Gamma_{k_m})=[i,j,a]$ and $[k,l,b]\in \Gamma_{k_m}$ such that $j< k$.
+Also, let $Min(\Gamma_{k_n})=[i',j',a]$ and $h'([k,l,b])=h([k,l,b])=[k',l',b]$.
+Assume by contradiction that $k'\leq j'$.
+Since $h$ is injective, there exists an element $[x,y,c]\in \Gamma_{k_m}$ incomparable to $Min(\Gamma_{k_m})$ which is mapped to an element $[x',y',c]$ greater than $Min(\Gamma_{k_n})$ (by definition, $\Gamma_{k_m}$ and $\Gamma_{k_n}$ have the same number of elements incomparable to $Min(\Gamma_{k_m})$ and respectively, $Min(\Gamma_{k_n})$).
+Since $[x,y,c]$ is incomparable to $[i,j,a]$, we obtain that $x\leq j< k$.
+Also, by the current assumptions, $k'<j'<x'$.
+Therefore, there exist two elements $[x,y,c]$ and $[k,l,b]$ such that the interval $[x,y]$ starts before $[k,l]$ which are mapped by $h$ to the elements $[x',y',c]$ and $[k',l',b]$ such that the interval $[x',y']$ starts after $[k',l']$.
+This contradicts the fact that $h$ witnesses $\Lambda_{k_m}\subseteqq \Lambda_{k_n}$.
+
+The properties of $Min(\Gamma_{k_m})$ and $Min(\Gamma_{k_n})$ imply that $h'$ also satisfies the third property in the definition of $\subseteqq$.
+
+Finally, since $\subseteqq$ is stronger than $\subseteq$, we get that $\subseteq$ is a wqo on width-bounded labeled interval orders.
+\end{proof}
 
@@
 
-Since the sequence of histories $h_{1}^n,h_{2}^n,\ldots$ is infinite there must exist
-
+Viewing histories as labeled interval orders, where the label of an element $o$ is the triple $(c(o),f(o),r(o))$,
+we get that the embedding order $\subseteq$ is a wqo on histories. Therefore, the infinite sequence
+$h_{1}^n,h_{2}^n,\ldots$  contains two elements $h_{i}^n$ and $h_j^n$ with $i<j$ such that $h_{i}^n\subseteq h_j^n$.
+This implies that $h_i^n\preceq h_j^n$, which ends our proof.
 \end{proof}

### 2015-03-22 patterns.tex

@@
 \begin{example}
+  \label{ex:patterns}
 
-  TODO EXAMPLE OF COMPLETE PATTERNS
-
-  \begin{figure}[t]
-    \centering
-    \includegraphics[scale=0.2]{figures/register-pattern-1},
-    \includegraphics[scale=0.2]{figures/register-pattern-2},
-    \includegraphics[scale=0.2]{figures/register-pattern-3}
-    \caption{Three histories which generate the complement of the atomic
-      register ADT kernel.}
-    \label{fig:register-patterns}
-  \end{figure}
+  The following four histories generate the complement of the atomic
+  single-value register ADT, witnessing either an unmatched read operation:
+\begin{verbatim}
+  [1:X] read => 1 (RO)  #
+\end{verbatim}
+  a read of an uninitialized register occurring after some write:
+\begin{verbatim}
+  [1:1] write(1)            #
+  [2:2] read => empty (RO)    #
+\end{verbatim}
+  a read which happens before its matching write operation:
+\begin{verbatim}
+  [1:2] read => 1 (RO)  #
+  [2:2] write(1)          #
+\end{verbatim}
+  and a read matching a write which is not the most recent:
+\begin{verbatim}
+  [1:1] write(1)        #
+  [2:2] write(2)          #
+  [3:1] read => 1 (RO)      #
+\end{verbatim}
+  Every sequential history not admitted by the atomic register ADT embeds at
+  least one of these four histories.
 

### 2015-03-22 nature.tex
More on nature.

@@
 
-TODO DEMONSTRATE FAITHFUL MATCHING SCHEMES, UNIQUE AND WIDTH-BOUNDED KERNELS,
-NORMALITY, AND PREDICTABILITY FOR EACH NATURALLY-OCCURRING ADT
-
-TODO EXPLAIN “ATOMIC”
+In this section we demonstrate that the premises used in the development of our
+symbolic ADT inference algorithm — i.e.,~unique and bounded-width kernels,
+faithful and normalizing matching schemes, and predictability — hold for the
+ADTs which are typically provided by concurrent object libraries. Furthermore,
+we demonstrate that the algorithm developed in this work computes precise
+symbolic representations for these ADTs which can be used in symbolic checkers
+for observational refinement~\cite{conf/pldi/EmmiEH15}. Our publicly available
+implementation\footnote{The URL has been suppressed for anonymity purposes.}
+enumerates the sequential histories\footnote{Our current implementation is
+limited to atomic ADTs. For non-atomic ADTs, the enumeration must cover all
+$k$-width histories, for some $k \in \mathbb{N}$.} not admitted by reference
+implementations of the undermentioned ADTs, keeping only the histories which
+are not generated by any other. In each case, our algorithm terminates in a
+matter of seconds.
 
@@
   \item read $\Rightarrow v$ returns the last-stored value $v$, or $\perp$
-  if no value has been stored.
+  if no value has yet been stored.
 
 \end{itemize}
+Its kernel is sequential.
+
 We say an execution of read and write methods \emph{writes unique values} if
@@
 \end{itemize}
-The kernel of atomic registers is the set of sequential histories given by
-sequences
-\begin{align*}
-  \bigcup_{0 \le k \le n < \omega}
-  \Big( \underline{\mathrm{read}:i} \Big)_{i=1}^{k}\ 
-  \Big( \underline{\mathrm{write}:i}\ (\mathrm{read}:i)^\ast \Big)_{i=k}^{n}
-\end{align*}
-in which the initial read operations match themselves, write operations match
-themselves, and each subsequent read operation matches the nearest preceding
-write operation.
 
@@
 
-THIS IS THE SET OF PATTERNS
-
+Our inference algorithm computes following set of histories to generate the
+complement of the atomic register ADT:
 \begin{verbatim}
@@
 \end{verbatim}
-
-The atomic register ADT is predictable \ldots
+As new basis histories are generated for $n=1$, $n=2$, and $n=3$ matches, the
+atomic register ADT is predictable.
 
@@
 TODO THERE ARE 24 PATTERNS FOR EACH OF THE 4 VARIATIONS
-
-NOTE LIMITATIONS: PRIORITY QUEUE, SEMAPHORE, RENDEZVOUS SYNCHRONIZER,
-EXCHANGER, BARRIER, …

### 2015-03-22 patterns.tex
Motivations for many sections.

--- 2015-03-22 patterns.tex
@@
-%!TEX root = draft.tex
-\section{Computing ADT Complements}
+\section{Computing Finite ADT Representations}
 \label{sec:patterns}
 
-TODO MOTIVATE THIS SECTION; ALSO, RETHINK THE TITLE OF THIS SECTION
+In this section we demonstrate that naturally-occurring ADTs can be precisely
+represented by finite sets of histories, despite the fact that these ADTs admit
+infinite sets of histories. This results is nontrivial, and relies on the
+identification of certain algebraic properties of the sets of histories
+admitted by ADTs. In particular, we find that these sets of histories are
+closed under the removal of certain operations, and that sets of histories
+adhere to a well-founded ordering under the relation induced by such removals.
 
@@
 
-TODO MOTIVATE WQO
-
-A \emph{well-quasi-ordering (wqo)} $R$ on a set $X$ is a reflexive, transitive
+Besides this closure property, the $\preceq$ relation enjoys a certain notion
+of well-foundedness (when restricted to certain class of histories): a
+\emph{well-quasi-ordering (wqo)} $R$ on a set $X$ is a reflexive, transitive
 binary relation on $X$ for which in every infinite sequence $x_0 x_1 \ldots$ of
@@
 \begin{example}
+  \label{ex:patterns}
 

### 2015-03-22 formula.tex
More motivation paragraphs.

@@
 
-TODO MOTIVATE THIS SECTION
+While Section~\ref{sec:patterns} demonstrates that (the complements) of
+naturally-occurring ADTs can be represented finitely, in this section we
+demonstrate that such finite representations have logical interpretations,
+allowing us to derive the formulas representing ADTs. In what follows, we
+describe how to obtain a formula which is satisfied by the histories which
+embed a given history, via $\preceq$. Then, using the finite set of histories
+which represent the complement of a given ADT, we represent the ADT itself as
+the conjunction of negations of these embedding formulas. The resulting formula
+is satisfied only on the histories which do not embed the generators of a given
+ADT’s complement.
 

### 2015-03-22 matching.tex
More motivation paragraphs.

@@
 
-TODO MOTIVATE THIS SECTION, AND ALSO ACCOUNT FOR THE READ-ONLY OPERATION
-LABELING TOO.
+Our solution to the ADT inference problem relies on identifying faithful
+matching schemes for ADT implementations. Though Section~\ref{sec:nature} shows
+such matching schemes exist for naturally-occurring ADTs, the next natural
+question with regard to automation is whether these matching schemes can be
+generated automatically.
 
-A \emph{language} $L = \tup{E,P,Q}$ is a set $E$ of executions, along with
-finite sequences $P = P_1 P_2 \ldots$ and $Q = Q_1 Q_2 \ldots$ of binary and
-ternary predicates $P_i(e,o_1)$ and $Q_i(e,o_1,o_2)$ ranging over the
-executions of $E$ and their operations. When $P$ is a $k$-length sequence of
-$n$-ary predicates, we write $P(x_1, \ldots, x_n)$ to denote the valuation
-sequence
+In this section we demonstrate that matching schemes can in fact be generated
+systematically with minimal manual specification by reduction to logical
+satisfiability. Essentially, we formulate matching schemes as uninterpreted
+functions in satisfiability queries constrained by the requirement that they
+\emph{normalize} an implementation’s executions, i.e.,~generate histories which
+are closed under the operation removals of Section~\ref{sec:patterns}. This
+requirement is given with respect to an enumeration of execution pairs in which
+the first is admitted by the given implementation, and the second is not, yet
+it is a projection of the first’s operations. Consequently, this prohibits any
+normalizing matching scheme from considering the projected operations a match.
+In what follows, we suppose the read-only component $R$ of matching schemes
+$\tup{M,R}$ is given, and focus on the generation of the per-execution matching
+functions $M$.
+
+The required manual specification includes identifying a class of executions to
+which an implementation should be subjected, and a set of predicates that are
+required in the logic of matching schemes. Formally, a \emph{language} $L =
+\tup{E,P,Q}$ is a set $E$ of executions, along with finite sequences $P = P_1
+P_2 \ldots$ and $Q = Q_1 Q_2 \ldots$ of binary and ternary predicates
+$P_i(e,o_1)$ and $Q_i(e,o_1,o_2)$ ranging over the executions of $E$ and their
+operations. When $P$ is a $k$-length sequence of $n$-ary predicates, we write
+$P(x_1, \ldots, x_n)$ to denote the valuation sequence
 \begin{align*}
@@
   the argument value to each write operation is unique. Consider the language
-  $L_\mathrm{reg}$
   whose executions write unique values, with the following predicates:
@@
   the value read. In the case such a write exists, it is unique in any
-  execution which writes unique values. The match is otherwise undefined.
+  execution which writes unique values. The match is otherwise undefined. This
+  scheme is exactly that which is specified in Example~\ref{ex:matching} of
+  Section~\ref{sec:histories}.
 
@@
 $H(\mathcal{I},M)$ is normal.
-
-\begin{example}
-
-  TODO SHOW THAT PREVIOUS SIMPLE MATCHING SCHEME NORMALIZES SOMETHING
-
-\end{example}
 

### 2015-03-23 related.tex
Added note for related work.
More related work.

@@
 representations, and work directly with given reference implementations, these
-approaches are intractable, since the number of linearizations is exponential
-in execution length.
+approaches are intractable as they elaborate an exponential\footnote{The number
+of possible linarizations is exponential in execution length.} number of
+linearizations.
 
@@
 conclusive evidence of a refinement violation in case of failure.
+
+The idea of inferring minimal finite representations of otherwise-infinite sets
+of implementation behaviors has also been proposed as an optimization to reduce
+the impact of state-space explosion in compositional
+verification~\cite{journals/ase/GiannakopoulouPB05}. This approach is based on
+learning minimal automata for regular languages, and requires the
+implementation of both language membership and equivalence
+queries~\cite{journals/iandc/Angluin87}. In our setting, it is unclear whether
+ADT implementations may admit regular characterizations in general, and
+furthermore how to automatically discharge language equivalence queries between
+candidate automata and the source code of reference implementations.
 

### 2015-03-23 discussion.tex
More on nature.
Discussion section.

--- 2015-03-07 nature.tex
@@
-\section{Naturally-Occurring ADTs}
-\label{sec:nature}
+\section{Discussion}
+\label{sec:discussion}
 
-TODO DEMONSTRATE FAITHFUL MATCHING SCHEMES, NORMALITY, AND PREDICTABILITY FOR
-EACH NATURALLY-OCCURRING ADT
+While the theoretical foundation proposed in this work covers the majority of
+naturally-occurring ADTs, it is worth mentioning a few conceptual limitations
+and possible solutions to overcome them.
 
-\subsection{The Atomic Queue}
+First, while our theoretical foundation does cover “concurrency-aware” ADT
+specifications~\cite{conf/podc/HemedR14}, i.e.,~non-atomic ADTs like the
+\emph{rendezvous synchronizer}, \emph{barrier}, and \emph{exchanger}, whose
+kernels have bounded width greater than $1$, our current implementation only
+handles atomic ADTs. In principle, this limitation is not fundamental. The key
+difference is in the enumeration of kernel histories. For a given ADT,
+determining whether a given history is in the kernel or its complement reduces
+to checking whether there exists an execution of which the history is an
+abstraction. For atomic ADTs, only a single sequential execution need be
+examined, in which no two operations overlap. For non-atomic ADTs, every
+possible interleaving of the internal actions of operations need be examined.
+While we expect this calculation to remain feasible given that ADT complements
+are normally represented with histories with few operations, we do expect it to
+incur a noticeable cost.
 
-\subsection{The Atomic Stack}
+Second, ADTs whose specifications rely on relations (besides equality) on
+method argument and return values cannot be expressed with our notion of
+histories based on matching schemes. A notable example of such an ADT is the
+\emph{priority queue}, whose \emph{dequeue} operations return the
+smallest/largest \emph{enqueue}d value which has not yet been dequeued. It
+appears that such ADTs are rare in practice. Overcoming this limitation would
+require a refinement to matching schemes which can track additional relations
+among data values.
 
-\subsection{The Atomic Set}
-
-\subsection{The Atomic Register}
-
-\subsection{The Atomic Lock}
-
-\subsection{The Atomic Semaphore}
-
-\subsection{The Rendezvous Synchronizer}
-
-\subsection{The Exchanger}
-
-\subsection{The Barrier}
-
-ANY OTHERS??
+Finally, in some cases it is unclear how to express matching schemes
+deterministically. For example, each \emph{wait} operation of \emph{semaphore}
+ADT should naturally match the \emph{notify} operation which enabled it.
+However, it is not possible to determine this based on operation labels alone,
+and it is not clear whether all implementations effectively keep track of this
+correspondence. Furthermore, choosing matches arbitrarily would be unsound, and
+result in inferring ADTs with artificial constraints, e.g.,~that the notify and
+wait operations behave according to a FIFO discipline, each wait matching the
+oldest unmatched notify. Overcoming this limitation may require introducing a
+notion of nondeterminism into matching schemes.

### 2015-03-23 nature.tex
Added lots about nature.

--- 2015-03-21 nature.tex
@@
 
-TODO DEMONSTRATE FAITHFUL MATCHING SCHEMES, UNIQUE AND WIDTH-BOUNDED KERNELS,
-NORMALITY, AND PREDICTABILITY FOR EACH NATURALLY-OCCURRING ADT
+\makeatletter
+\renewcommand{\verbatim@font}{\ttfamily\scriptsize}
+\makeatother
 
-TODO EXPLAIN “ATOMIC”
+In this section we demonstrate that the premises used in the development of our
+symbolic ADT inference algorithm — i.e.,~unique and bounded-width kernels,
+faithful and normalizing matching schemes, and predictability — hold for the
+ADTs which are typically provided by concurrent object libraries. Furthermore,
+we demonstrate that the algorithm developed in this work computes precise
+symbolic representations for these ADTs which can be used in symbolic checkers
+for observational refinement~\cite{conf/pldi/EmmiEH15}. Our publicly available
+implementation\footnote{The URL has been suppressed for anonymity purposes.}
+enumerates the sequential histories\footnote{Our current implementation is
+limited to atomic ADTs. For non-atomic ADTs, the enumeration must cover all
+$k$-width histories, for some $k \in \mathbb{N}$.} not admitted by reference
+implementations of the undermentioned ADTs, keeping only the histories which
+are not generated by any other. In each case, our algorithm terminates in a
+matter of seconds.
 
@@
 
-Atomic register objects implement an atomic single-value store, and provide two
+The atomic register implements an atomic single-value store, providing two
 methods:
@@
 
-  \item read $\Rightarrow v$ returns the last-stored value $v$, or $\perp$
-  if no value has been stored.
+  \item read $\Rightarrow v$ returns the last-stored value $v$, or the nil
+  value $\perp$ if no value has yet been stored.
 
 \end{itemize}
+As its name implies, its kernel is sequential.
+
 We say an execution of read and write methods \emph{writes unique values} if
@@
 \end{itemize}
-The kernel of atomic registers is the set of sequential histories given by
-sequences
-\begin{align*}
-  \bigcup_{0 \le k \le n < \omega}
-  \Big( \underline{\mathrm{read}:i} \Big)_{i=1}^{k}\ 
-  \Big( \underline{\mathrm{write}:i}\ (\mathrm{read}:i)^\ast \Big)_{i=k}^{n}
-\end{align*}
-in which the initial read operations match themselves, write operations match
-themselves, and each subsequent read operation matches the nearest preceding
-write operation.
-
-TODO ARGUE THAT THIS IS FAITHFUL
+This matching scheme is faithful since two executions with the same history
+are identical up to homomorphic renaming of data values, and the register ADT
+only relates data values via equality.
 
@@
 
-THIS IS THE SET OF PATTERNS
-\begin{align*}
-  & \mathrm{read}:\_ \\
-  & \underline{\mathrm{write}:1}\ \underline{\mathrm{read}:2} \\
-  & \mathrm{read}:1\ \underline{\mathrm{write}:1} \\
-  & \underline{\mathrm{write}:1}\ \underline{\mathrm{write}:2}\ \mathrm{read}:1
-\end{align*}
-The atomic register ADT is predictable \ldots
+Our inference algorithm computes following four histories to generate the
+complement of the atomic register ADT:
+\begin{verbatim}
+  [1:X] read => 1 (RO)  #          [1:2] read => 1 (RO)  #
+  ---                              [2:2] write(1)          #
+  [1:1] write(1)            #      ---
+  [2:2] read => empty (RO)    #    [1:1] write(1)        #
+\end{verbatim}
+As new generators are discovered for $n=1$ and $n=2$ matches only, the atomic
+register ADT is predictable.
 
-\subsection{The Atomic Queue}
+\subsection{The Atomic Queue \& The Atomic Stack}
 
-\subsection{The Atomic Stack}
+Atomic queues and stacks implement atomic collections of data values with
+first-in-first-out (FIFO) and last-in-first-out (LIFO) removal order,
+respectively, providing two methods:\footnote{These methods are also known as
+enqueue and dequeue, or push and pop.}
+\begin{itemize}
+
+  \item add$(v)$ adds the value $v$ to the collection, and
+
+  \item remove $\Rightarrow v$ returns the nil value $v = \mathord\perp$ if the
+  collection is empty, and otherwise removes and returns the least- or
+  most-recently added value $v$, respectively.
+
+\end{itemize}
+As their names imply, their kernels are sequential.
+
+An execution \emph{adds unique values} if the argument value to each add
+operation is unique. We give a faithful matching scheme over executions which
+add unique values as follows:
+\begin{itemize}
+
+  \item add$(v)$ operations match themselves, and
+
+  \item remove $\Rightarrow v$ operations are read-only, and match themselves
+  if $v = \mathord\perp$ is the nil value. If $v \neq \mathord\perp$, remove
+  $\Rightarrow v$ operations are not read-only, and match the unique add$(v)$
+  operation, if one exists, and otherwise have no match.
+
+\end{itemize}
+This matching scheme is faithful since two executions with the same history are
+identical up to homomorphic renaming of data values, and the queue and stack
+ADTs only relate data values via equality.
+
+It is not hard to see that the histories of atomic queues and stacks are
+normal. They are closed under removal of read-only operations: only empty
+removes, i.e.,~remove $\Rightarrow \mathord\perp$, are read-only. They are
+closed under the removal of duplicate operations: only non-empty removes can be
+duplicates, and such duplicates are not admitted in the first place, since each
+value is added only once. Similarly, unmatched operations, i.e.,~removes that
+return values that have not been added, are not admitted in the first place.
+Finally, since the removal of entire matches preserves the FIFO/LIFO behavior
+of the entire collection, and the correctness of empty returns, histories are
+also closed under match removal.
+
+We compute the following seven histories to generate the complement of the
+atomic queue ADT:
+\begin{verbatim}
+  [1:X] remove => 1  #           [1:1] add(1)       #
+  ---                            [2:1] remove => 1    #
+  [1:1] add(1)            #      [3:1] remove => 1      #
+  [2:2] remove => - (RO)    #    ---
+  ---                            [1:1] add(1)            #
+  [1:2] remove => 1  #           [2:2] remove => - (RO)    #
+  [2:2] add(1)         #         [3:1] remove => 1           #
+  ---                            ---
+  [1:1] add(1)       #           [1:1] add(1)       #
+  [2:2] add(2)         #         [2:2] add(2)         #
+  [3:2] remove => 2      #       [3:2] remove => 2      #
+                                 [4:1] remove => 1        #
+\end{verbatim}
+The histories computed for the atomic stack ADT are nearly identical,
+substituting only the bottom two histories for the following:
+\begin{verbatim}
+  [1:1] add(1)       #        [1:1] add(1)       #
+  [2:2] add(2)         #      [2:2] add(2)         #
+  [3:1] remove => 1      #    [3:1] remove => 1      #
+                              [4:2] remove => 2        #
+\end{verbatim}
+As new generators are discovered for $n=1$ and $n=2$ matches only, these ADTs
+are predictable.
 
@@
 
-\subsection{The Priority Queue (?)}
+Atomic sets implement collections which store one copy of each inserted data
+value no matter how many times the same value is inserted, until removed,
+providing three methods:
+\begin{itemize}
+
+  \item insert$(u) \Rightarrow v$ inserts the value $u$ to the collection,
+
+  \item remove$(u) \Rightarrow v$ removes $u$ from the collection, and
+
+  \item contains$(u) \Rightarrow v$ checks whether the set contains $u$.
+
+\end{itemize}
+Each operation returns the nil value $v = \mathord\perp$ when $u$ is not yet
+present, and otherwise returns $v = u$. As its name implies, its kernel is
+sequential.
+
+An execution \emph{inserts unique values} if the argument value to each insert
+operation which returns $\mathord\perp$ is unique. We give a faithful matching
+scheme over executions which insert unique values as follows:
+\begin{itemize}
+
+  \item any operation returning $v = \mathord\perp$ matches itself, and
+
+  \item any operation returning $v \neq \mathord\perp$ matches the unique
+  insert$(u) \Rightarrow \mathord\perp$ operation, if one exists, and otherwise
+  has no match.
+
+\end{itemize}
+This matching scheme is faithful since two executions with the same history are
+identical up to homomorphic renaming of data values, and sets only relates data
+values via equality.
+
+We compute the following nine histories to generate the complement of the
+atomic set ADT:
+\begin{verbatim}
+  [1:X] insert(1) => 1 (RO)  #        [1:1] insert(1) => -  #
+  ---                                 [2:1] remove(1) => 1    #
+  [1:X] remove(1) => 1  #             [3:1] remove(1) => 1      #
+  ---                                 ---
+  [1:X] contains(1) => 1 (RO)  #      [1:1] insert(1) => -         #
+  ---                                 [2:1] remove(1) => 1           #
+  [1:1] insert(1) => -       #        [3:1] contains(1) => 1 (RO)      #
+  [2:2] remove(1) => - (RO)    #      ---
+  ---                                 [1:1] insert(1) => -       #
+  [1:1] insert(1) => -         #      [2:2] remove(1) => - (RO)    #
+  [2:2] contains(1) => - (RO)    #    [3:1] remove(1) => 1           #
+\end{verbatim}
+As new generators are discovered for $n=1$ and $n=2$ matches only, the atomic
+set ADT is predictable.
 
@@
 
-\subsection{The Atomic Semaphore}
+Atomic locks implement resource-based mutual exclusion by providing two
+methods
+\begin{itemize}
 
-\subsection{The Rendezvous Synchronizer}
+  \item lock$(u) \Rightarrow v$ acquires the lock resource,
 
-\subsection{The Exchanger}
+  \item unlock $\Rightarrow v$ releases the lock resource.
 
-\subsection{The Barrier}
+\end{itemize}
+An operation returns the nil value $v = \mathord\perp$ when the lock is not
+currently held, indicating success for lock, and failure for unlock. Otherwise,
+the operations return the value $u = v$ passed as an argument to the last
+successful lock$(v)$ operation.
 
-ANY OTHERS??
+An execution \emph{uses unique keys} if the argument value to each lock
+operation is unique. We give a faithful matching scheme over executions using
+unique keys as follows:
+\begin{itemize}
+
+  \item any operation returning $v = \mathord\perp$ matches itself, and
+
+  \item any operation returning $v \neq \mathord\perp$ matches the unique
+  lock$(u) \Rightarrow \mathord\perp$ operation, if one exists, and otherwise
+  has no match.
+
+\end{itemize}
+It is easy to see that this matching scheme is faithful, and normalizing.
+
+We compute the following eleven histories to generate the complement of the
+atomic lock ADT:
+\begin{verbatim}
+  [1:X] unlock => 1  #             [1:X] lock(1) => 1 (RO)  #
+  ---                              ---
+  [1:1] lock(1) => -  #            [1:1] lock(1) => -  #
+  [2:2] lock(2) => -    #          [2:1] unlock => 1     #
+  ---                              [3:1] unlock => 1       #
+  [1:1] lock(1) => -      #        ---
+  [2:2] unlock => - (RO)    #      [1:1] lock(1) => -      #
+  ---                              [2:2] unlock => - (RO)    #
+  [1:1] lock(1) => -  #            [3:1] unlock => 1           #
+  [2:2] lock(2) => -    #          ---
+  [3:1] unlock => 1       #        [1:1] lock(1) => -  #
+  ---                              [2:2] lock(2) => -    #
+  [1:1] lock(1) => -  #            [3:1] unlock => 1       #
+  [2:2] lock(2) => -    #          [4:2] unlock => 2         #
+  [3:2] unlock => 2       #        ---
+  ---                              [1:1] lock(1) => -  #
+  [1:1] lock(1) => -       #       [2:2] lock(2) => -    #
+  [2:1] unlock => 1          #     [3:2] unlock => 2       #
+  [3:1] lock(2) => 1 (RO)      #   [4:1] unlock => 1         #
+\end{verbatim}
+As new generators are discovered for $n=1$ and $n=2$ matches only, the atomic
+lock ADT is predictable.
+
+\subsection{Work-Stealing Queue}
+
+TODO THERE ARE 24 PATTERNS FOR EACH OF THE 4 VARIATIONS

### 2015-03-24 histories.tex
Formula example.

@@
-%!TEX root = draft.tex
 \section{Histories of ADT Implementations}
@@
 
-TODO MOTIVATE THIS SECTION
+Implementations of abstract data types (ADTs) provide methods which are called,
+possibly concurrently, by client programs. We capture the possible logs of
+method calls and returns in executions of client programs abstractly, in order
+to facilitate the development of symbolic ADT representations.
 
-We fix an arbitrary infinite set $\mathbb{O}$ of operation identifiers, and
-given sets $\mathbb{M}$ and $\mathbb{V}$ of method names and values. A
-\emph{call action} binds an operation $o \in \mathbb{O}$ with a method name $p
-\in \mathbb{M}$ and argument value $v \in \mathbb{V}$, while a \emph{return
-action} binds an operation $o \in \mathbb{O}$ with a return value $v \in
-\mathbb{V}$. An \emph{execution} $e$ is a sequence of call and return actions
-where
+To this end, we fix an arbitrary infinite set $\mathbb{O}$ of operation
+identifiers, and given sets $\mathbb{M}$ and $\mathbb{V}$ of method names and
+values. A \emph{call action} binds an operation $o \in \mathbb{O}$ with a
+method name $p \in \mathbb{M}$ and argument value $v \in \mathbb{V}$, while a
+\emph{return action} binds an operation $o \in \mathbb{O}$ with a return value
+$v \in \mathbb{V}$. An \emph{execution} $e$ is a sequence of call and return
+actions where
 \begin{itemize}
@@
 \begin{example}
+  \label{ex:histories-of-executions}
 

### 2015-03-24 inference.tex

@@
 
-TODO MOTIVATE THIS SECTION
+In this section we formalize a notion of abstract data type and define the
+corresponding refinement and inference problems. These are the foundational
+problems addressed in this work.
 
@@
 \end{align*}
-where $\to = (\to_\mathrm{o} \cup \to_\mathrm{p} \cup \to_\mathrm{c})$. In
-Section~\ref{sec:nature} we demonstrate that naturally-occurring ADTs have
+where $\mathord\to = (\to_\mathrm{o} \cup \to_\mathrm{p} \cup \to_\mathrm{c})$.
+In Section~\ref{sec:nature} we demonstrate that naturally-occurring ADTs have
 unique, bounded-width kernels. For the remainder of this work, we assume ADTs
-have unique, bounded-width kernels, denoting \emph{the kernel} of $A$ by
-$\ker A$. When $\ker A$ has width $1$, we say $A$ is \emph{atomic}.
+have unique, bounded-width kernels, denoting \emph{the kernel} of $A$ by $\ker
+A$. When $\ker A$ has width $1$, we say $A$ is \emph{atomic}.
 
@@
 computing these automatically, we fix a language for symbolic representation. A
-\emph{history formula} $F$ is a first-order logic formula with
+\emph{history formula} is a first-order logic formula with
 \begin{itemize}
@@
   \begin{align*}
-    \forall x, y, z.\ 
-    & {\sf c}(x) \land {\sf f}(x) = \mathrm{write} \land {\sf m}(x) = x \\
-    & \land {\sf c}(y) \land {\sf f}(y) = \mathrm{write} \land {\sf m}(y) = y \\
-    & \land {\sf c}(z) \land {\sf f}(z) = \mathrm{read} \land {\sf m}(z) = x \\
-    & \land x < y 
-      \implies z < y
+    \forall x_1, x_2, x_3.\ 
+    & {\sf c}(x_1) \land {\sf f}(x_1) = \mathrm{write} \land {\sf m}(x_1) = x_1 \\
+    & \land {\sf c}(x_2) \land {\sf f}(x_2) = \mathrm{write} \land {\sf m}(x_2) = x_2 \\
+    & \land {\sf c}(x_3) \land {\sf f}(x_3) = \mathrm{read} \land {\sf m}(x_3) = x_1 \\
+    & \land x_1 < x_2 
+      \implies x_3 < x_2
   \end{align*}
@@
 
-  The \emph{symbolic ADT inference problem} is to compute a history formula $F$
+  The \emph{symbolic ADT inference problem} is to compute a history formula
   representing the ADT $H(\mathcal{I})$ of a given implementation $\mathcal{I}$.

### 2015-03-24 nature.tex

@@
 
-TODO THERE ARE 24 PATTERNS FOR EACH OF THE 4 VARIATIONS
+The work-stealing queue~\cite{journals/dc/HendlerLMS06} implements a collection
+of data values with first-in-first-out\footnote{There are actually four
+variations to the work-stealing queue, depending on the ends from which the
+take and steal operations remove values. While our approach works
+indifferently, below we focus on the variation where both remove the least
+recent.} (FIFO) removal order, proving three methods:
+\begin{itemize}
+
+  \item give$(v)$ adds the value $v$ to the queue,
+
+  \item take $\Rightarrow v$ removes the value $v$, and
+
+  \item steal $\Rightarrow v$ removes the value $v$.
+
+\end{itemize}
+Unlike atomic queues, the work-stealing queue permits values to be removed
+twice: once normally, via the take operation, and once exceptionally, via the
+steal operation. We give a faithful matching scheme over executions which add
+unique values analogously to the scheme for stacks and queues, in which give
+operations match themselves, while take and steal operations match the give
+operation which added their returned value, or themselves, in case the nil
+value is returned. Note that the kernel of work-stealing queues has width $2$,
+since a pairs of concurrent take and steal operations returning the same value
+are permitted, while sequentially they are not. As our implementation is
+currently limited to width-$1$ histories (see the discussion in
+Section~\ref{sec:discussion}), below we generate only the width-$1$ complement.
+
+We compute the following twenty-four histories to generate the complement of
+the work-stealing queue ADT:
+\begin{verbatim}
+  [1:_] take => 1  #             [1:_] steal => 1  #
+  ---                            ---
+  [1:1] give(1)         #        [1:1] give(1)          #
+  [2:2] take => - (RO)    #      [2:2] steal => - (RO)    #
+  ---                            ---
+  [1:2] take => 1  #             [1:2] steal => 1  #
+  [2:2] give(1)      #           [2:2] give(1)       #
+  ---                            ---
+  [1:1] give(1)    #             [1:1] give(1)     #
+  [2:2] give(2)      #           [2:2] give(2)       #
+  [3:2] take => 2      #         [3:2] steal => 2      #
+  ---                            ---
+  [1:1] give(1)    #             [1:1] give(1)     #
+  [2:1] take => 1    #           [2:1] take => 1     #
+  [3:1] take => 1      #         [3:1] steal => 1      #
+  ---                            ---
+  [1:1] give(1)         #        [1:1] give(1)         #
+  [2:2] take => - (RO)    #      [2:2] take => - (RO)    #
+  [3:1] take => 1           #    [3:1] steal => 1          #
+  ---                            ---
+  [1:1] give(1)     #            [1:1] give(1)     #
+  [2:1] steal => 1    #          [2:1] steal => 1    #
+  [3:1] take => 1       #        [3:1] steal => 1      #
+  ---                            ---
+  [1:1] give(1)          #       [1:1] give(1)          #
+  [2:2] steal => - (RO)    #     [2:2] steal => - (RO)    #
+  [3:1] take => 1            #   [3:1] steal => 1           #
+  ---                            ---
+  [1:2] take => 1   #            [1:3] take => 1   #
+  [2:2] give(1)       #          [2:3] steal => 1    #
+  [3:2] steal => 1      #        [3:3] give(1)         #
+  ---                            ---
+  [1:2] steal => 1  #            [1:3] steal => 1  #
+  [2:2] give(1)       #          [2:3] take => 1     #
+  [3:2] take => 1       #        [3:3] give(1)         #
+  ---                            ---
+  [1:1] give(1)    #             [1:1] give(1)     #
+  [2:2] give(2)      #           [2:2] give(2)       #
+  [3:2] take => 2      #         [3:2] take => 2       #
+  [4:1] take => 1        #       [4:1] steal => 1        #
+  ---                            ---
+  [1:1] give(1)     #            [1:1] give(1)     #
+  [2:2] give(2)       #          [2:2] give(2)       #
+  [3:2] steal => 2      #        [3:2] steal => 2      #
+  [4:1] take => 1         #      [4:1] steal => 1        #
+\end{verbatim}
+As new generators are discovered for $n=1$ and $n=2$ matches only, the
+work-stealing queue ADT is predictable.

### 2015-03-24 nature.tex
Register histories typo.

@@
 \begin{verbatim}
-  [1:X] read => 1 (RO)  #          [1:2] read => 1 (RO)  #
-  ---                              [2:2] write(1)          #
+  [1:X] read => 1 (RO)  #        [1:1] write(1)        #
+  ---                            [2:2] read => - (RO)    #
   [1:1] write(1)        #        ---
-  [2:2] read => empty (RO)    #    [1:1] write(1)        #
+  [2:2] write(2)          #      [1:2] read => 1 (RO)  #
+  [3:1] read => 1 (RO)      #    [2:2] write(1)          #
 \end{verbatim}

### 2015-03-24 draft.tex
Some algebra.
Added references to nature.
Added notion of “weight”.
Better Section 2.
Better organization.
Better titles.
Better definition of weight, includes duplicates.
Soft description of the formulas.
Split algorithm section in two.
Added a figure.
started wqo proof
Closure under unmatched-removal.
More on nature.
Updated intro.

@@
 \usepackage{microtype}
+\usepackage{graphicx}
 \usepackage{hyperref}
 \usepackage{amsmath}
+\usepackage{amssymb}
 \usepackage{amsthm}
@@
 \newtheorem{lemma}{Lemma}
+\newtheorem{example}{Example}
+\newtheorem{definition}{Definition}
+\DeclareMathOperator{\dom}{dom}
+\DeclareMathOperator{\img}{img}
+\DeclareMathOperator{\width}{width}
+\DeclareMathOperator{\weight}{weight}
+\DeclareMathOperator{\freq}{freq}
+
+\newcommand{\set}[1]{{\{{#1}\}}}
+\newcommand{\mset}[1]{{\{\!\{ #1 \}\!\}}}
+\newcommand{\tup}[1]{{\left\langle{#1}\right\rangle}}
+\renewcommand{\implies}{\Rightarrow}
 
@@
 
-\authorinfo{Michael Emmi}
-           {IMDEA Software Institute}
-           {michael.emmi@imdea.org}
-\authorinfo{Constantin Enea\and Jad Hamza}
-           {Université Paris Diderot}
-           {\{cenea,jhamza\}@univ-paris-diderot.fr}
+\authorinfo{}{}{}
+
+% \authorinfo{Michael Emmi}
+%            {IMDEA Software Institute}
+%            {michael.emmi@imdea.org}
+% \authorinfo{Constantin Enea}
+%            {Université Paris Diderot}
+%            {cenea@univ-paris-diderot.fr}
 
@@
 
-  Checking that efficient implementations of concurrent objects adhere to their
-  abstract data types (ADTs) can be made tractable when ADTs are given by
-  symbolic, logical representations. Obtaining these symbolic representations,
-  however, is a complex task requiring rare expertise. In practice, concurrent
-  objects are specified indirectly by imperative reference implementations.
+  Formal specifications are a vital component to the automated verification of
+  software systems. In the case of efficient implementations of concurrent
+  objects like atomic registers, queues, and locks, checking adherence to their
+  abstract data types (ADTs) is a difficult problem, yet can be made tractable
+  when the ADT specifications are given by symbolic, logical representations.
+  Obtaining these symbolic representations, however, is a complex task
+  requiring rare expertise. In practice, concurrent objects are specified
+  indirectly by imperative reference implementations.
 
@@
   be automatically generated from the executions of reference implementations.
-  Our approach exploits key features of naturally-occurring ADTs: violations
+  Our approach exploits two key features of naturally-occurring ADTs: violations
   can be decomposed into a small set of representative patterns, and these
@@
   of efficient symbolic verification algorithms without the burden of manual
-  specification.
+  specification. Furthermore, the concise ADT violation patterns we generate
+  are human-readable, and can serve as useful, formal documentation.
 
@@
 \input intro
+\input histories
 \input inference
-\input algebra
-\input algorithm
+\input patterns
+\input formula
 \input matching
-\input empirical
+\input nature
+\input discussion
 \input related

### 2015-03-24 intro.tex
Updated intro.

--- 2015-03-11 intro.tex
@@
 with more operations than a given violating execution is not necessarily a
-violation itself. For instance, an execution which contains only a single {\tt
-pop} operation returning the value {\tt 1} is a violation to the atomic stack
-ADT, whereas an execution containing an additional {\tt push(1)} operation,
-overlapping in time with the {\tt pop}, is not. Further complication arises
+violation itself. For instance, an execution which contains only a single
+pop operation returning the value $1$ is a violation to the atomic stack
+ADT, whereas an execution containing an additional push$(1)$ operation,
+overlapping in time with the pop, is not. Further complication arises
 from the infinite set of possible data values, i.e.,~method argument and return
@@
 without being sensitive to the data values themselves. For instance, a
-sequential execution in which {\tt 1} and {\tt 2} are pushed and subsequently
+sequential execution in which $1$ and $2$ are pushed and subsequently
 popped in the same order violates the atomic stack ADT. Yet, while replacing
-both values {\tt 1} and {\tt 2} with the value {\tt 1} results in a
-\emph{valid} stack execution, replacing them with {\tt 3} and {\tt 4} results
+both values $1$ and $2$ with the value $1$ results in a
+\emph{valid} stack execution, replacing them with $3$ and $4$ results
 in a violation.
@@
 \emph{matchings}. Intuitively, operations which refer to the same instances of
-values belong to the same matching. For example, a {\tt pop} operation
-returning the value {\tt 1} matches a preceding {\tt push(1)} operation. By
+values belong to the same matching. For example, a pop operation
+returning the value $1$ matches a preceding push$(1)$ operation. By
 comparing executions by the characteristics of their matchings, rather than the
@@
 naturally-occurring concurrent object ADTs are closed under the removal of
-matchings. For example, any execution of the atomic stack ADT using values {\tt
-1}, {\tt 2}, and {\tt 3} would remain a valid execution were all operations
-using the value {\tt 2} deleted. Conversely, any execution which extends a
+matchings. For example, any execution of the atomic stack ADT using values
+$1$, $2$, and $3$ would remain a valid execution were all operations
+using the value $2$ deleted. Conversely, any execution which extends a
 violating execution with additional matchings is itself a violation. This
@@
 
+Although our approach does require annotating the operations of concurrent
+object executions with a matching relation, and we demonstrate that these
+relations can be given for naturally-occurring ADTs, we also provide an
+automatic means for computing such \emph{matching schemes}. Again by sampling
+executions, we leverage automated symbolic reasoning engines to synthesize
+matching schemes for which given implementations are closed under the removal
+of matches. This further lowers the burden of automated verification. Rather
+than providing formal ADT specifications, or even matching schemes, users need
+only provide the predicates relevant in the logic of matching schemes, and we
+could automatically compute effective matching schemes, and ultimately,
+effective symbolic ADT representations.
+
 The contributions and outline of this work are as follows:
@@
 
+  \item An abstract notion of execution histories based on groups of matching
+  operations (§\ref{sec:histories}).
+
   \item The statement of the ADT inference problem (§\ref{sec:inference}).
 
-  \item Identification of algebraic properties allowing finite characterization
-  and computability of infinite ADT violations sets (§\ref{sec:algorithm}).
+  \item Identification of the algebraic properties allowing finite
+  characterization and computability of infinite ADT violations sets
+  (§\ref{sec:patterns}).
 
-  \item An algorithm to infer the matching functions required for our algebraic
+  \item The computability of symbolic ADT representations (§\ref{sec:formula}).
+
+  \item An algorithm to infer the matching schemes required for our algebraic
   characterization of ADTs (§\ref{sec:matching}).
 
-  \item Empirical validation that our ADT-inference approximation is complete
-  in practice, and effective for the detection of observational-refinement
-  violations~(§\ref{sec:empirical}).
+  \item An empirical study validating that naturally-occurring ADTs satisfy the
+  properties required for completeness of our inference algorithm, and that our
+  algorithm computes precise symbolic representations thereof
+  (§\ref{sec:nature}).
 
 \end{itemize}
-We conclude with a discussion of related work (§\ref{sec:related}).
+We conclude with a discussion of the limitations of our approach
+(§\ref{sec:discussion}) and related work (§\ref{sec:related}).
 
-As far as we are aware, this work is the first to suggest the automatic
+To the best of our knowledge, this work is the first to suggest the automatic
 generation of symbolic ADT representations for concurrent objects. By removing

### 2015-03-24 matching.tex
Quick typo fix.

@@
       \hline
-      0 & 0 & ${\rm w}(1)\ {\rm r}(1)\ {\rm r}(1)$ & not unique \\
-      0 & 1 & $\underline{{\rm w}(1)\ {\rm r}(1)}\ {\rm r}(1)$ & $\to_\mathrm{m}$ \\
-      1 & 0 & ${\rm w}(1)\ {\rm w}(2)\ {\rm r}(2)$ & not unique \\
-      1 & 1 & ${\rm w}(1)\ {\rm w}(2)$ & not unique \\
+      $0$ & $0$ & ${\rm w}(1)\ {\rm r}(1)\ {\rm r}(1)$ & not unique \\
+      $0$ & $1$ & $\underline{{\rm w}(1)\ {\rm r}(1)}\ {\rm r}(1)$ & $\to_\mathrm{m}$ \\
+      $1$ & $0$ & ${\rm w}(1)\ {\rm w}(2)\ {\rm r}(2)$ & not unique \\
+      $1$ & $1$ & ${\rm w}(1)\ {\rm w}(2)$ & not unique \\
     \end{tabular}

### 2015-03-24 intro.tex
Quick typo fixes.

@@
 
-The contributions and outline of this work are as follows:
+In summary, the contributions and outline of this work are:
 \begin{itemize}
@@
 \end{itemize}
-We conclude with a discussion of the limitations of our approach
+We conclude with a discussion about the limitations of our approach
 (§\ref{sec:discussion}) and related work (§\ref{sec:related}).

### 2015-03-24 matching.tex
Quick typo fix.

@@
       $1$ & $0$ & $0$ & ${\rm w}(1) \ {\rm w}(2) \ {\rm w}(3) \ {\rm r}(3)$ & not unique \\
-      $1$ & $0$ & $1$ & & \\
+      $1$ & $0$ & $1$ & — & — \\
       $1$ & $1$ & $0$ & ${\rm w}(1) \ {\rm w}(2) \ {\rm w}(3)$ & not unique \\
-      $1$ & $1$ & $1$ & & \\
+      $1$ & $1$ & $1$ & — & — \\
     \end{tabular}

### 2015-03-24 histories.tex
The last TODO.

@@
 $H(e,M,R) = H(e',M,R)$. A set of executions $E$ is \emph{data
-independent}\footnote{TODO NOTE THAT THIS IS NOT THE STANDARD DEFINITION…} when
-there exists a faithful matching scheme. By definition, our abstraction of
-executions into histories incurs no loss of precision for data-independent
-execution sets.
+independent}\footnote{Our definition of \emph{data independence} formalizes an
+existing informal notion, which stipulates that the implementation generating a
+set of executions does not predicate its actions on the data values passed as
+method arguments.} when there exists a faithful matching scheme. By definition,
+our abstraction of executions into histories incurs no loss of precision for
+data-independent execution sets.
 

### 2015-03-25 formula.tex
some final typos

--- 2015-03-22 formula.tex
@@
 
-TODO MOTIVATE THIS SECTION
+While Section~\ref{sec:patterns} demonstrates that (the kernel complements) of
+naturally-occurring ADTs can be represented finitely, in this section we
+demonstrate that such finite representations have logical interpretations,
+allowing us to derive the formulas representing ADTs. In what follows, we
+describe how to obtain a formula which is satisfied by the histories which
+embed a given history, via $\preceq$. Then, using the finite set of histories
+which represent the complement of a given ADT~\footnote{The complement of the kernel of an ADT $A$ is simply called the complement of $A$.}, we represent the ADT itself as
+the conjunction of negations of these embedding formulas. The resulting formula
+is satisfied only on the histories which do not embed the generators of a given
+ADT’s complement.
 
@@
 \begin{align*}
-  & \big(c(o) \implies {\sf c}(x)\big) \land {\sf f}(x) = f(o) \land {\sf r}(x) = r(o) \\
+  & \big(c(o) \implies {\sf c}(x)\big) \land {\sf f}(x) = f(o) \land {\sf r}(x) \Leftrightarrow r(o) \\
   & \land \big(o \in \dom(m)
@@
   \begin{align*}
-    \exists x_1.\ {\sf c}(x_1) \land {\sf f}(x_1) = \mathrm{read} \land {\sf um}(x_1).
+    \exists x_1.\ {\sf c}(x_1) \land {\sf f}(x_1) = \mathrm{read} \land {\sf um}(x_1) \land {\sf r}(x_1).
   \end{align*}
@@
   [1:1] write(1)            #
-  [2:2] read => empty (RO)    #    
+  [2:2] read => - (RO)    #    
 \end{verbatim}
@@
     & \quad \land {\sf c}(x_2) \land {\sf f}(x_2) = \mathrm{read}
-      \land \lnot{\sf um}(x_2) \land {\sf m}(x_2) = x_2 \\
+      \land \lnot{\sf um}(x_2) \land {\sf m}(x_2) = x_2\land {\sf r}(x_2) \\
     & \quad \land \big(\forall x.\ {\sf m}(x) = x_1 \implies {\sf r}(x) \lor \textsc{Identical}(x,x_1)\big) \\
@@
   The formula $\bigwedge_{h \in H} \neg \textsc{Embed}_h$ represents $A$,
-  for any set $H$ which generates the compliment of $A$.
+  for any set $H$ which generates the complement of $A$.
 

### 2015-03-25 histories.tex
some final typos

@@
 \end{itemize}
+The relation $<$ is an interval order because histories abstract executions where
+the set of call and return actions is totally ordered~\cite{conf/popl/BouajjaniEEH15}.
 Non-completed operations are \emph{pending}, and are maximal in the
@@
 \emph{match target}, and the inverse set $m^{-1}(o)$ of a target is a
-\emph{match}. We assume $o \in m^{-1}(o)$. A operation $o$ is \emph{unmatched}
+\emph{match}. We assume $o \in m^{-1}(o)$ for every $o$. A operation $o$ is \emph{unmatched}
 when $o$ is completed and $m(o)$ is undefined.
@@
     \item read $\Rightarrow v$ operations are read-only, and match themselves
-    when they return $v = \mathord\perp$, or a write$(v)$ operation, if one
+    when they return $v = -$, or a write$(v)$ operation, if one
     exists, and otherwise have no match,
@@
 
-  \item removing an order constraint (o),
+  \item unordering a pair of ordered operations (o),
 

### 2015-03-25 matching.tex
some final typos

@@
+%!TEX root = draft.tex
 \section{Computing Matching Schemes}
@@
 
-We automate the computation of matching schemes by considering the function
-${\sf g}$ underlying a simple matching scheme as an uninterpreted function in a
-logical satisfiability problem which uniquely determines the match of each
-operation, and constrain ${\sf g}$ to normalize the executions of an
+We automate the computation of matching schemes by constructing a logical
+formula that characterizes the boolean functions $G$ underlying a simple 
+matching scheme. These boolean functions are defined as the interpretation of 
+a function symbol ${\sf g}$. 
+%the function
+%symbol ${\sf g}$ underlying a simple matching scheme as an uninterpreted function in a
+%logical satisfiability problem which 
+The formula expresses the fact that the interpretations of ${\sf g}$ uniquely determine the match of each
+operation, and that they normalize the executions of an
 implementation. To this end, we fix an implementation $\mathcal{I}$ and a

### 2015-03-25 nature.tex
Updated intro.
some final typos

@@
+%!TEX root = draft.tex
 \section{Properties of Naturally-Occurring ADTs}
@@
   \item read $\Rightarrow v$ returns the last-stored value $v$, or the nil
-  value $\perp$ if no value has yet been stored.
+  value $-$ if no value has yet been stored.
 
@@
   \item read $\Rightarrow v$ operations are read-only, and match themselves if
-  $v = \perp$, or the unique write$(v)$ operation, if one exists, and otherwise 
+  $v = -$, or the unique write$(v)$ operation, if one exists, and otherwise 
   have no match.
@@
 
-It is easy to see that the histories of atomic registers normal. They are
-closed under removal of read-only and duplicate operations, since the each
+It is easy to see that the atomic register is normal. It is
+closed under removal of read-only and duplicate operations, since  each
 match can contain an arbitrary number of read operations. Since the histories
@@
 
-Our inference algorithm computes following four histories to generate the
+Our inference algorithm computes the following four histories to generate the
 complement of the atomic register ADT:
@@
 
-  \item remove $\Rightarrow v$ returns the nil value $v = \mathord\perp$ if the
+  \item remove $\Rightarrow v$ returns the nil value $v = -$ if the
   collection is empty, and otherwise removes and returns the least- or
@@
   \item remove $\Rightarrow v$ operations are read-only, and match themselves
-  if $v = \mathord\perp$ is the nil value. If $v \neq \mathord\perp$, remove
+  if $v = -$ is the nil value. If $v \neq -$, remove
   $\Rightarrow v$ operations are not read-only, and match the unique add$(v)$
@@
 
-It is not hard to see that the histories of atomic queues and stacks are
+It is not hard to see that the atomic queues and stacks are
 normal. They are closed under removal of read-only operations: only empty
-removes, i.e.,~remove $\Rightarrow \mathord\perp$, are read-only. They are
+removes, i.e.,~remove $\Rightarrow -$, are read-only. They are
 closed under the removal of duplicate operations: only non-empty removes can be
@@
 \end{itemize}
-Each operation returns the nil value $v = \mathord\perp$ when $u$ is not yet
+Each operation returns the nil value $v = -$ when $u$ is not yet
 present, and otherwise returns $v = u$. As its name implies, its kernel is
@@
 An execution \emph{inserts unique values} if the argument value to each insert
-operation which returns $\mathord\perp$ is unique. We give a faithful matching
+operation which returns $-$ is unique. We give a faithful matching
 scheme over executions which insert unique values as follows:
@@
 
-  \item any operation returning $v = \mathord\perp$ matches itself, and
+  \item any operation returning $v = -$ matches itself, and
 
-  \item any operation returning $v \neq \mathord\perp$ matches the unique
-  insert$(u) \Rightarrow \mathord\perp$ operation, if one exists, and otherwise
+  \item any operation returning $v \neq -$ matches the unique
+  insert$(v) \Rightarrow -$ operation, if one exists, and otherwise
   has no match.
@@
 \end{itemize}
-An operation returns the nil value $v = \mathord\perp$ when the lock is not
+An operation returns the nil value $v = -$ when the lock is not
 currently held, indicating success for lock, and failure for unlock. Otherwise,
 the operations return the value $u = v$ passed as an argument to the last
-successful lock$(v)$ operation.
+successful lock$(u)$ operation.
 
@@
 
-  \item any operation returning $v = \mathord\perp$ matches itself, and
+  \item any operation returning $v = -$ matches itself, and
 
-  \item any operation returning $v \neq \mathord\perp$ matches the unique
-  lock$(u) \Rightarrow \mathord\perp$ operation, if one exists, and otherwise
+  \item any operation returning $v \neq -$ matches the unique
+  lock$(v) \Rightarrow -$ operation, if one exists, and otherwise
   has no match.

### 2015-03-25 patterns.tex
some final typos

@@
+%!TEX root = draft.tex
 \section{Computing Finite ADT Representations}
@@
 labeled interval orders.
-
 Let $\Sigma$ be a finite alphabet. A labeled interval order is a triple $(A,\leq,\ell)$, where $(A,\leq)$ is an interval 
@@
 \begin{proof}
-By definition, for every interval order $(A,\leq,\ell)$, there exists a mapping $I$ from elements of $A$ to intervals over $\mathbb{N}$ such that for every $x,y\in A$, if $x\leq y$, then the interval $I(x)$ ends before the interval $I(y)$ (i.e., the upper bound of $I(x)$ is strictly smaller than the lower bound of $I(y)$).
+By definition, for every interval order $(A,\leq,\ell)$, there exists a mapping $I$ from elements of $A$ to intervals on $\mathbb{N}$ such that for every $x,y\in A$, if $x\leq y$, then the interval $I(x)$ ends before the interval $I(y)$ (i.e., the upper bound of $I(x)$ is strictly smaller than the lower bound of $I(y)$).
 %In general, there are more than one mappings $I_A$ that satisfy these constraints.
 In the following, we assume that the mapping $I_A$ is arbitrary but fixed.
-Therefore, in the following, we assume that an interval order is a multiset $\Gamma$ of triples of the form $[i,j,a]$ where $[i,j]$ is an interval on the integer line and $a$ is a symbol in $\Sigma$.
+Therefore, in the following, we assume that an interval order is a multiset $\Gamma$ of triples of the form $[i,j,a]$ where $[i,j]$ is an interval on $\mathbb{N}$ and $a$ is a symbol in $\Sigma$.
 
-We prove that another order on interval orders, denoted by $\subseteqq$, which is stronger than the embedding order $\subseteq$, is a wqo.
+We actually prove that another order on interval orders, denoted by $\subseteqq$, which is stronger than the embedding order $\subseteq$, is a wqo.
 The order $\subseteqq$ is defined by: $\Gamma_1\subseteqq \Gamma_2$ iff there exists an injective function $h:\Gamma_1\rightarrow \Gamma_2$ such that
 \begin{enumerate}
-	\item $h$ preserves the labeling, i.e., for any triple $[i,j,a]$, $h([i,j,a])=[i',j',a]$, for some $i'$, $j'$, 
+	\item $h$ preserves the labeling, i.e., for any triple $[i,j,a]$, $h([i,j,a])=[i',j',a]$, for some $i'$ and $j'$, 
 	\item $h$ preserves the ordering constraints, i.e., for every two triples $[i,j,a]$ and $[k,l,b]$ such that $j< k$, $h([i,j,a])=[i',j',a]$, $h([k,l,b])=[k',l',b]$, and $j'< k'$.
@@
 \end{enumerate}
+We say that $h$ \emph{witnesses} $\Gamma_1\subseteqq \Gamma_2$.
 
@@
 there exists no $i < j$ with $\Gamma_i\subseteqq \Gamma_j$.
-Also, assume that $\Gamma_1$ is the minimal size interval order that can start a bad sequence, $\Gamma_2$ is the minimal interval order that can continue a bad sequence starting with $\Gamma_1$, and so on.
+Also, assume that $\Gamma_1$ is the minimal size interval order that can start a bad sequence, $\Gamma_2$ is the minimal order that can continue a bad sequence starting with $\Gamma_1$, and so on.
 
 For each $\Gamma_k$, let $Min(\Gamma_k)$ be a triple $[i,j,a]$ such that 
-\begin{itemize}
-	\item $i$ is the minimal lower bound of an interval in $\Gamma_k$, and 
-	\item $j$ is the minimal upper bound of an interval in $\Gamma_k$ with lower bound $i$.
-\end{itemize}
+(1) $i$ is the minimal lower bound of an interval in $\Gamma_k$, and 
+(2) $j$ is the minimal upper bound of an interval in $\Gamma_k$ with lower bound $i$.
 %In general, this triple is not unique but it’s not important which one we choose.
-
-Also, let $Init(\Gamma_k) = (Min(\Gamma_k), P_k)$, where $P_k$ is the multiset of symbols labeling elements of $\Gamma_k$ that are incomparable to $Min(\Gamma_k)$.
+Also, let $Init(\Gamma_k) = (a_k, P_k)$, where $Min(\Gamma_k)=[i,j,a_k]$, for some $i$ and $j$, and $P_k$ is the multiset of symbols labeling elements of $\Gamma_k$ that are incomparable to $Min(\Gamma_k)$.
 Note that $P_k$ is bounded since we assume width-bounded interval orders.
 
-The infinite sequence $\Gamma_1$, $\Gamma_2$,$\ldots$ contains an infinite sequence $\Gamma_{k_1}$, $\Gamma_{k_2}$,$\ldots$ which, ignoring the integer interval, have the same value for $Init$.
-
+The infinite sequence $\Gamma_1$, $\Gamma_2$,$\ldots$ contains an infinite sequence $\Gamma_{k_1}$, $\Gamma_{k_2}$,$\ldots$ which have the same value for $Init$.
 For each $\Gamma_k$, let $\Lambda_k$ be the interval order obtained from $\Gamma_k$ by removing the triple $Min(\Gamma_k)$.
-
 By the minimality assumptions, the infinite sequence $\Gamma_1$, $\Gamma_2$,$\ldots$,$\Gamma_{k_1-1}$, $\Lambda_{k_1}$, $\Lambda_{k_2}$,$\ldots$ is not bad. 
@@
 Let $h$ be the injective function witnessing $\Lambda_{k_m}\subseteqq \Lambda_{k_n}$.
-We prove that the extension $h'$ of $h$ between $\Gamma_{k_m}$ and $\Gamma_{k_n}$, defined by $h'(Min(\Gamma_{k_m}))=Min(\Gamma_{k_n}$ and $h'(t)=h(t)$, for all $t\in \Lambda_{k_m}$ witnesses $\Gamma_{k_m}\subseteqq \Gamma_{k_n}$.
+We prove that the extension $h'$ of $h$ between $\Gamma_{k_m}$ and $\Gamma_{k_n}$, defined by $h'(Min(\Gamma_{k_m}))=Min(\Gamma_{k_n})$ and $h'(t)=h(t)$, for all $t\in \Lambda_{k_m}$, witnesses $\Gamma_{k_m}\subseteqq \Gamma_{k_n}$.
+
 Clearly, $h'$ preserves the labeling. 
-
 To prove that $h'$ preserves the ordering constraints, let $Min(\Gamma_{k_m})=[i,j,a]$ and $[k,l,b]\in \Gamma_{k_m}$ such that $j< k$.
@@
 Assume by contradiction that $k'\leq j'$.
-Since $h$ is injective, there exists an element $[x,y,c]\in \Gamma_{k_m}$ incomparable to $Min(\Gamma_{k_m})$ which is mapped to an element $[x',y',c]$ greater than $Min(\Gamma_{k_n})$ (by definition, $\Gamma_{k_m}$ and $\Gamma_{k_n}$ have the same number of elements incomparable to $Min(\Gamma_{k_m})$ and respectively, $Min(\Gamma_{k_n})$).
-Since $[x,y,c]$ is incomparable to $[i,j,a]$, we obtain that $x\leq j< k$.
-Also, by the current assumptions, $k'<j'<x'$.
-Therefore, there exist two elements $[x,y,c]$ and $[k,l,b]$ such that the interval $[x,y]$ starts before $[k,l]$ which are mapped by $h$ to the elements $[x',y',c]$ and $[k',l',b]$ such that the interval $[x',y']$ starts after $[k',l']$.
+Since $h$ is injective, there exists an element $[e,f,c]\in \Gamma_{k_m}$ incomparable to $Min(\Gamma_{k_m})$ which is mapped to an element $[e',f',c]$ greater than $Min(\Gamma_{k_n})$ (because, by definition, $\Gamma_{k_m}$ and $\Gamma_{k_n}$ have the same number of elements incomparable to $Min(\Gamma_{k_m})$ and respectively, $Min(\Gamma_{k_n})$).
+Since $[e,f,c]$ is incomparable to $[i,j,a]$, we obtain that $e\leq j< k$.
+Also, by the current assumptions, $k'\leq j'<e'$.
+Therefore, there exist two elements $[e,f,c]$ and $[k,l,b]$ such that the interval $[e,f]$ starts before $[k,l]$ which are mapped by $h$ to the elements $[e',f',c]$ and $[k',l',b]$ such that the interval $[e',f']$ starts after $[k',l']$.
 This contradicts the fact that $h$ witnesses $\Lambda_{k_m}\subseteqq \Lambda_{k_n}$.
 
-The properties of $Min(\Gamma_{k_m})$ and $Min(\Gamma_{k_n})$ imply that $h'$ also satisfies the third property in the definition of $\subseteqq$.
+The properties of $Min(\Gamma_{k_m})$ and $Min(\Gamma_{k_n})$ imply that $h'$ satisfies also the third property in the definition of $\subseteqq$.
 
@@
 \sigma(h)&=&(\set{\sigma(\mu): \mu\mbox{ a match}}, \set{\sigma(o): o\mbox{ read-only}}, \\
-&& \set{\sigma(o): o\mbox{ completed and unmatched}}, \\
-&& \mset{\sigma(o): o\mbox{ pending and unmatched}}).
+&& \set{\sigma(o): o\mbox{ unmatched}}, \\
+&& \mset{\sigma(o): o\mbox{ pending and non-matched}}).
 \end{align*}
@@
 (for matches, read-only and respectively, completed and unmatched operations) 
-and the multiset of signatures of the pending and unmatched operations.
+and the multiset of signatures of the pending and non-matched operations.
 
 %Assume by contradiction that $\preceq$ has an infinite antichain.
-Since the set of history signatures is bounded (because the set of methods is bounded)
-and the set of pending operations is also bounded (since we consider only bounded width histories),
+Since the set of history signatures is bounded (because the set of methods 
+and the set of pending operations are bounded),
 the sequence $h_1 h_2\ldots $ contains an infinite sequence of histories $h_1' h_2' \ldots$ that have 
@@
 \nu(h)&=&(\mset{\sigma(\mu): \mu\mbox{ a match of $h$}}, \mset{\sigma(o): o\mbox{ read-only}}, \\
-&& \mset{\sigma(o): o\mbox{ completed and unmatched}}).
+&& \mset{\sigma(o): o\mbox{ unmatched}}).
 \end{align*}
@@
 defined by $\alpha\leq\alpha'$ iff $\alpha(\sigma)\leq\alpha'(\sigma)$, for every $\sigma\in\Sigma$.
-Clearly, $\leq$ is a wqo and so is the component-wise extension of $\leq$ to history vectors.
+The relation $\leq$ is a wqo and so is the component-wise extension of $\leq$ to history vectors.
 Therefore, by known results, the sequence of histories $h_1' h_2' \ldots$ contains 
@@
 $\preceq$ is a wqo on $B$. Furthermore, when $A$ normal, it is closed under
-$\succeq$, and thus $B$ is closed under $\preceq$. Closure under a wqo implies
+$\succeq$, and thus $B$ is closed under $\preceq$~\footnote{Actually, $B$ is closed under $\preceq$ when restricted to width-bounded histories, i.e., if $h_1 \preceq h_2$, $h_1 \in B$, and $h_2$ is width-bounded, then $h_2\in B$.}.
+Closure under a wqo implies
 representation by a finite set. Formally, we say a set $X$ is \emph{finitely
@@
   [1:1] write(1)            #
-  [2:2] read => empty (RO)    #
+  [2:2] read => - (RO)    #
 \end{verbatim}

### 2015-03-25 histories.tex
example <= relation

@@
 actions where
+  \vspace{-1mm}
 \begin{itemize}
@@
   identifier.
-
+  \vspace{-1mm}
 \end{itemize}
@@
 Two operations $o_1, o_2 \in O$ are \emph{identical} when
+  \vspace{-1mm}
 \begin{itemize}
@@
   and $m(o_2)$ are undefined.
-
+  \vspace{-1mm}
 \end{itemize} 
@@
   in the history
+  \vspace{-1mm}
 \begin{verbatim}
@@
 \end{verbatim}
+  \vspace{-1mm}
   Operation~$1$ is a write operation which matches itself, and precedes a
@@
   execute concurrently
+    \vspace{-1mm}
 \begin{verbatim}
@@
 \end{verbatim}
+  \vspace{-1mm}
   Operation~$3$ is unmatched, indicated by the {\tt X}, and Operation~$4$ is
@@
   the matching scheme of Example~\ref{ex:matching} are $H(e_1)$:
+    \vspace{-1mm}
 \begin{verbatim}
@@
 \end{verbatim}
+  \vspace{-1mm}
   in which all three operations are sequential, $H(e_2)$:
+    \vspace{-1mm}
 \begin{verbatim}
@@
 \end{verbatim}
+  \vspace{-1mm}
   in which the first two operations overlap, and $H(e_3)$:
+    \vspace{-1mm}
 \begin{verbatim}
@@
 \end{verbatim}
+  \vspace{-1mm}
   in which the last two operations overlap.
@@
 \end{itemize}
+
+\begin{example}
+  The history
+    \vspace{-1mm}
+\begin{verbatim}
+    [1:1] write(a)        #
+    [2:2] write(b)          #
+    [3:2] read => b (RO)      #
+\end{verbatim}
+  \vspace{-1mm}
+  is related by $\to_\mathrm{o}$ to 
+    \vspace{-1mm}
+\begin{verbatim}
+    [1:1] write(a)        #
+    [2:2] write(b)        #
+    [3:2] read => b (RO)    #
+\end{verbatim}
+  \vspace{-1mm}
+which is related by $\to_\mathrm{c}$ to
+  \vspace{-1mm}
+\begin{verbatim}
+    [1:1] write(a)        #
+    [2:2] write(b)        #
+    [3:-] read* (RO)        #
+\end{verbatim}
+  \vspace{-1mm}
+which is related by $\to_\mathrm{p}$ to
+  \vspace{-1mm}
+\begin{verbatim}
+    [1:1] write(a)        #
+    [2:2] write(b)        #
+    [3:-] read* (RO)        #
+    [4:4] write(c)*         #
+\end{verbatim}
+  \vspace{-1mm}
+\end{example}
+
 A set of histories $H$ is \emph{closed} under a relation $\to$ when $h_2 \in H$
@@
 
+  \vspace{-1mm}
 \begin{lemma}
@@
   $\to_\mathrm{c}$.
-
+  \vspace{-1mm}
 \end{lemma}

### 2015-03-25 inference.tex
some final typos
example <= relation

@@
+%!TEX root = draft.tex
 \section{The Symbolic ADT Inference Problem}
@@
 set $H$ of histories generating $A$, i.e.,~such that
+  \vspace{-1mm}
 \begin{align*}
   A = \set{ h' : \exists h \in H.\ h \to^\ast h' }
+    \vspace{-1mm}
 \end{align*}
@@
 
+  \vspace{-1mm}
 \begin{definition}
@@
   refines $H(\mathcal{I}_2)$.
-
+  \vspace{-1mm}
 \end{definition}
@@
 \emph{history formula} is a first-order logic formula with
+  \vspace{-1mm}
 \begin{itemize}
@@
 
-  \item predicate symbols ${\sf c}$, ${\sf um}$, and ${\sf <}$ for completion,
-  non-matching, and order.
-
+  \item predicate symbols ${\sf c}$, ${\sf um}$, {\sf r}, and ${\sf <}$ for completion,
+  non-matching (operations which are not in the domain of the matching function), read-only, and order.
+  \vspace{-1mm}
 \end{itemize}
@@
   operation happens between a pair of matching write and read operations:
+  \vspace{-1mm}
   \begin{align*}
@@
       \implies x_3 < x_2
+        \vspace{-1mm}
   \end{align*}
+
   This is one of several requirements of atomic single-value register ADTs,

### 2015-03-25 patterns.tex
example <= relation

@@
 For every history $h$ and match signature $\sigma$, the \emph{$\sigma$-vector} of a history $h$ is the multiset
-\[
+$
 \nu_\sigma(h)=\mset{\nu(\mu): \sigma(\mu)=\sigma}
-\]
+$
 of match vectors of signature $\sigma$.
@@
 Let $\leq_v$ be an order relation on $\sigma$-vectors defined by $\nu_\sigma(h)\leq_v \nu_\sigma(h')$ iff
-\begin{itemize}
-	\item for every match vector $\nu$ in $\nu_\sigma(h)$ there exists a match vector $\nu'$ in $\nu_\sigma(h')$
-	such that $\nu\leq \nu'$ (here, $\leq$ is the order on multisets defined above).
-\end{itemize}
+for every match vector $\nu$ in $\nu_\sigma(h)$ there is a match vector $\nu'$ in $\nu_\sigma(h')$
+	s.t. $\nu\leq \nu'$ (here, $\leq$ is the order on multisets defined above).
 By known results, $\leq_v$ is a wqo.
@@
 \mathbb{N}$, we define
+  \vspace{-1mm}
 \begin{align*}
@@
   }
+    \vspace{-1mm}
 \end{align*}
@@
   single-value register ADT, witnessing either an unmatched read operation:
+  \vspace{-1mm}
 \begin{verbatim}
@@
 \end{verbatim}
+  \vspace{-1mm}
   a read of an uninitialized register occurring after some write:
@@
 \end{verbatim}
+    \vspace{-1mm}
   a read which happens before its matching write operation:
+      \vspace{-1mm}
 \begin{verbatim}
@@
 \end{verbatim}
+    \vspace{-1mm}
   and a read matching a write which is not the most recent:
+      \vspace{-1mm}
 \begin{verbatim}
@@
 \end{verbatim}
+    \vspace{-1mm}
   Every sequential history not admitted by the atomic register ADT embeds at

### 2015-03-25 patterns.tex
Fixed space hacks.

@@
 \mathbb{N}$, we define
-  \vspace{-1mm}
 \begin{align*}
@@
   }
-    \vspace{-1mm}
 \end{align*}
@@
   single-value register ADT, witnessing either an unmatched read operation:
-  \vspace{-1mm}
 \begin{verbatim}
@@
 \end{verbatim}
-  \vspace{-1mm}
   a read of an uninitialized register occurring after some write:
@@
 \end{verbatim}
-    \vspace{-1mm}
   a read which happens before its matching write operation:
-      \vspace{-1mm}
 \begin{verbatim}
@@
 \end{verbatim}
-    \vspace{-1mm}
   and a read matching a write which is not the most recent:
-      \vspace{-1mm}
 \begin{verbatim}
@@
 \end{verbatim}
-    \vspace{-1mm}
   Every sequential history not admitted by the atomic register ADT embeds at

### 2015-06-26 motivation.tex
Adding motivation section.

\section{Motivation}
\label{sec:motivation}

TODO DISCUSS SCALABLE VERIFICATION, MODULAR REASONING, INTERFACE SPECIFICATIONS,
DERIVING SPECIFICATIONS, CONCURRENCY, CONCURRENT ADTS, SYMBOLIC ADT SPECIFICATIONS

TODO CLIENT-SIDE REASONING

\begin{figure}
  \begin{minipage}{0.49\linewidth}
    \begin{verbatim}
assume w < x
( (q.add(<1,w>); q.add(<1,x>))
||(q.add(<2,w>); q.add(<2,x>)))
i, y := q.remove()
j, z := q.remove()
assert i == j ==> y < z
    \end{verbatim}
  \end{minipage}
  \hfill
  \begin{minipage}{0.49\linewidth}
    \begin{verbatim}
( (while * do q.add(<1,w++>))
||(while * do q.add(<2,x++>)))
while * do q.remove() end
i, y := q.remove()
j, z := q.remove()
assert i == j ==> y < z
    \end{verbatim}
  \end{minipage}
  \caption{A atomic queue clients with six operations (left), and with
  arbitrarily-many operations (right).}
  \label{fig:clients}
\end{figure}

\begin{align*}
  \forall w,x,y,z.\left(
    \begin{array}{l}
      {\sf add}(w) \land {\sf remove}(x) \land {\sf val}(w) = {\sf val}(x) \\
      \land\ {\sf add}(y) \land {\sf remove}(z) \land {\sf val}(y) = {\sf val}(z) \\
      \land\ {\sf before}(w,y) \implies {\sf before}(x,z)
    \end{array}
  \right)
\end{align*}

TODO IMPLEMENTATION-SIDE REASONING: EMP(6 < 4) AND TR(4 < 6)

\begin{verbatim}
    [1] add(1)        #
    [2] remove => 3     #####
    [3] remove => 1     #
    [4] add(2)            #
    [5] add(3)              #
    [6] remove => EMPTY       #
\end{verbatim}

\begin{align*}
  \forall x,y.\left(
    \begin{array}{l}
      {\sf remove}(x) \land {\sf val}(x) = {\sf empty} \\
      \land\ {\sf add}(y) \land {\sf unmatched}(y) \implies {\sf before}(x,y)
    \end{array}
  \right)
\end{align*}

TODO AUTOMATIC INFERENCE

### 2015-06-30 motivation.tex
Nearly complete draft of motivation section.

@@
 
-TODO DISCUSS SCALABLE VERIFICATION, MODULAR REASONING, INTERFACE SPECIFICATIONS,
-DERIVING SPECIFICATIONS, CONCURRENCY, CONCURRENT ADTS, SYMBOLIC ADT SPECIFICATIONS
+Effective scalable reasoning about nontrivial software implementations generally
+requires considering each software module separately, in isolation, using
+abstract specifications for other modules. When modules are objects whose
+methods may be called concurrently, their behavior is typically understood in
+terms of valid invocation sequences of atomic abstract data types (ADTs): any
+execution of overlapping method invocations can be \emph{linearized} into an
+invocation sequence admitted by the ADT~\cite{journals/toplas/HerlihyW90}. ADT
+specifications thus decouple reasoning about object implementations — is each
+concurrent execution linearizable? — from their clients’ invocations — does
+every valid linearization preserve client invariants?
 
-TODO CLIENT-SIDE REASONING
+Consider the two parallel programs in Figure~\ref{fig:clients} invoking the add
+and remove methods of an atomic queue implementation. The clients add increasing
+integer values $w$ and $x$ tagged with the integers $\set{1,2}$ indicating on
+which parallel branch each add occurs. Intuitively both programs are correct
+since values with the same tag are added in increasing order, and, crucially,
+the values of the queue ADT are removed in the same order in which they are
+added. Thus considering the sequences of values added by all six possible
+linearizations of the add operations for the first program:
+\begin{align*}
+  & \texttt{<1,w> <1,x> <2,w> <2,x>}, \\
+  & \texttt{<1,w> <2,w> <1,x> <2,x>}, \\
+  & \texttt{<1,w> <2,w> <2,x> <1,x>}, \\
+  & \texttt{<2,w> <1,w> <1,x> <2,x>}, \\
+  & \texttt{<2,w> <1,w> <2,x> <1,x>}, \\
+  & \texttt{<2,w> <2,x> <1,w> <1,x>}
+\end{align*}
+we see that comparison {\tt i == j} of tags for the first two removes only holds
+for the first and last linearizations, which remove $w$ then $x$ according to
+the queue ADT specification, implying that {\tt y < z} holds when {\tt i == j}
+holds.
 
@@
 
+ADTs are often defined in an imperative style using simple mathematical objects.
+For example, the atomic queue ADT can be specified via an abstract varaible $q$
+initialized to the empty sequence $\varepsilon$; add operations update $q$ to
+$q' = q \cdot x$ by appending the argument value $x$; when $q = x \cdot q'$,
+remove operations update $q$ to $q'$ and return $x$; otherwise, when $q =
+\varepsilon$, remove operations return a special value {\tt EMPTY}. Reasoning
+using such specifications requires the explicit consideration of possible
+linearizations one by one in order to determine valid return values, e.g.,~for
+the remove operations of queues.
+
+In this work we advocate \emph{symbolic} ADT specifications which facilitate the
+use of efficient reasoning algorithms, e.g.,~via symbolic reasoning engines. For
+simplicity we consider simple first-order languages whose variables represent
+\emph{operations}, i.e.,~method invocations, with predicates indicating the
+method name and argument/return values of each operation, along with the
+linearization order of operations. For example, the following formula partially
+specifies the atomic queue ADT: the removes matching a pair of add operations
+occur in the same order as their corresponding adds:
 \begin{align*}
@@
     \end{array}
-  \right)
+  \right).
 \end{align*}
+Such symbolic representations are more convenient than the aforementioned
+imperative style of ADT specification since they allow symbolic reasoning
+engines to prune redundant reasoning steps. For example, rather than enumerating
+the six linearizations above of the first program of Figure~\ref{fig:clients} to
+determine that the valid return values of remove operations are $w$ and $x$, a
+symbolic reasoning engine could eagerly rule out all but the first and last
+linearizations for which {\tt i == j} by looking at only the first two
+operations. Since the number of linearizations is generally exponential in the
+number of operations, such symbolic reasoning is of significant value.
 
-TODO IMPLEMENTATION-SIDE REASONING: EMP(6 < 4) AND TR(4 < 6)
+TODO THE SECOND CLIENT?
 
+Symbolic ADT specifications also facilitate reasoning about the correctness of
+object \emph{implementations} by avoiding the redundant reasoning inherent in
+the explicit enumeration of linearizations~\cite{conf/pldi/EmmiEH15}. For
+example, consider the execution log of an erroneous queue implementation
 \begin{verbatim}
@@
 \end{verbatim}
-
+where each operation is prefixed by a unique operation identifier; the “\#”
+symbol denotes the time span of each operation, from left to right, indicating
+that Operation~$2$ overlaps in time with Operations~$3$, $4$, and~$5$, and that
+all other operations are ordered in time. While this execution can be identified
+as a violation of the atomic queue ADT by considering every possible
+linearization of Operation~$2$, this approach is intractible in general, as
+the number of linearizations is exponential in the number of operations.
+However, consider the following formula partially specifying the atomic
+queue ADT: that the adds of values which have not yet been removed must be
+ordered after empty remove operations:
 \begin{align*}
@@
     \end{array}
-  \right)
+  \right).
 \end{align*}
+Combined with the transitivity of the {\sf before} relation, a symbolic
+reasoning engine could eagerly conclude that Operation~$4$ must be linearized
+both before Operation~$6$, due to transitivity, and after, due to the previous
+axiom. Such inconsistencies can be detected with great efficiency in
+practice~\cite{conf/pldi/EmmiEH15}.
 
-TODO AUTOMATIC INFERENCE
+While (symbolic) ADT specifications are crucial to scalable reasoning about
+nontrivial software implementations, writing effective speciications is often a
+challenge for programmers. Towards fuller automation of scalable reasoning
+techniques, we focus in this work on the automatic inference of symbolic
+specifications which can be used directly in both client-side and
+implementation-side reasoning of concurrent objects.

### 2015-07-06 algorithm.tex
Fixed formula precedence problem.
starting top-down presentation

--- 2015-03-13 algorithm.tex
@@
-\section{Computing Symbolic Representations}
-\label{sec:algorithm}
+%!TEX root = draft.tex
+\section{Solving the Symbolic ADT Inference Problem}
 
-TODO MOTIVATE THIS SECTION
+\begin{figure}
+\begin{lstlisting}
+Min := $\emptyset$
+while (true) {
+  $B_i$ := $ \set{ h \in B : \weight(h) = i }$;
+  MinTemp := Min $\cup$ Min($B_i$)
+  if ( MinTemp = Min )
+    break;
+  Min := MinTemp
+}
+return $\neg$Formula(Min);
+\end{lstlisting}
+\caption{The algorithm solving the symbolic ADT inference problem for a given ADT $A$.
+$B$ denotes the bounded complement of $\ker A$ and $\uparrow$\texttt{Min} denotes the upward closure w.r.t.
+$\preceq$ of a given set of histories.
+The procedure $\texttt{Min}(X)$ returns the $\preceq$-minimal elements of the set of histories $X$ and $\texttt{Formula}(X)$ returns a history formula which holds exactly for the histories in the upward closure of $X$.}
+\label{fig:algorithm}
+\end{figure}
 
-In addition to the relations $\to_\mathrm{o}$, $\to_\mathrm{p}$, and
-$\to_\mathrm{c}$ of Section~\ref{sec:histories} under which all ADTs are
-closed, the ADTs we consider in this work are also closed under additional
-relations which remove matches, and duplicate operations. The relations
-$\to_\mathrm{m}$ and $\to_\mathrm{d}$ relate two histories $h_1$ and $h_2$ when
-$h_2$ is obtained from $h_1$ by
-\begin{itemize}
+Figure~\ref{fig:algorithm} lists an algorithm for solving the symbolic ADT inference problem which is parametrized by 
+an order relation $\preceq$ between histories such that the complement of the ADT 
+is closed under $\preceq$. For a given set of elements $X$ and an order relation $\preceq$, the \emph{upward closure}
+of $X$ w.r.t. $\preceq$ is the set $\uparrow X=\set{y : \exists x\in X.\ x\preceq y}$. The algorithm performs an
+weight-increasing enumeration of $B$ while maintaining the
+$\preceq$-minimal elements, and stops whenever each history of weight $i$ is bigger (w.r.t. $\preceq$) than some
+element of the collected set of minimals.
 
-  \item removing a match (m), or
+The partial correctness of the algorithm (i.e., ignoring termination) relies on the input ADT being \emph{predictable}, i.e.,
+for all $i$,
+\begin{align*}
+B_{i+1} \succeq \texttt{Min}(B_{\leq i}) \mbox{ implies } \texttt{Min}(B) = \texttt{Min}(B_{\leq i}),
+\end{align*}
+where $B_{\leq i}=\bigcup_{1\leq j\leq i}B_j$.
 
-  \item removing a duplicate operation (d).
+\begin{theorem}\label{th:corr1}
+For any predictable ADT $A$, if the algorithm in Figure~\ref{fig:algorithm} terminates, then the returned formula represents $A$.
+\end{theorem}
+\begin{proof}
+%For a set $\Sigma$, let $\neg \Sigma$ denote its complement.
+Let $M$ be the value of \texttt{Min} at the end of the \texttt{while} loop.
+Since the complement of $A$ is closed under $\preceq$, and $M\subseteq B\subseteq  \neg \ker A \subseteq \neg A$, we have that $\uparrow M\subseteq \neg A$. 
+Since $A$ is predictable, we have that $B \subseteq\, \uparrow M$.
+The formula $\varphi$ returned by the algorithm holds exactly for the set of histories $X$ 
+which is the complement of $\uparrow M$. By the previous inclusions, we have that $A\subseteq X\subseteq \neg B$.
+Therefore, $\varphi$ holds for all histories in $\ker A$ and doesn't hold for histories in $B$. 
+\end{proof}
 
-\end{itemize}
-We say an ADT closed under $\to_\mathrm{m}$ and $\to_\mathrm{d}$ is
-\emph{normal}. In Section~\ref{sec:nature} we demonstrate that
-naturally-occurring ADTs are normal. Defining the relation $\succeq$ as the
-reflexive and transitive closure of the above relations,
-\begin{align*}
-  \succeq = (\to_\mathrm{o} \cup \to_\mathrm{p} \cup \to_\mathrm{c} \cup 
-  \to_\mathrm{m} \cup \to_\mathrm{d})^\ast
-\end{align*}
-closure under $\succeq$ follows immediately.
+\begin{remark}
+Note that for non-predictable ADTs, the value of $\texttt{Min}$ at the end of the \texttt{while} loop 
+is an under-approximation of $\texttt{Min}(B)$. 
+The resulting symbolic ADT representation is still satisfied by all histories in $\ker A$, therefore it would still be sound in only identifying
+\emph{actual} violations, yet incomplete in identifying \emph{all} violations.
+\end{remark}
 
-\begin{lemma}
+The following result states the conditions under which the algorithm in Figure~\ref{fig:algorithm} terminates.
 
-  Normal ADTs are closed under $\succeq$.
+\begin{theorem}\label{th:corr2}
+If the set of $\preceq$-minimal elements of any (potentially-infinite) set of histories is finite, then the algorithm in Figure~\ref{fig:algorithm} terminates.
+\end{theorem}
 
-\end{lemma}
-
-TODO MOTIVATE WQO
-
-A \emph{well-quasi-ordering (wqo)} $R$ on a set $X$ is a reflexive, transitive
-binary relation on $X$ for which in every infinite sequence $x_0 x_1 \ldots$ of
-elements from $X$, there exists indices $i < j$ such that $R(x_i,x_j)$.
-
-\begin{example}
-
-  Consider the infinite history sequence $h_1 h_2 \ldots$ where each $h_i$
-  contains $2i$ operations $o_1, o_1', \ldots, o_i, o_i'$ where each $o_j$ is a
-  completed {\tt push} operation matching itself, and each $o_j'$ is a pending
-  {\tt pop} operation with undefined matching. Because each successive $h_i$
-  has both more matches and more pending operations, no two histories of the
-  sequence are related by $\preceq$. Thus $\preceq$ is not a wqo.
-
-\end{example}
-
-The previous example demonstrates that $\preceq$ is not a wqo by allowing each
-history $h_i$ of the infinite sequence to contain more and more pending
-operations in order to ensure that $h_j \not\preceq h_i$ for every $j < i$.
-Curbing this ability by limiting the maximum amount of pending operations per
-history makes $\preceq$ a wqo. Although limiting to $k$ pending operations
-essentially limits us to width-$k$ histories, of executions with at most $k$
-operations parallel at any moment, e.g.,~of programs with at most $k$ threads,
-this restriction comes at no loss of completeness when considering only the
-histories of bounded-width ADT kernels, which is the subject of the remainder
-of this section.
-
-\begin{lemma}
-
-  $\preceq$ is a wqo on width-bounded histories.
-
-\end{lemma}
-
-For the remainder of this section, we fix an ADT $A$, and let $B$ be the
-complement of $\ker A$. When $\ker A$ has bounded width, so does $B$, and thus
-$\preceq$ is a wqo on $B$. Furthermore, when $A$ normal, it is closed under
-$\succeq$, and thus $B$ is closed under $\preceq$. Closure under a wqo implies
-representation by a finite set. Formally, we say a set $X$ is \emph{finitely
-representable} if there exists a finite set $Y$ and a relation $R \subseteq Y
-\times X$ such that $X = \set{ x : \exists y\in Y.\ R(y,x) }$. In our case, we
-obtain a finite set from which exactly the elements of $B$ are related by
-$\preceq$.
-
-\begin{lemma}
-
-  $B$ is finitely representable if $A$ is normal.
-
-\end{lemma}
-
-Computing $B$ thus amounts to computing its finite representation. In general,
-this is achieved by enumerating the elements of $B$ while maintaining the
-$\preceq$-minimal elements, and recognizing a condition under which all
-elements of $B$ are related to the current set of
-minimals~\cite{conf/lics/AbdullaCJT96, journals/tcs/FinkelS01}. In our case, we
-recognize a condition on the weight-increasing enumeration of $B$ which
-guarantees coverage by the current set of minimals. Formally, for all $i \in
-\mathbb{N}$, we define
-\begin{align*}
-  B_i &= \set{ h \in B : \weight(h) \le i } \\
-  B_i' &= \set{ h' \in B :
-    \exists h \in B_i.\ h \preceq h' \text{ and } \weight(h') \le i+1
-  }
-\end{align*}
-as, respectively, the histories of $B$ with at most $i$ matches and duplicates,
-and those derived from $B_i$ with at most $i\!+\!1$ matches and duplicates. We
-say that $A$ is \emph{predictable} if $B_i = B$ whenever $B_i' = B_{i+1}$. In
-Section~\ref{sec:nature} we demonstrate that naturally-occurring ADTs are
-predictable. Since the number of weight-$i$ histories is finite\footnote{As
-noted in Section~\ref{sec:histories}, we consider equality between histories up
-to renaming of operation identifiers.} for each $i \in \mathbb{N}$, each $B_i$
-is computable.
-
-\begin{lemma}
-
-  $B_i$ is computable, for all $i \in \mathbb{N}$, if $A$ is normal.
-
-\end{lemma}
-
-Predictability ensures that once $B_{i+1}$ does not alter the set of computed
-$\preceq$-minimals, with respect to those of $B_i$, then the minimals computed
-for $B_i$ are exactly those of $B$.
-
-\begin{lemma}
-
-  $B$ is computable if $A$ is normal and predictable.
-
-\end{lemma}
-
-Note that for non-predictable ADTs, computing $B_i$, for any $i \in
-\mathbb{N}$, under-approximates $B$. In practice, computing $B_i$ can be used
-to identify all violations which surface in histories with $i$ matches,
-overlooking violations which only surface with more than $i$ matches. The
-resulting symbolic ADT representation would still be sound in only identifying
-\emph{actual} violations, yet incomplete in identifying \emph{all} violations.
-
-TODO DESCRIBE HOW TO GET THE FORMULA $F$ FROM $B$, AND ARGUE THAT WE HAVE
-SOLVED THE PROBLEM.
-
-\begin{theorem}
-
-  “The symbolic ADT inference problem is computable for normal ADTs.”
-
-\end{theorem}
+The following section provides a concrete instantiation for the order relation $\preceq$ satisfying the requirements in 
+Theorem~\ref{th:corr1} and~\ref{th:corr2} except for predictability. 
+Although we don't provide a general result concerning predictability, we prove that it holds for many naturally-occurring ADTs (see Section~\ref{sec:nature}).

### 2015-07-06 algorithm.tex
starting top-down presentation

@@
 \end{lstlisting}
-\caption{The algorithm solving the symbolic ADT inference problem for a given ADT $A$.
-$B$ denotes the bounded complement of $\ker A$ and $\uparrow$\texttt{Min} denotes the upward closure w.r.t.
-$\preceq$ of a given set of histories.
+\caption{The algorithm solving the symbolic ADT inference problem for a given ADT $A$, where $B$ denotes the bounded complement of $\ker A$.
 The procedure $\texttt{Min}(X)$ returns the $\preceq$-minimal elements of the set of histories $X$ and $\texttt{Formula}(X)$ returns a history formula which holds exactly for the histories in the upward closure of $X$.}

### 2015-07-06 motivation.tex
Reworked the motivation.

--- 2015-06-26 motivation.tex
@@
 
-TODO DISCUSS SCALABLE VERIFICATION, MODULAR REASONING, INTERFACE SPECIFICATIONS,
-DERIVING SPECIFICATIONS, CONCURRENCY, CONCURRENT ADTS, SYMBOLIC ADT SPECIFICATIONS
-
-TODO CLIENT-SIDE REASONING
+Effective scalable reasoning about nontrivial software implementations generally
+requires considering each software module separately, in isolation, using
+abstract specifications for other modules. When modules are objects whose
+methods may be called concurrently, their behavior is typically understood in
+terms of valid invocation sequences of atomic abstract data types (ADTs): any
+execution of overlapping method invocations can be \emph{linearized} into an
+invocation sequence admitted by the ADT~\cite{journals/toplas/HerlihyW90}. For
+example, consider the execution history depicted in Figure~\ref{fig:clients} in
+which the add operations numbered $2$ and $3$ overlap with each other, and,
+respectively, with operations $1$ and $4$. Among the five possible ways to
+linearize these six operations, the linearization~$1$; $3$; $2$; $4$; $5$; $6$
+does correspond to a valid sequence of the atomic queue ADT, and so this history
+is linearizable. ADT specifications thus decouple reasoning about object
+implementations — is each concurrent execution linearizable? — from their
+clients’ invocations — do all valid linearizations preserve client invariants?
 
 \begin{figure}
-  \begin{minipage}{0.49\linewidth}
+  \begin{minipage}{0.43\linewidth}
+    \begin{verbatim}
+1: add(a)       #
+2: add(b)       ###
+3: add(c)         ###
+4: add(d)           #
+5: remove => a        #
+6: remove => c          #
+    \end{verbatim}
+  \end{minipage}
+  \hfill
+  \begin{minipage}{0.55\linewidth}
     \begin{verbatim}
@@
 (  (q.add(<1,w>); q.add(<1,x>))
-||(q.add(<2,w>); q.add(<2,x>)))
+|| (q.add(<2,w>); q.add(<2,x>)) )
 i, y := q.remove()
@@
   \end{minipage}
-  \hfill
-  \begin{minipage}{0.49\linewidth}
-    \begin{verbatim}
-( (while * do q.add(<1,w++>))
-||(while * do q.add(<2,x++>)))
-while * do q.remove() end
-i, y := q.remove()
-j, z := q.remove()
-assert i == j ==> y < z
-    \end{verbatim}
-  \end{minipage}
-  \caption{A atomic queue clients with six operations (left), and with
-  arbitrarily-many operations (right).}
+  \caption{An execution history with six numbered operations (left),
+    and a parallel program invoking six operations (right). The hash symbols
+    “\#” depict the time interval spanned by each operation horizontally.}
   \label{fig:clients}
@@
 
-\begin{align*}
-  \forall w,x,y,z.\left(
-    \begin{array}{l}
-      {\sf add}(w) \land {\sf remove}(x) \land {\sf val}(w) = {\sf val}(x) \\
-      \land\ {\sf add}(y) \land {\sf remove}(z) \land {\sf val}(y) = {\sf val}(z) \\
-      \land\ {\sf before}(w,y) \implies {\sf before}(x,z)
-    \end{array}
-  \right)
-\end{align*}
+Consider the parallel program in Figure~\ref{fig:clients} invoking the add and
+remove methods of an atomic queue implementation, adding increasing integer
+values $w$ and $x$ tagged with the integers $\set{1,2}$ indicating on which
+parallel branch each add occurs. Intuitively this program is correct since
+values with the same tag are added in increasing order, and, crucially, the
+values of the queue ADT are removed in the same order in which they are added.
+Among the six possible ways to linearize these operations, the comparison {\tt i ==
+j} of tags only holds for those two which begin with {\tt q.add(<1,$w$>);
+q.add(<1,$x$>)} and {\tt q.add(<2,$w$>); q.add(<2,$x$>)}. Since the queue ADT
+dictates that elements are removed in the order added, we conclude that $w$ and
+$x$ are removed in order when {\tt i == j}, and thus {\tt y < z} holds when {\tt
+i == j} holds.
 
-TODO IMPLEMENTATION-SIDE REASONING: EMP(6 < 4) AND TR(4 < 6)
+Although formal ADT specifications are indispensable for scalable program
+reasoning, formal-specification writing is a burden for which few programmers
+posses the required combination of expertise and willingness to overcome.
+According to current practices, programmers write obviously-correct ADT
+\emph{reference implementations}, e.g.,~an atomic queue whose methods are
+synchronized via a global lock, and refine them with more efficient fine-grained
+implementations, e.g.,~reducing synchronization overhead with specialized
+hardware instructions such as atomic compare and swap (CAS).
 
-\begin{verbatim}
-    [1] add(1)        #
-    [2] remove => 3     #####
-    [3] remove => 1     #
-    [4] add(2)            #
-    [5] add(3)              #
-    [6] remove => EMPTY       #
-\end{verbatim}
+Our goal in this work is thus the automated generation of formal ADT
+specifications, derived from reference implementations, which are suitable for
+automated reasoning. In particular, we aim to generate \emph{symbolic
+representations} of the history linearizations which are valid, according to
+ADTs given implicitly by reference implementations. We target declarative
+symbolic representations rather than imperative state-based representations in
+order to exploit efficient symbolic reasoning engines such as SMT solvers.
+Previous work demonstrates that symbolic representations can offer orders of
+magnitude improvements over imperative state-based representations by averting
+the need to enumerate linearizations explicitly in order to check their
+validity~\cite{conf/pldi/EmmiEH15}.
 
-\begin{align*}
-  \forall x,y.\left(
-    \begin{array}{l}
-      {\sf remove}(x) \land {\sf val}(x) = {\sf empty} \\
-      \land\ {\sf add}(y) \land {\sf unmatched}(y) \implies {\sf before}(x,y)
-    \end{array}
-  \right)
-\end{align*}
-
-TODO AUTOMATIC INFERENCE
+TODO ALGORITHM OVERVIEW

### 2015-07-06 motivation.tex
Abstract algorithm.

@@
-\section{Motivation}
+\section{Motivation \& Overview}
 \label{sec:motivation}
@@
   \caption{An execution history with six numbered operations (left),
-    and a parallel program invoking six operations (right). The hash symbols
-    “\#” depict the time interval spanned by each operation horizontally.}
+    and a parallel program invoking six operations (right). The “\#” symbols
+    depict the time intervals spanned by operations horizontally.}
   \label{fig:clients}
@@
 Among the six possible ways to linearize these operations, the comparison {\tt i ==
-j} of tags only holds for those two which begin with {\tt q.add(<1,$w$>);
-q.add(<1,$x$>)} and {\tt q.add(<2,$w$>); q.add(<2,$x$>)}. Since the queue ADT
-dictates that elements are removed in the order added, we conclude that $w$ and
-$x$ are removed in order when {\tt i == j}, and thus {\tt y < z} holds when {\tt
-i == j} holds.
+j} of tags only holds for those two beginning with
+\begin{quote}
+  \verb|q.add(<1,w>); q.add(<1,x>) | and \verb| q.add(<2,w>); q.add(<2,x>)|
+\end{quote}
+Since the queue ADT dictates that elements are removed in the order added, we
+conclude that $w$ and $x$ are removed in order when {\tt i == j}, and thus {\tt
+y < z} holds when {\tt i == j} holds.
 
@@
 posses the required combination of expertise and willingness to overcome.
-According to current practices, programmers write obviously-correct ADT
-\emph{reference implementations}, e.g.,~an atomic queue whose methods are
-synchronized via a global lock, and refine them with more efficient fine-grained
-implementations, e.g.,~reducing synchronization overhead with specialized
-hardware instructions such as atomic compare and swap (CAS).
+Typically programmers write simple ADT \emph{reference implementations},
+e.g.,~whose methods are synchronized via a global lock, and refine them with
+more efficient fine-grained implementations, e.g.,~reducing synchronization
+bottlenecks using specialized hardware instructions such as atomic compare and
+swap (CAS).
 
@@
 automated reasoning. In particular, we aim to generate \emph{symbolic
-representations} of the history linearizations which are valid, according to
-ADTs given implicitly by reference implementations. We target declarative
-symbolic representations rather than imperative state-based representations in
-order to exploit efficient symbolic reasoning engines such as SMT solvers.
-Previous work demonstrates that symbolic representations can offer orders of
-magnitude improvements over imperative state-based representations by averting
-the need to enumerate linearizations explicitly in order to check their
-validity~\cite{conf/pldi/EmmiEH15}.
+representations} of the valid history linearizations, according to ADTs given
+implicitly by reference implementations. We target declarative symbolic
+representations rather than imperative state-based representations in order to
+harness efficient symbolic reasoning algorithms: rather than enumerating
+linearizations explicitly, and checking their validity one by one, a symbolic
+reasoning engine may simultaneously rule out many possible linearizations,
+e.g.,~as in the example above. Previous work demonstrates that such symbolic
+reasoning can increase reasoning efficiency by orders of
+magnitude~\cite{conf/pldi/EmmiEH15}.
 
-TODO ALGORITHM OVERVIEW
+Our basic approach to inferring ADT specifications, as outlined by the abstract
+algorithm in Figure~\ref{fig:abstract}, is to identify a finite set of
+sequential execution histories which capture all of the reasons for which a
+sequence could be considered invalid, according to the ADT of a given reference
+implementation. These sequences thus serve as patterns indicating violations in
+the sequences which contain them. Thus the linearizations which exclude all
+violation patterns are considered valid.
+
+\begin{figure}
+  \begin{verbatim}
+procedure INFER_ADT_FROM_REFERENCE_IMPL(I)
+  PATTERNS := {}
+  for each sequential history H
+    if * then
+      break
+    else if H is EXECUTABLE with I then
+      continue
+    else if H is REDUNDANT with PATTERNS then
+      continue
+    else
+      add H to PATTERNS
+    end
+  end
+  return EXCLUSION of PATTERNS
+  \end{verbatim}
+  \caption{An abstract algorithm for symbolic ADT inference.}
+  \label{fig:abstract}
+\end{figure}
+
+The key technical obstacle in realizing the abstract algorithm of
+Figure~\ref{fig:abstract}, overcome in Section~\ref{sec:patterns}, is the
+classification of violations into a finite set of patterns, rendering the
+remaining invalid sequences redundant. The symbolic representation which
+excludes a given set of patterns is relatively straightforward to construct, and
+is given in Section~\ref{sec:formula}. Note however that the termination of this
+abstract algorithm is nondeterministic, and thus the inferred ADT specification
+is generally not the strongest possible in the sense that it may not exclude
+certain invalid linearizations. Nevertheless, specifications inferred by this
+algorithm are sound, in the sense that when modular program reasoning with
+inferred specifications succeeds, correctness follows. In any case, the
+specifications inferred by our refinement of this abstract algorithm in
+Section~\ref{sec:algorithm} are both sound and complete for the
+naturally-occurring ADTs of interest.

### 2015-07-06 motivation.tex
Patterns example in overview.

@@
 
+\begin{figure}
+  \begin{minipage}[b]{0.49\linewidth}
+    \begin{verbatim}
+[1:X] remove => 1  #
+---
+[1:1] add(1)           #
+[2:2] remove => empty    #
+---
+[1:2] remove => 1  #
+[2:2] add(1)         #
+---
+[1:1] add(1)       #
+[2:2] add(2)         #
+[3:2] remove => 2      #
+---
+[1:1] add(1)       #
+[2:1] remove => 1    #
+[3:1] remove => 1      #
+---
+[1:1] add(1)           #
+[2:2] remove => empty    #
+[3:1] remove => 1          #
+---
+[1:1] add(1)       #
+[2:2] add(2)         #
+[3:2] remove => 2      #
+[4:1] remove => 1        #
+    \end{verbatim}
+  \end{minipage}
+  \hfill
+  \begin{minipage}[b]{0.49\linewidth}
+    \begin{verbatim}
+[1:1] add(1)       #
+[2:X] remove => 2    #
+---
+[1:X] remove => 1  #
+[2:X] remove => 1    #
+---
+[1:X] remove => 1      #
+[2:2] remove => empty    #
+---
+[1:1] remove => empty  #
+[2:X] remove => 1        #
+---
+[1:1] add(1)       #
+[2:2] add(2)         #
+[3:X] remove => 3      #
+---
+[1:1] add(1)           #
+[2:2] add(2)             #
+[3:3] remove => empty      #
+---
+[1:1] add(1)       #
+[2:1] remove => 1    #
+[3:X] remove => 2      #
+    \end{verbatim}
+  \end{minipage}
+  \caption{Invalid sequences according to the atomic queue ADT. The histories
+  on the right are each redundant with some history on the left, and those on
+  the left constitute a complete set.}
+  \label{fig:patterns}
+\end{figure}
+
 The key technical obstacle in realizing the abstract algorithm of

### 2015-07-06 motivation.tex
Quick typo fix.

@@
   return EXCLUSION of PATTERNS
+end
   \end{verbatim}

### 2015-07-07 algorithm.tex
Abstract algorithm.
restructuring the technical part

@@
 }
-return $\neg$Formula(Min);
+return $F$(Min);
 \end{lstlisting}
@@
 
-Figure~\ref{fig:algorithm} lists an algorithm for solving the symbolic ADT inference problem which is parametrized by 
-an order relation $\preceq$ between histories such that the complement of the ADT 
-is closed under $\preceq$. For a given set of elements $X$ and an order relation $\preceq$, the \emph{upward closure}
-of $X$ w.r.t. $\preceq$ is the set $\uparrow X=\set{y : \exists x\in X.\ x\preceq y}$. The algorithm performs an
-weight-increasing enumeration of $B$ while maintaining the
+Figure~\ref{fig:algorithm} lists the algorithm for solving the symbolic ADT inference problem which is based on the relation $\preceq$ 
+defined in Section~\ref{sec:patterns}.
+%For a given set of elements $X$ and an order relation $\preceq$, the \emph{upward closure}
+%of $X$ w.r.t. $\preceq$ is the set $\uparrow X=\set{y : \exists x\in X.\ x\preceq y}$. 
+The algorithm performs an weight-increasing enumeration of $B$ while maintaining the
 $\preceq$-minimal elements, and stops whenever each history of weight $i$ is bigger (w.r.t. $\preceq$) than some
@@
 where $B_{\leq i}=\bigcup_{1\leq j\leq i}B_j$.
+Section~\ref{sec:nature} shows that many naturally-occurring ADTs are predictable.
+%Although we don't provide a general result concerning predictability, we prove that it holds for  (see ).
 
@@
 Let $M$ be the value of \texttt{Min} at the end of the \texttt{while} loop.
-Since the complement of $A$ is closed under $\preceq$, and $M\subseteq B\subseteq  \neg \ker A \subseteq \neg A$, we have that $\uparrow M\subseteq \neg A$. 
-Since $A$ is predictable, we have that $B \subseteq\, \uparrow M$.
-The formula $\varphi$ returned by the algorithm holds exactly for the set of histories $X$ 
-which is the complement of $\uparrow M$. By the previous inclusions, we have that $A\subseteq X\subseteq \neg B$.
-Therefore, $\varphi$ holds for all histories in $\ker A$ and doesn't hold for histories in $B$. 
+%By Lemma~\ref{lem:ADT_closure}, the complement of $A$ is closed under $\preceq$. Since 
+%$M\subseteq B\subseteq  \neg \ker A \subseteq \neg A$, the closure of $M$, denoted by $Cl(M)$, satisfies $Cl(M)\subseteq \neg A$. 
+Since $A$ is predictable, we have that $B \subseteq Cl(M)$. % where $Cl(M)$ denotes the closure of 
+Therefore, by Lemma~\ref{lem:formula1}, the returned formula represents $A$.
 \end{proof}
@@
 
-The following result states the conditions under which the algorithm in Figure~\ref{fig:algorithm} terminates.
+The termination of the algorithm in Figure~\ref{fig:algorithm} is a direct consequence of Lemma~\ref{lem:wqo}.
+Also, since the number of weight-$i$ histories is finite\footnote{As
+noted in Section~\ref{sec:histories}, we consider equality between histories up
+to renaming of operation identifiers.} for each $i \in \mathbb{N}$, each $B_i$
+is computable.
 
 \begin{theorem}\label{th:corr2}
-If the set of $\preceq$-minimal elements of any (potentially-infinite) set of histories is finite, then the algorithm in Figure~\ref{fig:algorithm} terminates.
+The algorithm in Figure~\ref{fig:algorithm} terminates.
 \end{theorem}
-
-The following section provides a concrete instantiation for the order relation $\preceq$ satisfying the requirements in 
-Theorem~\ref{th:corr1} and~\ref{th:corr2} except for predictability. 
-Although we don't provide a general result concerning predictability, we prove that it holds for many naturally-occurring ADTs (see Section~\ref{sec:nature}).

### 2015-07-07 formula.tex
restructuring the technical part

@@
 
-\begin{lemma}
+\begin{lemma}\label{lem:formula1}
 
@@
 negative embedding formulas $\set{ \lnot\textsc{Embed}_h : h \in H }$ from any
-set $H$ that generates the complement of $A$.
+set $H$ that \emph{generates} the bounded complement $B$ of $\ker A$, i.e.,
+$H\subseteq B\subseteq Cl_\preceq(H)$ where $Cl_\preceq(H)$ is the closure of $H$ w.r.t. $\preceq$.
 
-\begin{lemma}
+\begin{lemma}\label{lem:formula2}
 
-  The formula $\bigwedge_{h \in H} \neg \textsc{Embed}_h$ represents $A$,
-  for any set $H$ which generates the complement of $A$.
+  The formula $F(H)=\bigwedge_{h \in H} \neg \textsc{Embed}_h$ represents $A$,
+  for any set $H$ which generates the bounded complement of $\ker A$.
 
 \end{lemma}
+\begin{proof}
+For a set $\Sigma$, let $\neg \Sigma$ denote its complement.
+By Lemma~\ref{lem:ADT_closure}, the complement of $A$ is closed under $\preceq$. Since 
+$H\subseteq B\subseteq  \neg \ker A \subseteq \neg A$, we have that $Cl(H)\subseteq \neg A$. 
+By Lemma~\ref{lem:formula1}, the formula $F(H)$ holds exactly for the set of histories $\neg Cl(H)$. 
+%which is the complement of $Cl(M)$. 
+By the previous inclusions, we have that $A\subseteq \neg Cl(H)\subseteq \neg B$.
+Therefore, $F(H)$ holds for all histories in $\ker A$ 
+and doesn't hold for histories in $B$, which concludes the proof.
+\end{proof}
 
@@
 
-For normal and predictable ADTs, such sets are computable, and so our inference
-problem is also computable.
-
-\begin{theorem}
-
-  The symbolic ADT inference problem is computable for normal and predictable ADTs.
-
-\end{theorem}
+%For normal and predictable ADTs, such sets are computable, and so our inference
+%problem is also computable.
+%
+%\begin{theorem}
+%
+%  The symbolic ADT inference problem is computable for normal and predictable ADTs.
+%
+%\end{theorem}

### 2015-07-07 patterns.tex
restructuring the technical part

@@
 identification of certain algebraic properties of the sets of histories
-admitted by ADTs. In particular, we find that these sets of histories are
-closed under the removal of certain operations, and that sets of histories
-adhere to a well-founded ordering under the relation induced by such removals.
+admitted by ADTs, more precisely, that they are
+closed under the removal of certain operations. 
+%and that the bounded complement of their kernel has 
+%sets of histories
+%%adhere to a well-founded ordering under the relation induced by such removals.
+%have 
+%a finite set of minimal elements with respect to the relation induced by such removals.
 
@@
 
-\begin{lemma}
+\begin{lemma}\label{lem:ADT_closure}
 
@@
 Besides this closure property, the $\preceq$ relation enjoys a certain notion
-of well-foundedness (when restricted to certain class of histories): a
-\emph{well-quasi-ordering (wqo)} $R$ on a set $X$ is a reflexive, transitive
-binary relation on $X$ for which in every infinite sequence $x_0 x_1 \ldots$ of
-elements from $X$, there exists indices $i < j$ such that $R(x_i,x_j)$.
+of well-foundedness (when restricted to a certain class of histories): every potentially 
+infinite set of width-bounded histories has a finite set of minimal elements w.r.t. $\preceq$.
 
-\begin{example}
+\begin{lemma}\label{lem:wqo}
 
-  Consider the infinite history sequence $h_1 h_2 \ldots$ where each $h_i$
-  contains $2i$ operations $o_1, o_1', \ldots, o_i, o_i'$ where each $o_j$ is a
-  completed {\tt push} operation matching itself, and each $o_j'$ is a pending
-  {\tt pop} operation with undefined matching. Because each successive $h_i$
-  has both more matches and more pending operations, no two histories of the
-  sequence are related by $\preceq$. Thus $\preceq$ is not a wqo.
-
-\end{example}
-
-The previous example demonstrates that $\preceq$ is not a wqo by allowing each
-history $h_i$ of the infinite sequence to contain more and more pending
-operations in order to ensure that $h_j \not\preceq h_i$ for every $j < i$.
-Curbing this ability by limiting the maximum amount of pending operations per
-history makes $\preceq$ a wqo. Although limiting to $k$ pending operations
-essentially limits us to width-$k$ histories, of executions with at most $k$
-operations parallel at any moment, e.g.,~of programs with at most $k$ threads,
-this restriction comes at no loss of completeness when considering only the
-histories of bounded-width ADT kernels, which is the subject of the remainder
-of this section.
-
-We first prove that the embedding order, defined hereafter, is a wqo on width-bounded
-labeled interval orders.
-Let $\Sigma$ be a finite alphabet. A labeled interval order is a triple $(A,\leq,\ell)$, where $(A,\leq)$ is an interval 
-order, and $\ell:A\rightarrow \Sigma$ is a labeling function. The embedding order $\subseteq$ between labeled interval orders is defined as
-usual by:
-\begin{itemize}
-	\item $(A_1,\leq_1,\ell_1)\subseteq (A_2,\leq_2,\ell_2)$ iff there exists an injective function $h:A_1\rightarrow A_2$ that 
-	preserves the labeling, i.e., $\ell_1(x)=\ell_2(h(x))$, for every $x\in A_1$, and the order constraints, i.e.,
-	for every $x,y\in A_1$, if $x \leq_1 y$, then $h(x)\leq_2 h(y)$.
-\end{itemize}
-
-The width of an interval order $(A,\leq,\ell)$ is the maximum number of elements which are mutually unordered.
-
-\begin{lemma}
-$\subseteq$ is a wqo on width-bounded interval orders.
-\end{lemma}
-\begin{proof}
-By definition, for every interval order $(A,\leq,\ell)$, there exists a mapping $I$ from elements of $A$ to intervals on $\mathbb{N}$ such that for every $x,y\in A$, if $x\leq y$, then the interval $I(x)$ ends before the interval $I(y)$ (i.e., the upper bound of $I(x)$ is strictly smaller than the lower bound of $I(y)$).
-%In general, there are more than one mappings $I_A$ that satisfy these constraints.
-In the following, we assume that the mapping $I_A$ is arbitrary but fixed.
-Therefore, in the following, we assume that an interval order is a multiset $\Gamma$ of triples of the form $[i,j,a]$ where $[i,j]$ is an interval on $\mathbb{N}$ and $a$ is a symbol in $\Sigma$.
-
-We actually prove that another order on interval orders, denoted by $\subseteqq$, which is stronger than the embedding order $\subseteq$, is a wqo.
-The order $\subseteqq$ is defined by: $\Gamma_1\subseteqq \Gamma_2$ iff there exists an injective function $h:\Gamma_1\rightarrow \Gamma_2$ such that
-\begin{enumerate}
-	\item $h$ preserves the labeling, i.e., for any triple $[i,j,a]$, $h([i,j,a])=[i',j',a]$, for some $i'$ and $j'$, 
-	\item $h$ preserves the ordering constraints, i.e., for every two triples $[i,j,a]$ and $[k,l,b]$ such that $j< k$, $h([i,j,a])=[i',j',a]$, $h([k,l,b])=[k',l',b]$, and $j'< k'$.
-	\item for every two incomparable elements $x$ and $y$ of $\Gamma_1$, if the interval of $x$ starts before the interval of $y$, then the interval of $h(x)$ also starts before the interval of $h(y)$.
-Formally, for every two triples $[i,j,a]$ and $[k,l,b]$ such that $i\leq k$ (i.e., the first interval starts before the second interval), $h([i,j,a])=[i',j',a]$, $h([k,l,b])=[k',l',b]$, and $i'\leq k'$.
-\end{enumerate}
-We say that $h$ \emph{witnesses} $\Gamma_1\subseteqq \Gamma_2$.
-
-Assume $\Gamma_1$, $\Gamma_2$,$\ldots$ is a \emph{bad} sequence, i.e., an infinite sequence of interval orders s.t.
-there exists no $i < j$ with $\Gamma_i\subseteqq \Gamma_j$.
-Also, assume that $\Gamma_1$ is the minimal size interval order that can start a bad sequence, $\Gamma_2$ is the minimal order that can continue a bad sequence starting with $\Gamma_1$, and so on.
-
-For each $\Gamma_k$, let $Min(\Gamma_k)$ be a triple $[i,j,a]$ such that 
-(1) $i$ is the minimal lower bound of an interval in $\Gamma_k$, and 
-(2) $j$ is the minimal upper bound of an interval in $\Gamma_k$ with lower bound $i$.
-%In general, this triple is not unique but it’s not important which one we choose.
-Also, let $Init(\Gamma_k) = (a_k, P_k)$, where $Min(\Gamma_k)=[i,j,a_k]$, for some $i$ and $j$, and $P_k$ is the multiset of symbols labeling elements of $\Gamma_k$ that are incomparable to $Min(\Gamma_k)$.
-Note that $P_k$ is bounded since we assume width-bounded interval orders.
-
-The infinite sequence $\Gamma_1$, $\Gamma_2$,$\ldots$ contains an infinite sequence $\Gamma_{k_1}$, $\Gamma_{k_2}$,$\ldots$ which have the same value for $Init$.
-For each $\Gamma_k$, let $\Lambda_k$ be the interval order obtained from $\Gamma_k$ by removing the triple $Min(\Gamma_k)$.
-By the minimality assumptions, the infinite sequence $\Gamma_1$, $\Gamma_2$,$\ldots$,$\Gamma_{k_1-1}$, $\Lambda_{k_1}$, $\Lambda_{k_2}$,$\ldots$ is not bad. 
-Therefore, there exists $m<n$ such that $\Lambda_{k_m}\subseteqq \Lambda_{k_n}$.
-
-It remains to prove that:
-\begin{itemize}
-	\item if $Init(\Gamma_{k_m})=Init(\Gamma_{k_n})$ and $\Lambda_{k_m}\subseteqq \Lambda_{k_n}$, then $\Gamma_{k_m}\subseteqq \Gamma_{k_n}$.
-\end{itemize}
-
-Let $h$ be the injective function witnessing $\Lambda_{k_m}\subseteqq \Lambda_{k_n}$.
-We prove that the extension $h'$ of $h$ between $\Gamma_{k_m}$ and $\Gamma_{k_n}$, defined by $h'(Min(\Gamma_{k_m}))=Min(\Gamma_{k_n})$ and $h'(t)=h(t)$, for all $t\in \Lambda_{k_m}$, witnesses $\Gamma_{k_m}\subseteqq \Gamma_{k_n}$.
-
-Clearly, $h'$ preserves the labeling. 
-To prove that $h'$ preserves the ordering constraints, let $Min(\Gamma_{k_m})=[i,j,a]$ and $[k,l,b]\in \Gamma_{k_m}$ such that $j< k$.
-Also, let $Min(\Gamma_{k_n})=[i',j',a]$ and $h'([k,l,b])=h([k,l,b])=[k',l',b]$.
-Assume by contradiction that $k'\leq j'$.
-Since $h$ is injective, there exists an element $[e,f,c]\in \Gamma_{k_m}$ incomparable to $Min(\Gamma_{k_m})$ which is mapped to an element $[e',f',c]$ greater than $Min(\Gamma_{k_n})$ (because, by definition, $\Gamma_{k_m}$ and $\Gamma_{k_n}$ have the same number of elements incomparable to $Min(\Gamma_{k_m})$ and respectively, $Min(\Gamma_{k_n})$).
-Since $[e,f,c]$ is incomparable to $[i,j,a]$, we obtain that $e\leq j< k$.
-Also, by the current assumptions, $k'\leq j'<e'$.
-Therefore, there exist two elements $[e,f,c]$ and $[k,l,b]$ such that the interval $[e,f]$ starts before $[k,l]$ which are mapped by $h$ to the elements $[e',f',c]$ and $[k',l',b]$ such that the interval $[e',f']$ starts after $[k',l']$.
-This contradicts the fact that $h$ witnesses $\Lambda_{k_m}\subseteqq \Lambda_{k_n}$.
-
-The properties of $Min(\Gamma_{k_m})$ and $Min(\Gamma_{k_n})$ imply that $h'$ satisfies also the third property in the definition of $\subseteqq$.
-
-Finally, since $\subseteqq$ is stronger than $\subseteq$, we get that $\subseteq$ is a wqo on width-bounded labeled interval orders.
-\end{proof}
-
-\begin{lemma}
-
-  $\preceq$ is a wqo on width-bounded histories.
+  Sets of width-bounded histories have a finite set of $\preceq$-minimals.
 
@@
 
-\begin{proof}
-
-%We prove that $\preceq$ is well-founded, i.e., there is no infinite strictly decreasing sequence $h_1 \succ h_2\succ \ldots$\footnote{We define $\succ$ as usual by $h\succ h'$ iff $h\succeq h'$ and $h\neq h'$.},
-%and that $\preceq$ has no infinite antichain, i.e., there is no infinite set $\{h_1,h_2,\ldots\}$ of mutually incomparable elements 
-%$h_i\not\preceq h_j$ and $h_j\not\preceq h_i$ when $i\neq j$.
-%
-%Concerning well-foundedness, note that a given history $h_1$ can be weaken only finitely many times by removing matches, duplicates in a match, or happens-before ordering constraints, or by turning completed operations into pending operations.
-Since we restrict ourselves to width-bounded histories, $h_1$ can also be weaken only finitely many times by adding pending operations (adding a pending operation increases the width of a history by 1).
-%
-%TODO PROBLEM WITH PENDING OPS
-
-%To prove that $\preceq$ has no infinite antichain, 
-%we use the following definitions.
-Let $h_1 h_2\ldots $ be an infinite sequence of histories. We prove that there exists $i<j$ such that $h_i\preceq h_j$.
-
-The \emph{signature} of an operation $o$ in a history $h$ is the pair $\sigma(o)=(c(o),f(o))$.
-Then, the \emph{signature} of a match $\mu=\set{ o' \in O : m(o') = o}$ is the pair 
-\[
-\sigma(\mu)=(\sigma(o),\set{\sigma(o') : m(o')=o, o'\neq o})
-\]
-containing the signature of the match target and the set of the signatures of the other operations.
-The \emph{signature} of a history $h$ is the tuple 
-\begin{align*}
-\sigma(h)&=&(\set{\sigma(\mu): \mu\mbox{ a match}}, \set{\sigma(o): o\mbox{ read-only}}, \\
-&& \set{\sigma(o): o\mbox{ unmatched}}, \\
-&& \mset{\sigma(o): o\mbox{ pending and non-matched}}).
-\end{align*}
-
-The history signature contains sets of signatures 
-(for matches, read-only and respectively, completed and unmatched operations) 
-and the multiset of signatures of the pending and non-matched operations.
-
-%Assume by contradiction that $\preceq$ has an infinite antichain.
-Since the set of history signatures is bounded (because the set of methods 
-and the set of pending operations are bounded),
-the sequence $h_1 h_2\ldots $ contains an infinite sequence of histories $h_1' h_2' \ldots$ that have 
-the same signature.
-
-%
-%Also,
-
-The \emph{vector} of a history $h$ is the tuple
-\begin{align*}
-\nu(h)&=&(\mset{\sigma(\mu): \mu\mbox{ a match of $h$}}, \mset{\sigma(o): o\mbox{ read-only}}, \\
-&& \mset{\sigma(o): o\mbox{ unmatched}}).
-\end{align*}
-
-Let $\leq$ be an order relation on multisets $\alpha:\Sigma\rightarrow\mathbb{N}$ of (match or operation) signatures from a finite set $\Sigma$ 
-defined by $\alpha\leq\alpha'$ iff $\alpha(\sigma)\leq\alpha'(\sigma)$, for every $\sigma\in\Sigma$.
-The relation $\leq$ is a wqo and so is the component-wise extension of $\leq$ to history vectors.
-Therefore, by known results, the sequence of histories $h_1' h_2' \ldots$ contains 
-an infinite sequence $h_1'',h_2'',\ldots$ such that $\nu(h_1'')\leq \nu(h_2'')\leq \ldots$.
-
-The \emph{vector} of a match $\mu$ is the pair
-\[
-\nu(\mu)=(\sigma(o),\mset{\sigma(o') : m(o')=o, o'\neq o})
-\]
-containing the signature of the match target and the multiset of the signatures of the other operations.
-
-For every history $h$ and match signature $\sigma$, the \emph{$\sigma$-vector} of a history $h$ is the multiset
-$
-\nu_\sigma(h)=\mset{\nu(\mu): \sigma(\mu)=\sigma}
-$
-of match vectors of signature $\sigma$.
-
-Let $\leq_v$ be an order relation on $\sigma$-vectors defined by $\nu_\sigma(h)\leq_v \nu_\sigma(h')$ iff
-for every match vector $\nu$ in $\nu_\sigma(h)$ there is a match vector $\nu'$ in $\nu_\sigma(h')$
-	s.t. $\nu\leq \nu'$ (here, $\leq$ is the order on multisets defined above).
-By known results, $\leq_v$ is a wqo.
-
-Let $\sigma_1$, $\sigma_2$,$\ldots$, $\sigma_n$ be the match signatures defined over a set of methods $\mathbb{M}$.
-Since $\leq_v$ is a wqo, the sequence of histories $h_1'',h_2'',\ldots$ contains an 
-infinite sequence $h_{1}^1,h_{2}^1,\ldots$ such that 
-$\nu_{\sigma_1}(h_{1}^1)\leq_v \nu_{\sigma_1}(h_{2}^1)\leq_v \ldots$.
-Then, the sequence $h_{1}^1,h_{2}^1,\ldots$ contains 
-an infinite sequence $h_{1}^2,h_{2}^2,\ldots$ such that 
-$\nu_{\sigma_2}(h_{1}^2)\leq_v \nu_{\sigma_2}(h_{2}^2)\leq_v \ldots$.
-Applying a similar reasoning for the rest of the match signatures, 
-we obtain that $h_1'',h_2'',\ldots$ contains an 
-infinite sequence $h_{1}^n,h_{2}^n,\ldots$
-such that 
-$\nu_{\sigma}(h_{1}^n)\leq_v \nu_{\sigma}(h_{2}^n)\leq_v \ldots$, for every signature $\sigma$.
-Recall that we also have that
-$\nu(h_1^n)\leq\nu(h_2^n)\leq\ldots$ since $h_{1}^n,h_{2}^n,\ldots$ is a sub-sequence of $h_1'',h_2'',\ldots$.
-
-Viewing histories as labeled interval orders, where the label of an element $o$ is the triple $(c(o),f(o),r(o))$,
-we get that the embedding order $\subseteq$ is a wqo on histories. Therefore, the infinite sequence
-$h_{1}^n,h_{2}^n,\ldots$  contains two elements $h_{i}^n$ and $h_j^n$ with $i<j$ such that $h_{i}^n\subseteq h_j^n$.
-This implies that $h_i^n\preceq h_j^n$, which ends our proof.
-\end{proof}
-
-For the remainder of this section, we fix an ADT $A$, and let $B$ be the
+For the remainder of this section, we fix an ADT $A$, and let $B$ be the bounded
 complement of $\ker A$. When $\ker A$ has bounded width, so does $B$, and thus
-$\preceq$ is a wqo on $B$. Furthermore, when $A$ normal, it is closed under
-$\succeq$, and thus $B$ is closed under $\preceq$~\footnote{Actually, $B$ is closed under $\preceq$ when restricted to width-bounded histories, i.e., if $h_1 \preceq h_2$, $h_1 \in B$, and $h_2$ is width-bounded, then $h_2\in B$.}.
-Closure under a wqo implies
+$B$ has a finite set of $\preceq$-minimals. 
+Furthermore, when $A$ normal, it is closed under
+$\succeq$, and thus $B$ is closed under 
+$\preceq$~\footnote{Actually, $B$ is closed under $\preceq$ when restricted to width-bounded histories, i.e., if $h_1 \preceq h_2$, $h_1 \in B$, and $h_2$ is width-bounded, then $h_2\in B$.}.
+Closure under a relation satisfying Lemma~\ref{lem:wqo} implies
 representation by a finite set. Formally, we say a set $X$ is \emph{finitely
@@
 
-Computing $B$ thus amounts to computing its finite representation. In general,
-this is achieved by enumerating the elements of $B$ while maintaining the
-$\preceq$-minimal elements, and recognizing a condition under which all
-elements of $B$ are related to the current set of
-minimals~\cite{conf/lics/AbdullaCJT96, journals/tcs/FinkelS01}. In our case, we
-recognize a condition on the weight-increasing enumeration of $B$ which
-guarantees coverage by the current set of minimals. Formally, for all $i \in
-\mathbb{N}$, we define
-\begin{align*}
-  B_i &= \set{ h \in B : \weight(h) \le i } \\
-  B_i' &= \set{ h' \in B :
-    \exists h \in B_i.\ h \preceq h' \text{ and } \weight(h') \le i+1
-  }
-\end{align*}
-as, respectively, the histories of $B$ with at most $i$ matches and duplicates,
-and those derived from $B_i$ with at most $i\!+\!1$ matches and duplicates. We
-say that $A$ is \emph{predictable} if $B_i = B$ whenever $B_i' = B_{i+1}$. In
-Section~\ref{sec:nature} we demonstrate that naturally-occurring ADTs are
-predictable. Since the number of weight-$i$ histories is finite\footnote{As
-noted in Section~\ref{sec:histories}, we consider equality between histories up
-to renaming of operation identifiers.} for each $i \in \mathbb{N}$, each $B_i$
-is computable.
-
-\begin{lemma}
-
-  $B_i$ is computable, for all $i \in \mathbb{N}$, if $A$ is normal.
-
-\end{lemma}
-
-Predictability ensures that once $B_{i+1}$ does not alter the set of computed
-$\preceq$-minimals, with respect to those of $B_i$, then the minimals computed
-for $B_i$ are exactly those of $B$.
-
-\begin{lemma}
-
-  $B$ is computable if $A$ is normal and predictable.
-
-\end{lemma}
-
-Note that for non-predictable ADTs, computing $B_i$, for any $i \in
-\mathbb{N}$, under-approximates $B$. In practice, computing $B_i$ can be used
-to identify all violations which surface in histories with $i$ matches,
-overlooking violations which only surface with more than $i$ matches. The
-resulting symbolic ADT representation would still be sound in only identifying
-\emph{actual} violations, yet incomplete in identifying \emph{all} violations.
+%Computing $B$ thus amounts to computing its finite representation. In general,
+%this is achieved by enumerating the elements of $B$ while maintaining the
+%$\preceq$-minimal elements, and recognizing a condition under which all
+%elements of $B$ are related to the current set of
+%minimals~\cite{conf/lics/AbdullaCJT96, journals/tcs/FinkelS01}. In our case, we
+%recognize a condition on the weight-increasing enumeration of $B$ which
+%guarantees coverage by the current set of minimals. Formally, for all $i \in
+%\mathbb{N}$, we define
+%\begin{align*}
+%  B_i &= \set{ h \in B : \weight(h) \le i } \\
+%  B_i' &= \set{ h' \in B :
+%    \exists h \in B_i.\ h \preceq h' \text{ and } \weight(h') \le i+1
+%  }
+%\end{align*}
+%as, respectively, the histories of $B$ with at most $i$ matches and duplicates,
+%and those derived from $B_i$ with at most $i\!+\!1$ matches and duplicates. We
+%say that $A$ is \emph{predictable} if $B_i = B$ whenever $B_i' = B_{i+1}$. In
+%Section~\ref{sec:nature} we demonstrate that naturally-occurring ADTs are
+%predictable. Since the number of weight-$i$ histories is finite\footnote{As
+%noted in Section~\ref{sec:histories}, we consider equality between histories up
+%to renaming of operation identifiers.} for each $i \in \mathbb{N}$, each $B_i$
+%is computable.
+%
+%\begin{lemma}
+%
+%  $B_i$ is computable, for all $i \in \mathbb{N}$, if $A$ is normal.
+%
+%\end{lemma}
+%
+%Predictability ensures that once $B_{i+1}$ does not alter the set of computed
+%$\preceq$-minimals, with respect to those of $B_i$, then the minimals computed
+%for $B_i$ are exactly those of $B$.
+%
+%\begin{lemma}
+%
+%  $B$ is computable if $A$ is normal and predictable.
+%
+%\end{lemma}
+%
+%Note that for non-predictable ADTs, computing $B_i$, for any $i \in
+%\mathbb{N}$, under-approximates $B$. In practice, computing $B_i$ can be used
+%to identify all violations which surface in histories with $i$ matches,
+%overlooking violations which only surface with more than $i$ matches. The
+%resulting symbolic ADT representation would still be sound in only identifying
+%\emph{actual} violations, yet incomplete in identifying \emph{all} violations.
 

### 2015-07-07 algorithm.tex
restructuring the technical part

@@
 \caption{The algorithm solving the symbolic ADT inference problem for a given ADT $A$, where $B$ denotes the bounded complement of $\ker A$.
-The procedure $\texttt{Min}(X)$ returns the $\preceq$-minimal elements of the set of histories $X$ and $\texttt{Formula}(X)$ returns a history formula which holds exactly for the histories in the upward closure of $X$.}
+The procedure $\texttt{Min}(X)$ returns the $\preceq$-minimal elements of the set of histories $X$.
+%and $\texttt{Formula}(X)$ returns a history formula which holds exactly for the histories in the upward closure of $X$.
+} 
 \label{fig:algorithm}

### 2015-07-07 appendix.tex
restructuring the technical part

%!TEX root = draft.tex
\appendix
\section{Proof of Lemma~\ref{lem:wqo}}

The result follows from the fact that $\preceq$ is a well-quasi-ordering on width-bounded histories.
It is well-known that every potentially infinite set of elements has a finite set of minimals w.r.t. a well-quasi-ordering.

A \emph{well-quasi-ordering (wqo)} $R$ on a set $X$ is a reflexive, transitive
binary relation on $X$ for which in every infinite sequence $x_0 x_1 \ldots$ of
elements from $X$, there exists indices $i < j$ such that $R(x_i,x_j)$.

\begin{example}

  Consider the infinite history sequence $h_1 h_2 \ldots$ where each $h_i$
  contains $2i$ operations $o_1, o_1', \ldots, o_i, o_i'$ where each $o_j$ is a
  completed {\tt push} operation matching itself, and each $o_j'$ is a pending
  {\tt pop} operation with undefined matching. Because each successive $h_i$
  has both more matches and more pending operations, no two histories of the
  sequence are related by $\preceq$. Thus $\preceq$ is not a wqo.

\end{example}

The previous example demonstrates that $\preceq$ is not a wqo by allowing each
history $h_i$ of the infinite sequence to contain more and more pending
operations in order to ensure that $h_j \not\preceq h_i$ for every $j < i$.
Curbing this ability by limiting the maximum amount of pending operations per
history makes $\preceq$ a wqo. Although limiting to $k$ pending operations
essentially limits us to width-$k$ histories, of executions with at most $k$
operations parallel at any moment, e.g.,~of programs with at most $k$ threads,
this restriction comes at no loss of completeness when considering only the
histories of bounded-width ADT kernels, which is the subject of the remainder
of this section.

We first prove that the embedding order, defined hereafter, is a wqo on width-bounded
labeled interval orders.
Let $\Sigma$ be a finite alphabet. A labeled interval order is a triple $(A,\leq,\ell)$, where $(A,\leq)$ is an interval 
order, and $\ell:A\rightarrow \Sigma$ is a labeling function. The embedding order $\subseteq$ between labeled interval orders is defined as
usual by:
\begin{itemize}
	\item $(A_1,\leq_1,\ell_1)\subseteq (A_2,\leq_2,\ell_2)$ iff there exists an injective function $h:A_1\rightarrow A_2$ that 
	preserves the labeling, i.e., $\ell_1(x)=\ell_2(h(x))$, for every $x\in A_1$, and the order constraints, i.e.,
	for every $x,y\in A_1$, if $x \leq_1 y$, then $h(x)\leq_2 h(y)$.
\end{itemize}

The width of an interval order $(A,\leq,\ell)$ is the maximum number of elements which are mutually unordered.

\begin{lemma}
$\subseteq$ is a wqo on width-bounded interval orders.
\end{lemma}
\begin{proof}
By definition, for every interval order $(A,\leq,\ell)$, there exists a mapping $I$ from elements of $A$ to intervals on $\mathbb{N}$ such that for every $x,y\in A$, if $x\leq y$, then the interval $I(x)$ ends before the interval $I(y)$ (i.e., the upper bound of $I(x)$ is strictly smaller than the lower bound of $I(y)$). %In general, there are more than one mappings $I_A$ that satisfy these constraints. In the following, we assume that the mapping $I_A$ is arbitrary but fixed.
Therefore, in the following, we assume that an interval order is a multiset $\Gamma$ of triples of the form $[i,j,a]$ where $[i,j]$ is an interval on $\mathbb{N}$ and $a$ is a symbol in $\Sigma$.

We actually prove that another order on interval orders, denoted by $\subseteqq$, which is stronger than the embedding order $\subseteq$, is a wqo. The order $\subseteqq$ is defined by: $\Gamma_1\subseteqq \Gamma_2$ iff there exists an injective function $h:\Gamma_1\rightarrow \Gamma_2$ such that 
\begin{enumerate}
	\item $h$ preserves the labeling, i.e., for any triple $[i,j,a]$, $h([i,j,a])=[i',j',a]$, for some $i'$ and $j'$, 
	\item $h$ preserves the ordering constraints, i.e., for every two triples $[i,j,a]$ and $[k,l,b]$ such that $j< k$, $h([i,j,a])=[i',j',a]$, $h([k,l,b])=[k',l',b]$, and $j'< k'$.
	\item for every two incomparable elements $x$ and $y$ of $\Gamma_1$, if the interval of $x$ starts before the interval of $y$, then the interval of $h(x)$ also starts before the interval of $h(y)$. Formally, for every two triples $[i,j,a]$ and $[k,l,b]$ such that $i\leq k$ (i.e., the first interval starts before the second interval), $h([i,j,a])=[i',j',a]$, $h([k,l,b])=[k',l',b]$, and $i'\leq k'$.
\end{enumerate}
We say that $h$ \emph{witnesses} $\Gamma_1\subseteqq \Gamma_2$.

Assume $\Gamma_1$, $\Gamma_2$,$\ldots$ is a \emph{bad} sequence, i.e., an infinite sequence of interval orders s.t. there exists no $i < j$ with $\Gamma_i\subseteqq \Gamma_j$. Also, assume that $\Gamma_1$ is the minimal size interval order that can start a bad sequence, $\Gamma_2$ is the minimal order that can continue a bad sequence starting with $\Gamma_1$, and so on.

For each $\Gamma_k$, let $Min(\Gamma_k)$ be a triple $[i,j,a]$ such that 
(1) $i$ is the minimal lower bound of an interval in $\Gamma_k$, and 
(2) $j$ is the minimal upper bound of an interval in $\Gamma_k$ with lower bound $i$.
%In general, this triple is not unique but it’s not important which one we choose.
Also, let $Init(\Gamma_k) = (a_k, P_k)$, where $Min(\Gamma_k)=[i,j,a_k]$, for some $i$ and $j$, and $P_k$ is the multiset of symbols labeling elements of $\Gamma_k$ that are incomparable to $Min(\Gamma_k)$. Note that $P_k$ is bounded since we assume width-bounded interval orders.

The infinite sequence $\Gamma_1$, $\Gamma_2$,$\ldots$ contains an infinite sequence $\Gamma_{k_1}$, $\Gamma_{k_2}$,$\ldots$ which have the same value for $Init$.
For each $\Gamma_k$, let $\Lambda_k$ be the interval order obtained from $\Gamma_k$ by removing the triple $Min(\Gamma_k)$.
By the minimality assumptions, the infinite sequence $\Gamma_1$, $\Gamma_2$,$\ldots$,$\Gamma_{k_1-1}$, $\Lambda_{k_1}$, $\Lambda_{k_2}$,$\ldots$ is not bad. 
Therefore, there exists $m<n$ such that $\Lambda_{k_m}\subseteqq \Lambda_{k_n}$.

It remains to prove that:
\begin{itemize}
	\item if $Init(\Gamma_{k_m})=Init(\Gamma_{k_n})$ and $\Lambda_{k_m}\subseteqq \Lambda_{k_n}$, then $\Gamma_{k_m}\subseteqq \Gamma_{k_n}$.
\end{itemize}

Let $h$ be the injective function witnessing $\Lambda_{k_m}\subseteqq \Lambda_{k_n}$. We prove that the extension $h'$ of $h$ between $\Gamma_{k_m}$ and $\Gamma_{k_n}$, defined by $h'(Min(\Gamma_{k_m}))=Min(\Gamma_{k_n})$ and $h'(t)=h(t)$, for all $t\in \Lambda_{k_m}$, witnesses $\Gamma_{k_m}\subseteqq \Gamma_{k_n}$.

Clearly, $h'$ preserves the labeling. 
To prove that $h'$ preserves the ordering constraints, let $Min(\Gamma_{k_m})=[i,j,a]$ and $[k,l,b]\in \Gamma_{k_m}$ such that $j< k$. Also, let $Min(\Gamma_{k_n})=[i',j',a]$ and $h'([k,l,b])=h([k,l,b])=[k',l',b]$. Assume by contradiction that $k'\leq j'$. Since $h$ is injective, there exists an element $[e,f,c]\in \Gamma_{k_m}$ incomparable to $Min(\Gamma_{k_m})$ which is mapped to an element $[e',f',c]$ greater than $Min(\Gamma_{k_n})$ (because, by definition, $\Gamma_{k_m}$ and $\Gamma_{k_n}$ have the same number of elements incomparable to $Min(\Gamma_{k_m})$ and respectively, $Min(\Gamma_{k_n})$). 
Since $[e,f,c]$ is incomparable to $[i,j,a]$, we obtain that $e\leq j< k$. Also, by the current assumptions, $k'\leq j'<e'$. Therefore, there exist two elements $[e,f,c]$ and $[k,l,b]$ such that the interval $[e,f]$ starts before $[k,l]$ which are mapped by $h$ to the elements $[e',f',c]$ and $[k',l',b]$ such that the interval $[e',f']$ starts after $[k',l']$. This contradicts the fact that $h$ witnesses $\Lambda_{k_m}\subseteqq \Lambda_{k_n}$.

The properties of $Min(\Gamma_{k_m})$ and $Min(\Gamma_{k_n})$ imply that $h'$ satisfies also the third property in the definition of $\subseteqq$.

Finally, since $\subseteqq$ is stronger than $\subseteq$, we get that $\subseteq$ is a wqo on width-bounded labeled interval orders.
\end{proof}

\begin{lemma}

  $\preceq$ is a wqo on width-bounded histories.

\end{lemma}

\begin{proof}

%We prove that $\preceq$ is well-founded, i.e., there is no infinite strictly decreasing sequence $h_1 \succ h_2\succ \ldots$\footnote{We define $\succ$ as usual by $h\succ h'$ iff $h\succeq h'$ and $h\neq h'$.},
%and that $\preceq$ has no infinite antichain, i.e., there is no infinite set $\{h_1,h_2,\ldots\}$ of mutually incomparable elements 
%$h_i\not\preceq h_j$ and $h_j\not\preceq h_i$ when $i\neq j$.
%
%Concerning well-foundedness, note that a given history $h_1$ can be weaken only finitely many times by removing matches, duplicates in a match, or happens-before ordering constraints, or by turning completed operations into pending operations. Since we restrict ourselves to width-bounded histories, $h_1$ can also be weaken only finitely many times by adding pending operations (adding a pending operation increases the width of a history by 1).
%
%TODO PROBLEM WITH PENDING OPS

%To prove that $\preceq$ has no infinite antichain, 
%we use the following definitions.
Let $h_1 h_2\ldots $ be an infinite sequence of histories. We prove that there exists $i<j$ such that $h_i\preceq h_j$.

The \emph{signature} of an operation $o$ in a history $h$ is the pair $\sigma(o)=(c(o),f(o))$.
Then, the \emph{signature} of a match $\mu=\set{ o' \in O : m(o') = o}$ is the pair 
\[
\sigma(\mu)=(\sigma(o),\set{\sigma(o') : m(o')=o, o'\neq o})
\]
containing the signature of the match target and the set of the signatures of the other operations.
The \emph{signature} of a history $h$ is the tuple 
\begin{align*}
\sigma(h)&=&(\set{\sigma(\mu): \mu\mbox{ a match}}, \set{\sigma(o): o\mbox{ read-only}}, \\
&& \set{\sigma(o): o\mbox{ unmatched}}, \\
&& \mset{\sigma(o): o\mbox{ pending and non-matched}}).
\end{align*}

The history signature contains sets of signatures 
(for matches, read-only and respectively, completed and unmatched operations) 
and the multiset of signatures of the pending and non-matched operations.

%Assume by contradiction that $\preceq$ has an infinite antichain.
Since the set of history signatures is bounded (because the set of methods 
and the set of pending operations are bounded),
the sequence $h_1 h_2\ldots $ contains an infinite sequence of histories $h_1' h_2' \ldots$ that have 
the same signature.

%
%Also,

The \emph{vector} of a history $h$ is the tuple
\begin{align*}
\nu(h)&=&(\mset{\sigma(\mu): \mu\mbox{ a match of $h$}}, \mset{\sigma(o): o\mbox{ read-only}}, \\
&& \mset{\sigma(o): o\mbox{ unmatched}}).
\end{align*}

Let $\leq$ be an order relation on multisets $\alpha:\Sigma\rightarrow\mathbb{N}$ of (match or operation) signatures from a finite set $\Sigma$ 
defined by $\alpha\leq\alpha'$ iff $\alpha(\sigma)\leq\alpha'(\sigma)$, for every $\sigma\in\Sigma$.
The relation $\leq$ is a wqo and so is the component-wise extension of $\leq$ to history vectors.
Therefore, by known results, the sequence of histories $h_1' h_2' \ldots$ contains 
an infinite sequence $h_1'',h_2'',\ldots$ such that $\nu(h_1'')\leq \nu(h_2'')\leq \ldots$.

The \emph{vector} of a match $\mu$ is the pair
\[
\nu(\mu)=(\sigma(o),\mset{\sigma(o') : m(o')=o, o'\neq o})
\]
containing the signature of the match target and the multiset of the signatures of the other operations.

For every history $h$ and match signature $\sigma$, the \emph{$\sigma$-vector} of a history $h$ is the multiset
$
\nu_\sigma(h)=\mset{\nu(\mu): \sigma(\mu)=\sigma}
$
of match vectors of signature $\sigma$.

Let $\leq_v$ be an order relation on $\sigma$-vectors defined by $\nu_\sigma(h)\leq_v \nu_\sigma(h')$ iff
for every match vector $\nu$ in $\nu_\sigma(h)$ there is a match vector $\nu'$ in $\nu_\sigma(h')$
	s.t. $\nu\leq \nu'$ (here, $\leq$ is the order on multisets defined above).
By known results, $\leq_v$ is a wqo.

Let $\sigma_1$, $\sigma_2$,$\ldots$, $\sigma_n$ be the match signatures defined over a set of methods $\mathbb{M}$.
Since $\leq_v$ is a wqo, the sequence of histories $h_1'',h_2'',\ldots$ contains an 
infinite sequence $h_{1}^1,h_{2}^1,\ldots$ such that 
$\nu_{\sigma_1}(h_{1}^1)\leq_v \nu_{\sigma_1}(h_{2}^1)\leq_v \ldots$.
Then, the sequence $h_{1}^1,h_{2}^1,\ldots$ contains 
an infinite sequence $h_{1}^2,h_{2}^2,\ldots$ such that 
$\nu_{\sigma_2}(h_{1}^2)\leq_v \nu_{\sigma_2}(h_{2}^2)\leq_v \ldots$.
Applying a similar reasoning for the rest of the match signatures, 
we obtain that $h_1'',h_2'',\ldots$ contains an 
infinite sequence $h_{1}^n,h_{2}^n,\ldots$
such that 
$\nu_{\sigma}(h_{1}^n)\leq_v \nu_{\sigma}(h_{2}^n)\leq_v \ldots$, for every signature $\sigma$.
Recall that we also have that
$\nu(h_1^n)\leq\nu(h_2^n)\leq\ldots$ since $h_{1}^n,h_{2}^n,\ldots$ is a sub-sequence of $h_1'',h_2'',\ldots$.

Viewing histories as labeled interval orders, where the label of an element $o$ is the triple $(c(o),f(o),r(o))$,
we get that the embedding order $\subseteq$ is a wqo on histories. Therefore, the infinite sequence
$h_{1}^n,h_{2}^n,\ldots$  contains two elements $h_{i}^n$ and $h_j^n$ with $i<j$ such that $h_{i}^n\subseteq h_j^n$.
This implies that $h_i^n\preceq h_j^n$, which ends our proof.
\end{proof}

### 2015-07-07 motivation.tex
Presentation of modular reasoning.

@@
 is linearizable. ADT specifications thus decouple reasoning about object
-implementations — is each concurrent execution linearizable? — from their
-clients’ invocations — do all valid linearizations preserve client invariants?
+implementations from their client’s invocations:
+\begin{itemize}
+
+  \item Does every implementation execution have a valid linearization?
+
+  \item Does every valid linearizations preserve client invariants?
+
+\end{itemize}
+The former question depends only on a given object’s implementation, and the
+latter only on a given object’s clients.
 

### 2015-07-07 motivation.tex
Run of the abstract algorithm.

@@
 
-\begin{figure}
+\begin{figure}[t]
   \begin{minipage}{0.43\linewidth}
@@
 
-\begin{figure}
+\begin{figure}[t]
   \begin{verbatim}
@@
 
-\begin{figure}
+As an example of this algorithm at work, consider the sequential histories
+listed in Figure~\ref{fig:patterns}. These sequences arise from an enumeration
+of all $202$ possible method invocation sequences\footnote{Here we consider
+equivalence among sequences which are isomorphic up to renaming of data values,
+e.g.,~to avoid considering {\tt add(1); add(2); remove => 1} and {\tt add(3);
+add(2); remove => 3} as distinct. The notion of \emph{matching} introduced in
+Section~\ref{sec:histories} provides a clean formal treatment.} of length at
+most $4$ of an object with add and remove methods for which remove can return
+empty. The $31$ sequences which are executable by a correct reference
+implementation of an atomic queue implementation --- i.e.,~with matching return
+values for all invocations --- are discarded. The remaining $171$ sequences are
+invalid, $164$ of which are redundant with the seven {\tt PATTERNS} listed on the
+left-hand side of Figure~\ref{fig:patterns}. Seven of these redundant sequences
+are shown on the right-hand side of Figure~\ref{fig:patterns}. For example, the
+first five sequences on the right, as well as the last, are redundant with the
+first on the left since they each describe a violation in which a removed
+element was never added. The sixth sequence on the right is redundant with the
+second on the left since they both describe a violation in which remove returns
+empty before previously-added elements are removed. Finally, the exclusion of
+the computed pattern set can be expressed as a conjunction of formulas
+in first order logic describing the exclusion of each individual pattern. For
+example, the first pattern could be described by the formula
+\begin{align*}
+  \exists o.\
+    {\sf method}(o) = {\tt remove} \land {\sf unmatched}(o)
+\end{align*}
+describing an unmatched remove operation.
+
+\begin{figure}[t]
   \begin{minipage}[b]{0.49\linewidth}
@@
   on the right are each redundant with some history on the left, and those on
-  the left constitute a complete set.}
+  the left constitute a complete set. The “\#” symbols depict the time intervals
+  spanned by operations horizontally, and the “{\tt [$i$:$j$]}” notation refers
+  to the current line’s operation identifier $i$, and the identifier $j$ of
+  its “match-source” operation. The “{\tt X}” symbol denotes an absent match.}
   \label{fig:patterns}

### 2015-07-07 histories.tex
example <= relation
Fixed space hacks.
Improved histories section.

--- 2015-03-25 histories.tex
@@
   Consider the following three executions of a single-value register object
-  with read and write methods
+  with read and write methods:
   \begin{center}
-    \setlength{\tabcolsep}{1em}
-    \begin{tabular}{l|l|l}
-      $e_1$                           & $e_2$                           & $e_3$ \\
-      $1:\mathrm{write}(a)$           & $1:\mathrm{write}(a)$           & $1:\mathrm{write}(a)$ \\
-      $1:\mathbf{ret}$                & $2:\mathrm{write}(b)$           & $1:\mathbf{ret}$ \\
-      $2:\mathrm{write}(b)$           & $1:\mathbf{ret}$                & $2:\mathrm{write}(b)$ \\
-      $2:\mathbf{ret}$                & $2:\mathbf{ret}$                & $3:\mathrm{read}$ \\
-      $3:\mathrm{read}$               & $3:\mathrm{read}$               & $2:\mathbf{ret}$ \\
-      $3:\mathbf{ret} \Rightarrow a$  & $3:\mathbf{ret} \Rightarrow a$  & $3:\mathbf{ret} \Rightarrow a$ \\
-    \end{tabular}
+    \hfill
+    \begin{minipage}{0.3\linewidth}
+      $e_1$
+      \begin{verbatim}
+1: call write(a)
+1: return
+2: call write(b)
+2: return
+3: call read
+3: return => a
+      \end{verbatim}
+    \end{minipage}
+    \hfill
+    \begin{minipage}{0.3\linewidth}
+      $e_2$
+      \begin{verbatim}
+1: call write(a)
+2: call write(b)
+1: return
+2: return
+3: call read
+3: return => a
+      \end{verbatim}
+    \end{minipage}
+    \hfill
+    \begin{minipage}{0.3\linewidth}
+      $e_3$
+      \begin{verbatim}
+1: call write(a)
+1: return
+2: call write(b)
+3: call read
+2: return
+3: return => a
+      \end{verbatim}
+    \end{minipage}
+    \hfill
   \end{center}
-  in which actions are preceded by their operation identifiers. Executions
-  $e_2$ and $e_3$ are obtained from $e_1$ by permuting the return actions of
-  operations $1$ and $2$, respectively, forward, or symmetrically, the call
-  actions of operations $2$ and $3$, respectively, backward. Thus while the
-  operations of Execution $e_1$ are sequential, each following the previous in
-  time, those of $e_2$ and $e_3$ overlap. While $e_1$ should not be admitted
-  by an \emph{atomic} register, since the read of Operation~$3$ does not
-  return the most-recently-written value, both $e_2$ and $e_3$ could be
-  admitted, since there exist linearizations of the partially-ordered operations
-  such that Operation~$3$ does return the most-recently-written value.
+  Operation identifiers precede actions. Executions~$e_2$ and~$e_3$ are obtained
+  from $e_1$, respectively, by permuting the return actions of Operations~$1$
+  and~$2$ and the call actions of operations $2$ and $3$. Thus while the
+  operations of Execution~$e_1$ are sequential, each following the previous in
+  time, those of $e_2$ and $e_3$ overlap. While $e_1$ should not be admitted by
+  an atomic register, since the read of Operation~$3$ does not return the
+  most-recently-written value, both $e_2$ and $e_3$ should be admitted, since
+  Operation~$3$ returns the most-recently-written value in some linearization of
+  the overlapping operations.
 
@@
 
-In this work, histories are abstractions of executions which retain method
-names yet lose the exact argument and return values, and retain the relative
-order of operations yet lose the exact sequence of call and return actions.
+Histories abstract executions, retaining method
+names yet losing exact argument and return values, and retaining the relative
+order of operations, yet losing the exact sequence of call and return actions.
 Formally, a \emph{history} is a tuple $h = \tup{O,<,c,f,m,r}$ where
@@
 
+\begin{example}
+
+  By unordering the first two operations of the history
+\begin{verbatim}
+    [1:1] write(a)        #
+    [2:2] write(b)          #
+    [3:2] read => b (RO)      #
+\end{verbatim}
+  we derive the $\to_\mathrm{o}$-related history
+\begin{verbatim}
+    [1:1] write(a)        #
+    [2:2] write(b)        #
+    [3:2] read => b (RO)    #
+\end{verbatim}
+  from which we can derive the $\to_\mathrm{c}$-related history
+\begin{verbatim}
+    [1:1] write(a)        #
+    [2:2] write(b)        #
+    [3:_] read* (RO)        #
+\end{verbatim}
+  by making Operation~$3$ pending, and from which we can derive the
+  $\to_\mathrm{p}$-related history
+\begin{verbatim}
+    [1:1] write(a)        #
+    [2:2] write(b)        #
+    [3:_] read* (RO)        #
+    [4:4] write(c)*         #
+\end{verbatim}
+  by adding an additional pending operation.
+
+\end{example}
+
+Since implementations only dictate methods' executions between their respective
+calls and returns, for any execution they admit, they must also admit executions
+with weaker inter-operation ordering, in which calls may happen earlier, and/or
+returns later. Thus any weakening of an admitted history must also be admitted.
+
 \begin{lemma}

### 2015-07-07 histories.tex
Better section titles.

@@
 %!TEX root = draft.tex
-\section{Histories of ADT Implementations}
+\section{Implementations \& Histories}
 \label{sec:histories}
@@
 
-  By unordering the first two operations of the history
+  By un-ordering the first two operations of the history
 \begin{verbatim}
@@
 
-Since implementations only dictate methods' executions between their respective
-calls and returns, for any execution they admit, they must also admit executions
-with weaker inter-operation ordering, in which calls may happen earlier, and/or
-returns later. Thus any weakening of an admitted history must also be admitted.
+As these weakening operations align with the environment-capturing closure
+properties on executions, the set of histories of an implementation is also
+closed.
 

### 2015-07-07 inference.tex
Fixed space hacks.
Adding motivation section.
starting top-down presentation
Rebranded ADTs as the kernels.

@@
 
-An \emph{abstract data type (ADT)} $A$ is the set $H(\mathcal{I})$ of histories
-of an implementation $\mathcal{I}$. A \emph{kernel} of an ADT $A$ is a minimal
-set $H$ of histories generating $A$, i.e.,~such that
-  \vspace{-1mm}
+A \emph{kernel} of a set $H$ is a minimal set $K$ generating $H$ via the
+relation $\mathord\to = (\to_\mathrm{o} \cup \to_\mathrm{p} \cup
+\to_\mathrm{c})$, i.e.,
 \begin{align*}
-  A = \set{ h' : \exists h \in H.\ h \to^\ast h' }
-    \vspace{-1mm}
+  H = \set{ h' : \exists h \in K.\ h \to^\ast h' }.
 \end{align*}
-where $\mathord\to = (\to_\mathrm{o} \cup \to_\mathrm{p} \cup \to_\mathrm{c})$.
-In Section~\ref{sec:nature} we demonstrate that naturally-occurring ADTs have
-unique, bounded-width kernels. For the remainder of this work, we assume ADTs
-have unique, bounded-width kernels, denoting \emph{the kernel} of $A$ by $\ker
-A$. When $\ker A$ has width $1$, we say $A$ is \emph{atomic}.
+An \emph{abstract data type (ADT)} $A$ is the kernel of the set $H(\mathcal{I})$
+of histories of some implementation $\mathcal{I}$, and $H(A)$ denotes the
+closure of $A$ under $\to$. While the kernels of arbitrary-width history sets
+may not be unique, those with \emph{atomic} kernels, i.e.,~of width-$1$, are
+guaranteed to be unique. Section~\ref{sec:nature} demonstrates that the
+histories of naturally-occurring implementations have bounded-width and
+unique ADTs, which we assume for the remainder of this work.
 
@@
 
-  \vspace{-1mm}
 \begin{definition}
 
-  We say an implementation $\mathcal{I}$ \emph{refines} an ADT $A$ when
-  $H(\mathcal{I}) \subseteq A$. We say an implementation $\mathcal{I}_1$
-  \emph{refines} another implementation $\mathcal{I}_2$ when $\mathcal{I}_1$
-  refines $H(\mathcal{I}_2)$.
-  \vspace{-1mm}
+  An implementation $\mathcal{I}_1$ \emph{refines} another implementation
+  $\mathcal{I}_2$ when $H(\mathcal{I}_1) \subseteq H(\mathcal{I}_2)$. An
+  implementation $\mathcal{I}$ refines an ADT $A$ when $\mathcal{I}$ refines
+  $H(A)$.
+
 \end{definition}
@@
 \emph{history formula} is a first-order logic formula with
-  \vspace{-1mm}
 \begin{itemize}
@@
   non-matching (operations which are not in the domain of the matching function), read-only, and order.
-  \vspace{-1mm}
+
 \end{itemize}
@@
   operation happens between a pair of matching write and read operations:
-  \vspace{-1mm}
   \begin{align*}
@@
       \implies x_3 < x_2
-        \vspace{-1mm}
   \end{align*}
-
-  This is one of several requirements of atomic single-value register ADTs,
-  which is satisfied by certain linearizations of histories $H(e_2)$ and
-  $H(e_3)$ of Example~\ref{ex:histories-of-executions}, yet not $H(e_1)$.
+  This is one of several requirements of atomic single-value register ADTs.
+  It is satisfied by certain linearizations of the histories $H(e_2)$ and
+  $H(e_3)$ from Example~\ref{ex:histories-of-executions}, yet not $H(e_1)$.
 
@@
 
-The \emph{complement} of a history set $H$ of width $k \in \mathbb{N} \cup
+The \emph{bounded complement} of a history set $H$ of width $k \in \mathbb{N} \cup
 \set{\omega}$ is the set of histories of width at most $k$ which are excluded
-from $H$. Let $A$ be an ADT and $B$ the complement of $\ker A$. We say that
+from $H$. Let $A$ be an ADT and $B$ its bounded complement. We say that
 a history formula $F$ \emph{represents} $A$ when
@@
 
-  \item $h \models F$ for all $h \in \ker A$, and
+  \item $h \models F$ for all $h \in A$, and
 
@@
 \end{itemize}
-The inference problem is thus to compute a formula representing an ADT.
+ADT inference is to compute a formula representing an ADT.
 
@@
   The \emph{symbolic ADT inference problem} is to compute a history formula
-  representing the ADT $H(\mathcal{I})$ of a given implementation $\mathcal{I}$.
+  representing the ADT $\ker H(\mathcal{I})$ of a given (reference) implementation
+  $\mathcal{I}$.
 

### 2015-07-07 appendix.tex
Section titles + restructuring the restructuring.

@@
 
-The result follows from the fact that $\preceq$ is a well-quasi-ordering on width-bounded histories.
 It is well-known that every potentially infinite set of elements has a finite set of minimals w.r.t. a well-quasi-ordering.
-
-A \emph{well-quasi-ordering (wqo)} $R$ on a set $X$ is a reflexive, transitive
-binary relation on $X$ for which in every infinite sequence $x_0 x_1 \ldots$ of
-elements from $X$, there exists indices $i < j$ such that $R(x_i,x_j)$.
-
-\begin{example}
-
-  Consider the infinite history sequence $h_1 h_2 \ldots$ where each $h_i$
-  contains $2i$ operations $o_1, o_1', \ldots, o_i, o_i'$ where each $o_j$ is a
-  completed {\tt push} operation matching itself, and each $o_j'$ is a pending
-  {\tt pop} operation with undefined matching. Because each successive $h_i$
-  has both more matches and more pending operations, no two histories of the
-  sequence are related by $\preceq$. Thus $\preceq$ is not a wqo.
-
-\end{example}
-
-The previous example demonstrates that $\preceq$ is not a wqo by allowing each
-history $h_i$ of the infinite sequence to contain more and more pending
-operations in order to ensure that $h_j \not\preceq h_i$ for every $j < i$.
-Curbing this ability by limiting the maximum amount of pending operations per
-history makes $\preceq$ a wqo. Although limiting to $k$ pending operations
-essentially limits us to width-$k$ histories, of executions with at most $k$
-operations parallel at any moment, e.g.,~of programs with at most $k$ threads,
-this restriction comes at no loss of completeness when considering only the
-histories of bounded-width ADT kernels, which is the subject of the remainder
-of this section.
 

### 2015-07-07 formula.tex
Section titles + restructuring the restructuring.

@@
 %!TEX root = draft.tex
-\section{Computing Symbolic Representations}
+\section{Symbolic ADT Representations}
 \label{sec:formula}
@@
 embed a given history, via $\preceq$. Then, using the finite set of histories
-which represent the complement of a given ADT~\footnote{The complement of the kernel of an ADT $A$ is simply called the complement of $A$.}, we represent the ADT itself as
+which represent the complement of a given ADT, we represent the ADT itself as
 the conjunction of negations of these embedding formulas. The resulting formula

### 2015-07-07 patterns.tex
Section titles + restructuring the restructuring.

@@
 %!TEX root = draft.tex
-\section{Computing Finite ADT Representations}
+\section{Finite ADT Representations}
 \label{sec:patterns}
@@
 represented by finite sets of histories, despite the fact that these ADTs admit
-infinite sets of histories. This results is nontrivial, and relies on the
-identification of certain algebraic properties of the sets of histories
-admitted by ADTs, more precisely, that they are
-closed under the removal of certain operations. 
-%and that the bounded complement of their kernel has 
-%sets of histories
-%%adhere to a well-founded ordering under the relation induced by such removals.
-%have 
-%a finite set of minimal elements with respect to the relation induced by such removals.
+infinite sets of histories. This results relies on the identification of certain
+algebraic properties of the sets of histories admitted by ADTs. In particular,
+we find that sets of histories admitted by ADTs are closed under the removal
+of certain operations, and that these sets adhere to a well-founded ordering
+under the relation induced by such removals.
 
 In addition to the relations $\to_\mathrm{o}$, $\to_\mathrm{p}$, and
-$\to_\mathrm{c}$ of Section~\ref{sec:histories} under which all ADTs are
-closed, the ADTs we consider in this work are also closed under additional
-relations which remove read-only operations, unmatched operations, entire
-matches, and duplicate operations. The relations $\to_\mathrm{r}$,
-$\to_\mathrm{u}$, $\to_\mathrm{m}$ and $\to_\mathrm{d}$ relate two histories
-$h_1$ and $h_2$ when $h_2$ is obtained from $h_1$ by
+$\to_\mathrm{c}$ of Section~\ref{sec:histories} under which all implementation
+history sets are closed, the histories of ADT implementations we consider in
+this work are also closed under additional relations which remove read-only
+operations, unmatched operations, entire matches, and duplicate operations. The
+relations $\to_\mathrm{r}$, $\to_\mathrm{u}$, $\to_\mathrm{m}$ and
+$\to_\mathrm{d}$ relate two histories $h_1$ and $h_2$ when $h_2$ is obtained
+from $h_1$ by
 \begin{itemize}
@@
 \end{itemize}
-We say an ADT closed under $\to_\mathrm{r}$, $\to_\mathrm{u}$, $\to_\mathrm{m}$
-and $\to_\mathrm{d}$ is \emph{normal}. In Section~\ref{sec:nature} we
-demonstrate that naturally-occurring ADTs are normal. Defining the relation
-$\succeq$ as the reflexive and transitive closure of the above relations,
+We say an ADT whose histories are closed under $\to_\mathrm{r}$,
+$\to_\mathrm{u}$, $\to_\mathrm{m}$ and $\to_\mathrm{d}$ is \emph{normal}. In
+Section~\ref{sec:nature} we demonstrate that naturally-occurring ADTs are
+normal. Defining the relation $\succeq$ as the reflexive and transitive closure
+of the above relations,
 \begin{align*}
@@
 
-\begin{lemma}\label{lem:ADT_closure}
+\begin{lemma}
+  \label{lem:ADT_closure}
 
@@
 
-Besides this closure property, the $\preceq$ relation enjoys a certain notion
-of well-foundedness (when restricted to a certain class of histories): every potentially 
-infinite set of width-bounded histories has a finite set of minimal elements w.r.t. $\preceq$.
+Besides this closure property, the $\preceq$ relation enjoys a certain notion of
+well-foundedness when restricted to bounded-width histories: the set of
+$\preceq$-minimal elements of any (potentially infinite) history set is finite.
+This property is what enables us to represent infinite sets of invalid ADT
+histories with a finite set of minimal examples. This property is captured
+formally with wqos: a \emph{well-quasi-ordering (wqo)} $R$ on a set $X$ is a
+reflexive, transitive binary relation on $X$ for which in every infinite
+sequence $x_0 x_1 \ldots$ of elements from $X$, there exists indices $i < j$
+such that $R(x_i,x_j)$.
 
-\begin{lemma}\label{lem:wqo}
+\begin{example}
 
-  Sets of width-bounded histories have a finite set of $\preceq$-minimals.
+  Consider the infinite history sequence $h_1 h_2 \ldots$ where each $h_i$
+  contains $2i$ operations $o_1, o_1', \ldots, o_i, o_i'$ where each $o_j$ is a
+  completed {\tt push} operation matching itself, and each $o_j'$ is a pending
+  {\tt pop} operation with undefined matching. Because each successive $h_i$
+  has both more matches and more pending operations, no two histories of the
+  sequence are related by $\preceq$. Thus $\preceq$ is not a wqo.
+
+\end{example}
+
+The previous example demonstrates that $\preceq$ is not a wqo by allowing each
+history $h_i$ of the infinite sequence to contain more and more pending
+operations in order to ensure that $h_j \not\preceq h_i$ for every $j < i$.
+Curbing this ability by limiting the maximum amount of pending operations per
+history makes $\preceq$ a wqo. Although limiting to $k$ pending operations
+essentially limits us to width-$k$ histories, of executions with at most $k$
+operations parallel at any moment, e.g.,~of programs with at most $k$ threads,
+this restriction comes at no loss of completeness when considering only the
+histories of bounded-width ADTs, which is the subject of the remainder
+of this section.
+
+\begin{lemma}
+  \label{lem:wqo}
+
+  $\preceq$ is a wqo on width-bounded histories.
 
@@
 
-For the remainder of this section, we fix an ADT $A$, and let $B$ be the bounded
-complement of $\ker A$. When $\ker A$ has bounded width, so does $B$, and thus
-$B$ has a finite set of $\preceq$-minimals. 
-Furthermore, when $A$ normal, it is closed under
-$\succeq$, and thus $B$ is closed under 
-$\preceq$~\footnote{Actually, $B$ is closed under $\preceq$ when restricted to width-bounded histories, i.e., if $h_1 \preceq h_2$, $h_1 \in B$, and $h_2$ is width-bounded, then $h_2\in B$.}.
-Closure under a relation satisfying Lemma~\ref{lem:wqo} implies
-representation by a finite set. Formally, we say a set $X$ is \emph{finitely
-representable} if there exists a finite set $Y$ and a relation $R \subseteq Y
-\times X$ such that $X = \set{ x : \exists y\in Y.\ R(y,x) }$. In our case, we
-obtain a finite set from which exactly the elements of $B$ are related by
-$\preceq$.
+For the remainder of this section, we fix an ADT $A$, and let $B$ be its bounded
+complement. When $A$ has bounded width, so does $B$, and thus $\preceq$ is a wqo
+on $B$. Furthermore, when $A$ normal, it is closed under $\succeq$, and thus $B$
+is closed under $\preceq$~\footnote{Actually, $B$ is closed under $\preceq$ when
+restricted to width-bounded histories, i.e., if $h_1 \preceq h_2$, $h_1 \in B$,
+and $h_2$ is width-bounded, then $h_2\in B$.}. Closure under a relation
+satisfying Lemma~\ref{lem:wqo} implies representation by a finite set. Formally,
+we say a set $X$ is \emph{finitely representable} if there exists a finite set
+$Y$ and a relation $R \subseteq Y \times X$ such that $X = \set{ x : \exists
+y\in Y.\ R(y,x) }$. In our case, we obtain a finite set from which exactly the
+elements of $B$ are related by $\preceq$.
 

### 2015-07-07 draft.tex
Fixed space hacks.
Updated to POPL format.
Adding motivation section.
Updated the title with “Symbolic”.
starting top-down presentation
Abstract algorithm.
restructuring the technical part
Updated the abstract.

@@
-\documentclass[10pt,preprint,nocopyrightspace,natbib,authoryear]{sigplanconf}
+\documentclass[9pt,preprint,nocopyrightspace,natbib,authoryear]{sigplanconf}
 
@@
 \usepackage{amsthm}
+\usepackage{listings}
+
+\lstset{
+  basicstyle=\ttfamily,
+  mathescape
+}
 
 \newtheorem{theorem}{Theorem}
+\newtheorem{remark}{Remark}
 \newtheorem{lemma}{Lemma}
@@
 \renewcommand{\implies}{\Rightarrow}
+
+\makeatletter
+\renewcommand{\verbatim@font}{\normalfont\ttfamily\scriptsize}
+\makeatother
 
@@
 
-\title{Abstract Data Type Inference}
+\title{Symbolic Abstract Data Type Inference}
 % \subtitle{}
@@
 
-  Formal specifications are a vital component to the automated verification of
+  Formal specifications are a vital component to the scalable verification of
   software systems. In the case of efficient implementations of concurrent
-  objects like atomic registers, queues, and locks, checking adherence to their
-  abstract data types (ADTs) is a difficult problem, yet can be made tractable
-  when the ADT specifications are given by symbolic, logical representations.
-  Obtaining these symbolic representations, however, is a complex task
-  requiring rare expertise. In practice, concurrent objects are specified
-  indirectly by imperative reference implementations.
+  objects like atomic registers, queues, and locks, symbolic representations
+  of their abstract data types (ADTs) enable efficient modular reasoning,
+  decoupling clients from implementations. Writing adequate formal
+  specifications, however, is a complex task requiring rare expertise. In
+  practice, concurrent objects are specified indirectly by reference
+  implementations.
 
@@
 \input intro
+\input motivation
 \input histories
@@
 \input formula
+\input algorithm
 \input matching
@@
 
+\input appendix
+
 % % The bibliography should be embedded for final submission.

### 2015-07-08 draft.tex
Convincing intro.

@@
 
-  Formal specifications are a vital component to the scalable verification of
+  Formal specification is a vital ingredient to scalable verification of
   software systems. In the case of efficient implementations of concurrent
-  objects like atomic registers, queues, and locks, symbolic representations
-  of their abstract data types (ADTs) enable efficient modular reasoning,
-  decoupling clients from implementations. Writing adequate formal
+  objects like atomic registers, queues, and locks, symbolic formal
+  representations of their abstract data types (ADTs) enable efficient modular
+  reasoning, decoupling clients from implementations. Writing adequate formal
   specifications, however, is a complex task requiring rare expertise. In
-  practice, concurrent objects are specified indirectly by reference
-  implementations.
+  practice, programmers write reference implementations as informal
+  specifications.
 

### 2015-07-08 intro.tex
Convincing intro.

@@
 
-Many modern software systems rely on efficient implementations of abstract data
-types (ADTs) like atomic collections, semaphores, and locks. Efficient
-implementation is difficult though, because efficiency entails maximizing
-concurrency, and concurrency jeopardizes atomicity. Consequently,
-implementations are vulnerable to subtle and elusive bugs compromising
-adherence to their ADTs~\cite{journals/cacm/staff11}, thus necessitating
-automated verification. Adherence of an implementation to an ADT is captured
-formally by a property called \emph{observational refinement}: any observable
-behavior over the method call and return actions of the implementation is also
-admitted by the the ADT.
+Effective scalable reasoning about nontrivial software implementations generally
+requires considering each software module separately, in isolation, using
+abstract specifications for other modules. When modules are objects whose
+methods may be called concurrently, their behavior is typically understood in
+terms of invocation sequences of abstract data types (ADTs): an execution with
+overlapping method invocations is considered valid when those invocations can be
+\emph{linearized} into a sequence admitted by the
+ADT~\cite{journals/toplas/HerlihyW90}. For example, consider the execution
+history depicted in Figure~\ref{fig:clients} in which the add operations
+numbered $2$ and $3$ overlap with each other, and, respectively, with operations
+$1$ and $4$. This execution is valid with respect to the atomic queue ADT
+because among the five possible ways to linearize its six operations, the
+sequence~$1$, $3$, $2$, $4$, $5$, $6$ is admitted. ADT specifications thus
+decouple reasoning about object implementations from their clients’ invocations:
+\begin{itemize}
 
-Although checking observational refinement is a fundamentally hard
-problem\footnote{See Section~\ref{sec:related} for further discussion.} for
-arbitrary ADTs~\cite{journals/siamcomp/GibbonsK97, conf/esop/BouajjaniEEH13},
-recent work demonstrates that implementations of naturally-occurring ADTs like
-atomic collections, semaphores, and locks can be checked efficiently given
-effective symbolic ADT representations~\cite{conf/popl/BouajjaniEEH15,
-journals/arxiv/BouajjaniEEH15, conf/pldi/EmmiEH15}, e.g.,~specified in
-first-order logic. However, writing accurate symbolic specifications is an
-expected challenge for ADT designers. In practice, ADTs are specified
-indirectly as the observable behaviors of possibly-abstract imperative
-\emph{reference implementations}. Mapping these reference implementations to
-logical formulæ describing the same sets of behaviors is a complex task
-requiring rare expertise.
+  \item Does every implementation execution have a valid linearization?
+
+  \item Does every valid linearizations preserve client invariants?
+
+\end{itemize}
+The former question depends only on a given object’s implementation, and the
+latter only on a given object’s clients.
+
+\begin{figure}[t]
+  \begin{minipage}{0.43\linewidth}
+    \begin{verbatim}
+1: add(a)       #
+2: add(b)       ###
+3: add(c)         ###
+4: add(d)           #
+5: remove => a        #
+6: remove => c          #
+    \end{verbatim}
+  \end{minipage}
+  \hfill
+  \begin{minipage}{0.55\linewidth}
+    \begin{verbatim}
+assume w < x
+(  (q.add(<1,w>); q.add(<1,x>))
+|| (q.add(<2,w>); q.add(<2,x>)) )
+i, y := q.remove()
+j, z := q.remove()
+assert i == j ==> y < z
+    \end{verbatim}
+  \end{minipage}
+  \caption{An execution history with six numbered operations (left),
+    and a parallel program invoking six operations (right). The “\#” symbols
+    depict the time intervals spanned by operations horizontally.}
+  \label{fig:clients}
+\end{figure}
+
+Although formal ADT specifications are indispensable for scalable program
+reasoning, formal-specification writing is a burden for which few programmers
+posses the required combination of expertise and willingness to overcome.
+Typically programmers write simple ADT \emph{reference implementations},
+e.g.,~whose methods are synchronized via a global lock, and refine them with
+more efficient fine-grained implementations, e.g.,~reducing synchronization
+bottlenecks using specialized hardware instructions such as atomic compare and
+swap (CAS).
+
+Our goal in this work is thus the automated generation of formal ADT
+specifications, derived from reference implementations, which are suitable for
+automated reasoning. In particular, we aim to generate \emph{symbolic
+representations} of valid invocation sequences for ADTs which are given
+implicitly by reference implementations. We target declarative symbolic
+representations rather than imperative state-based representations in order to
+harness efficient symbolic reasoning algorithms: rather than enumerating
+linearizations explicitly, and checking their validity one by one, a symbolic
+reasoning engine may simultaneously rule out many possible linearizations.
+Previous work demonstrates that such symbolic reasoning can increase
+efficiency by orders of magnitude~\cite{conf/pldi/EmmiEH15}.
 
@@
 generated automatically from the executions of reference implementations,
-enabling the application of efficient symbolic refinement-checking algorithms
+enabling the application of efficient symbolic reasoning algorithms
 without the burden of writing formal specifications manually. Our approach
@@
 hypothesis that violation patterns manifest in executions with few operations,
-we propose an under-approximating algorithm which extracts the patterns
-observed in all violating executions up to a given number of operations. In
-theory, for an arbitrary ADT, this is clearly incomplete: any violation which
-only surfaces with a greater number of operations would not be captured, thus
-resulting in a symbolic ADT representation which can fail to identify
-violations — though still guarantees never to classify a valid execution as a
-violation. Empirically, however, we demonstrate that our hypothesis holds: the
-patterns emerging from executions with few operations are complete in
+we propose an under-approximating algorithm which extracts the patterns observed
+in all violating executions up to a given number of operations. In theory, for
+an arbitrary ADT, this is clearly incomplete: any violation which only surfaces
+with a greater number of operations would not be captured, thus resulting in a
+symbolic ADT representation which can fail to identify violations — though still
+guarantees never to classify a valid execution as a violation, and is thus sound
+for program reasoning. Empirically, however, we demonstrate that our hypothesis
+holds: the patterns emerging from executions with few operations are complete in
 characterizing all violations of naturally-occurring concurrent object ADTs,
@@
 object executions with a matching relation, and we demonstrate that these
-relations can be given for naturally-occurring ADTs, we also provide an
+relations are easy to provide for naturally-occurring ADTs, we also outline an
 automatic means for computing such \emph{matching schemes}. Again by sampling
@@
 
-  \item The statement of the ADT inference problem (§\ref{sec:inference}).
+  \item The statement of the symbolic ADT inference problem (§\ref{sec:inference}).
 
-  \item Identification of the algebraic properties allowing finite
-  characterization and computability of infinite ADT violations sets
-  (§\ref{sec:patterns}).
+  \item Identification of the algebraic properties allowing a finite
+  characterization of infinite ADT violation sets (§\ref{sec:patterns}).
 
-  \item The computability of symbolic ADT representations (§\ref{sec:formula}).
+  \item The symbolic representation of ADT violations (§\ref{sec:formula}).
+
+  \item The computability of symbolic ADT representations (§\ref{sec:algorithm}).
 
@@
 To the best of our knowledge, this work is the first to suggest the automatic
-generation of symbolic ADT representations for concurrent objects. By removing
-the burden of writing formal ADT specifications manually, this work broadens
-the scope of efficient symbolic refinement-checking algorithms to
-newly-designed ADTs, apart from those few traditionally studied in the
-literature.
+generation of ADT specifications. By removing the burden of writing formal ADT
+specifications manually, this work broadens the scope of modular program
+reasoning using efficient symbolic algorithms to newly-designed ADTs, apart from
+those few traditionally studied in the literature.

### 2015-07-08 motivation.tex
Comment on the possible queue violations.
Convincing intro.

@@
 \label{sec:motivation}
-
-Effective scalable reasoning about nontrivial software implementations generally
-requires considering each software module separately, in isolation, using
-abstract specifications for other modules. When modules are objects whose
-methods may be called concurrently, their behavior is typically understood in
-terms of valid invocation sequences of atomic abstract data types (ADTs): any
-execution of overlapping method invocations can be \emph{linearized} into an
-invocation sequence admitted by the ADT~\cite{journals/toplas/HerlihyW90}. For
-example, consider the execution history depicted in Figure~\ref{fig:clients} in
-which the add operations numbered $2$ and $3$ overlap with each other, and,
-respectively, with operations $1$ and $4$. Among the five possible ways to
-linearize these six operations, the linearization~$1$; $3$; $2$; $4$; $5$; $6$
-does correspond to a valid sequence of the atomic queue ADT, and so this history
-is linearizable. ADT specifications thus decouple reasoning about object
-implementations from their client’s invocations:
-\begin{itemize}
-
-  \item Does every implementation execution have a valid linearization?
-
-  \item Does every valid linearizations preserve client invariants?
-
-\end{itemize}
-The former question depends only on a given object’s implementation, and the
-latter only on a given object’s clients.
-
-\begin{figure}[t]
-  \begin{minipage}{0.43\linewidth}
-    \begin{verbatim}
-1: add(a)       #
-2: add(b)       ###
-3: add(c)         ###
-4: add(d)           #
-5: remove => a        #
-6: remove => c          #
-    \end{verbatim}
-  \end{minipage}
-  \hfill
-  \begin{minipage}{0.55\linewidth}
-    \begin{verbatim}
-assume w < x
-(  (q.add(<1,w>); q.add(<1,x>))
-|| (q.add(<2,w>); q.add(<2,x>)) )
-i, y := q.remove()
-j, z := q.remove()
-assert i == j ==> y < z
-    \end{verbatim}
-  \end{minipage}
-  \caption{An execution history with six numbered operations (left),
-    and a parallel program invoking six operations (right). The “\#” symbols
-    depict the time intervals spanned by operations horizontally.}
-  \label{fig:clients}
-\end{figure}
 
@@
 
-Although formal ADT specifications are indispensable for scalable program
-reasoning, formal-specification writing is a burden for which few programmers
-posses the required combination of expertise and willingness to overcome.
-Typically programmers write simple ADT \emph{reference implementations},
-e.g.,~whose methods are synchronized via a global lock, and refine them with
-more efficient fine-grained implementations, e.g.,~reducing synchronization
-bottlenecks using specialized hardware instructions such as atomic compare and
-swap (CAS).
-
-Our goal in this work is thus the automated generation of formal ADT
-specifications, derived from reference implementations, which are suitable for
-automated reasoning. In particular, we aim to generate \emph{symbolic
-representations} of the valid history linearizations, according to ADTs given
-implicitly by reference implementations. We target declarative symbolic
-representations rather than imperative state-based representations in order to
-harness efficient symbolic reasoning algorithms: rather than enumerating
-linearizations explicitly, and checking their validity one by one, a symbolic
-reasoning engine may simultaneously rule out many possible linearizations,
-e.g.,~as in the example above. Previous work demonstrates that such symbolic
-reasoning can increase reasoning efficiency by orders of
-magnitude~\cite{conf/pldi/EmmiEH15}.
+TODO SEGUE
 
@@
 values for all invocations --- are discarded. The remaining $171$ sequences are
-invalid, $164$ of which are redundant with the seven {\tt PATTERNS} listed on the
-left-hand side of Figure~\ref{fig:patterns}. Seven of these redundant sequences
-are shown on the right-hand side of Figure~\ref{fig:patterns}. For example, the
-first five sequences on the right, as well as the last, are redundant with the
-first on the left since they each describe a violation in which a removed
-element was never added. The sixth sequence on the right is redundant with the
-second on the left since they both describe a violation in which remove returns
-empty before previously-added elements are removed. Finally, the exclusion of
-the computed pattern set can be expressed as a conjunction of formulas
-in first order logic describing the exclusion of each individual pattern. For
-example, the first pattern could be described by the formula
+invalid, $164$ of which are redundant with the seven {\tt PATTERNS} listed on
+the left-hand side of Figure~\ref{fig:patterns}. Seven of these redundant
+sequences are shown on the right-hand side of Figure~\ref{fig:patterns}. For
+example, the first five sequences on the right, as well as the last, are
+redundant with the first on the left since they each describe a violation in
+which a removed element was never added. The sixth sequence on the right is
+redundant with the second on the left since they both describe a violation in
+which remove returns empty before previously-added elements are removed. Other
+represented violations include removing an element before it is added, and
+removing elements in the opposite order in which they were added. Finally, the
+exclusion of the computed pattern set can be expressed as a conjunction of
+formulas in first order logic describing the exclusion of each individual
+pattern. For example, the first pattern could be described by the formula
 \begin{align*}

### 2015-07-08 histories.tex
Better overview.

@@
 
-Implementations of abstract data types (ADTs) provide methods which are called,
-possibly concurrently, by client programs. We capture the possible logs of
-method calls and returns in executions of client programs abstractly, in order
-to facilitate the development of symbolic ADT representations.
+Abstract data types (ADT) implementations provide methods which can be invoked
+concurrently by threads of client programs. We capture the possible histories of
+call and return actions in the executions of client programs as
+partially-ordered method invocations. While many ADTs are “atomic”
+(e.g.,~queues, locks) in the sense that their ADTs are specified as sets of
+sequential histories, many (e.g.,~rendezvous synchronizers, barriers) are
+non-atomic~\cite{conf/podc/HemedR14}. While much of the following development
+could be simplified for atomic ADTs by considering invocation sequences rather
+than concurrent invocation histories, we maintain generality so that our results
+apply to non-atomic ADTs as well.
 
@@
 \end{itemize}
-The relation $<$ is an interval order because histories abstract executions where
-the set of call and return actions is totally ordered~\cite{conf/popl/BouajjaniEEH15}.
-Non-completed operations are \emph{pending}, and are maximal in the
-happens-before order.
+The relation $<$ is an interval order because executions’ call and return
+actions are totally ordered~\cite{conf/popl/BouajjaniEEH15}. Non-completed
+operations are \emph{pending}, and are maximal in the happens-before order.
 

### 2015-07-08 intro.tex
Better overview.

@@
 \end{figure}
+
+\begin{example}
+
+  Consider the parallel program in Figure~\ref{fig:clients} invoking the add and
+  remove methods of an atomic queue implementation, adding increasing integer
+  values $w$ and $x$ tagged with the integers $\set{1,2}$ indicating on which
+  parallel branch each add occurs. Intuitively this program is correct since
+  values with the same tag are added in increasing order, and, crucially, the
+  values of the queue ADT are removed in the same order in which they are added.
+  Among the six possible ways to linearize these operations, the comparison {\tt i ==
+  j} of tags only holds for those two beginning with
+  \begin{quote}
+    \verb|q.add(<1,w>); q.add(<1,x>) | and \verb| q.add(<2,w>); q.add(<2,x>)|
+  \end{quote}
+  Since the queue ADT dictates that elements are removed in the order added, we
+  conclude that $w$ and $x$ are removed in order when {\tt i == j}, and thus {\tt
+  y < z} holds when {\tt i == j} holds.
+
+\end{example}
 

### 2015-07-08 overview.tex
Better overview.

--- 2015-07-08 motivation.tex
@@
-\section{Motivation \& Overview}
-\label{sec:motivation}
-
-Consider the parallel program in Figure~\ref{fig:clients} invoking the add and
-remove methods of an atomic queue implementation, adding increasing integer
-values $w$ and $x$ tagged with the integers $\set{1,2}$ indicating on which
-parallel branch each add occurs. Intuitively this program is correct since
-values with the same tag are added in increasing order, and, crucially, the
-values of the queue ADT are removed in the same order in which they are added.
-Among the six possible ways to linearize these operations, the comparison {\tt i ==
-j} of tags only holds for those two beginning with
-\begin{quote}
-  \verb|q.add(<1,w>); q.add(<1,x>) | and \verb| q.add(<2,w>); q.add(<2,x>)|
-\end{quote}
-Since the queue ADT dictates that elements are removed in the order added, we
-conclude that $w$ and $x$ are removed in order when {\tt i == j}, and thus {\tt
-y < z} holds when {\tt i == j} holds.
-
-TODO SEGUE
+\section{Overview of an ADT Inference Algorithm}
+\label{sec:overview}
 
@@
 empty. The $31$ sequences which are executable by a correct reference
-implementation of an atomic queue implementation --- i.e.,~with matching return
-values for all invocations --- are discarded. The remaining $171$ sequences are
+implementation of an atomic queue — i.e.,~with consistent return
+values for each invocation — are discarded. The remaining $171$ sequences are
 invalid, $164$ of which are redundant with the seven {\tt PATTERNS} listed on
@@
 \end{align*}
-describing an unmatched remove operation.
+describing an unmatched remove operation. The formula which excludes all
+violation patterns is satisfied by an invocation sequence if and only if that
+sequence is admitted by the given reference implementation’s implicit ADT.
 
@@
   to the current line’s operation identifier $i$, and the identifier $j$ of
-  its “match-source” operation. The “{\tt X}” symbol denotes an absent match.}
+  its matching operation. Adds and empty removes match themselves, and the
+  “{\tt X}” symbol denotes an absent match.}
   \label{fig:patterns}
@@
 Section~\ref{sec:algorithm} are both sound and complete for the
-naturally-occurring ADTs of interest.
+naturally-occurring ADTs of interest. The remainder of this article develops
+the technical machinery required to realize this abstract inference algorithm.

### 2015-07-08 algorithm.tex
Added back the section label that Constantin deleted. * He is so clumsy with version control!
Section titles + restructuring the restructuring.
Tried to clean up the f**king mess of the algorithm section.

--- 2015-07-07 algorithm.tex
@@
 %!TEX root = draft.tex
-\section{Solving the Symbolic ADT Inference Problem}
+\section{A Symbolic ADT Inference Algorithm}
+\label{sec:algorithm}
 
-\begin{figure}
-\begin{lstlisting}
-Min := $\emptyset$
-while (true) {
-  $B_i$ := $ \set{ h \in B : \weight(h) = i }$;
-  MinTemp := Min $\cup$ Min($B_i$)
-  if ( MinTemp = Min )
-    break;
-  Min := MinTemp
+Algorithm~\ref{alg:inference} lists the algorithm for solving the symbolic ADT
+inference problem which is based on the relation $\preceq$ defined in
+Section~\ref{sec:patterns}. The algorithm performs an weight-increasing
+enumeration of $B$ while maintaining the $\preceq$-minimal elements, and stops
+whenever no new minimals are found at a given weight.
+
+\begin{algorithm}[t]
+  \SetAlgoLined
+  \SetKwInOut{Input}{Input}
+  \Input{A reference implementation $\mathcal{I}$}
+  \KwResult{A formula representing the ADT of $\mathcal{I}$}
+  weight $\gets$ $0$ \;
+  patterns $\gets$ $\emptyset$ \;
+  \Repeat{$B_\text{weight} = B_{\text{weight}-1}$}{
+    weight $\gets$ weight + $1$ \;
+    patterns $\gets$ $\min B_\text{weight}$ \;
   }
-return $F$(Min);
-\end{lstlisting}
-\caption{The algorithm solving the symbolic ADT inference problem for a given ADT $A$, where $B$ denotes the bounded complement of $\ker A$.
-The procedure $\texttt{Min}(X)$ returns the $\preceq$-minimal elements of the set of histories $X$ and $\texttt{Formula}(X)$ returns a history formula which holds exactly for the histories in the upward closure of $X$.}
-\label{fig:algorithm}
-\end{figure}
+  \Return $F(\text{patterns})$
+  \caption{Symbolic ADT inference.}
+  \label{alg:inference}
+\end{algorithm}
 
-Figure~\ref{fig:algorithm} lists the algorithm for solving the symbolic ADT inference problem which is based on the relation $\preceq$ 
-defined in Section~\ref{sec:patterns}.
-%For a given set of elements $X$ and an order relation $\preceq$, the \emph{upward closure}
-%of $X$ w.r.t. $\preceq$ is the set $\uparrow X=\set{y : \exists x\in X.\ x\preceq y}$. 
-The algorithm performs an weight-increasing enumeration of $B$ while maintaining the
-$\preceq$-minimal elements, and stops whenever each history of weight $i$ is bigger (w.r.t. $\preceq$) than some
-element of the collected set of minimals.
+The partial correctness of Algorithm~\ref{alg:inference} (i.e.,~ignoring
+termination) relies on the input ADT being \emph{predictable}, i.e.,~for all
+$i \in \mathbb{N}$,
+\begin{align*}
+  B_{i+1} \succeq \min \bigcup_{0 < j \le i} b_j
+  \text{ implies } \min B = \min \bigcup_{0 < j \le i} b_j
+\end{align*}
+Section~\ref{sec:nature} shows that many naturally-occurring ADTs are predictable.
 
-The partial correctness of the algorithm (i.e., ignoring termination) relies on the input ADT being \emph{predictable}, i.e.,
-for all $i$,
-\begin{align*}
-B_{i+1} \succeq \texttt{Min}(B_{\leq i}) \mbox{ implies } \texttt{Min}(B) = \texttt{Min}(B_{\leq i}),
-\end{align*}
-where $B_{\leq i}=\bigcup_{1\leq j\leq i}B_j$.
-Section~\ref{sec:nature} shows that many naturally-occurring ADTs are predictable.
-%Although we don't provide a general result concerning predictability, we prove that it holds for  (see ).
+\begin{theorem}
+  \label{thm:algorithm}
 
-\begin{theorem}\label{th:corr1}
-For any predictable ADT $A$, if the algorithm in Figure~\ref{fig:algorithm} terminates, then the returned formula represents $A$.
+  Algorithm~\ref{alg:inference} terminates. If the input implementation’s
+  ADT is predictable, then the returned formula represents it.
+
 \end{theorem}
+
 \begin{proof}
-%For a set $\Sigma$, let $\neg \Sigma$ denote its complement.
-Let $M$ be the value of \texttt{Min} at the end of the \texttt{while} loop.
-%By Lemma~\ref{lem:ADT_closure}, the complement of $A$ is closed under $\preceq$. Since 
-%$M\subseteq B\subseteq  \neg \ker A \subseteq \neg A$, the closure of $M$, denoted by $Cl(M)$, satisfies $Cl(M)\subseteq \neg A$. 
-Since $A$ is predictable, we have that $B \subseteq Cl(M)$. % where $Cl(M)$ denotes the closure of 
-Therefore, by Lemma~\ref{lem:formula1}, the returned formula represents $A$.
+  Termination is a direct consequence of Lemma~\ref{lem:wqo}, and since the
+  number of histories of any weight $i \in \mathbb{N}$ is finite\footnote{
+    As noted in Section~\ref{sec:histories}, we consider equality between
+    histories up to renaming of operation identifiers.
+  }
+  each $B_i$ is computable.
+
+  Furthermore, when $A$ is predictable, we
+  have $B \subseteq \text{patterns}^+$, where $\text{patterns}^+$ is the
+  upward closure of $\text{patterns}$ under $\preceq$. Thus by
+  Lemma~\ref{lem:formula1}, the returned formula represents $A$.
 \end{proof}
 
-\begin{remark}
-Note that for non-predictable ADTs, the value of $\texttt{Min}$ at the end of the \texttt{while} loop 
-is an under-approximation of $\texttt{Min}(B)$. 
-The resulting symbolic ADT representation is still satisfied by all histories in $\ker A$, therefore it would still be sound in only identifying
-\emph{actual} violations, yet incomplete in identifying \emph{all} violations.
-\end{remark}
-
-The termination of the algorithm in Figure~\ref{fig:algorithm} is a direct consequence of Lemma~\ref{lem:wqo}.
-Also, since the number of weight-$i$ histories is finite\footnote{As
-noted in Section~\ref{sec:histories}, we consider equality between histories up
-to renaming of operation identifiers.} for each $i \in \mathbb{N}$, each $B_i$
-is computable.
-
-\begin{theorem}\label{th:corr2}
-The algorithm in Figure~\ref{fig:algorithm} terminates.
-\end{theorem}
+Note that for non-predictable ADTs, the value of $\text{patterns}^+$ upon
+termination of Algorithm~\ref{alg:inference} is an under-approximation of the
+bounded ADT complement $B$. The resulting symbolic ADT representation is still
+satisfied by all histories in $A$, therefore it would still be sound for modular
+program reasoning, identifying only actual violations, and implying the
+correctness of client programs which do not depend on the stronger criteria
+which excludes unidentified violations. However, the under-approximation may be
+incomplete in identifying all violations, and in proving client programs which
+depend on the stronger criteria which excludes them all.

### 2015-07-08 inference.tex
Section titles + restructuring the restructuring.
Better overview.
Tried to clean up the f**king mess of the algorithm section.

@@
 
-A \emph{kernel} of a set $H$ is a minimal set $K$ generating $H$ via the
-relation $\mathord\to = (\to_\mathrm{o} \cup \to_\mathrm{p} \cup
-\to_\mathrm{c})$, i.e.,
+Let $H^*$ denote the \emph{closure} of a history set $H$ under the relation
+$\mathord\to = (\to_\mathrm{o} \cup \to_\mathrm{p} \cup \to_\mathrm{c})$, i.e.,
 \begin{align*}
-  H = \set{ h' : \exists h \in K.\ h \to^\ast h' }.
+  H^* = \set{ h' : \exists h \in H.\ h \to^\ast h' }.
 \end{align*}
-An \emph{abstract data type (ADT)} $A$ is the kernel of the set $H(\mathcal{I})$
-of histories of some implementation $\mathcal{I}$, and $H(A)$ denotes the
-closure of $A$ under $\to$. While the kernels of arbitrary-width history sets
-may not be unique, those with \emph{atomic} kernels, i.e.,~of width-$1$, are
-guaranteed to be unique. Section~\ref{sec:nature} demonstrates that the
-histories of naturally-occurring implementations have bounded-width and
-unique ADTs, which we assume for the remainder of this work.
+A set $K$ \emph{generates} $H$ when $K^* = H$, and a \emph{kernel} of a set $H$
+is a minimal set $K$ which generates $H$. While the kernels of arbitrary sets of
+histories need not be unique, the kernels of sets which have sequential kernels
+are unique. Furthermore, Section~\ref{sec:nature} demonstrates that the
+histories of naturally-occurring implementations have unique kernels, which we
+assume for the remainder of this work. An \emph{abstract data type (ADT)} $A$ is
+the kernel of the set $H(\mathcal{I})$  of histories of some implementation
+$\mathcal{I}$. We say that \emph{$A$ is the ADT of $\mathcal{I}$} when $A^* =
+H(\mathcal{I})$.
 
-We define \emph{observational refinement} between implementations and ADTs as
-history-set inclusion. Although this refinement is typically defined with
-respect to the admissibility of program executions, recent work demonstrates
-that these definitions are equivalent~\cite{conf/popl/BouajjaniEEH15}.
+ADTs and reference implementations serve as specifications to more efficient
+implementations in the sense that they limit the set of histories that efficient
+implementation may admit. This notion of refinement ensures that client program
+(safety) properties which hold using the ADT or reference implementation also
+hold using refined implementations~\cite{conf/popl/BouajjaniEEH15}.
 
@@
   $\mathcal{I}_2$ when $H(\mathcal{I}_1) \subseteq H(\mathcal{I}_2)$. An
-  implementation $\mathcal{I}$ refines an ADT $A$ when $\mathcal{I}$ refines
-  $H(A)$.
+  implementation $\mathcal{I}$ refines an ADT $A$ when $H(\mathcal{I})
+  \subseteq A^*$.
 
@@
 
-Recent works demonstrate efficient algorithms\footnote{In time polynomial in
-the number of operations, per execution.} for checking observational
-refinement~\cite{conf/popl/BouajjaniEEH15, conf/pldi/EmmiEH15}, yet rely on
+Recent works demonstrate efficient\footnote{In time polynomial in the number of
+operations, per execution.} refinement-checking
+algorithms~\cite{conf/popl/BouajjaniEEH15, conf/pldi/EmmiEH15} yet rely on
 hand-written symbolic ADT representations. In order to frame the problem of
@@
   The \emph{symbolic ADT inference problem} is to compute a history formula
-  representing the ADT $\ker H(\mathcal{I})$ of a given (reference) implementation
-  $\mathcal{I}$.
+  representing the ADT of a given implementation.
 
 \end{definition}
+
+Computing a history formula representing the ADT of a reference implementation
+$\mathcal{I}$ thus enables efficient modular program reasoning without the
+burden of writing precise formal specifications for $\mathcal{I}$.

### 2015-07-09 discussion.tex
some final typos
Removed kernel talk.

@@
 specifications~\cite{conf/podc/HemedR14}, i.e.,~non-atomic ADTs like the
-\emph{rendezvous synchronizer}, \emph{barrier}, and \emph{exchanger}, whose
-kernels have bounded width greater than $1$, our current implementation only
+\emph{rendezvous synchronizer}, \emph{barrier}, and \emph{exchanger}, which
+have bounded width greater than $1$, our current implementation only
 handles atomic ADTs. In principle, this limitation is not fundamental. The key
-difference is in the enumeration of kernel histories. For a given ADT,
-determining whether a given history is in the kernel or its complement reduces
+difference is in the enumeration ADT histories. For a given ADT,
+determining whether a given history is admitted or not reduces
 to checking whether there exists an execution of which the history is an
@@
 \emph{priority queue}, whose \emph{dequeue} operations return the
-smallest/largest \emph{enqueue}d value which has not yet been dequeued. It
-appears that such ADTs are rare in practice. Overcoming this limitation would
+smallest/largest \emph{enqueue}d value which has not yet been dequeued.
+%It appears that such ADTs are rare in practice.
+Overcoming this limitation would
 require a refinement to matching schemes which can track additional relations

### 2015-07-09 formula.tex
Removed kernel talk.

@@
 
-While Section~\ref{sec:patterns} demonstrates that (the kernel complements) of
+While Section~\ref{sec:patterns} demonstrates that (the complements) of
 naturally-occurring ADTs can be represented finitely, in this section we

### 2015-07-09 nature.tex
Fixed space hacks.
Section titles + restructuring the restructuring.
Removed kernel talk.

@@
 %!TEX root = draft.tex
-\section{Properties of Naturally-Occurring ADTs}
+\section{Naturally-Occurring ADTs}
 \label{sec:nature}
 
-\makeatletter
-\renewcommand{\verbatim@font}{\ttfamily\scriptsize}
-\makeatother
-
 In this section we demonstrate that the premises used in the development of our
-symbolic ADT inference algorithm — i.e.,~unique and bounded-width kernels,
+symbolic ADT inference algorithm — i.e.,~uniqueness and bounded-width,
 faithful and normalizing matching schemes, and predictability — hold for the
@@
 \end{itemize}
-As its name implies, its kernel is sequential.
+As its name implies, the atomic register ADT is sequential.
 
@@
 of atomic registers do not contain unmatched operations, they are also closed
-under their removal. Finally, since the kernel contains an arbitrary number of
-sequenced matches, atomic register histories are also closed under match
+under their removal. Finally, since the sequences contain an arbitrary number of
+matches, atomic register histories are also closed under match
 removal.
@@
 \end{itemize}
-As their names imply, their kernels are sequential.
+As their names imply, these ADTs are sequential.
 
@@
 Each operation returns the nil value $v = -$ when $u$ is not yet
-present, and otherwise returns $v = u$. As its name implies, its kernel is
-sequential.
+present, and otherwise returns $v = u$. As its name implies, the atomic set
+ADT is sequential.
 
@@
 operation which added their returned value, or themselves, in case the nil
-value is returned. Note that the kernel of work-stealing queues has width $2$,
+value is returned. Note that the work-stealing queue ADT has width $2$,
 since a pairs of concurrent take and steal operations returning the same value

### 2015-07-09 related.tex
Removed kernel talk.

@@
 suggest patterns expressed as \emph{finite} languages, which recognize
-violations in ADT \emph{kernels} — i.e.,~the sequential executions of atomic
+violations in ADTs — i.e.,~the sequential executions of atomic
 objects, or the “concurrency-aware” executions of “concurrency-aware”

### 2015-07-09 overview.tex
Proper algorithm rendering.

@@
 Our basic approach to inferring ADT specifications, as outlined by the abstract
-algorithm in Figure~\ref{fig:abstract}, is to identify a finite set of
+algorithm in Algorithm~\ref{alg:abstract}, is to identify a finite set of
 sequential execution histories which capture all of the reasons for which a
@@
 
-\begin{figure}[t]
-  \begin{verbatim}
-procedure INFER_ADT_FROM_REFERENCE_IMPL(I)
-  PATTERNS := {}
-  for each sequential history H
-    if * then
+\begin{algorithm}[t]
+  \SetAlgoLined
+  \SetKwInOut{Input}{input}
+  \Input{A reference implementation $\mathcal{I}$}
+  \KwResult{A formula representing the ADT of $\mathcal{I}$}
+  patterns $\gets$ $\emptyset$ \;
+  \For{each sequential history $h$}{
+    \uIf{$*$}{
       break
-    else if H is EXECUTABLE with I then
+    }
+    \uElseIf{$h$ is \emph{executable} with $\mathcal{I}$}{
       continue
-    else if H is REDUNDANT with PATTERNS then
+    }
+    \uElseIf{$h$ is \emph{redundant} with patterns}{
       continue
-    else
-      add H to PATTERNS
-    end
-  end
-  return EXCLUSION of PATTERNS
-end
-  \end{verbatim}
-  \caption{An abstract algorithm for symbolic ADT inference.}
-  \label{fig:abstract}
-\end{figure}
+    }
+    \Else{
+      add $h$ to patterns
+    }
+  }
+  \Return \emph{exclusion} of patterns
+  \caption{Abstract algorithm for symbolic ADT inference.}
+  \label{alg:abstract}
+\end{algorithm}
 
@@
 The key technical obstacle in realizing the abstract algorithm of
-Figure~\ref{fig:abstract}, overcome in Section~\ref{sec:patterns}, is the
+Algorithm~\ref{alg:abstract}, overcome in Section~\ref{sec:patterns}, is the
 classification of violations into a finite set of patterns, rendering the

### 2015-07-09 algorithm.tex
Algorithm cleanup.

@@
 
-Algorithm~\ref{alg:inference} lists the algorithm for solving the symbolic ADT
-inference problem which is based on the relation $\preceq$ defined in
-Section~\ref{sec:patterns}. The algorithm performs an weight-increasing
-enumeration of $B$ while maintaining the $\preceq$-minimal elements, and stops
-whenever no new minimals are found at a given weight.
+Algorithm~\ref{alg:inference} solves the symbolic ADT inference problem by
+computing a finite representation of an ADT complement $B$ using the $\preceq$
+relation of Section~\ref{sec:patterns}. In general, this is achieved by
+enumerating the elements of $B$ while maintaining the $\preceq$-minimal
+elements, and recognizing a condition under which all the elements of $B$ are
+related to the current set of minimals~\cite{conf/lics/AbdullaCJT96,
+journals/tcs/FinkelS01}. In our case, we stratify our enumeration of $B$ by the
+relative sizes of its histories. Formally, we say the \emph{weight} of a history
+is the maximum among operation frequencies and the number of matches. We then
+define
+\begin{align*}
+  & B_i = \set{ h \in B : \weight(h) \le i } \\
+  & B_i' = \set{ h \in B : \exists h' \in B_i.\ h' \preceq h \text{ and } \weight(h) \le i+1 }
+\end{align*}
+respectively as the histories of $B$ with at most $i$ matches and duplicates,
+and those derived from $B_i$ with at most $i+1$ matches and duplicates. We say
+that an ADT with complement $B$ is \emph{predictable} if $B_i^* = B$ whenever
+$B_i' = B_{i+1}$, i.e.,~if all histories of $B$ are represented by $B_i$ whenever
+all histories of $B_{i+1}$ are represented by $B_i$. Algorithm~\ref{alg:inference}
+then performs a weight-increasing enumeration of $B$, collecting
+$\preceq$-minimals from smaller-weight violations before advancing to greater
+weights. When no violation is found at a given weight, the algorithm terminates.
+This algorithm is guaranteed to terminate since $\preceq$ is a
+wqo~\cite{conf/lics/AbdullaCJT96, journals/tcs/FinkelS01}. Furthermore, this
+algorithm is sound for arbitrary ADTs, and complete for predictable ADTs.
+Many naturally-occurring ADTs are predictable — in fact all of the examples we
+know of are predictable, as demonstrated in Section~\ref{sec:nature}.
 
@@
   \SetKwInOut{Input}{Input}
-  \Input{A reference implementation $\mathcal{I}$}
+  \Input{A reference implementation $\mathcal{I}$ of width $k$}
   \KwResult{A formula representing the ADT of $\mathcal{I}$}
-  weight $\gets$ $0$ \;
   patterns $\gets$ $\emptyset$ \;
-  \Repeat{$B_\text{weight} = B_{\text{weight}-1}$}{
-    weight $\gets$ weight + $1$ \;
-    patterns $\gets$ $\min B_\text{weight}$ \;
+  $w$ $\gets$ $0$ \;
+  \Repeat{none-found}{
+    none-found $\gets$ true \;
+    \For{each $k$-width history $h$ with weight $w$}{
+      \uIf{$h$ is \emph{executable} with $\mathcal{I}$}{
+        continue
+      }
+      \uElseIf{$\exists h' \in \text{patterns}.\ h' \preceq h$}{
+        continue
+      }
+      \Else{
+        add $h$ to patterns \;
+        none-found $\gets$ false
+      }
+    }
+    $w$ $\gets$ $w$ + $1$ \;
   }
@@
 \end{algorithm}
-
-The partial correctness of Algorithm~\ref{alg:inference} (i.e.,~ignoring
-termination) relies on the input ADT being \emph{predictable}, i.e.,~for all
-$i \in \mathbb{N}$,
-\begin{align*}
-  B_{i+1} \succeq \min \bigcup_{0 < j \le i} b_j
-  \text{ implies } \min B = \min \bigcup_{0 < j \le i} b_j
-\end{align*}
-Section~\ref{sec:nature} shows that many naturally-occurring ADTs are predictable.
 
@@
   each $B_i$ is computable.
-
-  Furthermore, when $A$ is predictable, we
-  have $B \subseteq \text{patterns}^+$, where $\text{patterns}^+$ is the
-  upward closure of $\text{patterns}$ under $\preceq$. Thus by
-  Lemma~\ref{lem:formula1}, the returned formula represents $A$.
+  When $A$ is predictable, then $B = \text{patterns}^*$, and thus by
+  Lemma~\ref{lem:formula1}, the returned formula $F(\text{patterns})$
+  represents $A$.
 \end{proof}
 
-Note that for non-predictable ADTs, the value of $\text{patterns}^+$ upon
-termination of Algorithm~\ref{alg:inference} is an under-approximation of the
-bounded ADT complement $B$. The resulting symbolic ADT representation is still
-satisfied by all histories in $A$, therefore it would still be sound for modular
-program reasoning, identifying only actual violations, and implying the
-correctness of client programs which do not depend on the stronger criteria
-which excludes unidentified violations. However, the under-approximation may be
-incomplete in identifying all violations, and in proving client programs which
-depend on the stronger criteria which excludes them all.
+For non-predictable ADTs, the value of $\text{patterns}^*$ upon termination of
+Algorithm~\ref{alg:inference} is an under-approximation of the bounded ADT
+complement $B$. The resulting symbolic ADT representation is still satisfied by
+all histories in $A$, therefore it would still be sound for modular program
+reasoning, identifying only actual violations, and implying the correctness of
+client programs which do not depend on the stronger criteria which excludes
+unidentified violations. However, the under-approximation may be incomplete in
+identifying all violations, and in proving client programs which depend on the
+stronger criteria which excludes them all.

### 2015-07-09 formula.tex
Algorithm cleanup.

@@
 
-We obtain a formula representing an ADT $A$ by taking the conjunction of the
+We obtain a formula representing an ADT by taking the conjunction of the
 negative embedding formulas $\set{ \lnot\textsc{Embed}_h : h \in H }$ from any
-set $H$ that \emph{generates} the bounded complement $B$ of $\ker A$, i.e.,
-$H\subseteq B\subseteq Cl_\preceq(H)$ where $Cl_\preceq(H)$ is the closure of $H$ w.r.t. $\preceq$.
+set $H$ that generates its bounded complement $B$, i.e.,~whose closure under
+$\preceq$, denoted by $H^*$ henceforth, is equal to $B$. We define the
+\emph{exclusion formula} of a set of histories $H$ as $F(H) = \bigwedge_{h \in
+H} \neg \textsc{Embed}_h$.
 
-\begin{lemma}\label{lem:formula2}
+\begin{lemma}
+  \label{lem:formula2}
 
-  The formula $F(H)=\bigwedge_{h \in H} \neg \textsc{Embed}_h$ represents $A$,
-  for any set $H$ which generates the bounded complement of $\ker A$.
+  Let $B$ be the bounded complement of a bounded-width ADT $A$, and $H$
+  a set of histories. $F(H)$ represents $A$ if $H^* = B$.
 
 \end{lemma}
+
 \begin{proof}
-For a set $\Sigma$, let $\neg \Sigma$ denote its complement.
-By Lemma~\ref{lem:ADT_closure}, the complement of $A$ is closed under $\preceq$. Since
-$H\subseteq B\subseteq  \neg \ker A \subseteq \neg A$, we have that $Cl(H)\subseteq \neg A$.
-By Lemma~\ref{lem:formula1}, the formula $F(H)$ holds exactly for the set of histories $\neg Cl(H)$.
-%which is the complement of $Cl(M)$.
-By the previous inclusions, we have that $A\subseteq \neg Cl(H)\subseteq \neg B$.
-Therefore, $F(H)$ holds for all histories in $\ker A$
-and doesn't hold for histories in $B$, which concludes the proof.
+
+  For a set $\Sigma$, let $\neg \Sigma$ denote its complement. By
+  Lemma~\ref{lem:ADT_closure}, the complement of $A$ is closed under $\preceq$.
+  Since $H\subseteq B\subseteq  \neg A \subseteq \neg A$, we have that
+  $H^* \subseteq \neg A$. By Lemma~\ref{lem:formula1}, the formula $F(H)$ holds
+  exactly for the set of histories $\neg H^*$. By the previous inclusions, we
+  have that $A\subseteq \neg H^* \subseteq \neg B$. Therefore, $F(H)$ holds for
+  all histories in $A$ and doesn't hold for histories in $B$.
 \end{proof}

### 2015-07-09 overview.tex
Algorithm cleanup.

@@
 algorithm are sound, in the sense that when modular program reasoning with
-inferred specifications succeeds, correctness follows. In any case, the
-specifications inferred by our refinement of this abstract algorithm in
-Section~\ref{sec:algorithm} are both sound and complete for the
-naturally-occurring ADTs of interest. The remainder of this article develops
+inferred specifications succeeds, correctness follows. In any case, in
+Section~\ref{sec:algorithm} we give conditions under which our refinement
+of this abstract algorithm is both sound and complete, and show that these
+conditions hold for naturally-occurring ADTs, thus resulting in sound and
+complete symbolic ADT specifications. The remainder of this article develops
 the technical machinery required to realize this abstract inference algorithm.

### 2015-07-09 appendix.tex
Space.

@@
 %!TEX root = draft.tex
-\appendix
-\section{Proof of Lemma~\ref{lem:wqo}}
-
-It is well-known that every potentially infinite set of elements has a finite set of minimals w.r.t. a well-quasi-ordering.
+\section{Proof that $\preceq$ is a Well-Quasi-Order (Lemma~\ref{lem:wqo})}
 
@@
 
-We actually prove that another order on interval orders, denoted by $\subseteqq$, which is stronger than the embedding order $\subseteq$, is a wqo.
+We prove that another order $\subseteqq$, stronger than the embedding order $\subseteq$, is a wqo.
 The order $\subseteqq$ is defined by: $\Gamma_1\subseteqq \Gamma_2$ iff there exists an injective function $h:\Gamma_1\rightarrow \Gamma_2$ such that
@@
 
-It remains to prove that:
-\begin{itemize}
-	\item if $Init(\Gamma_{k_m})=Init(\Gamma_{k_n})$ and $\Lambda_{k_m}\subseteqq \Lambda_{k_n}$, then $\Gamma_{k_m}\subseteqq \Gamma_{k_n}$.
-\end{itemize}
-
-Let $h$ be the injective function witnessing $\Lambda_{k_m}\subseteqq \Lambda_{k_n}$.
-We prove that the extension $h'$ of $h$ between $\Gamma_{k_m}$ and $\Gamma_{k_n}$, defined by $h'(Min(\Gamma_{k_m}))=Min(\Gamma_{k_n})$ and $h'(t)=h(t)$, for all $t\in \Lambda_{k_m}$, witnesses $\Gamma_{k_m}\subseteqq \Gamma_{k_n}$.
+It remains to prove that $\Gamma_{k_m}\subseteqq \Gamma_{k_n}$ when
+$Init(\Gamma_{k_m})=Init(\Gamma_{k_n})$ and $\Lambda_{k_m}\subseteqq
+\Lambda_{k_n}$. Let $h$ be the injective function witnessing
+$\Lambda_{k_m}\subseteqq \Lambda_{k_n}$. We prove that the extension $h'$ of $h$
+between $\Gamma_{k_m}$ and $\Gamma_{k_n}$, defined by
+$h'(Min(\Gamma_{k_m}))=Min(\Gamma_{k_n})$ and $h'(t)=h(t)$, for all $t\in
+\Lambda_{k_m}$, witnesses $\Gamma_{k_m}\subseteqq \Gamma_{k_n}$.
 
@@
 \end{align*}
-
-The history signature contains sets of signatures
-(for matches, read-only and respectively, completed and unmatched operations)
-and the multiset of signatures of the pending and non-matched operations.
-
+The history signature contains sets of signatures (for matches, read-only,
+completed, and unmatched operations) and the multiset of signatures of the
+pending and non-matched operations.
 %Assume by contradiction that $\preceq$ has an infinite antichain.

### 2015-07-09 matching.tex
Section titles + restructuring the restructuring.
Space.

@@
 %!TEX root = draft.tex
-\section{Computing Matching Schemes}
+\section{Matching Scheme Inference}
 \label{sec:matching}
@@
 Any such pair of executions can be used to rule out several possibilities,
-for instance, that the operations removed from $e$ to obtain $e'$:
+e.g.,~that the operations removed from $e$ to obtain $e'$:
 \begin{itemize}
@@
 
-In this case, the reason for unsatisfiability can be used as a counterexample
-to refine the given language, for example, by adding additional predicates.
+The reason for unsatisfiability can be used as a counterexample to refine the
+given language, e.g.,~by adding additional predicates.
 
@@
 
-On the other hand, the satisfiability of $\textsc{Normalizes}_F$ does not
+Satisfiability of $\textsc{Normalizes}_F$ does not
 necessarily lead to a unique matching scheme, since $\textsc{Normalize}_F$ can

### 2015-07-09 nature.tex
Space.

@@
 
-It is not hard to see that the atomic queues and stacks are
+Atomic queues and stacks are
 normal. They are closed under removal of read-only operations: only empty
@@
 Each operation returns the nil value $v = -$ when $u$ is not yet
-present, and otherwise returns $v = u$. As its name implies, the atomic set
-ADT is sequential.
+present, and otherwise returns $v = u$. The atomic set ADT is sequential.
 
@@
 \end{itemize}
-It is easy to see that this matching scheme is faithful, and normalizing.
+This matching scheme is faithful, and normalizing.
 
@@
 twice: once normally, via the take operation, and once exceptionally, via the
-steal operation. We give a faithful matching scheme over executions which add
-unique values analogously to the scheme for stacks and queues, in which give
+steal operation. A faithful matching scheme over executions which add
+unique values is analogous to the scheme for stacks and queues: give
 operations match themselves, while take and steal operations match the give

### 2015-07-09 inference.tex
Reference to proof.

@@
 
-Let $H^*$ denote the \emph{closure} of a history set $H$ under the relation
-$\mathord\to = (\to_\mathrm{o} \cup \to_\mathrm{p} \cup \to_\mathrm{c})$, i.e.,
-\begin{align*}
-  H^* = \set{ h' : \exists h \in H.\ h \to^\ast h' }.
-\end{align*}
-A set $K$ \emph{generates} $H$ when $K^* = H$, and a \emph{kernel} of a set $H$
-is a minimal set $K$ which generates $H$. While the kernels of arbitrary sets of
-histories need not be unique, the kernels of sets which have sequential kernels
-are unique. Furthermore, Section~\ref{sec:nature} demonstrates that the
-histories of naturally-occurring implementations have unique kernels, which we
-assume for the remainder of this work. An \emph{abstract data type (ADT)} $A$ is
-the kernel of the set $H(\mathcal{I})$  of histories of some implementation
-$\mathcal{I}$. We say that \emph{$A$ is the ADT of $\mathcal{I}$} when $A^* =
-H(\mathcal{I})$.
+A Set $K$ \emph{generates} $H$ when the closure of $K$ under the relation
+$\mathord\to = (\to_\mathrm{o} \cup \to_\mathrm{p} \cup \to_\mathrm{c})$
+is equal to $H$, i.e., $H = \set{ h : \exists h' \in K.\ h' \to^\ast h }$.
+A \emph{kernel} of a set $H$ is a minimal set generating $H$. While the kernels
+of arbitrary sets need not be unique, the kernels of sets which
+have sequential kernels are unique. Furthermore, Section~\ref{sec:nature}
+demonstrates that the histories of naturally-occurring implementations have
+unique kernels, which we assume for the remainder of this work.
+An \emph{abstract data type (ADT)} $A$ is the kernel of the set $H(\mathcal{I})$
+of histories of some implementation $\mathcal{I}$; we say that \emph{$A$ is the
+ADT of $\mathcal{I}$}.
 

### 2015-07-09 patterns.tex
Reference to proof.

@@
 
-The previous example demonstrates that $\preceq$ is not a wqo by allowing each
-history $h_i$ of the infinite sequence to contain more and more pending
-operations in order to ensure that $h_j \not\preceq h_i$ for every $j < i$.
-Curbing this ability by limiting the maximum amount of pending operations per
-history makes $\preceq$ a wqo. Although limiting to $k$ pending operations
-essentially limits us to width-$k$ histories, of executions with at most $k$
-operations parallel at any moment, e.g.,~of programs with at most $k$ threads,
-this restriction comes at no loss of completeness when considering only the
-histories of bounded-width ADTs, which is the subject of the remainder
-of this section.
+This example demonstrates that $\preceq$ is not a wqo by allowing each history
+$h_i$ of the infinite sequence to contain more and more pending operations in
+order to ensure that $h_j \not\preceq h_i$ for every $j < i$. Curbing this
+ability by limiting the maximum amount of pending operations per history makes
+$\preceq$ a wqo. Although limiting to $k$ pending operations essentially limits
+us to width-$k$ histories, of executions with at most $k$ operations parallel at
+any moment, e.g.,~of programs with at most $k$ threads, this restriction comes
+at no loss of completeness when considering only the histories of bounded-width
+ADTs, which is the subject of the remainder of this section.
 
@@
 
-  $\preceq$ is a wqo on width-bounded histories.
+  $\preceq$ is a wqo on bounded-width histories.\footnote{The proof of
+  Lemma~\ref{lem:wqo} appears in Appendix~\ref{sec:proof}.}
 
@@
 on $B$. Furthermore, when $A$ normal, it is closed under $\succeq$, and thus $B$
-is closed under $\preceq$~\footnote{Actually, $B$ is closed under $\preceq$ when
-restricted to width-bounded histories, i.e., if $h_1 \preceq h_2$, $h_1 \in B$,
-and $h_2$ is width-bounded, then $h_2\in B$.}. Closure under a relation
+is closed\footnote{Actually, $B$ is closed under $\preceq$ when restricted to
+width-bounded histories, i.e., if $h_1 \preceq h_2$, $h_1 \in B$, and $h_2$ is
+width-bounded, then $h_2\in B$.} under $\preceq$. Closure under a relation
 satisfying Lemma~\ref{lem:wqo} implies representation by a finite set. Formally,

### 2015-10-25 draft.tex
Better overview.
Tried to clean up the f**king mess of the algorithm section.
Space.
Applied SIGPLAN formatting guidelines.

@@
-\documentclass[9pt,preprint,nocopyrightspace,natbib,authoryear]{sigplanconf}
+\documentclass[9pt,natbib,authoryear]{sigplanconf}
 
 \usepackage[utf8]{inputenc}
+\usepackage[T1]{fontenc}
 \usepackage{microtype}
@@
 \usepackage{listings}
+\usepackage[ruled]{algorithm2e}
 
@@
 \newtheorem{theorem}{Theorem}
-\newtheorem{remark}{Remark}
 \newtheorem{lemma}{Lemma}
@@
 
-\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country}
-\copyrightyear{20yy}
-\copyrightdata{978-1-nnnn-nnnn-n/yy/mm}
-\doi{nnnnnnn.nnnnnnn}
+% \conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country}
+% \copyrightyear{20yy}
+% \copyrightdata{978-1-nnnn-nnnn-n/yy/mm}
+% \doi{nnnnnnn.nnnnnnn}
+\toappear{}
 
@@
 
-\authorinfo{}{}{}
+\authorinfo{Michael Emmi}
+           {IMDEA Software Institute, Spain}
+           {michael.emmi@imdea.org}
 
-% \authorinfo{Michael Emmi}
-%            {IMDEA Software Institute}
-%            {michael.emmi@imdea.org}
-% \authorinfo{Constantin Enea}
-%            {Université Paris Diderot}
-%            {cenea@univ-paris-diderot.fr}
+\authorinfo{Constantin Enea}
+           {Université Paris Diderot, France}
+           {cenea@univ-paris-diderot.fr}
 
@@
 \input intro
-\input motivation
+\input overview
 \input histories
@@
 
-% \appendix
+\appendix
+\input appendix
 
@@
 \bibliography{violin}
-
-\input appendix
 

### 2015-10-25 histories.tex
Algorithm cleanup.
Fixed typos from Reviewer A.
Corrections from Reviewer B.

@@
 \emph{match target}, and the inverse set $m^{-1}(o)$ of a target is a
-\emph{match}. We assume $o \in m^{-1}(o)$ for every $o$. A operation $o$ is \emph{unmatched}
-when $o$ is completed and $m(o)$ is undefined.
+\emph{match}. We assume $o \in m^{-1}(o)$ for every $o \in \img(m)$.
+An operation $o$ is \emph{unmatched} when $o$ is completed and $m(o)$ is undefined.
 
@@
 mutually unordered, and the \emph{width} of a history set is the maximum width
-of its elements. Width-$1$ histories are \emph{sequential}. The \emph{weight}
-of a history is the maximum among operation frequencies and the number of
-matches. Since the operation identifiers of a history have no intrinsic
-meaning, we consider equality between histories up to renaming of operation
-identifiers.
+of its elements. Width-$1$ histories are \emph{sequential}. Since the operation
+identifiers of a history have no intrinsic meaning, we consider equality between
+histories up to renaming of operation identifiers.
 
@@
 We denote the set $\set{H(e,M,R) : e \in E}$ of histories of an execution set
-$E$ by $H(E,M,R)$. When the matching scheme $\tup{M,E}$ is clear from the
+$E$ by $H(E,M,R)$. When the matching scheme $\tup{M,R}$ is clear from the
 context, we abbreviate $H(e,M,R)$ and $H(E,M,R)$ by $H(e)$ and $H(E)$.

### 2015-10-25 inference.tex
Corrections from Reviewer B.

@@
 
-A Set $K$ \emph{generates} $H$ when the closure of $K$ under the relation
+A Set $K$ \emph{generates} $H$ when the closure $K^*$ of $K$ under the relation
 $\mathord\to = (\to_\mathrm{o} \cup \to_\mathrm{p} \cup \to_\mathrm{c})$

### 2015-10-25 patterns.tex
Corrections from Reviewer B.

@@
 
-Besides this closure property, the $\preceq$ relation enjoys a certain notion of
+Besides this closure property, the inverse $\preceq$ relation enjoys a certain notion of
 well-foundedness when restricted to bounded-width histories: the set of
@@
 complement. When $A$ has bounded width, so does $B$, and thus $\preceq$ is a wqo
-on $B$. Furthermore, when $A$ normal, it is closed under $\succeq$, and thus $B$
+on $B$. Furthermore, when $A$ is normal, it is closed under $\succeq$, and thus $B$
 is closed\footnote{Actually, $B$ is closed under $\preceq$ when restricted to

### 2015-10-25 related.tex
Corrections from Reviewer B.

@@
 Inference of symbolic abstract data types is a recently-pertinent problem
-arising from the the emergence of efficient symbolic refinement-checking
+arising from the emergence of efficient symbolic refinement-checking
 algorithms~\cite{conf/popl/BouajjaniEEH15, conf/pldi/EmmiEH15} which require

### 2015-11-02 discussion.tex
Corrections from Reviewer B.
Paragraph formatting.

@@
 handles atomic ADTs. In principle, this limitation is not fundamental. The key
-difference is in the enumeration ADT histories. For a given ADT,
+difference is in the enumeration of ADT histories. For a given ADT,
 determining whether a given history is admitted or not reduces
@@
 Second, ADTs whose specifications rely on relations (besides equality) on
-method argument and return values cannot be expressed with our notion of
-histories based on matching schemes. A notable example of such an ADT is the
+method argument and return values cannot be expressed with our matching-scheme
+based notion of histories. A notable example is the
 \emph{priority queue}, whose \emph{dequeue} operations return the
 smallest/largest \emph{enqueue}d value which has not yet been dequeued.
-%It appears that such ADTs are rare in practice.
 Overcoming this limitation would
@@
 and it is not clear whether all implementations effectively keep track of this
-correspondence. Furthermore, choosing matches arbitrarily would be unsound, and
+correspondence. Choosing matches arbitrarily would be unsound, and
 result in inferring ADTs with artificial constraints, e.g.,~that the notify and
 wait operations behave according to a FIFO discipline, each wait matching the
-oldest unmatched notify. Overcoming this limitation may require introducing a
-notion of nondeterminism into matching schemes.
+oldest unmatched notify. Overcoming this limitation may require introducing
+nondeterminism into matching schemes.

### 2015-11-02 histories.tex
Paragraph formatting.

@@
 \end{itemize}
-The relation $<$ is an interval order because executions’ call and return
+The relation $<$ is an interval order since call and return
 actions are totally ordered~\cite{conf/popl/BouajjaniEEH15}. Non-completed
@@
 
-The \emph{width} of a history is the maximum number of its operations which are
-mutually unordered, and the \emph{width} of a history set is the maximum width
-of its elements. Width-$1$ histories are \emph{sequential}. Since the operation
-identifiers of a history have no intrinsic meaning, we consider equality between
-histories up to renaming of operation identifiers.
+The \emph{width} of a history is its maximum number of mutually-unordered
+operations; the \emph{width} of a history set is the maximum width
+of its elements. Width-$1$ histories are \emph{sequential}. Since operation
+identifiers have no intrinsic meaning, we consider equality between
+histories up to their renaming.
 

### 2015-11-02 intro.tex
Quick typo fix.
Removed kernel talk.
Paragraph formatting.

@@
 
-  \item Does every implementation execution have a valid linearization?
+  \item Is there a valid linearization for each implementation execution?
 
-  \item Does every valid linearizations preserve client invariants?
+  \item Does every valid linearization preserve client invariants?
 
@@
 The fundamental challenge is in identifying the algebraic properties of ADTs
-which allow us to characterize an infinite set of violating executions with a
+which characterize an infinite set of violating executions with a
 finite set of patterns. This characterization is non-trivial since an execution
@@
 \end{itemize}
-We conclude with a discussion about the limitations of our approach
+We conclude by discussing limitations
 (§\ref{sec:discussion}) and related work (§\ref{sec:related}).

### 2015-11-02 nature.tex
Margins.
Paragraph formatting.

@@
-%!TEX root = draft.tex
 \section{Naturally-Occurring ADTs}
@@
 for observational refinement~\cite{conf/pldi/EmmiEH15}. Our publicly available
-implementation\footnote{The URL has been suppressed for anonymity purposes.}
+implementation\footnote{\url{https://github.com/imdea-software/violin}}
 enumerates the sequential histories\footnote{Our current implementation is
@@
 are not generated by any other. In each case, our algorithm terminates in a
-matter of seconds.
+matter of seconds with a list of human-readable ADT violation patterns.
 

### 2015-11-02 overview.tex
Paragraph formatting.

@@
   \end{minipage}
-  \caption{Invalid sequences according to the atomic queue ADT. The histories
-  on the right are each redundant with some history on the left, and those on
+  \caption{Invalid sequences according to the atomic queue ADT. Histories
+  on the right are each redundant with some on the left; those on
   the left constitute a complete set. The “\#” symbols depict the time intervals
@@
 
-The key technical obstacle in realizing the abstract algorithm of
-Algorithm~\ref{alg:abstract}, overcome in Section~\ref{sec:patterns}, is the
-classification of violations into a finite set of patterns, rendering the
+The key technical obstacle in realizing Algorithm~\ref{alg:abstract},
+which we overcome in Section~\ref{sec:patterns}, is
+classifying violations into a finite set of patterns, rendering the
 remaining invalid sequences redundant. The symbolic representation which

### 2015-11-02 algorithm.tex
Within the page limits.

@@
-%!TEX root = draft.tex
 \section{A Symbolic ADT Inference Algorithm}
@@
 
-Algorithm~\ref{alg:inference} solves the symbolic ADT inference problem by
+Algorithm~\ref{alg:inference} solves symbolic ADT inference by
 computing a finite representation of an ADT complement $B$ using the $\preceq$
-relation of Section~\ref{sec:patterns}. In general, this is achieved by
-enumerating the elements of $B$ while maintaining the $\preceq$-minimal
+relation of Section~\ref{sec:patterns}. This is generally achieved by
+enumerating $B$ while maintaining the $\preceq$-minimal
 elements, and recognizing a condition under which all the elements of $B$ are
@@
 journals/tcs/FinkelS01}. In our case, we stratify our enumeration of $B$ by the
-relative sizes of its histories. Formally, we say the \emph{weight} of a history
+relative sizes of its histories. Formally, the \emph{weight} of a history
 is the maximum among operation frequencies and the number of matches. We then

### 2015-11-02 formula.tex
Fixed typos from Reviewer A.
Corrections from Reviewer B.
Within the page limits.

@@
-%!TEX root = draft.tex
 \section{Symbolic ADT Representations}
@@
 
-While Section~\ref{sec:patterns} demonstrates that (the complements) of
+While Section~\ref{sec:patterns} demonstrates that (the complements of)
 naturally-occurring ADTs can be represented finitely, in this section we
-demonstrate that such finite representations have logical interpretations,
-allowing us to derive the formulas representing ADTs. In what follows, we
-describe how to obtain a formula which is satisfied by the histories which
-embed a given history, via $\preceq$. Then, using the finite set of histories
+demonstrate that those representations have logical interpretations,
+allowing us to derive formulas representing ADTs. In what follows we
+describe how to obtain formulas satisfied by the histories which
+embed a given history via $\preceq$. Then, using the finite set of histories
 which represent the complement of a given ADT, we represent the ADT itself as
@@
 o_n }$. Without loss of generality, we assume the match targets $\set{o_1,
-\ldots, o_k}$ of $h$ are indexed consecutively from $1$ to $k$, for some $k \le
-n$. For each $o \in O$, we define the macro $\textsc{Embed}_o(x,Y,z)$:
+\ldots, o_k}$ of $h$ are indexed consecutively from $1$ to $k$ for some $k \le
+n$. For each $o \in O$, the macro $\textsc{Embed}_o(x,Y,z)$:
 \begin{align*}
@@
 \end{align*}
-capturing the correspondence between the operation $o$ and the logical variable
+captures the correspondence between the operation $o$ and the logical variable
 $x$ representing $o$. The variables $Y$ and $z$ represent the operations
@@
   \forall x.
+  \lnot{\sf um}(x) \land
   {\sf m}(x) = z \implies
@@
   \end{align*}
-  The $\textsc{Embed}$ formula for second history,
+  The $\textsc{Embed}$ formula for the second history,
 \begin{verbatim}
@@
 set $H$ that generates its bounded complement $B$, i.e.,~whose closure under
-$\preceq$, denoted by $H^*$ henceforth, is equal to $B$. We define the
-\emph{exclusion formula} of a set of histories $H$ as $F(H) = \bigwedge_{h \in
+$\preceq$, denoted by $H^*$ henceforth, is equal to $B$. The
+\emph{exclusion formula} of a set of histories $H$ is $F(H) = \bigwedge_{h \in
 H} \neg \textsc{Embed}_h$.
@@
 
-  For a set $\Sigma$, let $\neg \Sigma$ denote its complement. By
+  Let $\neg \Sigma$ denote the complement of a set $\Sigma$. By
   Lemma~\ref{lem:ADT_closure}, the complement of $A$ is closed under $\preceq$.
-  Since $H\subseteq B\subseteq  \neg A \subseteq \neg A$, we have that
+  Since $H^* \subseteq B\subseteq \neg A$, we know
   $H^* \subseteq \neg A$. By Lemma~\ref{lem:formula1}, the formula $F(H)$ holds
-  exactly for the set of histories $\neg H^*$. By the previous inclusions, we
+  exactly for the histories $\neg H^*$. By the previous inclusions, we
   have that $A\subseteq \neg H^* \subseteq \neg B$. Therefore, $F(H)$ holds for
-  all histories in $A$ and doesn't hold for histories in $B$.
+  $A$ and not for $B$.
 \end{proof}
@@
 \end{example}
-
-%For normal and predictable ADTs, such sets are computable, and so our inference
-%problem is also computable.
-%
-%\begin{theorem}
-%
-%  The symbolic ADT inference problem is computable for normal and predictable ADTs.
-%
-%\end{theorem}

### 2015-11-02 matching.tex
Corrections from Reviewer B.
Within the page limits.

@@
-%!TEX root = draft.tex
 \section{Matching Scheme Inference}
@@
 The required manual specification includes identifying a class of executions to
-which an implementation should be subjected, and a set of predicates that are
+which an implementation should be exposed, and a set of predicates
 required in the logic of matching schemes. Formally, a \emph{language} $L =
@@
 P_2 \ldots$ and $Q = Q_1 Q_2 \ldots$ of binary and ternary predicates
-$P_i(e,o_1)$ and $Q_i(e,o_1,o_2)$ ranging over the executions of $E$ and their
+$P_i(e,o_1)$ and $Q_i(e,o_1,o_2)$ ranging over executions and their
 operations. When $P$ is a $k$-length sequence of $n$-ary predicates, we write
@@
   We define the function $G(x_\mathsf{w}, y_\mathsf{w}, z_\mathsf{v})$ over
-  valuations to the predicates above to be satisfied if and only if:
-  \begin{align*}
-    x_\mathsf{w} \land z_\mathsf{v}
-  \end{align*}
+  valuations to the predicates above to be satisfied if and only if
+  $x_\mathsf{w} \land z_\mathsf{v}$.
   Intuitively, this defines a simple matching scheme for which write
@@
   execution which writes unique values. The match is otherwise undefined. This
-  scheme is exactly that which is specified in Example~\ref{ex:matching} of
+  is the scheme specified in Example~\ref{ex:matching} of
   Section~\ref{sec:histories}.
@@
   \bigwedge_{o_1,o_2 \in O_e} \Big(
-    \textsc{IsMatch}_{e,o_2,o_2} \implies
+    \textsc{IsMatch}_{e,o_1,o_2} \implies
     \bigwedge_{o_1' \neq o_1} \lnot \textsc{IsMatch}_{e,o_1',o_2}
@@
   Example~\ref{ex:register-language}, yet this time without the predicate ${\sf
-  v}(e,o_1,o_2)$. The following table lists all possible predicate valuations
-  $\tup{x_{\sf w}, y_{\sf w}}$, and for each valuation a valid execution which
-  excludes the positive valuation of ${\sf g}(x_{\sf w}, y_{\sf w})$ in the
-  satisfaction of $\textsc{Normalize}_F$.
+  v}(e,o_1,o_2)$. The following table lists valid executions excluding each
+  of the possible predicate valuations $\tup{x_{\sf w}, y_{\sf w}}$
+  in the satisfaction of $\textsc{Normalize}_F$.
   \begin{center}
@@
   \end{center}
-  Thus any enumeration $F$ including the executions above excludes every
-  possible valuation of $\tup{x_{\sf w}, y_{\sf w}}$, resulting in an
+  Thus any enumeration $F$ including the executions above results in an
   unsatisfiable $\textsc{Normalize}_F$.

### 2015-11-02 draft.tex
Final version candidate.

@@
 \bibliographystyle{abbrvnat}
-\bibliography{violin}
 
-% % The bibliography should be embedded for final submission.
-%
-% \begin{thebibliography}{}
-% \softraggedright
-%
-% \bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
-% P. Q. Smith, and X. Y. Jones. ...reference text...
-%
-% \end{thebibliography}
+% \bibliography{dblp,books}
+
+\begin{thebibliography}{26}
+\providecommand{\natexlab}[1]{#1}
+\providecommand{\url}[1]{\texttt{#1}}
+\expandafter\ifx\csname urlstyle\endcsname\relax
+  \providecommand{\doi}[1]{doi: #1}\else
+  \providecommand{\doi}{doi: \begingroup \urlstyle{rm}\Url}\fi
+
+\bibitem[Abdulla et~al.(1996)Abdulla, Cerans, Jonsson, and
+  Tsay]{conf/lics/AbdullaCJT96}
+P.~A. Abdulla, K.~Cerans, B.~Jonsson, and Y.~Tsay.
+\newblock General decidability theorems for infinite-state systems.
+\newblock In \emph{Proceedings, 11th Annual {IEEE} Symposium on Logic in
+  Computer Science, New Brunswick, New Jersey, USA, July 27-30, 1996}, pages
+  313--321. {IEEE} Computer Society, 1996.
+\newblock \url{http://dx.doi.org/10.1109/LICS.1996.561359}.
+
+\bibitem[Abdulla et~al.(2013)Abdulla, Haziza, Hol{\'{\i}}k, Jonsson, and
+  Rezine]{conf/tacas/AbdullaHHJR13}
+P.~A. Abdulla, F.~Haziza, L.~Hol{\'{\i}}k, B.~Jonsson, and A.~Rezine.
+\newblock An integrated specification and verification technique for highly
+  concurrent data structures.
+\newblock In \emph{Tools and Algorithms for the Construction and Analysis of
+  Systems - 19th International Conference, {TACAS} 2013, Held as Part of the
+  European Joint Conferences on Theory and Practice of Software, {ETAPS} 2013,
+  Rome, Italy, March 16-24, 2013. Proceedings}, volume 7795 of \emph{Lecture
+  Notes in Computer Science}, pages 324--338. Springer, 2013.
+\newblock \url{http://dx.doi.org/10.1007/978-3-642-36742-7_23}.
+
+\bibitem[Amit et~al.(2007)Amit, Rinetzky, Reps, Sagiv, and
+  Yahav]{conf/cav/AmitRRSY07}
+D.~Amit, N.~Rinetzky, T.~W. Reps, M.~Sagiv, and E.~Yahav.
+\newblock Comparison under abstraction for verifying linearizability.
+\newblock In \emph{Computer Aided Verification, 19th International Conference,
+  {CAV} 2007, Berlin, Germany, July 3-7, 2007, Proceedings}, volume 4590 of
+  \emph{Lecture Notes in Computer Science}, pages 477--490. Springer, 2007.
+\newblock \url{http://dx.doi.org/10.1007/978-3-540-73368-3_49}.
+
+\bibitem[Angluin(1987)]{journals/iandc/Angluin87}
+D.~Angluin.
+\newblock Learning regular sets from queries and counterexamples.
+\newblock \emph{Inf. Comput.}, 75\penalty0 (2):\penalty0 87--106, 1987.
+\newblock \url{http://dx.doi.org/10.1016/0890-5401(87)90052-6}.
+
+\bibitem[Bouajjani et~al.(2015{\natexlab{a}})Bouajjani, Emmi, Enea, and
+  Hamza]{conf/icalp/BouajjaniEEH15}
+A.~Bouajjani, M.~Emmi, C.~Enea, and J.~Hamza.
+\newblock On reducing linearizability to state reachability.
+\newblock In \emph{Automata, Languages, and Programming - 42nd International
+  Colloquium, {ICALP} 2015, Kyoto, Japan, July 6-10, 2015, Proceedings, Part
+  {II}}, volume 9135 of \emph{Lecture Notes in Computer Science}, pages
+  95--107. Springer, 2015{\natexlab{a}}.
+\newblock \url{http://dx.doi.org/10.1007/978-3-662-47666-6_8}.
+
+\bibitem[Bouajjani et~al.(2015{\natexlab{b}})Bouajjani, Emmi, Enea, and
+  Hamza]{conf/popl/BouajjaniEEH15}
+A.~Bouajjani, M.~Emmi, C.~Enea, and J.~Hamza.
+\newblock Tractable refinement checking for concurrent objects.
+\newblock In \emph{Proceedings of the 42nd Annual {ACM} {SIGPLAN-SIGACT}
+  Symposium on Principles of Programming Languages, {POPL} 2015, Mumbai, India,
+  January 15-17, 2015}, pages 651--662. {ACM}, 2015{\natexlab{b}}.
+\newblock \url{http://doi.acm.org/10.1145/2676726.2677002}.
+
+\bibitem[Burckhardt et~al.(2010)Burckhardt, Dern, Musuvathi, and
+  Tan]{conf/pldi/BurckhardtDMT10}
+S.~Burckhardt, C.~Dern, M.~Musuvathi, and R.~Tan.
+\newblock Line-up: a complete and automatic linearizability checker.
+\newblock In \emph{Proceedings of the 2010 {ACM} {SIGPLAN} Conference on
+  Programming Language Design and Implementation, {PLDI} 2010, Toronto,
+  Ontario, Canada, June 5-10, 2010}, pages 330--340. {ACM}, 2010.
+\newblock \url{http://doi.acm.org/10.1145/1806596.1806634}.
+
+\bibitem[Burnim et~al.(2011)Burnim, Necula, and Sen]{conf/asplos/BurnimNS11}
+J.~Burnim, G.~C. Necula, and K.~Sen.
+\newblock Specifying and checking semantic atomicity for multithreaded
+  programs.
+\newblock In \emph{Proceedings of the 16th International Conference on
+  Architectural Support for Programming Languages and Operating Systems,
+  {ASPLOS} 2011, Newport Beach, CA, USA, March 5-11, 2011}, pages 79--90.
+  {ACM}, 2011.
+\newblock \url{http://doi.acm.org/10.1145/1950365.1950377}.
+
+\bibitem[Dodds et~al.(2015)Dodds, Haas, and Kirsch]{conf/popl/DoddsHK15}
+M.~Dodds, A.~Haas, and C.~M. Kirsch.
+\newblock A scalable, correct time-stamped stack.
+\newblock In \emph{Proceedings of the 42nd Annual {ACM} {SIGPLAN-SIGACT}
+  Symposium on Principles of Programming Languages, {POPL} 2015, Mumbai, India,
+  January 15-17, 2015}, pages 233--246. {ACM}, 2015.
+\newblock \url{http://doi.acm.org/10.1145/2676726.2676963}.
+
+\bibitem[Dragoi et~al.(2013)Dragoi, Gupta, and Henzinger]{conf/cav/DragoiGH13}
+C.~Dragoi, A.~Gupta, and T.~A. Henzinger.
+\newblock Automatic linearizability proofs of concurrent objects with
+  cooperating updates.
+\newblock In \emph{Computer Aided Verification - 25th International Conference,
+  {CAV} 2013, Saint Petersburg, Russia, July 13-19, 2013. Proceedings}, volume
+  8044 of \emph{Lecture Notes in Computer Science}, pages 174--190. Springer,
+  2013.
+\newblock \url{http://dx.doi.org/10.1007/978-3-642-39799-8_11}.
+
+\bibitem[Emmi et~al.(2015)Emmi, Enea, and Hamza]{conf/pldi/EmmiEH15}
+M.~Emmi, C.~Enea, and J.~Hamza.
+\newblock Monitoring refinement via symbolic reasoning.
+\newblock In \emph{Proceedings of the 36th {ACM} {SIGPLAN} Conference on
+  Programming Language Design and Implementation, Portland, OR, USA, June
+  15-17, 2015}, pages 260--269. {ACM}, 2015.
+\newblock \url{http://doi.acm.org/10.1145/2737924.2737983}.
+
+\bibitem[Finkel and Schnoebelen(2001)]{journals/tcs/FinkelS01}
+A.~Finkel and P.~Schnoebelen.
+\newblock Well-structured transition systems everywhere!
+\newblock \emph{Theor. Comput. Sci.}, 256\penalty0 (1-2):\penalty0 63--92,
+  2001.
+\newblock \url{http://dx.doi.org/10.1016/S0304-3975(00)00102-X}.
+
+\bibitem[Fishburn(1985)]{books/Fishburn85}
+P.~C. Fishburn.
+\newblock \emph{Interval Orders and Interval Graphs: A Study of Partially
+  Ordered Sets}.
+\newblock Wiley-Interscience Series in Discrete Mathematics. John Wiley \& Sons
+  Inc, 1985.
+
+\bibitem[Giannakopoulou et~al.(2005)Giannakopoulou, Pasareanu, and
+  Barringer]{journals/ase/GiannakopoulouPB05}
+D.~Giannakopoulou, C.~S. Pasareanu, and H.~Barringer.
+\newblock Component verification with automatically generated assumptions.
+\newblock \emph{Autom. Softw. Eng.}, 12\penalty0 (3):\penalty0 297--320, 2005.
+\newblock \url{http://dx.doi.org/10.1007/s10515-005-2641-y}.
+
+\bibitem[Hemed and Rinetzky(2014)]{conf/podc/HemedR14}
+N.~Hemed and N.~Rinetzky.
+\newblock Brief announcement: concurrency-aware linearizability.
+\newblock In \emph{{ACM} Symposium on Principles of Distributed Computing,
+  {PODC} '14, Paris, France, July 15-18, 2014}, pages 209--211. {ACM}, 2014.
+\newblock \url{http://doi.acm.org/10.1145/2611462.2611513}.
+
+\bibitem[Hendler et~al.(2006)Hendler, Lev, Moir, and
+  Shavit]{journals/dc/HendlerLMS06}
+D.~Hendler, Y.~Lev, M.~Moir, and N.~Shavit.
+\newblock A dynamic-sized nonblocking work stealing deque.
+\newblock \emph{Distributed Computing}, 18\penalty0 (3):\penalty0 189--207,
+  2006.
+\newblock \url{http://dx.doi.org/10.1007/s00446-005-0144-5}.
+
+\bibitem[Henzinger et~al.(2013)Henzinger, Sezgin, and
+  Vafeiadis]{conf/concur/HenzingerSV13}
+T.~A. Henzinger, A.~Sezgin, and V.~Vafeiadis.
+\newblock Aspect-oriented linearizability proofs.
+\newblock In \emph{{CONCUR} 2013 - Concurrency Theory - 24th International
+  Conference, {CONCUR} 2013, Buenos Aires, Argentina, August 27-30, 2013.
+  Proceedings}, volume 8052 of \emph{Lecture Notes in Computer Science}, pages
+  242--256. Springer, 2013.
+\newblock \url{http://dx.doi.org/10.1007/978-3-642-40184-8_18}.
+
+\bibitem[Herlihy and Wing(1990)]{journals/toplas/HerlihyW90}
+M.~Herlihy and J.~M. Wing.
+\newblock Linearizability: {A} correctness condition for concurrent objects.
+\newblock \emph{{ACM} Trans. Program. Lang. Syst.}, 12\penalty0 (3):\penalty0
+  463--492, 1990.
+\newblock \url{http://doi.acm.org/10.1145/78969.78972}.
+
+\bibitem[Liang and Feng(2013)]{conf/pldi/LiangF13}
+H.~Liang and X.~Feng.
+\newblock Modular verification of linearizability with non-fixed linearization
+  points.
+\newblock In \emph{{ACM} {SIGPLAN} Conference on Programming Language Design
+  and Implementation, {PLDI} '13, Seattle, WA, USA, June 16-19, 2013}, pages
+  459--470. {ACM}, 2013.
+\newblock \url{http://doi.acm.org/10.1145/2462156.2462189}.
+
+\bibitem[Liu et~al.(2009)Liu, Chen, Liu, and Sun]{conf/fm/LiuCLS09}
+Y.~Liu, W.~Chen, Y.~A. Liu, and J.~Sun.
+\newblock Model checking linearizability via refinement.
+\newblock In \emph{{FM} 2009: Formal Methods, Second World Congress, Eindhoven,
+  The Netherlands, November 2-6, 2009. Proceedings}, volume 5850 of
+  \emph{Lecture Notes in Computer Science}, pages 321--337. Springer, 2009.
+\newblock \url{http://dx.doi.org/10.1007/978-3-642-05089-3_21}.
+
+\bibitem[O'Hearn et~al.(2010)O'Hearn, Rinetzky, Vechev, Yahav, and
+  Yorsh]{conf/podc/OHearnRVYY10}
+P.~W. O'Hearn, N.~Rinetzky, M.~T. Vechev, E.~Yahav, and G.~Yorsh.
+\newblock Verifying linearizability with hindsight.
+\newblock In \emph{Proceedings of the 29th Annual {ACM} Symposium on Principles
+  of Distributed Computing, {PODC} 2010, Zurich, Switzerland, July 25-28,
+  2010}, pages 85--94. {ACM}, 2010.
+\newblock \url{http://doi.acm.org/10.1145/1835698.1835722}.
+
+\bibitem[Shacham et~al.(2011)Shacham, Bronson, Aiken, Sagiv, Vechev, and
+  Yahav]{conf/oopsla/ShachamBASVY11}
+O.~Shacham, N.~G. Bronson, A.~Aiken, M.~Sagiv, M.~T. Vechev, and E.~Yahav.
+\newblock Testing atomicity of composed concurrent operations.
+\newblock In \emph{Proceedings of the 26th Annual {ACM} {SIGPLAN} Conference on
+  Object-Oriented Programming, Systems, Languages, and Applications, {OOPSLA}
+  2011, part of {SPLASH} 2011, Portland, OR, USA, October 22 - 27, 2011}, pages
+  51--64. {ACM}, 2011.
+\newblock \url{http://doi.acm.org/10.1145/2048066.2048073}.
+
+\bibitem[Vafeiadis(2010)]{conf/cav/Vafeiadis10}
+V.~Vafeiadis.
+\newblock Automatically proving linearizability.
+\newblock In \emph{Computer Aided Verification, 22nd International Conference,
+  {CAV} 2010, Edinburgh, UK, July 15-19, 2010. Proceedings}, volume 6174 of
+  \emph{Lecture Notes in Computer Science}, pages 450--464. Springer, 2010.
+\newblock \url{http://dx.doi.org/10.1007/978-3-642-14295-6_40}.
+
+\bibitem[Wing and Gong(1993)]{journals/jpdc/WingG93}
+J.~M. Wing and C.~Gong.
+\newblock Testing and verifying concurrent objects.
+\newblock \emph{J. Parallel Distrib. Comput.}, 17\penalty0 (1-2):\penalty0
+  164--182, 1993.
+\newblock \url{http://dx.doi.org/10.1006/jpdc.1993.1015}.
+
+\bibitem[Zhang et~al.(2013)Zhang, Chattopadhyay, and Wang]{conf/kbse/ZhangCW13}
+L.~Zhang, A.~Chattopadhyay, and C.~Wang.
+\newblock Round-up: Runtime checking quasi linearizability of concurrent data
+  structures.
+\newblock In \emph{2013 28th {IEEE/ACM} International Conference on Automated
+  Software Engineering, {ASE} 2013, Silicon Valley, CA, USA, November 11-15,
+  2013}, pages 4--14. {IEEE}, 2013.
+\newblock \url{http://dx.doi.org/10.1109/ASE.2013.6693061}.
+
+\bibitem[Zhang(2011)]{conf/icse/Zhang11a}
+S.~J. Zhang.
+\newblock Scalable automatic linearizability checking.
+\newblock In \emph{Proceedings of the 33rd International Conference on Software
+  Engineering, {ICSE} 2011, Waikiki, Honolulu , HI, USA, May 21-28, 2011},
+  pages 1185--1187. {ACM}, 2011.
+\newblock \url{http://doi.acm.org/10.1145/1985793.1986037}.
+
+\end{thebibliography}
 

### 2015-11-02 related.tex
Final version candidate.

@@
 generalized to ADTs which can be expressed with a particular form of recursive
-definition~\cite{journals/arxiv/BouajjaniEEH15}. These works suggest patterns
+definition~\cite{conf/icalp/BouajjaniEEH15}. These works suggest patterns
 which directly recognize violations in \emph{concurrent} executions,

### 2015-11-04 draft.tex
Addressing Constantin’s comments.

@@
            {Université Paris Diderot, France}
-           {cenea@univ-paris-diderot.fr}
+           {cenea@liafa.univ-paris-diderot.fr}
 
@@
 % \acks …
+% \section*{Acknowledgements}
+
+\acks This work is supported in part by the VECOLIB project (ANR-14-CE28-0018),
+the N-GREENS Software project (Ref. S2013/ICE-2731), and AMAROUT-II
+(EU-FP7-COFUND-291803).
 

### 2015-11-04 histories.tex
Addressing Constantin’s comments.

@@
     \item read $\Rightarrow v$ operations are read-only, and match themselves
-    when they return $v = -$, or a write$(v)$ operation, if one
+    when they return $v = -$ (empty), or a write$(v)$ operation, if one
     exists, and otherwise have no match,

### 2015-11-04 inference.tex
Addressing Constantin’s comments.

--- 2015-07-09 inference.tex
@@
 
-A Set $K$ \emph{generates} $H$ when the closure of $K$ under the relation
+A set $K$ \emph{generates} $H$ when the closure $K^*$ of $K$ under the relation
 $\mathord\to = (\to_\mathrm{o} \cup \to_\mathrm{p} \cup \to_\mathrm{c})$

### 2015-11-04 intro.tex
Addressing Constantin’s comments.

@@
 reasoning, formal-specification writing is a burden for which few programmers
-posses the required combination of expertise and willingness to overcome.
+possess the required combination of expertise and willingness to overcome.
 Typically programmers write simple ADT \emph{reference implementations},

### 2015-11-04 nature.tex
Addressing Constantin’s comments.

@@
 first-in-first-out (FIFO) and last-in-first-out (LIFO) removal order,
-respectively, providing two methods:\footnote{These methods are also known as
-enqueue and dequeue, or push and pop.}
+respectively, providing two methods:
 \begin{itemize}
@@
 An execution \emph{inserts unique values} if the argument value to each insert
-operation which returns $-$ is unique. We give a faithful matching
+operation returning $-$ is unique. We give a faithful matching
 scheme over executions which insert unique values as follows:
@@
 value is returned. Note that the work-stealing queue ADT has width $2$,
-since a pairs of concurrent take and steal operations returning the same value
+since pairs of concurrent take and steal operations returning the same value
 are permitted, while sequentially they are not. As our implementation is
@@
 \begin{verbatim}
-  [1:_] take => 1  #             [1:_] steal => 1  #
+  [1:X] take => 1  #             [1:X] steal => 1  #
   ---                            ---

### 2015-11-04 patterns.tex
Addressing Constantin’s comments.

@@
-%!TEX root = draft.tex
 \section{Finite ADT Representations}
@@
 represented by finite sets of histories, despite the fact that these ADTs admit
-infinite sets of histories. This results relies on the identification of certain
+infinite sets of histories. This result relies on the identification of certain
 algebraic properties of the sets of histories admitted by ADTs. In particular,
@@
 \end{verbatim}
-  Every sequential history not admitted by the atomic register ADT embeds at
-  least one of these four histories.
+  Every sequential history not admitted by the atomic register ADT embeds
+  (via $\succeq$) at least one of these four histories.
 
